%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Unit 14: Order of Differential Equations}
\label{chap:unit14}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%===============================================================
\section{Order of Differential Equations}
\label{sec:order-of-differential-equations}
%%===============================================================

Introduction

% [Image: Visualization of heat transfer in a pump casing, created by solving
the heat equation. Heat is being generated internally in the casing and
being cooled at the boundary, providing a steady state temperature
distribution.]{width=""350""}
In mathematics, a \textbf{differential equation} is an equation that relates
one or more functions and their derivatives. In applications, the
functions generally represent physical quantities, the derivatives
represent their rates of change, and the differential equation defines a
relationship between the two. Such relations are common; therefore,
differential equations play a prominent role in many disciplines
including engineering, physics, economics, and biology.

Mainly the study of differential equations consists of the study of
their solutions (the set of functions that satisfy each equation), and
of the properties of their solutions. Only the simplest differential
equations are solvable by explicit formulas; however, many properties of
solutions of a given differential equation may be determined without
computing them exactly.

Often when a closed-form expression for the solutions is not available,
solutions may be approximated numerically using computers. The theory of
dynamical systems puts emphasis on qualitative analysis of systems
described by differential equations, while many numerical methods have
been developed to determine solutions with a given degree of accuracy.
The order of a differential equation is determined by the highest-order
derivative. The higher the order of the differential equation, the more
arbitrary constants need to be added to the general solution. A
first-order equation will have one, a second-order two, and so on. The
degree of a differential equation, similarly, is determined by the
highest exponent on any variables involved.

Examples:

\begin{itemize}
    \item $y'' + xy' - x^3y = \sin{x}$ is second order (or \""of order 2\"").
    \item $x't + x = t^2$ is first order.
    \item $y''' + 3y'' + 3y' + y = x^2$ is of order 3.
\end{itemize}

\paragraph{Types}


Differential equations can be divided into several types. Apart from
describing the properties of the equation itself, these classes of
differential equations can help inform the choice of approach to a
solution. Commonly used distinctions include whether the equation is
ordinary or partial, linear or non-linear, and homogeneous or
heterogeneous. This list is far from exhaustive; there are many other
properties and subclasses of differential equations which can be very
useful in specific contexts.

\paragraph{Ordinary differential equations}


An ordinary differential equation (\textit{ODE}) is an equation containing an
unknown function of one real or complex variable
`{{mvar|x}}`{=mediawiki}, its derivatives, and some given functions of
`{{mvar|x}}`{=mediawiki}. The unknown function is generally represented
by a variable (often denoted `{{mvar|y}}`{=mediawiki}), which,
therefore, \textit{depends} on `{{mvar|x}}`{=mediawiki}. Thus
`{{mvar|x}}`{=mediawiki} is often called the independent variable of the
equation. The term \""\textit{ordinary}\"" is used in contrast with the term
partial differential equation, which may be with respect to \textit{more than}
one independent variable.

Linear differential equations are the differential equations that are
linear in the unknown function and its derivatives. Their theory is well
developed, and in many cases one may express their solutions in terms of
integrals.

Most ODEs that are encountered in physics are linear. Therefore, most
special functions may be defined as solutions of linear differential
equations (see Holonomic function).

As, in general, the solutions of a differential equation cannot be
expressed by a closed-form expression, numerical methods are commonly
used for solving differential equations on a computer.

\paragraph{Partial differential equations}


A partial differential equation (\textit{PDE}) is a differential equation that
contains unknown multivariable functions and their partial derivatives.
(This is in contrast to ordinary differential equations, which deal with
functions of a single variable and their derivatives.) PDEs are used to
formulate problems involving functions of several variables, and are
either solved in closed form, or used to create a relevant computer
model.

PDEs can be used to describe a wide variety of phenomena in nature such
as sound, heat, electrostatics, electrodynamics, fluid flow, elasticity,
or quantum mechanics. These seemingly distinct physical phenomena can be
formalized similarly in terms of PDEs. Just as ordinary differential
equations often model one-dimensional dynamical systems, partial
differential equations often model multidimensional systems. Stochastic
partial differential equations generalize partial differential equations
for modeling randomness.

\paragraph{Non-linear differential equations}


A \textbf{non-linear differential equation} is a differential equation that
is not a linear equation in the unknown function and its derivatives
(the linearity or non-linearity in the arguments of the function are not
considered here). There are very few methods of solving nonlinear
differential equations exactly; those that are known typically depend on
the equation having particular symmetries. Nonlinear differential
equations can exhibit very complicated behavior over extended time
intervals, characteristic of chaos. Even the fundamental questions of
existence, uniqueness, and extendability of solutions for nonlinear
differential equations, and well-posedness of initial and boundary value
problems for nonlinear PDEs are hard problems and their resolution in
special cases is considered to be a significant advance in the
mathematical theory (cf. Navier--Stokes existence and smoothness).
However, if the differential equation is a correctly formulated
representation of a meaningful physical process, then one expects it to
have a solution.

Linear differential equations frequently appear as approximations to
nonlinear equations. These approximations are only valid under
restricted conditions. For example, the harmonic oscillator equation is
an approximation to the nonlinear pendulum equation that is valid for
small amplitude oscillations (see below).

\paragraph{Equation order}


Differential equations are described by their order, determined by the
term with the highest derivatives. An equation containing only first
derivatives is a \textit{first-order differential equation}, an equation
containing the second derivative is a *second-order differential
equation*, and so on. Differential equations that describe natural
phenomena almost always have only first and second order derivatives in
them, but there are some exceptions, such as the thin film equation,
which is a fourth order partial differential equation.

\paragraph{Examples}


In the first group of examples \textit{u} is an unknown function of \textit{x}, and
\textit{c} and \textit{?} are constants that are supposed to be known. Two broad
classifications of both ordinary and partial differential equations
consist of distinguishing between \textit{linear} and \textit{nonlinear} differential
equations, and between \textit{homogeneous} differential equations and
\textit{heterogeneous} ones.

\begin{itemize}
    \item Heterogeneous first-order linear constant coefficient ordinary
\end{itemize}
    differential equation:

:   

    :   $\frac{du}{dx} = cu+x^2.$

\begin{itemize}
    \item Homogeneous second-order linear ordinary differential equation:
\end{itemize}

:   

    :   $\frac{d^2u}{dx^2} - x\frac{du}{dx} + u = 0.$

\begin{itemize}
    \item Homogeneous second-order linear constant coefficient ordinary
\end{itemize}
    differential equation describing the harmonic oscillator:

:   

    :   $\frac{d^2u}{dx^2} + \omega^2u = 0.$

\begin{itemize}
    \item Heterogeneous first-order nonlinear ordinary differential equation:
\end{itemize}

:   

    :   $\frac{du}{dx} = u^2 + 4.$

\begin{itemize}
    \item Second-order nonlinear (due to sine function) ordinary differential
\end{itemize}
    equation describing the motion of a pendulum of length \textit{L}:

:   

    :   $L\frac{d^2u}{dx^2} + g\sin u = 0.$

In the next group of examples, the unknown function \textit{u} depends on two
variables \textit{x} and \textit{t} or \textit{x} and \textit{y}.

\begin{itemize}
    \item Homogeneous first-order linear partial differential equation:
\end{itemize}

:   

    :   $\frac{\partial u}{\partial t} + t\frac{\partial u}{\partial x} = 0.$

\begin{itemize}
    \item Homogeneous second-order linear constant coefficient partial
\end{itemize}
    differential equation of elliptic type, the Laplace equation:

:   

    :   $\frac{\partial^2 u}{\partial x^2} + \frac{\partial^2 u}{\partial y^2} = 0.$

\begin{itemize}
    \item Homogeneous third-order non-linear partial differential equation :
\end{itemize}

:   

    :   $\frac{\partial u}{\partial t} = 6u\frac{\partial u}{\partial x} - \frac{\partial^3 u}{\partial x^3}.$

\paragraph{Existence of solutions}


Solving differential equations is not like solving algebraic equations.
Not only are their solutions often unclear, but whether solutions are
unique or exist at all are also notable subjects of interest.

For first order initial value problems, the Peano existence theorem
gives one set of circumstances in which a solution exists. Given any
point $(a,b)$ in the xy-plane, define some rectangular region $Z$, such
that $Z = [l,m]\times[n,p]$ and $(a,b)$ is in the interior of $Z$. If we
are given a differential equation $\frac{dy}{dx} = g(x,y)$ and the
condition that $y=b$ when $x=a$, then there is locally a solution to
this problem if $g(x,y)$ and $\frac{\partial g}{\partial x}$ are both
continuous on $Z$. This solution exists on some interval with its center
at $a$. The solution may not be unique. (See Ordinary differential
equation for other results.)

However, this only helps us with first order initial value problems.
Suppose we had a linear initial value problem of the nth order:

$$f_{n}(x)\frac{d^n y}{dx^n} + \cdots + f_{1}(x)\frac{d y}{dx} + f_{0}(x)y = g(x)$$
such that

$$y(x_{0})=y_{0}, y'(x_{0}) = y'_{0}, y''(x_{0}) = y''_{0}, \ldots$$

For any nonzero $f_{n}(x)$, if $\{f_{0},f_{1},\ldots\}$ and $g$ are
continuous on some interval containing $x_{0}$, $y$ is unique and
exists.

\paragraph{Connection to difference equations}


The theory of differential equations is closely related to the theory of
difference equations, in which the coordinates assume only discrete
values, and the relationship involves values of the unknown function or
functions and values at nearby coordinates. Many methods to compute
numerical solutions of differential equations or study the properties of
differential equations involve the approximation of the solution of a
differential equation by the solution of a corresponding difference
equation.

\paragraph{Applications}


The study of differential equations is a wide field in pure and applied
mathematics, physics, and engineering. All of these disciplines are
concerned with the properties of differential equations of various
types. Pure mathematics focuses on the existence and uniqueness of
solutions, while applied mathematics emphasizes the rigorous
justification of the methods for approximating solutions. Differential
equations play an important role in modeling virtually every physical,
technical, or biological process, from celestial motion, to bridge
design, to interactions between neurons. Differential equations such as
those used to solve real-life problems may not necessarily be directly
solvable, i.e. do not have closed form solutions. Instead, solutions can
be approximated using numerical methods.

Many fundamental laws of physics and chemistry can be formulated as
differential equations. In biology and economics, differential equations
are used to model the behavior of complex systems. The mathematical
theory of differential equations first developed together with the
sciences where the equations had originated and where the results found
application. However, diverse problems, sometimes originating in quite
distinct scientific fields, may give rise to identical differential
equations. Whenever this happens, mathematical theory behind the
equations can be viewed as a unifying principle behind diverse
phenomena. As an example, consider the propagation of light and sound in
the atmosphere, and of waves on the surface of a pond. All of them may
be described by the same second-order partial differential equation, the
wave equation, which allows us to think of light and sound as forms of
waves, much like familiar waves in the water. Conduction of heat, the
theory of which was developed by Joseph Fourier, is governed by another
second-order partial differential equation, the heat equation. It turns
out that many diffusion processes, while seemingly different, are
described by the same equation; the Black--Scholes equation in finance
is, for instance, related to the heat equation.

\paragraph{Resources}


\begin{itemize}
    \item \href{https://courses.lumenlearning.com/boundless-calculus/chapter/differential-equations/}{Differential
\end{itemize}
    Equations},
    Lumen Learning
\begin{itemize}
    \item \href{https://www.analyzemath.com/calculus/Differential_Equations/order_linearity.html}{Order and Linearity of Differential
\end{itemize}
    Equations},
    Analyze Math

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Differential_equation}{Differential equation,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license"

%%===============================================================
\section{Solutions of Differential Equations}
\label{sec:solutions-of-differential-equations}
%%===============================================================

"A solution of a differential equation is an expression of the dependent
variable that satisfies the relation established in the differential
equation. For example, the solution of $y' + y - 2 = 4x$ will be some
equation y = f(x) such that y and its first derivative, y\', satisfy the
relation $y' + y - 2 = 4x$. The general solution of a differential
equation will have one or more arbitrary constants, depending on the
order of the original differential equation (the solution of a first
order diff. eq. will have one arbitrary constant, a second order one
will have two, etc.).

Examples:

\begin{itemize}
    \item $y' = 2x$. Through simple integration, we can calculate the general
\end{itemize}
    solution of this equation to be $y = x^2 + C$, where C is an
    arbitrary constant.
\begin{itemize}
    \item $y' - y = 0$. The G.S. is $y = Ce^{x}$. $y' = Ce^{x}$, so
\end{itemize}
    $y' - y = 0 \to Ce^{x} - Ce^{x} = 0$, so this solution satisfies the
    relationship for all arbitrary constants C.
\begin{itemize}
    \item $y'' + y' - 2y = 0$. The G.S. is $y = Ce^{x} + De^{-2x}$.
\end{itemize}
    $y' = Ce^{x} - 2De^{-2x}$ and $y'' = Ce^{x} + 4De^{-2x}$, so
    $0 = y'' + y' - 2y$ becomes
    $0 = Ce^{x} + 4De^{-2x} + Ce^{x} - 2De^{-2x} - 2(Ce^{x} + De^{-2x}) = Ce^{x} + Ce^{x} - 2Ce^{x} + 4De^{-2x} - 2De^{-2x} - 2De^{-2x}) = 0$.

The particular solution of a differential equation can be solved if we
have enough points to solve for the arbitrary constants (see initial
value problems for more).

\paragraph{Existence of solutions}


Solving differential equations is not like solving algebraic equations.
Not only are their solutions often unclear, but whether solutions are
unique or exist at all are also notable subjects of interest.

For first order initial value problems, the Peano existence theorem
gives one set of circumstances in which a solution exists. Given any
point $(a,b)$ in the xy-plane, define some rectangular region $Z$, such
that $Z = [l,m]\times[n,p]$ and $(a,b)$ is in the interior of $Z$. If we
are given a differential equation $\frac{dy}{dx} = g(x,y)$ and the
condition that $y=b$ when $x=a$, then there is locally a solution to
this problem if $g(x,y)$ and $\frac{\partial g}{\partial x}$ are both
continuous on $Z$. This solution exists on some interval with its center
at $a$. The solution may not be unique.

However, this only helps us with first order initial value problems.
Suppose we had a linear initial value problem of the nth order:

$$f_{n}(x)\frac{d^n y}{dx^n} + \cdots + f_{1}(x)\frac{d y}{dx} + f_{0}(x)y = g(x)$$
such that

$$y(x_{0})=y_{0}, y'(x_{0}) = y'_{0}, y''(x_{0}) = y''_{0}, \ldots$$

For any nonzero $f_{n}(x)$, if $\{f_{0},f_{1},\ldots\}$ and $g$ are
continuous on some interval containing $x_{0}$, $y$ is unique and
exists.

\paragraph{Resources}


\begin{itemize}
    \item \href{http://www.maths.gla.ac.uk/\textasciitilde{}cc/2x/2005_2xnotes/2x_chap5.pdf}{Differential
\end{itemize}
    Equations},
    University of Glascow
\begin{itemize}
    \item \href{https://www.api.simply.science/index.php/math/calculus/differential-equations/types-of-differential-equations/9869-general-and-particular-solutions-of-a-differential-equation}{General and Particular
\end{itemize}
    Solutions},
    Simply Math

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Differential_equation}{Differential equation,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license"

%%===============================================================
\section{Initial Value Problem (IVP)}
\label{sec:initial-value-problem-ivp}
%%===============================================================

Definition

An \textbf{initial value problem} is a differential equation

$$y'(t) = f(t, y(t))$$ with
$f\colon \Omega \subset \mathbb{R} \times \mathbb{R}^n \to \mathbb{R}^n$
where $\Omega$ is an open set of $\mathbb{R} \times \mathbb{R}^n$,
together with a point in the domain of $f$

$$(t_0, y_0) \in \Omega,$$ called the initial condition.

A \textbf{solution} to an initial value problem is a function $y$ that is a
solution to the differential equation and satisfies

$$y(t_0) = y_0.$$

In higher dimensions, the differential equation is replaced with a
family of equations $y_i'(t)=f_i(t, y_1(t), y_2(t), \dotsc)$, and $y(t)$
is viewed as the vector $(y_1(t), \dotsc, y_n(t))$, most commonly
associated with the position in space. More generally, the unknown
function $y$ can take values on infinite dimensional spaces, such as
Banach spaces or spaces of distributions.

Initial value problems are extended to higher orders by treating the
derivatives in the same way as an independent function, e.g.
$y''(t)=f(t,y(t),y'(t))$.

\paragraph{Examples}


With initial value problems, we are given a differential equation, and
one or more points (depending on the order of the equation) to solve the
constants in the general solution. For a first order differential
equation we need 1 point $(x, y(x))$, for a second order equation we
need 2 points (typically $(x_1, y(x_1))$ and either $(x_2, y(x_2))$ or
$(x_2, y'(x_2))$), and so on.

A simple example is to solve $y'(t) = 0.85 y(t)$ and $y(0) = 19$. We are
trying to find a formula for $y(t)$ that satisfies these two equations.

Rearrange the equation so that $y$ is on the left hand side

:   $\frac{y'(t)}{y(t)} = 0.85$

Now integrate both sides with respect to $t$ (this introduces an unknown
constant $B$).

:   $\int \frac{y'(t)}{y(t)}\,dt = \int 0.85\,dt$
:   $\ln |y(t)| = 0.85t + B$

Eliminate the logarithm with exponentiation on both sides

:   $| y(t) | = e^Be^{0.85t}$

Let $C$ be a new unknown constant, $C = \pm e^B$, so

:   $y(t) = Ce^{0.85t}$

Now we need to find a value for $C$. Use $y(0) = 19$ as given at the
start and substitute 0 for $t$ and 19 for $y$

:   $19 = C e^{0.85 \cdot 0}$
:   $C = 19$

this gives the final solution of $y(t) = 19e^{0.85t}$.

Second example

The solution of

:   $y'+3y=6t+5,\qquad y(0)=3$

can be found to be

:   $y(t)=2e^{-3t}+2t+1. \,$

Indeed,

:   $\begin{aligned}
    y'+3y \&= \tfrac{d}{dt} (2e^{-3t}+2t+1)+3(2e^{-3t}+2t+1) \\\          \&= (-6e^{-3t}+2)+(6e^{-3t}+6t+3) \\\          \&= 6t+5.
    \end{aligned}$

Some more simple examples of initial value problems:

\begin{itemize}
    \item $y' = 2x$, $y(2) = 0$. With this point and the general solution
\end{itemize}
    $y = x^2 + C$, we can calculate the constant C to be -4. Thus the
    particular solution is $y = x^2 - 4$.
\begin{itemize}
    \item $y' - y = 0$, $y(0) = 3$. $3 = Ce^{0} \implies C = 3$, so the
\end{itemize}
    particular solution is $y = 3e^{x}$.
\begin{itemize}
    \item $y'' + y' - 2y = 0$, $y(0) = 2$, $y'(0) = -1$. So,
\end{itemize}
    $2 = Ce^{0} + De^{0} = C + D$ and $-1 = Ce^{0} - 2De^{0} = C - 2D$.
    Thus C = 1 and D = 1, and the particular solution is
    $y = e^{x} + e^{-2x}$.

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Initial_value_problem}{Initial value problem,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license"

%%===============================================================
\section{Cauchy Problem}
\label{sec:cauchy-problem}
%%===============================================================

"A \textbf{Cauchy problem} in mathematics asks for the solution of a partial
differential equation that satisfies certain conditions that are given
on a hypersurface in the domain. A Cauchy problem can be an initial
value problem or a boundary value problem. It is named after
Augustin-Louis Cauchy.

\paragraph{Formal statement}


For a partial differential equation defined on \textbf{R}^\textit{n+1}^ and a smooth
manifold \textit{S} ? \textbf{R}^\textit{n+1}^ of dimension \textit{n} (\textit{S} is called the Cauchy
surface), the Cauchy problem consists of finding the unknown functions
$u_1,\dots,u_N$ of the differential equation with respect to the
independent variables $t,x_1,\dots,x_n$ that satisfies

$$\begin{aligned}&\frac{\partial^{n_i}u_i}{\partial t^{n_i}} = F_i\left(t,x_1,\dots,x_n,u_1,\dots,u_N,\dots,\frac{\partial^k u_j}{\partial t^{k_0}\partial x_1^{k_1}\dots\partial x_n^{k_n}},\dots\right) \\\&\text{for } i,j = 1,2,\dots,N;\, k_0+k_1+\dots+k_n=k\leq n_j;\, k_0<n_j
\end{aligned}$$ subject to the condition, for some value $t=t_0$,

$$\frac{\partial^k u_i}{\partial t^k}=\phi_i^{(k)}(x_1,\dots,x_n)
\quad \text{for } k=0,1,2,\dots,n_i-1$$

where $\phi_i^{(k)}(x_1,\dots,x_n)$ are given functions defined on the
surface $S$ (collectively known as the \textbf{Cauchy data} of the problem).
The derivative of order zero means that the function itself is
specified.

\paragraph{Cauchy--Kowalevski theorem}


The Cauchy--Kowalevski theorem states that *If all the functions $F_i$
are analytic in some neighborhood of the point
$(t^0,x_1^0,x_2^0,\dots,\phi_{j,k_0,k_1,\dots,k_n}^0,\dots)$, and if all
the functions $\phi_j^{(k)}$ are analytic in some neighborhood of the
point $(x_1^0,x_2^0,\dots,x_n^0)$, then the Cauchy problem has a unique
analytic solution in some neighborhood of the point
$(t^0,x_1^0,x_2^0,\dots,x_n^0)$*.

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Cauchy_problem}{Cauchy problem,
\end{itemize}
    Wikipedia} under a CC
    BY-SA license"

%%===============================================================
\section{Separation of Variables (1st Order)}
\label{sec:separation-of-variables-1st-order}
%%===============================================================

"In mathematics, \textbf{separation of variables} (also known as the **Fourier
method**) is any of several methods for solving ordinary and partial
differential equations, in which algebra allows one to rewrite an
equation so that each of two variables occurs on a different side of the
equation.

% [Image: Solve proportional first order differential equation by separation of
variables.]

% [Image: Solve linear first order differential equation by separation of
variables.]

\paragraph{Ordinary differential equations (ODE)}


Suppose a differential equation can be written in the form

$$\frac{d}{dx} f(x) = g(x)h(f(x))$$

which we can write more simply by letting $y = f(x)$:

$$\frac{dy}{dx}=g(x)h(y).$$

As long as \textit{h}(\textit{y}) ? 0, we can rearrange terms to obtain:

$${dy \over h(y)} = g(x) \, dx,$$

so that the two variables \textit{x} and \textit{y} have been separated. \textit{dx} (and
\textit{dy}) can be viewed, at a simple level, as just a convenient notation,
which provides a handy mnemonic aid for assisting with manipulations. A
formal definition of \textit{dx} as a differential (infinitesimal) is somewhat
advanced.

\paragraph{Alternative notation}


Those who dislike Leibniz\'s notation may prefer to write this as

$$\frac{1}{h(y)} \frac{dy}{dx} = g(x),$$

but that fails to make it quite as obvious why this is called
\""separation of variables\"". Integrating both sides of the equation with
respect to $x$, we have

$$\int \frac{1}{h(y)} \frac{dy}{dx} \, dx = \int g(x) \, dx,$$ \textbf{(A1)}

:   or equivalently,

$$\int \frac{1}{h(y)} \, dy = \int g(x) \, dx$$

because of the substitution rule for integrals.

If one can evaluate the two integrals, one can find a solution to the
differential equation. Observe that this process effectively allows us
to treat the derivative $\frac{dy}{dx}$ as a fraction which can be
separated. This allows us to solve separable differential equations more
conveniently, as demonstrated in the example below.

(Note that we do not need to use two constants of integration, in
equation (A1) as in

$$\int \frac{1}{h(y)} \, dy + C_1 = \int g(x) \, dx + C_2,$$ because a
single constant $C = C_2 - C_1$ is equivalent.)

\paragraph{Example}


Population growth is often modeled by the differential equation

:   $\frac{dP}{dt}=kP\left(1-\frac{P}{K}\right)$

where $P$ is the population with respect to time $t$, $k$ is the rate of
growth, and $K$ is the carrying capacity of the environment.

Separation of variables may be used to solve this differential equation.

:   $\begin{aligned}
    \& \frac{dP}{dt}=kP\left(1-\frac{P}{K}\right) \\\\[5pt]
    \& \int\frac{dP}{P\left(1-\frac{P}{K}\right)}=\int k\,dt
    \end{aligned}$

To evaluate the integral on the left side, we simplify the fraction

$$\begin{aligned} \frac{1}{P\left(1-\frac{P}{K}\right)}=\frac{K}{P\left(K-P\right)}
\hspace{45em} \end{aligned}$$

and then, we decompose the fraction into partial fractions

:   $\frac{K}{P(K-P)}=\frac{1}{P}+\frac{1}{K-P}$

Thus we have

:   $\begin{aligned}
    \& \int\left(\frac{1}{P}+\frac{1}{K-P}\right) dP=\int k\,dt \\\\[6pt]
    \& \ln|P|-\ln|K-P|=kt+C \\\\[6pt]
    \& \ln|K-P|-\ln|P|=-kt-C \\\\[6pt]
    \& \ln\left|\cfrac{K-P}{P}\right|=-kt-C \\\\[6pt]
    \& \left|\dfrac{K-P}{P}\right|=e^{-kt-C} \\\\[6pt]
    \& \left|\dfrac{K-P}{P}\right|=e^{-C}e^{-kt} \\\\[6pt]
    \& \frac{K-P}{P}=\pm e^{-C}e^{-kt}
    \end{aligned}$

Let $A=\pm e^{-C}$.

$$\begin{aligned}
& \frac{K-P}{P}=Ae^{-kt} \\\\[6pt]
& \frac{K}{P}-1=Ae^{-kt} \\\\[6pt]
& \frac{K}{P}=1+Ae^{-kt} \\\\[6pt]
& \frac{P}{K}=\frac{1}{1+Ae^{-kt}} \\\\[6pt]
& P=\frac{K}{1+Ae^{-kt}}
\end{aligned}$$

Therefore, the solution to the logistic equation is

:   $P(t)=\frac{K}{1+Ae^{-kt}}$

To find $A$, let $t=0$ and $P\left(0\right)=P_0$. Then we have

:   $P_0=\frac{K}{1+Ae^0}$

Noting that $e^0=1$, and solving for \textit{A} we get

:   $A=\frac{K-P_0}{P_0}.$

\paragraph{Generalization of separable ODEs to the nth order}


Much like one can speak of a separable first-order ODE, one can speak of
a separable second-order, third-order or \textit{n}th-order ODE. Consider the
separable first-order ODE:

$$\frac{dy}{dx}=f(y)g(x)$$ The derivative can alternatively be written
the following way to underscore that it is an operator working on the
unknown function, \textit{y}:

$$\frac{dy}{dx}=\frac{d}{dx}(y)$$ Thus, when one separates variables for
first-order equations, one in fact moves the \textit{dx} denominator of the
operator to the side with the \textit{x} variable, and the \textit{d}(\textit{y}) is left on
the side with the \textit{y} variable. The second-derivative operator, by
analogy, breaks down as follows:

$$\frac{d^2y}{dx^2} = \frac{d}{dx}\left(\frac{dy}{dx}\right) = \frac{d}{dx}\left(\frac{d}{dx}(y)\right)$$
The third-, fourth- and \textit{n}th-derivative operators break down in the
same way. Thus, much like a first-order separable ODE is reducible to
the form

$$\frac{dy}{dx}=f(y)g(x)$$ a separable second-order ODE is reducible to
the form

$$\frac{d^2y}{dx^2}=f\left(y'\right)g(x)$$ and an nth-order separable
ODE is reducible to

$$\frac{d^ny}{dx^n}=f\!\left(y^{(n-1)}\right)g(x)$$

\paragraph{Example}


Consider the simple nonlinear second-order differential
equation$$y''=(y')^2.$$This equation is an equation only of \textit{y\'\'} and
\textit{y\'}, meaning it is reducible to the general form described above and
is, therefore, separable. Since it is a second-order separable equation,
collect all \textit{x} variables on one side and all \textit{y\'} variables on the
other to get$$\frac{d(y')}{(y')^2}=dx.$$Now, integrate the right side
with respect to \textit{x} and the left with respect to
\textit{y}\'$$\int \frac{d(y')}{(y')^2}=\int dx.$$This
gives$-\frac{1}{y'}=x+C_1,$which simplifies
to$$y'=-\frac{1}{x+C_1}~.$$This is now a simple integral problem that
gives the final answer$$y=C_2-\ln|x+C_1|.$$

\paragraph{Partial differential equations}


The method of separation of variables is also used to solve a wide range
of linear partial differential equations with boundary and initial
conditions, such as the heat equation, wave equation, Laplace equation,
Helmholtz equation and biharmonic equation.

The analytical method of separation of variables for solving partial
differential equations has also been generalized into a computational
method of decomposition in invariant structures that can be used to
solve systems of partial differential equations.

\paragraph{Example: homogeneous case}


Consider the one-dimensional heat equation. The equation is

$$\frac{\partial u}{\partial t} - \alpha\frac{\partial^{2}u}{\partial x^{2}} = 0$$
\textbf{(1)}

The variable \textit{u} denotes temperature. The boundary condition is
homogeneous, that is

$$u\big|_{x=0}=u\big|_{x=L}=0$$ \textbf{(2)}

Let us attempt to find a solution which is not identically zero
satisfying the boundary conditions but with the following property: \textit{u}
is a product in which the dependence of \textit{u} on \textit{x}, \textit{t} is separated,
that is:

$$u(x,t) = X(x) T(t).$$ \textbf{(3)}

Substituting \textit{u} back into equation \textbf{(1)} and using the product rule,

$$\frac{T'(t)}{\alpha T(t)} = \frac{X''(x)}{X(x)}.$$ \textbf{(4)}

Since the right hand side depends only on \textit{x} and the left hand side
only on \textit{t}, both sides are equal to some constant value -\textit{?}. Thus:

$$T'(t) = - \lambda \alpha T(t),$$ \textbf{(5)}

and

$$X''(x) = - \lambda X(x).$$ \textbf{(6)}

-\textit{?} here is the eigenvalue for both differential operators, and
\textit{T}(\textit{t}) and \textit{X}(\textit{x}) are corresponding eigenfunctions.

We will now show that solutions for \textit{X}(\textit{x}) for values of \textit{?} = 0
cannot occur:

Suppose that \textit{?} \< 0. Then there exist real numbers \textit{B}, \textit{C} such that

$$X(x) = B e^{\sqrt{-\lambda} \, x} + C e^{-\sqrt{-\lambda} \, x}.$$

From \textbf{(2)} we get

$$X(0) = 0 = X(L),$$ \textbf{(7)}

and therefore \textit{B} = 0 = \textit{C} which implies \textit{u} is identically 0.

Suppose that \textit{?} = 0. Then there exist real numbers \textit{B}, \textit{C} such that

$$X(x) = Bx + C.$$

From \textbf{(7)} we conclude in the same manner as in 1 that \textit{u} is
identically 0.

Therefore, it must be the case that \textit{?} \> 0. Then there exist real
numbers \textit{A}, \textit{B}, \textit{C} such that

$$T(t) = A e^{-\lambda \alpha t},$$ and

$$X(x) = B \sin(\sqrt{\lambda} \, x) + C \cos(\sqrt{\lambda} \, x).$$

From \textbf{(7)} we get \textit{C} = 0 and that for some positive integer \textit{n},

$$\sqrt{\lambda} = n \frac{\pi}{L}.$$

This solves the heat equation in the special case that the dependence of
\textit{u} has the special form of \textbf{(3)}.

In general, the sum of solutions to \textbf{(1)} which satisfy the boundary
conditions \textbf{(2)} also satisfies \textbf{(1)} and \textbf{(3)}. Hence a complete
solution can be given as

$$u(x,t) = \sum_{n = 1}^{\infty} D_n \sin \frac{n\pi x}{L} \exp\left(-\frac{n^2 \pi^2 \alpha t}{L^2}\right),$$

where \textit{D}\textasciitilde{}\textit{n}\textasciitilde{} are coefficients determined by initial condition.

Given the initial condition

$$u\big|_{t=0}=f(x),$$

we can get

$$f(x) = \sum_{n = 1}^{\infty} D_n \sin \frac{n\pi x}{L}.$$

This is the sine series expansion of \textit{f}(\textit{x}). Multiplying both sides
with $\sin \frac{n\pi x}{L}$ and integrating over \[0, \textit{L}\] result in

$$D_n = \frac{2}{L} \int_0^L f(x) \sin \frac{n\pi x}{L} \, dx.$$

This method requires that the eigenfunctions \textit{X}, here
$\left\{\sin \frac{n\pi x}{L}\right\}_{n=1}^{\infty}$, are orthogonal
and complete. In general this is guaranteed by Sturm-Liouville theory.

\paragraph{Example: nonhomogeneous case}


Suppose the equation is nonhomogeneous,

$$\frac{\partial u}{\partial t}-\alpha\frac{\partial^{2}u}{\partial x^{2}}=h(x,t)$$
\textbf{(8)}

with the boundary condition the same as \textbf{(2)}.

Expand \textit{h}(\textit{x,t}), \textit{u}(\textit{x},\textit{t}) and \textit{f}(\textit{x}) into

$$h(x,t)=\sum_{n=1}^{\infty}h_{n}(t)\sin\frac{n\pi x}{L},$$ \textbf{(9)}

$$u(x,t)=\sum_{n=1}^{\infty}u_{n}(t)\sin\frac{n\pi x}{L},$$ \textbf{(10)}

:   $f(x)=\sum_{n=1}^{\infty}b_{n}\sin\frac{n\pi x}{L},$ \textbf{(11)}

where \textit{h}\textasciitilde{}\textit{n}\textasciitilde{}(\textit{t}) and \textit{b}\textasciitilde{}\textit{n}\textasciitilde{} can be calculated by integration, while
\textit{u}\textasciitilde{}\textit{n}\textasciitilde{}(\textit{t}) is to be determined.

Substitute \textbf{(9)} and \textbf{(10)} back to \textbf{(8)} and considering the
orthogonality of sine functions we get

:   $u'_{n}(t)+\alpha\frac{n^{2}\pi^{2}}{L^{2}}u_{n}(t)=h_{n}(t),$

which are a sequence of linear differential equations that can be
readily solved with, for instance, Laplace transform, or Integrating
factor. Finally, we can get

:   $u_{n}(t)=e^{-\alpha\frac{n^{2}\pi^{2}}{L^{2}} t} \left (b_{n}+\int_{0}^{t}h_{n}(s)e^{\alpha\frac{n^{2}\pi^{2}}{L^{2}} s} \, ds \right).$

If the boundary condition is nonhomogeneous, then the expansion of
\textbf{(9)} and \textbf{(10)} is no longer valid. One has to find a function \textit{v}
that satisfies the boundary condition only, and subtract it from \textit{u}.
The function \textit{u-v} then satisfies homogeneous boundary condition, and
can be solved with the above method.

\paragraph{Example: mixed derivatives}


For some equations involving mixed derivatives, the equation does not
separate as easily as the heat equation did in the first example above,
but nonetheless separation of variables may still be applied. Consider
the two-dimensional biharmonic equation

$$\frac{\partial^4 u}{\partial x^4} + 2\frac{\partial^4 u}{\partial x^2\partial y^2} + \frac{\partial^4 u}{\partial y^4} = 0.$$

Proceeding in the usual manner, we look for solutions of the form

$$u(x,y) = X(x)Y(y)$$

and we obtain the equation

$$\frac{X^{(4)}(x)}{X(x)} + 2\frac{X''(x)}{X(x)}\frac{Y''(y)}{Y(y)} + \frac{Y^{(4)}(y)}{Y(y)} = 0.$$

Writing this equation in the form

$$E(x) + F(x)G(y) + H(y) = 0,$$

we see that the derivative with respect to \textit{x} and \textit{y} eliminates the
first and last terms, so that

$$F'(x)G'(y) = 0,$$

i.e. either \textit{F}(\textit{x}) or \textit{G}(\textit{y}) must be a constant, say -?. This
further implies that either $-E(x)=F(x)G(y)+H(y)$ or
$-H(y)=E(x)+F(x)G(y)$ are constant. Returning to the equation for \textit{X}
and \textit{Y}, we have two cases

$$\begin{aligned}
    X''(x) &= -\lambda_1X(x) \\\    X^{(4)}(x) &= \mu_1X(x) \\\    Y^{(4)}(y) - 2\lambda_1Y''(y) &= -\mu_1Y(y)
\end{aligned}$$

and

$$\begin{aligned}
    Y''(y) &= -\lambda_2Y(y) \\\    Y^{(4)}(y) &= \mu_2Y(y) \\\    X^{(4)}(x) - 2\lambda_2X''(x) &= -\mu_2X(x)
\end{aligned}$$

which can each be solved by considering the separate cases for
$\lambda_i0$ and noting that
$\mu_i=\lambda_i^2$.

\paragraph{Curvilinear coordinates}


In orthogonal curvilinear coordinates, separation of variables can still
be used, but in some details different from that in Cartesian
coordinates. For instance, regularity or periodic condition may
determine the eigenvalues in place of boundary conditions. See spherical
harmonics for example.

\paragraph{Applicability}


\textbf{Partial differential equations}

For many PDEs, such as the wave equation, Helmholtz equation and
Schrodinger equation, the applicability of separation of variables is a
result of the spectral theorem. In some cases, separation of variables
may not be possible. Separation of variables may be possible in some
coordinate systems but not others, and which coordinate systems allow
for separation depends on the symmetry properties of the equation. Below
is an outline of an argument demonstrating the applicability of the
method to certain linear equations, although the precise method may
differ in individual cases (for instance in the biharmonic equation
above).

Consider an initial boundary value problem for a function $u(x,t)$ on
$D = \{(x,t): x \in [0,l], t \geq 0 \}$ in two variables:

$$(Tu)(x,t) = (Su)(x,t)$$

where $T$ is a differential operator with respect to $x$ and $S$ is a
differential operator with respect to $t$ with boundary data:

$$(Tu)(0,t) = (Tu)(a) = 0$$ for $t \geq 0$

$$(Su)(x,0)=h(x)$$ for $0 \leq x \leq a$

where $h$ is a known function.

We look for solutions of the form $u(x,t) = f(x) g(t)$. Dividing the PDE
through by $f(x)g(t)$ gives

$$\frac{Tf}{f} = \frac{Sg}{g}$$

The right hand side depends only on $x$ and the left hand side only on
$t$ so both must be equal to a constant $K$, which gives two ordinary
differential equations

$$Tf = Kf, Sg = Kg$$

which we can recognize as eigenvalue problems for the operators for $T$
and $S$. If $T$ is a compact, self-adjoint operator on the space
$L^2[0,l]$ along with the relevant boundary conditions, then by the
Spectral theorem there exists a basis for $L^2[0,l]$ consisting of
eigenfunctions for $T$. Let the spectrum of $T$ be $E$ and let
$f_{\lambda}$ be an eigenfunction with eigenvalue $\lambda \in E$. Then
for any function which at each time $t$ is square-integrable with
respect to $x$, we can write this function as a linear combination of
the $f_{\lambda}$. In particular, we know the solution $u$ can be
written as

$$u(x,t) = \sum_{\lambda \in E} c_{\lambda}(t)f_{\lambda}(x)$$

For some functions $c_{\lambda}(t)$. In the separation of variables,
these functions are given by solutions to $Sg = Kg$

Hence, the spectral theorem ensures that the separation of variables
will (when it is possible) find all the solutions.

For many differential operators, such as $\frac{d^2}{dx^2}$, we can show
that they are self-adjoint by integration by parts. While these
operators may not be compact, their inverses (when they exist) may be,
as in the case of the wave equation, and these inverses have the same
eigenfunctions and eigenvalues as the original operator (with the
possible exception of zero).

\paragraph{Matrices}


The matrix form of the separation of variables is the Kronecker sum.

As an example we consider the 2D discrete Laplacian on a regular grid:

$$L = \mathbf{D_{xx}}\oplus\mathbf{D_{yy}}=\mathbf{D_{xx}}\otimes\mathbf{I}+\mathbf{I}\otimes\mathbf{D_{yy}}, \,$$

where $\mathbf{D_{xx}}$ and $\mathbf{D_{yy}}$ are 1D discrete Laplacians
in the \textit{x}- and \textit{y}-directions, correspondingly, and $\mathbf{I}$ are
the identities of appropriate sizes. See the main article Kronecker sum
of discrete Laplacians for details.

\paragraph{Software}


Some mathematical programs are able to do separation of variables: Xcas
among others.

\paragraph{Resources}


\begin{itemize}
    \item \href{https://tutorial.math.lamar.edu/Classes/DE/Separable.aspx}{Separation of Variables
\end{itemize}
    Notes}.
    Produced by Paul Dawkins, Lamar University

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Separation_of_variables}{Separation of variables,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license"

%%===============================================================
\section{Homogeneous Differential Equations (1st Order)}
\label{sec:homogeneous-differential-equations-1st-order}
%%===============================================================

"A differential equation can be \textbf{homogeneous} in either of two
respects.

A first order differential equation is said to be homogeneous if it may
be written

$$f(x,y) \, dy = g(x,y) \, dx,$$ where `{{mvar|f}}`{=mediawiki} and
`{{mvar|g}}`{=mediawiki} are homogeneous functions of the same degree of
`{{mvar|x}}`{=mediawiki} and `{{mvar|y}}`{=mediawiki}. In this case, the
change of variable `{{math|1=''y'' = ''ux''}}`{=mediawiki} leads to an
equation of the form

$$\frac{dx}{x} = h(u) \, du,$$ which is easy to solve by integration of
the two members.

Otherwise, a differential equation is homogeneous if it is a homogeneous
function of the unknown function and its derivatives. In the case of
linear differential equations, this means that there are no constant
terms. The solutions of any linear ordinary differential equation of any
order may be deduced by integration from the solution of the homogeneous
equation obtained by removing the constant term.

\paragraph{Homogeneous first-order differential equations}


A first-order ordinary differential equation in the form:

$$M(x,y)\,dx + N(x,y)\,dy = 0$$

is a homogeneous type if both functions
`{{math|''M''(''x'', ''y'')}}`{=mediawiki} and
`{{math|''N''(''x'', ''y'')}}`{=mediawiki} are homogeneous functions of
the same degree `{{mvar|n}}`{=mediawiki}. That is, multiplying each
variable by a parameter `{{math|?}}`{=mediawiki}, we find

$$M(\lambda x, \lambda y) = \lambda^n M(x,y) \quad \text{and} \quad N(\lambda x, \lambda y) = \lambda^n N(x,y)\,.$$

Thus,

$$\frac{M(\lambda x, \lambda y)}{N(\lambda x, \lambda y)} = \frac{M(x,y)}{N(x,y)}\,.$$

\paragraph{Solution method}


In the quotient $\frac{M(tx,ty)}{N(tx,ty)} = \frac{M(x,y)}{N(x,y)}$, we
can let $t = \frac{1}{x}$ to simplify this quotient to a function
`{{mvar|f}}`{=mediawiki} of the single variable $\frac{y}{x}$:

$$\frac{M(x,y)}{N(x,y)} = \frac{M(tx,ty)}{N(tx,ty)} = \frac{M(1,y/x)}{N(1,y/x)}=f(y/x)\,.$$
That is

$$\frac{dy}{dx} = -f(y/x).$$

Introduce the change of variables
`{{math|1=''y'' = ''ux''}}`{=mediawiki}; differentiate using the product
rule:

$$\frac{dy}{dx}=\frac{d(ux)}{dx} = x\frac{du}{dx} + u\frac{dx}{dx} = x\frac{du}{dx} + u.$$

This transforms the original differential equation into the separable
form

:   $x\frac{du}{dx} = -f(u) - u,$

or

:   $\frac 1x\frac{dx}{du} = \frac {-1}{f(u) + u},$

which can now be integrated directly: `{{math|ln ''x''}}`{=mediawiki}
equals the antiderivative of the right-hand side (see ordinary
differential equation).

\paragraph{Special case}


A first order differential equation of the form
(`{{mvar|a}}`{=mediawiki}, `{{mvar|b}}`{=mediawiki},
`{{mvar|c}}`{=mediawiki}, `{{mvar|e}}`{=mediawiki},
`{{mvar|f}}`{=mediawiki}, `{{mvar|g}}`{=mediawiki} are all constants)

$$\left(ax + by + c\right) dx + \left(ex + fy + g\right) dy = 0$$ where
`{{math|''af'' ? ''be''}}`{=mediawiki} can be transformed into a
homogeneous type by a linear transformation of both variables
(`{{mvar|a}}`{=mediawiki} and `{{mvar|ß}}`{=mediawiki} are constants):

$$t = x + \alpha; \; \; z = y + \beta \,.$$

\paragraph{Homogeneous linear differential equations}


A linear differential equation is \textbf{homogeneous} if it is a homogeneous
linear equation in the unknown function and its derivatives. It follows
that, if `{{math|''f''(''x'')}}`{=mediawiki} is a solution, so is
`{{math|''cf''(''x'')}}`{=mediawiki}, for any (non-zero) constant
`{{mvar|c}}`{=mediawiki}. In order for this condition to hold, each
nonzero term of the linear differential equation must depend on the
unknown function or any derivative of it. A linear differential equation
that fails this condition is called \textbf{inhomogeneous.}

A linear differential equation can be represented as a linear operator
acting on `{{math|''y''(''x'')}}`{=mediawiki} where
`{{mvar|x}}`{=mediawiki} is usually the independent variable and
`{{mvar|y}}`{=mediawiki} is the dependent variable. Therefore, the
general form of a linear homogeneous differential equation is

$$L(y) = 0$$

where `{{mvar|L}}`{=mediawiki} is a differential operator, a sum of
derivatives (defining the \""0th derivative\"" as the original,
non-differentiated function), each multiplied by a function
`{{math|''f''''i''}}`{=mediawiki} of
`{{mvar|x}}`{=mediawiki}:

$$L = \sum_{i=0}^n f_i(x)\frac{d^i}{dx^i} \, ,$$ where
`{{math|''f''''i''}}`{=mediawiki} may be constants, but not
all `{{math|''f''''i''}}`{=mediawiki} may be zero.

For example, the following linear differential equation is homogeneous:

$$\sin(x) \frac{d^2y}{dx^2} + 4 \frac{dy}{dx} + y = 0 \,,$$

whereas the following two are inhomogeneous:

$$2 x^2 \frac{d^2y}{dx^2} + 4 x \frac{dy}{dx} + y = \cos(x) \,;$$

$$2 x^2 \frac{d^2y}{dx^2} - 3 x \frac{dy}{dx} + y = 2 \,.$$ The
existence of a constant term is a sufficient condition for an equation
to be inhomogeneous, as in the above example.

\paragraph{Second order equations}


In order to show how these equations are solved, lets start with the
most basic case - a second order equation

$$A\frac{d^2y}{dx^2}+B\frac{dy}{dx}+Cy=0$$

where A, B, and C are constants.

\paragraph{The Auxiliary Quadratic}


For a DE

$$a\frac{d^2y}{dx^2}+b\frac{dy}{dx}+cy=0,$$

Make the following substitution:

$$y=e^{mx}\,$$

This gives also

$$\frac{dy}{dx}=me^{mx}\,$$

$$\frac{d^2y}{dx^2}=m^2e^{mx}\,$$

The DE is now

$$am^2e^{mx}+bme^{mx}+ce^{mx}=0 \,$$

Dividing by $e^{mx}$ gives (note$$e^{mx}$$ can never equal zero)

$$am^2+bm+c=0 \,$$

This is the \textbf{auxiliary quadratic} (AQ) of the DE. There are four
classes of outcomes to the auxiliary quadratic:

1.  $b^2-4ac > 0, \,$, giving two distinct, real, roots.
2.  $b^2-4ac = 0, \,$, giving two coincident real, roots.
3.  $b^2-4ac < 0, \,$, giving complex roots.

:   \textit{a}: Purely imaginary roots.
:   \textit{b}: Complex-conjugate pair.

The method of solution of the DE depends on the class of AQ.

\paragraph{Class 1: Distinct, Real Roots}


Consider the DE

$$\frac{d^2y}{dx^2}+\frac{dy}{dx}-6y=0$$

The AQ is

$$m^2+m-6=0 \,$$

$$(m+3)(m-2)=0 \,$$

This gives us the following roots:

$$m=2, \ m=-3$$

Going back to the substitution we made to obtain the AQ, we have

$$y=e^{2x},\ y=e^{-3x}$$

as two distinct solutions to the DE. According to the superposition
principle, the general solution is therefore

$$y=Ae^{2x}+Be^{-3x} \,$$

\paragraph{General Solution to Class 1 DEs}


:   Generalizing, for the Second Order DE

$$a\frac{d^2y}{dx^2}+b\frac{dy}{dx}+cy=0$$

:   with the auxiliary quadratic given by

$$am^2+bm+c=0 \,$$

:   with roots \textit{a} and \textit{ß}, the general solution is

$$y=Ae^{\alpha x}+Be^{\beta x} \,$$

\paragraph{Class 2: Coincident, Real Roots}


Consider the DE

$$\frac{d^2y}{dx^2}-4\frac{dy}{dx}+4y=0$$

The AQ is

$$m^2-4m+4=0 \,$$

$$(m-2)(m-2)=0 \,$$

so, $e^{2x}\,$ is a solution. However, we cannot have it as both
solutions as the factor of two produced will be absorbed into the
constant, leaving us with only one constant, and therefore a DE without
a full solution.

For the other solution we will use the Method of Reduction of Order. To
do so we assume that it is in the form of:

$$y_2=u(x)y_1=u(x)e^{2x}\,$$ At the end we will check if our assumption
is correct. We will now substitute this equation and solve for u(x)

$$y_2''-4y_2'+4y_2=0\,$$

$$u''(x)e^{2x}+2u'(x)e^{2x}+2u'(x)e^{2x}+4u(x)e^{2x}-4(u'(x)e^{2x}+2u(x)e^{2x})+4u(x)e^{2x}=0\,$$

$$u''(x)e^{2x}=0\,$$

$$e^{2x}\,$$ is always non-zero so the only way the product can equal
zero is if:

$$u''(x)=0\,$$ Integrating twice offers

$$u(x)=a_1x + a_2\,$$ Therefore

$$y_2=(a_1x+a_2)y_1\,$$

$$y_2=(a_1x+a_2)e^{2x}\,$$ Our general solution is:

$$y=C_1y_1+C_2y_2\,$$

$$y=C_1e^{2x}+C_2(a_1x+a_2)e^{2x}\,$$

$$y=(C_1+a_2)e^{2x}+C_2a_1xe^{2x}\,$$ Because each constant is arbitrary
we can simply write

$$y=C_1e^{2x}+C_2xe^{2x}\,$$

The Method of Reduction of Order can be used on different equations and
u(x) does not always equal x. You can see below that $y=xe^{x}$ is a
valid solution.

$$y=xe^{2x}\,$$

$$\frac{dy}{dx}=2xe^{2x}+e^{2x}=e^{2x}(2x+1)\,$$

$$\frac{d^2y}{dx^2}=2e^{2x}+2e^{2x}(2x+1)=e^{2x}(4x+4)\,$$

To check substitute these into the original DE:

$$e^{2x}(4x+4)-4e^{2x}(2x+1)+4xe^{2x}=0 \,$$

$$4xe^{2x}+4e^{2x}-8xe^{2x}-4e^{2x}+4xe^{2x}=0 \,$$

$$0=0\,$$

Therefore, $y=xe^{2x}\,$ is a solution as well.

\paragraph{General Solution to Class 2 DEs}


:   Generalizing, for the Second Order DE

$$a\frac{d^2y}{dx^2}+b\frac{dy}{dx}+cy=0$$

:   with the coincident root \textit{a} of the AQ, the general solution is

$$y=(A+Bx)e^{\alpha x} \,$$

\paragraph{Class 3a: Purely Imaginary Roots}


To have complex roots, the AQ must have a discriminant less than zero,
so

$$b^2-4ac < 0 \,$$

Also, for the solution to be purely imaginary, the value of \textit{b} must be
exactly zero.

Therefore,

$$4ac>0\,$$

This means that \textit{a} and \textit{c} have to have the same signs: either \textit{a} and
\textit{c} are both positive or they are both negative. If we consider our
general second-order DE:

$$a\frac{d^2y}{dx^2}+b\frac{dy}{dx}+cy=0$$

Setting \textit{b} to zero gives

$$a\frac{d^2y}{dx^2}+cy=0$$

Dividing through by \textit{a} gives

$$\frac{d^2y}{dx^2}+\frac{c}{a}y=0$$.

Therefore, the \textit{y} term is always positive, and this can be represented
by

$$\frac{d^2y}{dx^2}+\omega^2y=0$$.

(I\'m using \textit{?} here as it is used for simple harmonic motion, which is
the primary use of this DE). There are now two paths to the solution of
the DE. The first relies on us spotting that we can use the cyclical
nature of trig. functions when derived. Substitute the following

$$y=\cos \omega x \,$$

$$\frac{dy}{dx}=-\omega \sin \omega x\,$$

$$\frac{d^2y}{dx^2}=-\omega^2 \cos \omega x\,$$

And check in our DE:

$$\frac{d^2y}{dx^2}+\omega^2 y=-\omega^2 \cos \omega x+\omega^2 \cos \omega x=0$$

This checks out, so $\cos \omega x\,$ is a solution. A similar result
holds true for the substitution using $y=\sin \omega x \,$.

Our solutions are therefore

$$y=\cos \omega x \,$$

$$y=\sin \omega x \,$$

So the general solution is

$$y=A\cos \omega x+B\sin \omega x. \,$$

The other method of solving this equation is to use Euler\'s Formula:

$$e^{\omega ix}=\cos \omega x +i \sin x \,$$

:   and

$$e^{-\omega ix}=\cos \omega x -i \sin x \,$$

From our original DE, we have an AQ of

$$m^2+\omega^2=0 \,$$

giving us roots of

$$m=\pm i \omega \,$$

so the general solution, similar to the Class 1 DEs, is

$$y=Ae^{\omega ix}+Be^{-\omega ix} \,$$

$$y=A(\cos \omega x +i \sin \omega x)+B(\cos \omega x -i \sin \omega x)\,$$

$$y=(A+B)\cos \omega x+i(A-B)\sin \omega x \,$$

Since A and B are arbitrary, we can set new constants for convenience,
letting our new A equal A+B and our new B equal \textit{i}(A-B).

Thus we have as our general solution

$$y=A\cos \omega x+B\sin \omega x \,$$

\paragraph{General Solution to Class 3a DEs}


:   Generalizing, for the Second Order DE

$$\frac{d^2y}{dx^2}+\omega^2 y=0$$

:   the general solution is

$$y=A\cos \omega x+B\sin \omega x \,$$

\paragraph{Class 3b: Complex Conjugate Roots}


Since it is a proven theorem that complex roots of polynomials always
occur in conjugate pairs, the only remaining class of AQ is the one with
complex conjugates for solutions.

Given that the solutions are complex, we know that in the AQ

$$am^2+bm+c=0 \,$$

$$b^2-4ac,<0,\ b \ne 0 \,$$ (see Class 3a).

The roots of this are in the form

$$(p \pm iq)\,$$

The general solution is then

$$y=Ae^{(p+iq)x}+Be^{(p-iq)x}\,$$

$$y=e^{px} \left( Ae^{iqx}+Be^{-iqx}\right)\,$$

From Euler\'s Formulas, we can now get

$$y=e^{px}\left( A\left(\cos q x + i \sin q x \right)+B\left(\cos q x - i \sin q x \right) \right)\,$$

$$y=e^{px}\left( (A+B)\cos q x+i(A-B)\sin q x \right)\,$$

As A and B are arbitrary, we can collapse them as in Class 3a, so that
we have the general solution

$$y=e^{px}\left( A\cos q x+B\sin q x \right)\,$$

\paragraph{General Solution to Class 3b DEs}


:   Generalizing, for the Second Order DE

$$a\frac{d^2y}{dx^2}+b\frac{dy}{dx}+cy=0$$

:   with an AQ with roots

$$m=p \pm iq,$$

:   The general solution is

$$y=e^{px}\left( A\cos q x+B\sin q x \right)\,$$

We have now covered all possible types of homogeneous second-order
differential equation, and we didn\'t even have to integrate anything!
We will now have a look at higher order equivalents.

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Homogeneous_differential_equation}{Homogeneous differential equation,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license
\begin{itemize}
    \item \href{https://en.wikibooks.org/wiki/Ordinary_Differential_Equations/Homogenous_1}{Homogeneous 1, Wikibooks: Ordinary Differential
\end{itemize}
    Equations}
    under a CC BY-SA license"

%%===============================================================
\section{Linear Differential Equations (1st Order)}
\label{sec:linear-differential-equations-1st-order}
%%===============================================================

"In mathematics, a \textbf{linear differential equation} is a differential
equation that is defined by a linear polynomial in the unknown function
and its derivatives, that is an equation of the form

$$a_0(x)y + a_1(x)y' + a_2(x)y'' \cdots + a_n(x)y^{(n)} = b(x)$$ where
\textit{a}\textasciitilde{}0\textasciitilde{}(\textit{x}), ..., \textit{a}\textasciitilde{}\textit{n}\textasciitilde{}(\textit{x}) and `{{math|''b''(''x'')}}`{=mediawiki}
are arbitrary differentiable functions that do not need to be linear,
and `{{math|''y''', …, ''y''(''n'') }}`{=mediawiki} are the
successive derivatives of an unknown function `{{mvar|y}}`{=mediawiki}
of the variable `{{mvar|x}}`{=mediawiki}.

Such an equation is an ordinary differential equation (ODE). A *linear
differential equation* may also be a linear partial differential
equation (PDE), if the unknown function depends on several variables,
and the derivatives that appear in the equation are partial derivatives.

A linear differential equation or a system of linear equations such that
the associated homogeneous equations have constant coefficients may be
solved by quadrature, which means that the solutions may be expressed in
terms of integrals. This is also true for a linear equation of order
one, with non-constant coefficients. An equation of order two or higher
with non-constant coefficients cannot, in general, be solved by
quadrature. For order two, Kovacic\'s algorithm allows deciding whether
there are solutions in terms of integrals, and computing them if any.

The solutions of linear differential equations with polynomial
coefficients are called holonomic functions. This class of functions is
stable under sums, products, differentiation, integration, and contains
many usual functions and special functions such as exponential function,
logarithm, sine, cosine, inverse trigonometric functions, error
function, Bessel functions and hypergeometric functions. Their
representation by the defining differential equation and initial
conditions allows making algorithmic (on these functions) most
operations of calculus, such as computation of antiderivatives, limits,
asymptotic expansion, and numerical evaluation to any precision, with a
certified error bound.

\paragraph{Basic terminology}


The highest order of derivation that appears in a (linear) differential
equation is the \textit{order} of the equation. The term
`{{math|''b''(''x'')}}`{=mediawiki}, which does not depend on the
unknown function and its derivatives, is sometimes called the *constant
term* of the equation (by analogy with algebraic equations), even when
this term is a non-constant function. If the constant term is the zero
function, then the differential equation is said to be \textit{homogeneous}, as
it is a homogeneous polynomial in the unknown function and its
derivatives. The equation obtained by replacing, in a linear
differential equation, the constant term by the zero function is the
\textit{associated homogeneous equation}. A differential equation has *constant
coefficients* if only constant functions appear as coefficients in the
associated homogeneous equation.

A \textit{solution} of a differential equation is a function that satisfies the
equation. The solutions of a homogeneous linear differential equation
form a vector space. In the ordinary case, this vector space has a
finite dimension, equal to the order of the equation. All solutions of a
linear differential equation are found by adding to a particular
solution any solution of the associated homogeneous equation.

\paragraph{Linear differential operator}


A \textit{basic differential operator} of order `{{mvar|i}}`{=mediawiki} is a
mapping that maps any differentiable function to its
`{{mvar|i}}`{=mediawiki}th derivative, or, in the case of several
variables, to one of its partial derivatives of order
`{{mvar|i}}`{=mediawiki}. It is commonly denoted

$$\frac{d^i}{dx^i}$$ in the case of univariate functions, and

$$\frac{\partial^{i_1+\cdots +i_n}}{\partial x_1^{i_1}\cdots \partial x_n^{i_n}}$$
in the case of functions of `{{mvar|n}}`{=mediawiki} variables. The
basic differential operators include the derivative of order 0, which is
the identity mapping.

A \textbf{linear differential operator} (abbreviated, in this article, as
\textit{linear operator} or, simply, \textit{operator}) is a linear combination of
basic differential operators, with differentiable functions as
coefficients. In the univariate case, a linear operator has thus the
form

$$a_0(x)+a_1(x)\frac{d}{dx} + \cdots +a_n(x)\frac{d^n}{dx^n},$$ where
`{{math|''a''0(''x''), …, ''a''''n''(''x'')}}`{=mediawiki}
are differentiable functions, and the nonnegative integer
`{{mvar|n}}`{=mediawiki} is the \textit{order} of the operator (if
`{{math|''a''''n''(''x'')}}`{=mediawiki} is not the zero
function).

Let `{{mvar|L}}`{=mediawiki} be a linear differential operator. The
application of `{{mvar|L}}`{=mediawiki} to a function
`{{mvar|f}}`{=mediawiki} is usually denoted
`{{math|''Lf''}}`{=mediawiki} or `{{math|''Lf''(''X'')}}`{=mediawiki},
if one needs to specify the variable (this must not be confused with a
multiplication). A linear differential operator is a linear operator,
since it maps sums to sums and the product by a scalar to the product by
the same scalar.

As the sum of two linear operators is a linear operator, as well as the
product (on the left) of a linear operator by a differentiable function,
the linear differential operators form a vector space over the real
numbers or the complex numbers (depending on the nature of the functions
that are considered). They form also a free module over the ring of
differentiable functions.

The language of operators allows a compact writing for differentiable
equations: if

$$L=a_0(x)+a_1(x)\frac{d}{dx} + \cdots +a_n(x)\frac{d^n}{dx^n},$$ is a
linear differential operator, then the equation

$$a_0(x)y +a_1(x)y' + a_2(x)y'' +\cdots +a_n(x)y^{(n)}=b(x)$$ may be
rewritten

$$Ly=b(x).$$

There may be several variants to this notation; in particular the
variable of differentiation may appear explicitly or not in
`{{mvar|y}}`{=mediawiki} and the right-hand and of the equation, such as
`{{math|1=''Ly''(''x'') = ''b''(''x'')}}`{=mediawiki} or
`{{math|1=''Ly'' = ''b''}}`{=mediawiki}.

The \textit{kernel} of a linear differential operator is its kernel as a linear
mapping, that is the vector space of the solutions of the (homogeneous)
differential equation `{{math|1=''Ly'' = 0}}`{=mediawiki}.

In the case of an ordinary differential operator of order
`{{mvar|n}}`{=mediawiki}, Carathéodory\'s existence theorem implies
that, under very mild conditions, the kernel of `{{mvar|L}}`{=mediawiki}
is a vector space of dimension `{{mvar|n}}`{=mediawiki}, and that the
solutions of the equation
`{{math|1=''Ly''(''x'') = ''b''(''x'')}}`{=mediawiki} have the form

$$S_0(x) + c_1S_1(x) + \cdots +c_nS_n(x),$$ where
`{{math|''c''1, …, ''c''''n''}}`{=mediawiki} are
arbitrary numbers. Typically, the hypotheses of Carathéodory\'s theorem
are satisfied in an interval `{{mvar|I}}`{=mediawiki}, if the functions
`{{math|''b'', ''a''0, …, ''a''''n''}}`{=mediawiki}
are continuous in `{{mvar|I}}`{=mediawiki}, and there is a positive real
number `{{mvar|k}}`{=mediawiki} such that \|\textit{a}\textasciitilde{}\textit{n}\textasciitilde{}(\textit{x})\| \> \textit{k} for
every `{{mvar|x}}`{=mediawiki} in `{{mvar|I}}`{=mediawiki}.

\paragraph{Homogeneous equation with constant coefficients}


A homogeneous linear differential equation has \textit{constant coefficients}
if it has the form

$$a_0y + a_1y' + a_2y'' + \cdots + a_n y^{(n)} = 0$$ where
`{{math|''a''1, …, ''a''''n''}}`{=mediawiki} are
(real or complex) numbers. In other words, it has constant coefficients
if it is defined by a linear operator with constant coefficients.

The study of these differential equations with constant coefficients
dates back to Leonhard Euler, who introduced the exponential function
`{{math|''e''''x''}}`{=mediawiki}, which is the unique
solution of the equation `{{math|1=''f''' = ''f''}}`{=mediawiki} such
that `{{math|1=''f''(0) = 1}}`{=mediawiki}. It follows that the
`{{mvar|n}}`{=mediawiki}th derivative of
`{{math|''e''''cx'' }}`{=mediawiki} is
`{{math|''c''''n''''e''''cx''}}`{=mediawiki}, and
this allows solving homogeneous linear differential equations rather
easily.

Let

$$a_0y + a_1y' + a_2y'' + \cdots + a_ny^{(n)} = 0$$ be a homogeneous
linear differential equation with constant coefficients (that is
`{{math|''a''0, …, ''a''''n''}}`{=mediawiki} are
real or complex numbers).

Searching solutions of this equation that have the form
`{{math|''e''''ax''}}`{=mediawiki} is equivalent to searching
the constants `{{mvar|a}}`{=mediawiki} such that

$$a_0e^{\alpha x} + a_1\alpha e^{\alpha x} + a_2\alpha^2 e^{\alpha x}+\cdots + a_n\alpha^n e^{\alpha x} = 0.$$
Factoring out `{{math|''e''''ax''}}`{=mediawiki} (which is
never zero), shows that `{{mvar|a}}`{=mediawiki} must be a root of the
\textit{characteristic polynomial}

$$a_0 + a_1t + a_2t^2 + \cdots + a_nt^n$$ of the differential equation,
which is the left-hand side of the characteristic equation

$$a_0 + a_1t + a_2t^2 + \cdots + a_nt^n = 0.$$

When these roots are all distinct, one has `{{mvar|n}}`{=mediawiki}
distinct solutions that are not necessarily real, even if the
coefficients of the equation are real. These solutions can be shown to
be linearly independent, by considering the Vandermonde determinant of
the values of these solutions at
`{{math|1=''x'' = 0, …, ''n'' – 1}}`{=mediawiki}. Together they form a
basis of the vector space of solutions of the differential equation
(that is, the kernel of the differential operator).

+----------------------------------------------------------------------+
| Example                                                              |
+======================================================================+
| $$y''''-2y'''+2y''-2y'+y=0$$ has the characteristic equation         |
|                                                                      |
| :   $z^4-2z^3+2z^2-2z+1=0.$                                          |
|                                                                      |
| This has zeros, `{{mvar|i}}`{=mediawiki},                            |
| `{{math|-''i''}}`{=mediawiki}, and `{{math|1}}`{=mediawiki}          |
| (multiplicity 2). The solution basis is thus                         |
|                                                                      |
| :   $e^{ix}, \; e^{-ix}, \; e^x, \; xe^x.$                              |
|                                                                      |
| A real basis of solution is thus                                     |
|                                                                      |
| :   $\cos x, \; \sin x, \; e^x, \; xe^x.$                               |
+----------------------------------------------------------------------+

In the case where the characteristic polynomial has only simple roots,
the preceding provides a complete basis of the solutions vector space.
In the case of multiple roots, more linearly independent solutions are
needed for having a basis. These have the form

$$x^ke^{\alpha x},$$ where `{{mvar|k}}`{=mediawiki} is a nonnegative
integer, `{{mvar|a}}`{=mediawiki} is a root of the characteristic
polynomial of multiplicity `{{mvar|m}}`{=mediawiki}, and
`{{math|''k'' < ''m''}}`{=mediawiki}. For proving that these functions
are solutions, one may remark that if `{{mvar|a}}`{=mediawiki} is a root
of the characteristic polynomial of multiplicity
`{{mvar|m}}`{=mediawiki}, the characteristic polynomial may be factored
as
`{{math|''P''(''t'') (''t'' - ''a'')''m''}}`{=mediawiki}.
Thus, applying the differential operator of the equation is equivalent
with applying first `{{mvar|m}}`{=mediawiki} times the operator
$\frac{d}{dx} - \alpha$, and then the operator that has
`{{mvar|P}}`{=mediawiki} as characteristic polynomial. By the
exponential shift theorem,

$$\left(\frac{d}{dx}-\alpha\right)\left(x^ke^{\alpha x}\right)= kx^{k-1}e^{\alpha x},$$

and thus one gets zero after `{{math|''k'' + 1}}`{=mediawiki}
application of $\frac{d}{dx} - \alpha$.

As, by the fundamental theorem of algebra, the sum of the multiplicities
of the roots of a polynomial equals the degree of the polynomial, the
number of above solutions equals the order of the differential equation,
and these solutions form a base of the vector space of the solutions.

In the common case where the coefficients of the equation are real, it
is generally more convenient to have a basis of the solutions consisting
of real-valued functions. Such a basis may be obtained from the
preceding basis by remarking that, if
`{{math|''a'' + ''ib''}}`{=mediawiki} is a root of the characteristic
polynomial, then `{{math|''a'' – ''ib''}}`{=mediawiki} is also a root,
of the same multiplicity. Thus a real basis is obtained by using
Euler\'s formula, and replacing
`{{math|''x''''k''''e''(''a''+''ib'')''x''}}`{=mediawiki}
and
`{{math|''x''''k''''e''(''a''-''ib'')''x''}}`{=mediawiki}
by
`{{math|''x''''k''''e''''ax'' cos(''bx'')}}`{=mediawiki}
and
`{{math|''x''''k''''e''''ax'' sin(''bx'')}}`{=mediawiki}.

\paragraph{Second-order case}


A homogeneous linear differential equation of the second order may be
written

$$y'' + ay' + by = 0,$$ and its characteristic polynomial is

$$r^2 + ar + b.$$

If `{{mvar|a}}`{=mediawiki} and `{{mvar|b}}`{=mediawiki} are real, there
are three cases for the solutions, depending on the discriminant
`{{math|1=''D'' = ''a''2 - 4''b''}}`{=mediawiki}. In all
three cases, the general solution depends on two arbitrary constants
`{{math|''c''1}}`{=mediawiki} and
`{{math|''c''2}}`{=mediawiki}.

\begin{itemize}
    \item If `{{math|''D'' > 0}}`{=mediawiki}, the characteristic polynomial
\end{itemize}
    has two distinct real roots `{{mvar|a}}`{=mediawiki}, and
    `{{mvar|ß}}`{=mediawiki}. In this case, the general solution is

$$c_1 e^{\alpha x} + c_2 e^{\beta x}.$$

\begin{itemize}
    \item If `{{math|1=''D'' = 0}}`{=mediawiki}, the characteristic polynomial
\end{itemize}
    has a double root `{{math|-''a''/2}}`{=mediawiki}, and the general
    solution is

$$(c_1 + c_2 x) e^{-ax/2}.$$

\begin{itemize}
    \item If `{{math|''D'' < 0}}`{=mediawiki}, the characteristic polynomial
\end{itemize}
    has two complex conjugate roots
    `{{math|''a'' ± ''ßi''}}`{=mediawiki}, and the general solution is

$$c_1 e^{(\alpha + \beta i)x} + c_2 e^{(\alpha - \beta i)x},$$

:   which may be rewritten in real terms, using Euler\'s formula as
    $$e^{\alpha x}  (c_1\cos(\beta x) + c_2 \sin(\beta x)).$$

Finding the solution `{{math|''y''(''x'')}}`{=mediawiki} satisfying
`{{math|1=''y''(0) = ''d''1}}`{=mediawiki} and
`{{math|1=''y'''(0) = ''d''2}}`{=mediawiki}, one equates the
values of the above general solution at `{{math|0}}`{=mediawiki} and its
derivative there to `{{math|''d''1}}`{=mediawiki} and
`{{math|''d''2}}`{=mediawiki}, respectively. This results in
a linear system of two linear equations in the two unknowns
`{{math|''c''1}}`{=mediawiki} and
`{{math|''c''2}}`{=mediawiki}. Solving this system gives the
solution for a so-called Cauchy problem, in which the values at
`{{math|0}}`{=mediawiki} for the solution of the DEQ and its derivative
are specified.

\paragraph{Non-homogeneous equation with constant coefficients}


A non-homogeneous equation of order `{{mvar|n}}`{=mediawiki} with
constant coefficients may be written

$$y^{(n)}(x) + a_1 y^{(n-1)}(x) + \cdots + a_{n-1} y'(x)+ a_ny(x) = f(x),$$
where `{{math|''a''1, …, ''a''''n''}}`{=mediawiki}
are real or complex numbers, `{{mvar|f}}`{=mediawiki} is a given
function of `{{mvar|x}}`{=mediawiki}, and `{{mvar|y}}`{=mediawiki} is
the unknown function (for sake of simplicity,
\""`{{math|(''x'')}}`{=mediawiki}\"" will be omitted in the following).

There are several methods for solving such an equation. The best method
depends on the nature of the function `{{mvar|f}}`{=mediawiki} that
makes the equation non-homogeneous. If `{{mvar|f}}`{=mediawiki} is a
linear combination of exponential and sinusoidal functions, then the
exponential response formula may be used. If, more generally,
`{{mvar|f}}`{=mediawiki} is a linear combination of functions of the
form `{{math|''x''''n''''e''''ax''}}`{=mediawiki},
`{{math|''x''''n'' cos(''ax'')}}`{=mediawiki}, and
`{{math|''x''''n'' sin(''ax'')}}`{=mediawiki}, where
`{{mvar|n}}`{=mediawiki} is a nonnegative integer, and
`{{mvar|a}}`{=mediawiki} a constant (which need not be the same in each
term), then the method of undetermined coefficients may be used. Still
more general, the annihilator method applies when
`{{mvar|f}}`{=mediawiki} satisfies a homogeneous linear differential
equation, typically, a holonomic function.

The most general method is the variation of constants, which is
presented here.

The general solution of the associated homogeneous equation

$$y^{(n)} + a_1 y^{(n-1)} + \cdots + a_{n-1} y'+ a_ny = 0$$ is

$$y=u_1y_1+\cdots+ u_ny_n,$$ where
`{{math|(''y''1, …, ''y''''n'')}}`{=mediawiki} is
a basis of the vector space of the solutions and
`{{math|''u''1, …, ''u''''n''}}`{=mediawiki} are
arbitrary constants. The method of variation of constants takes its name
from the following idea. Instead of considering
`{{math|''u''1, …, ''u''''n''}}`{=mediawiki} as
constants, they can considered as unknown functions that have to be
determined for making `{{mvar|y}}`{=mediawiki} a solution of the
non-homogeneous equation. For this purpose, one adds the constraints

$$\begin{aligned}
0 &= u'_1y_1 + u'_2y_2 + \cdots+u'_ny_n \\\0 &= u'_1y'_1 + u'_2y'_2 + \cdots + u'_n y'_n \\\  & \; \;\vdots \\\0 &= u'_1y^{(n-2)}_1+u'_2y^{(n-2)}_2 + \cdots + u'_n y^{(n-2)}_n,
\end{aligned}$$ which imply (by product rule and induction)

$$y^{(i)} = u_1 y_1^{(i)} + \cdots + u_n y_n^{(i)}$$ for
`{{math|1=''i'' = 1, …, ''n'' – 1}}`{=mediawiki}, and

$$y^{(n)} = u_1 y_1^{(n)} + \cdots + u_n y_n^{(n)} +u'_1y_1^{(n-1)}+u'_2y_2^{(n-1)}+\cdots+u'_ny_n^{(n-1)}.$$

Replacing in the original equation `{{mvar|y}}`{=mediawiki} and its
derivatives by these expressions, and using the fact that
`{{math|''y''1, …, ''y''''n''}}`{=mediawiki} are
solutions of the original homogeneous equation, one gets

$$f=u'_1y_1^{(n-1)} + \cdots + u'_ny_n^{(n-1)}.$$

This equation and the above ones with `{{math|0}}`{=mediawiki} as
left-hand side form a system of `{{mvar|n}}`{=mediawiki} linear
equations in
`{{math|''u'''1, …, ''u'''''n''}}`{=mediawiki}
whose coefficients are known functions (`{{mvar|f}}`{=mediawiki}, the
$y_i$, and their derivatives). This system can be solved by any method
of linear algebra. The computation of antiderivatives gives
`{{math|''u''1, …, ''u''''n''}}`{=mediawiki}, and
then
`{{math|1=''y'' = ''u''1''y''1 + ? + ''u''''n''''y''''n''}}`{=mediawiki}.

As antiderivatives are defined up to the addition of a constant, one
finds again that the general solution of the non-homogeneous equation is
the sum of an arbitrary solution and the general solution of the
associated homogeneous equation.

\paragraph{First-order linear equation}


The general form of a linear ordinary differential equation of order 1,
after dividing out the coefficient of
`{{math|''y'''(''x'')}}`{=mediawiki}, is:

$$y'(x) = f(x) y(x) + g(x).$$

If the equation is homogeneous, i.e.
`{{math|1=''g''(''x'') = 0}}`{=mediawiki}, one may rewrite and
integrate:

$$\frac{y'}{y}= f, \qquad \log y = k +F,$$ where
`{{mvar|k}}`{=mediawiki} is an arbitrary constant of integration and
$F=\textstyle\int f\,dx$ is any antiderivative of
`{{mvar|f}}`{=mediawiki}. Thus, the general solution of the homogeneous
equation is

$$y=ce^F,$$ where `{{math|1=''c'' = ''e''''k''}}`{=mediawiki}
is an arbitrary constant.

For the general non-homogeneous equation, one may multiply it by the
reciprocal `{{math|''e''-''F''}}`{=mediawiki} of a solution
of the homogeneous equation. This gives

$$y'e^{-F}-yfe^{-F}= ge^{-F}.$$ As
$-fe^{-F} = \frac{d}{dx} \left(e^{-F}\right),$ the product rule allows
rewriting the equation as

$$\frac{d}{dx}\left(ye^{-F}\right)= ge^{-F}.$$ Thus, the general
solution is

$$y=ce^F + e^F\int ge^{-F}dx,$$ where `{{mvar|c}}`{=mediawiki} is a
constant of integration, and `{{mvar|F}}`{=mediawiki} is any
antiderivative of `{{mvar|f}}`{=mediawiki} (changing of antiderivative
amounts to change the constant of integration).

\paragraph{Example}


Solving the equation

:   $y'(x) + \frac{y(x)}{x} = 3x.$

The associated homogeneous equation $y'(x) + \frac{y(x)}{x} = 0$ gives

$$\frac{y'}{y}=-\frac{1}{x},$$ that is

$$y=\frac{c}{x}.$$

Dividing the original equation by one of these solutions gives

:   $xy'+y=3x^2.$

That is

$$(xy)'=3x^2,$$

$$xy=x^3 +c,$$ and

$$y(x)=x^2+c/x.$$ For the initial condition

:   $y(1)=\alpha,$

one gets the particular solution

$$y(x)=x^2+\frac{\alpha-1}{x}.$$

\paragraph{Making Linear Equations from Non-Linear Equations}


Sometimes a non-linear equation, which is not solvable like this, can be
made linear, and more easily solvable, by applying a substitution.

\paragraph{Example}


$$y\frac{dy}{dx}+xy^2=5x$$

:   Let\'s make the following substitution:

$$v=y^2 \,$$

$$\frac{dv}{dx}=2y\frac{dy}{dx}$$

:   Plugging in, we get

$$\frac{1}{2}\frac{dv}{dx}+xv=5x$$

$$\frac{dv}{dx}+2xv=10x$$

:   We can then solve as a linear equation in \textit{v}, using the
    step-by-step method above:

<!-- -->

:   \textbf{Step 1}: Find the integrating factor:

$$e^{\int P(x)dx}$$

$$e^{\int 2x dx}=e^{x^2+C}$$

$$e^{\int P(x)dx}=Ce^{x^2}$$

:   Letting C=1 for convenience, we get $e^{x^2}$ as our integrating
    factor.

<!-- -->

:   \textbf{Step 2}: Multiply through

$$e^{x^2}\frac{dv}{dx}+e^{x^2}xv=10e^{x^2}x$$

:   \textbf{Step 3}: Recognize that the left hand is
    $\frac{d}{dx} e^{\int P(x)dx}v$

$$\frac{d}{dx} e^{x^2}v=10e^{x^2}x$$

:   \textbf{Step 4}: Integrate both sides w.r.t.\textit{x}.

$$\int (\frac{d}{dx} e^{x^2}v)dx=\int 10e^{x^2}xdx$$

$$e^{x^2}v=5e^{x^2}+C$$

:   \textbf{Step 5}: Solve for \textit{v}.

$$v=5+\frac{C}{e^{x^2}}$$

:   Now that we have \textit{v}, solve for \textit{y}.

$$v=y^2 \,$$

$$y^2=5+\frac{C}{e^{x^2}} \,$$

\paragraph{Higher order linear equations}


A linear ordinary equation of order one with variable coefficients may
be solved by quadrature, which means that the solutions may be expressed
in terms of integrals. This is not the case for order at least two. This
is the main result of Picard--Vessiot theory which was initiated by
Émile Picard and Ernest Vessiot, and whose recent developments are
called differential Galois theory.

The impossibility of solving by quadrature can be compared with the
Abel--Ruffini theorem, which states that an algebraic equation of degree
at least five cannot, in general, be solved by radicals. This analogy
extends to the proof methods and motivates the denomination of
differential Galois theory.

Similarly to the algebraic case, the theory allows deciding which
equations may be solved by quadrature, and if possible solving them.
However, for both theories, the necessary computations are extremely
difficult, even with the most powerful computers.

Nevertheless, the case of order two with rational coefficients has been
completely solved by Kovacic\'s algorithm.

\paragraph{Cauchy--Euler equation}


Cauchy--Euler equations are examples of equations of any order, with
variable coefficients, that can be solved explicitly. These are the
equations of the form

$$x^n y^{(n)}(x) + a_{n-1} x^{n-1} y^{(n-1)}(x) + \cdots + a_0 y(x) = 0,$$
where $a_0, \ldots, a_{n-1}$ are constant coefficients.

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Linear_differential_equation}{Linear differential equation,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license
\begin{itemize}
    \item \href{https://en.wikibooks.org/wiki/Ordinary_Differential_Equations/First_Order_Linear_1}{First Order Linear 1, Wikibooks: Differential
\end{itemize}
    Equations}
    under a CC BY-SA license"

%%===============================================================
\section{Integrating Factor}
\label{sec:integrating-factor}
%%===============================================================

The Method of Integrating Factors

Let ``{=html}$p$``{=html} and
``{=html}$g$``{=html} be functions of
``{=html}$t$``{=html} and consider the
following first order differential equation:



$\begin{aligned} \quad \frac{dy}{dt} + p(t) y = g(t) \end{aligned}$



If we multiply the both sides of the equation above by the function
``{=html}$\mu (t)$``{=html} we get
that:



$\begin{aligned} \quad \mu (t) \frac{dy}{dt} + \mu (t) p(t) y = \mu (t) g(t) \end{aligned}$



If we can guarantee that
``{=html}$\mu ' (t) = \mu (t) p(t)$``{=html},
then notice that by applying the product rule for differentiation that
we get:
``{=html}$\frac{d}{dt} \left ( \mu (t) y \right ) = \mu (t) \frac{dy}{dt} + \mu ' (t) y$``{=html}
and substituting
``{=html}$\mu ' (t) = \mu (t) y$``{=html}
and we get that
``{=html}$\frac{d}{dt} \left ( \mu (t) y \right ) = \mu (t) \frac{dy}{dt} + \mu (t) p(t) y$``{=html}
which is exactly the lefthand side of the equation above. Thus we get
that:



$\begin{aligned} \frac{d}{dt} \left ( \mu (t) y \right ) = \mu (t) g(t) \\\\ \end{aligned}$



The above differential equation can be solved by integrating both sides
of the equation with respect to
``{=html}$t$``{=html} and isolating
``{=html}$y$``{=html}. The question now
arises on how we can find such a function
``{=html}$\mu (t)$``{=html}.

> 
>
> ``{=html}Definition:``{=html} If
> ``{=html}$\frac{dy}{dt} + p(t) y = g(t)$``{=html}
> is a first order differential equation, then
> ``{=html}$\mu (t)$``{=html} is called
> an ``{=html}Integrating Factor``{=html} if for
> ``{=html}$\mu (t) \frac{dy}{dt} + \mu (t) p(t) y = \mu (t) g(t)$``{=html}
> we have that
> ``{=html}$\mu ' (t) = \mu (t) p(t)$``{=html}.
>
> 

The following proposition will give us a formula for obtaining the
integrating factor for differential equations in the form
``{=html}$\frac{dy}{dt} + p(t) y = g(t)$``{=html}.

> 
>
> ``{=html}Proposition 1:``{=html} If
> ``{=html}$\frac{dy}{dt} + p(t) y = g(t)$``{=html}
> is a differential equation, then an integrating factor
> ``{=html}$\mu (t)$``{=html} of this
> equation is given by the formula
> ``{=html}$\mu (t) = e^{\int p(t) \; dt}$``{=html}.
>
> 

\begin{itemize}
    \item ``{=html}Proof:``{=html} We want to find
\end{itemize}
    ``{=html}$\mu (t)$``{=html} such
    that
    ``{=html}$\mu ' (t) = \mu (t) p(t)$``{=html}.
    We can rewrite this equation as as
    ``{=html}$\frac{\mu '(t)}{\mu (t)} = p(t)$``{=html}
    and then:



$\begin{aligned} \quad \frac{ \mu '(t)}{ \mu (t)} = p(t) \\\\ \quad \frac{d}{dt} \ln \mid \mu (t) \mid = p(t) \\\\ \quad \int \frac{d}{dt} \ln \mid \mu (t) \mid \; dt = \int p(t) \; dt \\\\ \quad \ln \mid \mu (t) \mid = \int p(t) \; dt \\\\ \quad \mu (t) = \pm e^{\int p(t) \; dt} \end{aligned}$



\begin{itemize}
    \item Since we only need one integrating factor to solve differential
\end{itemize}
    equations in the form
    ``{=html}$\frac{dy}{dt} + p(t) y = g(t)$``{=html},
    we can more generally note that
    ``{=html}$\mu (t) = e^{\int p(t) \; dt}$``{=html}
    is an integrating factor of this differential equation.
    ``{=html}$\blacksquare$``{=html}

``{=html}Notice that from proposition 1 that integrating factors
``{=html}$\mu (t)$``{=html} are not
unique. In fact, there are infinitely many integrating factors. This can
be see when evaluating the indefinite integral in
``{=html}$\mu (t) = e^{\int p(t) \; dt}$``{=html}
which will result in getting
``{=html}$\mu (t) = e^{P(t) + C}$``{=html}
where ``{=html}$P$``{=html} is any
antiderivative if ``{=html}$p$``{=html}
and where ``{=html}$C$``{=html} is a
constant. We will always use the simplest integrating factor in solving
differential equations of this type.``{=html}

Let\'s now look at some examples of applying the method of integrating
factors.

\paragraph{Example 1}


``{=html}Find all solutions to the differential equation
``{=html}$\frac{dy}{dt} + \frac{2y}{t} = \frac{\sin t}{t^2}$``{=html}.``{=html}

We first notice that our differential equation is in the appropriate
form
``{=html}$\frac{dy}{dt} + p(t) y = g(t)$``{=html}
where
``{=html}$p(t) = \frac{2}{t}$``{=html}
and
``{=html}$g(t) = \frac{\sin t}{t^2}$``{=html}.
We compute our integrating factor as:



$\begin{aligned} \quad \mu (t) = e^{\int p(t) \; dt} = e^{\int \frac{2}{t} \; dt} = e^{2 \ln (t)} = e^{\ln (t^2)} = t^2 \end{aligned}$



Thus we have that for
``{=html}$C$``{=html} as a constant:



$\begin{aligned} \quad \mu (t) \frac{dy}{dt} + \mu (t) p(t) y = \mu (t) g(t) \\\\ \quad t^2 \frac{dy}{dt} + 2t y = \sin t \\\\ \quad \frac{d}{dt} \left ( t^2 y \right ) = \sin t \\\\ \quad \int \frac{d}{dt} \left ( t^2 y \right ) \; dt = \int \sin t \; dt \\\\ \quad t^2 y = -\cos t + C \\\\ \quad y = \frac{-\cos t}{t^2} + \frac{C}{t^2} \end{aligned}$



\paragraph{Example 2}


``{=html}Find all solutions to the differential equation
``{=html}$\frac{dy}{dt} - \frac{y}{t} + te^{-t} = 0$``{=html}.``{=html}

We first rewrite our differential equation as
``{=html}$\frac{dy}{dt} - \frac{y}{t} = - te^{-t}$``{=html}.
We note that in this form we have
``{=html}$p(t) = - \frac{1}{t}$``{=html}
and
``{=html}$g(t) = -t e^{-t}$``{=html}.
We now find an integrating factor:



$\begin{aligned} \quad \mu (t) = e^{\int p(t) \; dt} = e^{\int -\frac{1}{t} \; dt} = e^{-\ln t} = e^{\ln \left ( \frac{1}{t} \right)} = \frac{1}{t} \end{aligned}$



Thus we have that for
``{=html}$C$``{=html} as a constant:



$\begin{aligned} \quad \mu (t) \frac{dy}{dt} + \mu (t) p(t) y = \mu (t) g(t) \\\\ \quad \frac{1}{t} \frac{dy}{dt} - \frac{1}{t^2} y = -e^{-t} \\\\ \quad \frac{d}{dt} \left ( \frac{y}{t} \right ) = -e^{-t} \\\\ \quad \int \frac{d}{dt} \left ( \frac{y}{t} \right ) \; dt = -\int e^{-t} \; dt \\\\ \quad \frac{y}{t} = e^{-t} + C \\\\ \quad y = te^{-t} + tC \end{aligned}$



\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{http://mathonline.wikidot.com/the-method-of-integrating-factors}{The Method of Integrating Factors,
\end{itemize}
    mathonline.wikidot.com}
    under a CC BY-SA license"

%%===============================================================
\section{Bernoulli Equations (1st Order)}
\label{sec:bernoulli-equations-1st-order}
%%===============================================================

Bernoulli Equations (1st Order)

%%===============================================================
\section{Exact Differential Equations (1st Order)}
\label{sec:exact-differential-equations-1st-order}
%%===============================================================

Definition

Given a simply connected and open subset \textit{D} of \textbf{R}^2^ and two
functions \textit{I} and \textit{J} which are continuous on \textit{D}, an implicit
first-order ordinary differential equation of the form

:   $I(x, y)\, dx + J(x, y)\, dy = 0,$

is called an \textbf{exact differential equation} if there exists a
continuously differentiable function \textit{F}, called the **potential
function**, so that

$$\frac{\partial F}{\partial x} = I$$ and

$$\frac{\partial F}{\partial y} = J.$$

An exact equation may also be presented in the following form:

$$I(x, y) + J(x, y) \, y'(x) = 0$$ where the same constraints on \textit{I} and
\textit{J} apply for the differential equation to be exact.

The nomenclature of \""exact differential equation\"" refers to the exact
differential of a function. For a function
$F(x_0, x_1,...,x_{n-1},x_n)$, the exact or total derivative with
respect to $x_0$ is given by

$$\frac{dF}{dx_0}=\frac{\partial F}{\partial x_0}+\sum_{i=1}^{n}\frac{\partial F}{\partial x_i}\frac{dx_i}{dx_0}.$$

\paragraph{Example}


The function $F:\mathbb{R}^{2}\to\mathbb{R}$ given by

$$F(x,y) = \frac{1}{2}(x^2 + y^2)+c$$

is a potential function for the differential equation

$$x\,dx + y\,dy = 0.\,$$

\paragraph{Existence of potential functions}


In physical applications the functions \textit{I} and \textit{J} are usually not only
continuous but even continuously differentiable. Schwarz\'s Theorem then
provides us with a necessary criterion for the existence of a potential
function. For differential equations defined on simply connected sets
the criterion is even sufficient and we get the following theorem:

Given a differential equation of the form (for example, when \textit{F} has
zero slope in the x and y direction at \textit{F}(\textit{x},\textit{y})):

:   $I(x, y)\, dx + J(x, y)\, dy = 0,$

with \textit{I} and \textit{J} continuously differentiable on a simply connected and
open subset \textit{D} of \textbf{R}^2^ then a potential function \textit{F} exists if and
only if

$$\frac{\partial I}{\partial y}(x, y) = \frac{\partial J}{\partial x}(x, y).$$

\paragraph{Solutions to exact differential equations}


Given an exact differential equation defined on some simply connected
and open subset \textit{D} of \textbf{R}^2^ with potential function \textit{F}, a
differentiable function \textit{f} with (\textit{x}, \textit{f}(\textit{x})) in \textit{D} is a solution if
and only if there exists real number \textit{c} so that

$$F(x, f(x)) = c.$$

For an initial value problem

$$y(x_0) = y_0$$ we can locally find a potential function by

$$\begin{aligned}
F(x,y) &= \int_{x_0}^x I(t,y_0) dt + \int_{y_0}^y J(x,t) dt \\\&= \int_{x_0}^x I(t,y_0) dt + \int_{y_0}^y \left[ J(x_0,t) +  \int_{x_0}^{x} \frac{\partial I}{\partial y}(u, t)\, du \right]dt.
\end{aligned}$$

Solving

$$F(x,y) = c$$ for \textit{y}, where \textit{c} is a real number, we can then
construct all solutions.

\paragraph{Second order exact differential equations}


The concept of exact differential equations can be extended to second
order equations. Consider starting with the first-order exact equation:

$$I\left(x,y\right)+J\left(x,y\right){dy \over dx}=0$$

Since both functions $I\left(x,y\right)$, $J\left(x,y\right)$ are
functions of two variables, implicitly differentiating the multivariate
function yields

$${dI \over dx} +\left({ dJ\over dx}\right){dy \over dx}+{d^2y \over dx^2}\left(J\left(x,y\right)\right)=0$$

Expanding the total derivatives gives that

$${dI \over dx}={\partial I\over\partial x}+{\partial I\over\partial y}{dy \over dx}$$

and that

$${dJ \over dx}={\partial J\over\partial x}+{\partial J\over\partial y}{dy \over dx}$$

Combining the ${dy \over dx}$ terms gives

$${\partial I\over\partial x}+{dy \over dx}\left({\partial I\over\partial y}+{\partial J\over\partial x}+{\partial J\over\partial y}{dy \over dx}\right)+{d^2y \over dx^2}\left(J\left(x,y\right)\right)=0$$

If the equation is exact, then
${\partial J\over\partial x}={\partial I\over\partial y}$. Additionally,
the total derivative of $J\left(x,y\right)$ is equal to its implicit
ordinary derivative ${dJ \over dx}$. This leads to the rewritten
equation

$${\partial I\over\partial x}+{dy \over dx}\left({\partial J\over\partial x}+{dJ \over dx}\right)+{d^2y \over dx^2}\left(J\left(x,y\right)\right)=0$$

Now, let there be some second-order differential equation

$$f\left(x,y\right)+g\left(x,y,{dy \over dx}\right){dy \over dx}+{d^2y \over dx^2}\left(J\left(x,y\right)\right)=0$$

If ${\partial J\over\partial x}={\partial I\over\partial y}$ for exact
differential equations, then

$$\int \left({\partial I\over\partial y}\right)dy=\int \left({\partial J\over\partial x}\right)dy$$

and

$$\int \left({\partial I\over\partial y}\right)dy=\int \left({\partial J\over\partial x}\right)dy=I\left(x,y\right)-h\left(x\right)$$

where $h\left(x\right)$ is some arbitrary function only of $x$ that was
differentiated away to zero upon taking the partial derivative of
$I\left(x,y\right)$ with respect to $y$. Although the sign on
$h\left(x\right)$ could be positive, it is more intuitive to think of
the integral\'s result as $I\left(x,y\right)$ that is missing some
original extra function $h\left(x\right)$ that was partially
differentiated to zero.

Next, if

$${dI\over dx}={\partial I\over\partial x}+{\partial I\over\partial y}{dy \over dx}$$

then the term ${\partial I\over\partial x}$ should be a function only of
$x$ and $y$, since partial differentiation with respect to $x$ will hold
$y$ constant and not produce any derivatives of $y$. In the second order
equation

$$f\left(x,y\right)+g\left(x,y,{dy \over dx}\right){dy \over dx}+{d^2y \over dx^2}\left(J\left(x,y\right)\right)=0$$

only the term $f\left(x,y\right)$ is a term purely of $x$ and $y$. Let
${\partial I\over\partial x}=f\left(x,y\right)$. If
${\partial I\over\partial x}=f\left(x,y\right)$, then

$$f\left(x,y\right)={ dI\over dx}-{\partial I\over\partial y}{dy \over dx}$$

Since the total derivative of $I\left(x,y\right)$ with respect to $x$ is
equivalent to the implicit ordinary derivative ${dI \over dx}$ , then

$$f\left(x,y\right)+{\partial I\over\partial y}{dy \over dx}={dI \over dx}={d \over dx}\left(I\left(x,y\right)-h\left(x\right)\right)+{dh\left(x\right) \over dx}$$

So,

$${dh\left(x\right) \over dx}=f\left(x,y\right)+{\partial I\over\partial y}{dy \over dx}-{d \over dx}\left(I\left(x,y\right)-h\left(x\right)\right)$$

and

$$h\left(x\right)=\int\left(f\left(x,y\right)+{\partial I\over\partial y}{dy \over dx}-{d \over dx}\left(I\left(x,y\right)-h\left(x\right)\right)\right)dx$$

Thus, the second order differential equation

$$f\left(x,y\right)+g\left(x,y,{dy \over dx}\right){dy \over dx}+{d^2y \over dx^2}\left(J\left(x,y\right)\right)=0$$

is exact only if
$g\left(x,y,{dy \over dx}\right)={ dJ\over dx}+{\partial J\over\partial x}={dJ \over dx}+{\partial J\over\partial x}$
and only if the below expression

$$\int\left(f\left(x,y\right)+{\partial I\over\partial y}{dy \over dx}-{d \over dx}\left(I\left(x,y\right)-h\left(x\right)\right)\right)dx=\int \left(f\left(x,y\right)-{\partial \left(I\left(x,y\right)-h\left(x\right)\right)\over\partial x}\right)dx$$

is a function solely of $x$. Once $h\left(x\right)$ is calculated with
its arbitrary constant, it is added to
$I\left(x,y\right)-h\left(x\right)$ to make $I\left(x,y\right)$. If the
equation is exact, then we can reduce to the first order exact form
which is solvable by the usual method for first-order exact equations.

$$I\left(x,y\right)+J\left(x,y\right){dy \over dx}=0$$

Now, however, in the final implicit solution there will be a $C_1x$ term
from integration of $h\left(x\right)$ with respect to $x$ twice as well
as a $C_2$, two arbitrary constants as expected from a second-order
equation.

\paragraph{Example}


Given the differential equation

$$\left(1-x^2\right)y''-4xy'-2y=0$$

one can always easily check for exactness by examining the $y''$ term.
In this case, both the partial and total derivative of $1-x^2$ with
respect to $x$ are $-2x$, so their sum is $-4x$, which is exactly the
term in front of $y'$. With one of the conditions for exactness met, one
can calculate that

$$\int \left(-2x\right)dy=I\left(x,y\right)-h\left(x\right)=-2xy$$

Letting $f\left(x,y\right)=-2y$, then

$$\int \left(-2y-2xy'-{d \over dx}\left(-2xy \right)\right)dx=\int \left(-2y-2xy'+2xy'+2y\right)dx=\int \left(0\right)dx=h\left(x\right)$$

So, $h\left(x\right)$ is indeed a function only of $x$ and the second
order differential equation is exact. Therefore, $h\left(x\right)=C_1$
and $I\left(x,y\right)=-2xy+C_1$. Reduction to a first-order exact
equation yields

$$-2xy+C_1+\left(1-x^2\right)y'=0$$

Integrating $I\left(x,y\right)$ with respect to $x$ yields

$$-x^2y+C_1x+i\left(y\right)=0$$

where $i\left(y\right)$ is some arbitrary function of $y$.
Differentiating with respect to $y$ gives an equation correlating the
derivative and the $y'$ term.

$$-x^2+i'\left(y\right)=1-x^2$$

So, $i\left(y\right)=y+C_2$ and the full implicit solution becomes

$$C_1x+C_2+y-x^2y=0$$

Solving explicitly for $y$ yields

$$y= \frac{C_1x+C_2}{1-x^2}$$

\paragraph{Higher order exact differential equations}


The concepts of exact differential equations can be extended to any
order. Starting with the exact second order equation

$${d^2y \over dx^2}\left(J\left(x,y\right)\right)+{dy \over dx}\left({dJ \over dx}+{\partial J\over\partial x}\right)+f\left(x,y\right)=0$$

it was previously shown that equation is defined such that

$$f\left(x,y\right)={dh\left(x\right) \over dx}+{d \over dx}\left(I\left(x,y\right)-h\left(x\right)\right)-{\partial J\over\partial x}{dy \over dx}$$

Implicit differentiation of the exact second-order equation $n$ times
will yield an $\left(n+2\right)$th order differential equation with new
conditions for exactness that can be readily deduced from the form of
the equation produced. For example, differentiating the above
second-order differential equation once to yield a third-order exact
equation gives the following form

$${d^3y \over dx^3}\left(J\left(x,y\right)\right)+{d^2y \over dx^2}{dJ \over dx}+{d^2y \over dx^2}\left({dJ \over dx}+{\partial J\over\partial x}\right)+{dy \over dx}\left({d^2J \over dx^2}+{d \over dx}\left({\partial J\over\partial x}\right)\right)+{df\left(x,y\right) \over dx}=0$$

where

$${df\left(x,y\right) \over dx}={d^2h\left(x\right) \over dx^2}+{d^2 \over dx^2}\left(I\left(x,y\right)-h\left(x\right)\right)-{d^2y \over dx^2}{\partial J\over\partial x}-{dy \over dx}{d \over dx}\left({\partial J\over\partial x}\right)=F\left(x,y,{dy \over dx}\right)$$
and where $F\left(x,y,{dy \over dx}\right)$ is a function only of $x,y$
and ${dy \over dx}$. Combining all ${dy \over dx}$ and
${d^2y \over dx^2}$ terms not coming from
$F\left(x,y,{dy \over dx}\right)$ gives

$${d^3y \over dx^3}\left(J\left(x,y\right)\right)+{d^2y \over dx^2}\left(2{dJ \over dx}+{\partial J\over\partial x}\right)+{dy \over dx}\left({d^2J \over dx^2}+{d \over dx}\left({\partial J\over\partial x}\right)\right)+F\left(x,y,{dy \over dx}\right)=0$$

Thus, the three conditions for exactness for a third-order differential
equation are: the ${d^2y \over dx^2}$ term must be
$2{dJ \over dx}+{\partial J\over\partial x}$, the ${dy \over dx}$ term
must be
${d^2J \over dx^2}+{d \over dx}\left({\partial J\over\partial x}\right)$
and

$$F\left(x,y,{dy \over dx}\right)-{d^2 \over dx^2}\left(I\left(x,y\right)-h\left(x\right)\right)+{d^2y \over dx^2}{\partial J\over\partial x}+{dy \over dx}{d \over dx}\left({\partial J\over\partial x}\right)$$

must be a function solely of $x$.

\paragraph{Example}


Consider the nonlinear third-order differential equation

$$yy'''+3y'y''+12x^2=0$$

If $J\left(x,y\right)=y$, then
$y''\left(2{dJ \over dx}+{\partial J\over\partial x}\right)$ is $2y'y''$
and
$y'\left({d^2J \over dx^2}+{d \over dx}\left({\partial J\over\partial x}\right)\right)=y'y''$which
together sum to $3y'y''$. Fortunately, this appears in our equation. For
the last condition of exactness,

$$F\left(x,y,{dy \over dx}\right)-{d^2 \over dx^2}\left(I\left(x,y\right)-h\left(x\right)\right)+{d^2y \over dx^2}{\partial J\over\partial x}+{dy \over dx}{d \over dx}\left({\partial J\over\partial x}\right)=12x^2-0+0+0=12x^2$$

which is indeed a function only of $x$. So, the differential equation is
exact. Integrating twice yields that
$h\left(x\right)=x^4+C_1x+C_2=I\left(x,y\right)$. Rewriting the equation
as a first-order exact differential equation yields

$$x^4+C_1x+C_2+yy'=0$$

Integrating $I\left(x,y\right)$ with respect to $x$ gives that
${x^5\over 5}+C_1x^2+C_2x+i\left(y\right)=0$. Differentiating with
respect to $y$ and equating that to the term in front of $y'$ in the
first-order equation gives that $i'\left(y\right)=y$ and that
$i\left(y\right)={y^2\over 2}+C_3$. The full implicit solution becomes

$${x^5\over 5}+C_1x^2+C_2x+C_3+{y^2\over 2}=0$$

The explicit solution, then, is

$$y=\pm\sqrt{C_1x^2+C_2x+C_3-\frac{2x^5}{5}}$$

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Exact_differential_equation}{Exact differential equation,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license"

%%===============================================================
\section{Linear Independence of Functions}
\label{sec:linear-independence-of-functions}
%%===============================================================

"If we have an $n^{\mathrm{th}}$ order linear homogenous differential
equation
$\frac{d^ny}{dt^n} + p_1(t) \frac{d^{n-1}y}{dt^{n-1}} + ... + p_{n-1}(t) \frac{dy}{dt} + p_n(t)y = 0$
where $p_1$, $p_2$, ..., $p_n$ are continuous on an open interval $I$
and if $y = y_1(t)$, $y = y_2(t)$, ..., $y = y_n(t)$ are solutions to
this differential equation, then provided that
$W(y_1, y_2, ..., y_n) \neq 0$ for at least one point $t \in I$, then
$y_1$, $y_2$, ..., $y_n$ form a fundamental set of solutions to this
differential equation - that is, for constants $C_1$, $C_2$, ..., $C_n$,
then every solution to this differential equation can be written in the
form:

$\begin{aligned} \quad y = C_1y_1(t) + C_2y_2(t) + ... + C_ny_n(t) \end{aligned}$

We will now look at the connection between the solutions $y_1$, $y_2$,
..., $y_n$ forming a fundamental set of solutions and the linear
independence/dependence of such solutions. We first define linear
independence and linear dependence below.



``{=html}Definition:``{=html} The functions $f_1$,
$f_2$, ..., $f_n$ are said to be ``{=html}Linearly
Independent``{=html} on an interval $I$ if for constants $k_1$,
$k_2$, ..., $k_n$ we have that
$k_1f_1(t) + k_2f_2(t) + ... + k_nf_n(t) = 0$ implies that
$k_1 = k_2 = ... = k_n = 0$ for all $t \in I$. This set of functions is
said to be ``{=html}Linearly Dependent``{=html} if
$k_1f_1(t) + k_2f_2(t) + ... + k_nf_n(t) = 0$ where $k_1$, $k_2$, ...,
$k_n$ are not all zero for all $t \in I$.



Perhaps the simplest linearly independent sets of functions is that set
that contains $f_1(t) = 1$, $f_2(t) = t$, and $f_3(t) = t^2$. Let $k_1$,
$k_2$, and $k_3$ be constants and consider the following equation:

$\begin{aligned} \quad k_1f_1(t) + k_2f_2(t) + k_3f_3(t) = 0 \\\\ \quad k_1 + k_2t + k_3t^2 = 0 \end{aligned}$

It\'s not hard to see that equation above is satisfied if and only if
the constants $k_1 = k_2 = k_3 = 0$.

For another example, consider the functions $f_1(t) = \sin t$ and
$f_2(t) = \sin (t + \pi)$ defined on all of $\mathbb{R}$. This set of
functions is not linearly independent. To show this, let $k_1$ and $k_2$
be constants and consider the following equation:

$\begin{aligned} \quad k_1f_1(t) + k_2f_2(t) = 0 \\\\ \quad k_1 \sin t + k_2 \sin (t + \pi) = 0 \end{aligned}$

Now choose $t = \pi$. Then we have that:

$\begin{aligned} \quad k_1 \sin \pi + k_2 \sin 2\pi = 0 \\\\ \end{aligned}$

But the above equation is true for any choice of constants $k_1$ and
$k_2$ since $\sin \pi = \sin 2\pi = 0$, and thus $f_1$ and $f_2$ do not
form a linearly independent set on all of $\mathbb{R}$.

From the concept of linear independence/dependence, we obtain the
following theorem on fundamental sets of solutions for $n^{\mathrm{th}}$
order linear homogenous differential equations.



``{=html}Theorem 1:``{=html} Let
$\frac{d^ny}{dt^n} + p_1(t) \frac{d^{n-1}y}{dt^{n-1}} + ... + p_{n-1}(t) \frac{dy}{dt} + p_n(t)y = 0$
be an $n^{\mathrm{th}}$ order linear homogenous differential equation.
If $y = y_1(t)$, $y = y_2(t)$, ..., $y = y_n(t)$ are solutions to this
differential equation then $y_1$, $y_2$, ..., $y_n$ form a fundamental
set of solutions to this differential equation on the open interval $I$
if and only if $y_1$, $y_2$, ..., $y_n$ are linearly dependent on $I$.



\begin{itemize}
    \item ``{=html}Proof:``{=html} Consider the following
\end{itemize}
    $n^{\mathrm{th}}$ order linear homogenous differential equation:

$\begin{aligned} \quad \frac{d^ny}{dt^n} + p_1(t) \frac{d^{n-1}y}{dt^{n-1}} + ... + p_{n-1}(t) \frac{dy}{dt} + p_n(t)y = 0 \end{aligned}$

\begin{itemize}
    \item $\Rightarrow$ Suppose that $y = y_1(t)$, $y = y_2(t)$, ...,
\end{itemize}
    $y = y_n(t)$ form a fundamental set of solutions to this
    differential equation on the open interval $I$. Then this implies
    that for all $t_0 \in I$ we have that :

$\begin{aligned} \quad W(y_1, y_2, ..., y_n)  \bigg|_{t_0} \neq 0 \end{aligned}$

\begin{itemize}
    \item Thus this implies that following system of equations have only
\end{itemize}
    trivial solution $k_1 = k_2 = ... = k_n = 0$:

$\begin{aligned} \quad k_1y_1(t) + k_2y_2(t) + ... + k_ny_n(t) = 0 \\\\ \quad k_1y_1'(t) + k_2y_2'(t) + ... + k_ny_n'(t) = 0 \\\\ \quad \quad \quad \quad \quad \quad \vdots \quad \quad \quad \quad \quad \quad \\\\ \quad k_1y_1^{(n-2)}(t) + k_2y_2^{(n-2)}(t) + ... + k_ny_n^{(n-2)}(t) = 0 \\\\ \quad k_1y_1^{(n-1)}(t) + k_2y_2^{(n-1)}(t) + ... + k_ny_n^{(n-1)}(t) = 0 \end{aligned}$

\begin{itemize}
    \item Thus the equation $k_1y_1(t) + k_2y_2(t) + ... + k_ny_n(t) = 0$
\end{itemize}
    implies that $k_1 = k_2 = ... = k_n = 0$. Thus $y_1$, $y_2$, ...,
    $y_n$ are linearly independent on $I$.

<!-- -->

\begin{itemize}
    \item $\Leftarrow$ We will prove the converse of Theorem 1 by
\end{itemize}
    contradiction. Suppose that $y_1$, $y_2$, ..., $y_n$ are linearly
    independent on $I$, and assume that instead $y_1$, $y_2$, ..., $y_n$
    do NOT form a fundamental set of solutions on $I$. Then for some
    $t_0 \in I$, the Wronskian
    $W(y_1, y_2, ..., y_n)  \bigg|_{t_0} = 0$. Thus the system of
    equations above does not have only the trivial solution. Let the
    constants $k_1^*$, $k_2^*$, ..., $k_n^*$ be a nontrivial solution to
    this system. Define $\phi(t)$ as:

$\begin{aligned} \quad \phi(t) = k_1^* y_1(t) + k_2^* y_2(t) + ... + k_n^* y_n(t) \end{aligned}$

\begin{itemize}
    \item Note that $y = \phi(t)$ satisfies the initial conditions
\end{itemize}
    $y(t_0) = 0$, $y'(t_0) = 0$, ..., $y^{(n-1)} (t_0) = 0$, and
    $\phi(t)$ satisfies our $n^{\mathrm{th}}$ order linear homogenous
    differential equation because $\phi(t)$ is a linear combination of
    the solutions $y_1$, $y_2$, ..., $y_n$.

<!-- -->

\begin{itemize}
    \item Now note that the function $y = 0$ also satisfies the differential
\end{itemize}
    equation and the initial conditions. By the existence/uniqueness
    theorem for $n^{\mathrm{th}}$ order linear homogenous differential
    equations, this implies that $\phi(t) = 0$ for all $t \in I$, so:

$\begin{aligned} \quad 0 = k_1^* y_1(t) + k_2^* y_2(t) + ... + k_n^* y_n(t) \end{aligned}$

\begin{itemize}
    \item But $y_1$, $y_2$, ..., $y_n$ are linearly independent which implies
\end{itemize}
    that $k_1^* = k_2^* = ... = k_n^* = 0$. Thus $k_1^*$, $k_2^*$, ...,
    $k_n^*$ is a trivial solution to the system above, which is a
    contradiction. Therefore our assumption that $y_1$, $y_2$, ...,
    $y_n$ do not form a fundamental set of solutions was false.
    $\blacksquare$

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{http://mathonline.wikidot.com/linear-independence-dependence-of-a-set-of-functions}{Linear Independence Dependence of a Set of Functions,
\end{itemize}
    }
    under a CC BY-SA license"

%%===============================================================
\section{Linear Dependence of Functions}
\label{sec:linear-dependence-of-functions}
%%===============================================================

"If we have an $n^{\mathrm{th}}$ order linear homogenous differential
equation
$\frac{d^ny}{dt^n} + p_1(t) \frac{d^{n-1}y}{dt^{n-1}} + ... + p_{n-1}(t) \frac{dy}{dt} + p_n(t)y = 0$
where $p_1$, $p_2$, ..., $p_n$ are continuous on an open interval $I$
and if $y = y_1(t)$, $y = y_2(t)$, ..., $y = y_n(t)$ are solutions to
this differential equation, then provided that
$W(y_1, y_2, ..., y_n) \neq 0$ for at least one point $t \in I$, then
$y_1$, $y_2$, ..., $y_n$ form a fundamental set of solutions to this
differential equation - that is, for constants $C_1$, $C_2$, ..., $C_n$,
then every solution to this differential equation can be written in the
form:

$\begin{aligned} \quad y = C_1y_1(t) + C_2y_2(t) + ... + C_ny_n(t) \end{aligned}$

We will now look at the connection between the solutions $y_1$, $y_2$,
..., $y_n$ forming a fundamental set of solutions and the linear
independence/dependence of such solutions. We first define linear
independence and linear dependence below.



``{=html}Definition:``{=html} The functions $f_1$,
$f_2$, ..., $f_n$ are said to be ``{=html}Linearly
Independent``{=html} on an interval $I$ if for constants $k_1$,
$k_2$, ..., $k_n$ we have that
$k_1f_1(t) + k_2f_2(t) + ... + k_nf_n(t) = 0$ implies that
$k_1 = k_2 = ... = k_n = 0$ for all $t \in I$. This set of functions is
said to be ``{=html}Linearly Dependent``{=html} if
$k_1f_1(t) + k_2f_2(t) + ... + k_nf_n(t) = 0$ where $k_1$, $k_2$, ...,
$k_n$ are not all zero for all $t \in I$.



Perhaps the simplest linearly independent sets of functions is that set
that contains $f_1(t) = 1$, $f_2(t) = t$, and $f_3(t) = t^2$. Let $k_1$,
$k_2$, and $k_3$ be constants and consider the following equation:

$\begin{aligned} \quad k_1f_1(t) + k_2f_2(t) + k_3f_3(t) = 0 \\\\ \quad k_1 + k_2t + k_3t^2 = 0 \end{aligned}$

It\'s not hard to see that equation above is satisfied if and only if
the constants $k_1 = k_2 = k_3 = 0$.

For another example, consider the functions $f_1(t) = \sin t$ and
$f_2(t) = \sin (t + \pi)$ defined on all of $\mathbb{R}$. This set of
functions is not linearly independent. To show this, let $k_1$ and $k_2$
be constants and consider the following equation:

$\begin{aligned} \quad k_1f_1(t) + k_2f_2(t) = 0 \\\\ \quad k_1 \sin t + k_2 \sin (t + \pi) = 0 \end{aligned}$

Now choose $t = \pi$. Then we have that:

$\begin{aligned} \quad k_1 \sin \pi + k_2 \sin 2\pi = 0 \\\\ \end{aligned}$

But the above equation is true for any choice of constants $k_1$ and
$k_2$ since $\sin \pi = \sin 2\pi = 0$, and thus $f_1$ and $f_2$ do not
form a linearly independent set on all of $\mathbb{R}$.

From the concept of linear independence/dependence, we obtain the
following theorem on fundamental sets of solutions for $n^{\mathrm{th}}$
order linear homogenous differential equations.



``{=html}Theorem 1:``{=html} Let
$\frac{d^ny}{dt^n} + p_1(t) \frac{d^{n-1}y}{dt^{n-1}} + ... + p_{n-1}(t) \frac{dy}{dt} + p_n(t)y = 0$
be an $n^{\mathrm{th}}$ order linear homogenous differential equation.
If $y = y_1(t)$, $y = y_2(t)$, ..., $y = y_n(t)$ are solutions to this
differential equation then $y_1$, $y_2$, ..., $y_n$ form a fundamental
set of solutions to this differential equation on the open interval $I$
if and only if $y_1$, $y_2$, ..., $y_n$ are linearly dependent on $I$.



\begin{itemize}
    \item ``{=html}Proof:``{=html} Consider the following
\end{itemize}
    $n^{\mathrm{th}}$ order linear homogenous differential equation:

$\begin{aligned} \quad \frac{d^ny}{dt^n} + p_1(t) \frac{d^{n-1}y}{dt^{n-1}} + ... + p_{n-1}(t) \frac{dy}{dt} + p_n(t)y = 0 \end{aligned}$

\begin{itemize}
    \item $\Rightarrow$ Suppose that $y = y_1(t)$, $y = y_2(t)$, ...,
\end{itemize}
    $y = y_n(t)$ form a fundamental set of solutions to this
    differential equation on the open interval $I$. Then this implies
    that for all $t_0 \in I$ we have that :

$\begin{aligned} \quad W(y_1, y_2, ..., y_n)  \bigg|_{t_0} \neq 0 \end{aligned}$

\begin{itemize}
    \item Thus this implies that following system of equations have only
\end{itemize}
    trivial solution $k_1 = k_2 = ... = k_n = 0$:

$\begin{aligned} \quad k_1y_1(t) + k_2y_2(t) + ... + k_ny_n(t) = 0 \\\\ \quad k_1y_1'(t) + k_2y_2'(t) + ... + k_ny_n'(t) = 0 \\\\ \quad \quad \quad \quad \quad \quad \vdots \quad \quad \quad \quad \quad \quad \\\\ \quad k_1y_1^{(n-2)}(t) + k_2y_2^{(n-2)}(t) + ... + k_ny_n^{(n-2)}(t) = 0 \\\\ \quad k_1y_1^{(n-1)}(t) + k_2y_2^{(n-1)}(t) + ... + k_ny_n^{(n-1)}(t) = 0 \end{aligned}$

\begin{itemize}
    \item Thus the equation $k_1y_1(t) + k_2y_2(t) + ... + k_ny_n(t) = 0$
\end{itemize}
    implies that $k_1 = k_2 = ... = k_n = 0$. Thus $y_1$, $y_2$, ...,
    $y_n$ are linearly independent on $I$.

<!-- -->

\begin{itemize}
    \item $\Leftarrow$ We will prove the converse of Theorem 1 by
\end{itemize}
    contradiction. Suppose that $y_1$, $y_2$, ..., $y_n$ are linearly
    independent on $I$, and assume that instead $y_1$, $y_2$, ..., $y_n$
    do NOT form a fundamental set of solutions on $I$. Then for some
    $t_0 \in I$, the Wronskian
    $W(y_1, y_2, ..., y_n)  \bigg|_{t_0} = 0$. Thus the system of
    equations above does not have only the trivial solution. Let the
    constants $k_1^*$, $k_2^*$, ..., $k_n^*$ be a nontrivial solution to
    this system. Define $\phi(t)$ as:

$\begin{aligned} \quad \phi(t) = k_1^* y_1(t) + k_2^* y_2(t) + ... + k_n^* y_n(t) \end{aligned}$

\begin{itemize}
    \item Note that $y = \phi(t)$ satisfies the initial conditions
\end{itemize}
    $y(t_0) = 0$, $y'(t_0) = 0$, ..., $y^{(n-1)} (t_0) = 0$, and
    $\phi(t)$ satisfies our $n^{\mathrm{th}}$ order linear homogenous
    differential equation because $\phi(t)$ is a linear combination of
    the solutions $y_1$, $y_2$, ..., $y_n$.

<!-- -->

\begin{itemize}
    \item Now note that the function $y = 0$ also satisfies the differential
\end{itemize}
    equation and the initial conditions. By the existence/uniqueness
    theorem for $n^{\mathrm{th}}$ order linear homogenous differential
    equations, this implies that $\phi(t) = 0$ for all $t \in I$, so:

$\begin{aligned} \quad 0 = k_1^* y_1(t) + k_2^* y_2(t) + ... + k_n^* y_n(t) \end{aligned}$

\begin{itemize}
    \item But $y_1$, $y_2$, ..., $y_n$ are linearly independent which implies
\end{itemize}
    that $k_1^* = k_2^* = ... = k_n^* = 0$. Thus $k_1^*$, $k_2^*$, ...,
    $k_n^*$ is a trivial solution to the system above, which is a
    contradiction. Therefore our assumption that $y_1$, $y_2$, ...,
    $y_n$ do not form a fundamental set of solutions was false.
    $\blacksquare$

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{http://mathonline.wikidot.com/linear-independence-dependence-of-a-set-of-functions}{Linear Independence Dependence of a Set of Functions,
\end{itemize}
    }
    under a CC BY-SA license"

%%===============================================================
\section{Wronskian}
\label{sec:wronskian}
%%===============================================================

Wronskian

Let L(y)=0 be a homogeneous equation of degree n, and let
$y_1,y_2,...,y_n$ be linearly independent solutions of this equation.

The general solution is then $y=C_1y_1+C_2y_2+...+C_ny_n$.

Suppose instead that the solutions are not linearly independent. Then
there exists values for $C_1,C_2,...,C_n$ not all zero such that
$C_1y_1+C_2y_2+...+C_ny_n$=0.

Then the following equations also hold true:

$C_1y_1'+C_2y_2'+...+C_ny_n'=0$$C_1y_1''+C_2y_2''+...+C_ny_n''=0$\...$C_1y_1^{(n-1)}+C_2y_2^{(n-1)}+...+C_ny_n^{(n-1)}=0$

When there is to be a non-trivial solution to this system of homogeneous
linear equations, the column vectors corresponding to each coefficient
are linearly dependent. This is equivalent to saying that the following
determinant, called the \textbf{Wronskian} is 0:

$\begin{Vmatrix}
y_1 \& y_2 \& ... \& y_n \\\y_1' \& y_2' \& ... \& y_n' \\\y_1'' \& y_2'' \& ... \& y_n'' \\\... \& ... \& ... \& ... \\\y_1^{(n-1)} \& y_2^{(n-1)} \& ... \& y_n^{(n-1)} \\\\end{Vmatrix} = 0$

Thus, if the solutions are dependent, then their Wronskian is equal to
0, and conversely, and conversely if the Wronskian is equal to 0, then
the columns of that matrix must be linearly dependent, indicating that
the solutions are linearly dependent.

\paragraph{Resources}


\begin{itemize}
    \item \href{https://tutorial.math.lamar.edu/Classes/DE/Wronskian.aspx}{Notes on the
\end{itemize}
    Wronskian}.
    Produced by Paul Dawkins, Lamar University

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikibooks.org/wiki/Ordinary_Differential_Equations/Linear_Equations}{Linear Equations, Wikibooks: Ordinary Differential
\end{itemize}
    Equations}
    under a CC BY-SA license"

%%===============================================================
\section{Reduction of the Order}
\label{sec:reduction-of-the-order}
%%===============================================================

"\textbf{Reduction of order} is a technique in mathematics for solving
second-order linear ordinary differential equations. It is employed when
one solution $y_1(x)$ is known and a second linearly independent
solution $y_2(x)$ is desired. The method also applies to \textit{n}-th order
equations. In this case the ansatz will yield an (\textit{n}-1)-th order
equation for $v$.

\paragraph{Second-order linear ordinary differential equations}


\paragraph{An example}


Consider the general, homogeneous, second-order linear constant
coefficient ordinary differential equation. (ODE)

$$a y''(x) + b y'(x) + c y(x) = 0,$$

where $a, b, c$ are real non-zero coefficients. Two linearly independent
solutions for this ODE can be straightforwardly found using
characteristic equations except for the case when the discriminant,
$b^2 - 4 a c$, vanishes. In this case,

$$a y''(x) + b y'(x) + \frac{b^2}{4a} y(x) = 0,$$

from which only one solution,

$$y_1(x) = e^{-\frac{b}{2a} x},$$

can be found using its characteristic equation.

The method of reduction of order is used to obtain a second linearly
independent solution to this differential equation using our one known
solution. To find a second solution we take as a guess

$$y_2(x) = v(x) y_1(x)$$

where $v(x)$ is an unknown function to be determined. Since $y_2(x)$
must satisfy the original ODE, we substitute it back in to get

$$a \left( v'' y_1 + 2 v' y_1' + v y_1'' \right) + b \left( v' y_1 + v y_1' \right) + \frac{b^2}{4a} v y_1 = 0.$$

Rearranging this equation in terms of the derivatives of $v(x)$ we get

$$\left(a y_1 \right) v'' + \left( 2 a y_1' + b y_1 \right) v' + \left( a y_1'' + b y_1' + \frac{b^2}{4a} y_1 \right) v = 0.$$

Since we know that $y_1(x)$ is a solution to the original problem, the
coefficient of the last term is equal to zero. Furthermore, substituting
$y_1(x)$ into the second term\'s coefficient yields (for that
coefficient)

$$2 a \left( - \frac{b}{2a} e^{-\frac{b}{2a} x} \right) + b e^{-\frac{b}{2a} x} = \left( -b + b \right) e^{-\frac{b}{2a} x} = 0.$$

Therefore, we are left with

$$a y_1 v'' = 0.$$

Since $a$ is assumed non-zero and $y_1(x)$ is an exponential function
(and thus always non-zero), we have

$$v'' = 0.$$

This can be integrated twice to yield

$$v(x) = c_1 x + c_2$$

where $c_1, c_2$ are constants of integration. We now can write our
second solution as

$$y_2(x) = ( c_1 x + c_2 ) y_1(x) = c_1 x y_1(x) + c_2 y_1(x).$$

Since the second term in $y_2(x)$ is a scalar multiple of the first
solution (and thus linearly dependent) we can drop that term, yielding a
final solution of

$$y_2(x) = x y_1(x) = x e^{-\frac{b}{2 a} x}.$$

Finally, we can prove that the second solution $y_2(x)$ found via this
method is linearly independent of the first solution by calculating the
Wronskian

$$W(y_1,y_2)(x) = \begin{vmatrix} y_1 & x y_1 \\\\ y_1' & y_1 + x y_1' \end{vmatrix} = y_1 ( y_1 + x y_1' ) - x y_1 y_1' = y_1^{2} + x y_1 y_1' - x y_1 y_1' = y_1^{2} = e^{-\frac{b}{a}x} \neq 0.$$

Thus $y_2(x)$ is the second linearly independent solution we were
looking for.

\paragraph{General method}


Given the general non-homogeneous linear differential equation

$$y'' + p(t)y' + q(t)y = r(t)$$

and a single solution $y_1(t)$ of the homogeneous equation \[$r(t)=0$\],
let us try a solution of the full non-homogeneous equation in the form:

$$y_2 = v(t)y_1(t)$$

where $v(t)$ is an arbitrary function. Thus

$$y_2' = v'(t)y_1(t) + v(t)y_1'(t)$$

and

$$y_2'' = v''(t)y_1(t) + 2v'(t)y_1'(t) + v(t)y_1''(t).$$

If these are substituted for $y$, $y'$, and $y''$ in the differential
equation, then

$$y_1(t)\,v'' + (2y_1'(t)+p(t)y_1(t))\,v' + (y_1''(t)+p(t)y_1'(t)+q(t)y_1(t))\,v = r(t).$$

Since $y_1(t)$ is a solution of the original homogeneous differential
equation, $y_1''(t)+p(t)y_1'(t)+q(t)y_1(t)=0$, so we can reduce to

$$y_1(t)\,v'' + (2y_1'(t)+p(t)y_1(t))\,v' = r(t)$$

which is a first-order differential equation for $v'(t)$ (reduction of
order). Divide by $y_1(t)$, obtaining

$$v''+\left(\frac{2y_1'(t)}{y_1(t)}+p(t)\right)\,v'=\frac{r(t)}{y_1(t)}.$$

The integrating factor is
$\mu(t)=e^{\int(\frac{2y_1'(t)}{y_1(t)}+p(t))dt}=y_1^2(t)e^{\int p(t) dt}$.

Multiplying the differential equation by the integrating factor
$\mu(t)$, the equation for $v(t)$ can be reduced to

$$\frac{d}{dt}\left(v'(t) y_1^2(t) e^{\int p(t) dt}\right) = y_1(t)r(t)e^{\int p(t) dt}.$$

After integrating the last equation, $v'(t)$ is found, containing one
constant of integration. Then, integrate $v'(t)$ to find the full
solution of the original non-homogeneous second-order equation,
exhibiting two constants of integration as it should:

$$y_2(t) = v(t)y_1(t).$$

\paragraph{Resources}


\begin{itemize}
    \item \href{https://tutorial.math.lamar.edu/Classes/DE/ReductionofOrder.aspx}{Reduction of the Order
\end{itemize}
    Notes}.
    Produced by Paul Dawkins, Lamar University

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Reduction_of_order}{Reduction of order,
\end{itemize}
    Wikipedia} under a
    CC BY-SA license"

%%===============================================================
\section{Linear Homogeneous Equations}
\label{sec:linear-homogeneous-equations}
%%===============================================================

"A differential equation can be \textbf{homogeneous} in either of two
respects.

A first order differential equation is said to be homogeneous if it may
be written

$$f(x,y) \, dy = g(x,y) \, dx,$$ where `{{mvar|f}}`{=mediawiki} and
`{{mvar|g}}`{=mediawiki} are homogeneous functions of the same degree of
`{{mvar|x}}`{=mediawiki} and `{{mvar|y}}`{=mediawiki}. In this case, the
change of variable `{{math|1=''y'' = ''ux''}}`{=mediawiki} leads to an
equation of the form

$$\frac{dx}{x} = h(u) \, du,$$ which is easy to solve by integration of
the two members.

Otherwise, a differential equation is homogeneous if it is a homogeneous
function of the unknown function and its derivatives. In the case of
linear differential equations, this means that there are no constant
terms. The solutions of any linear ordinary differential equation of any
order may be deduced by integration from the solution of the homogeneous
equation obtained by removing the constant term.

\paragraph{Homogeneous first-order differential equations}


A first-order ordinary differential equation in the form:

$$M(x,y)\,dx + N(x,y)\,dy = 0$$

is a homogeneous type if both functions
`{{math|''M''(''x'', ''y'')}}`{=mediawiki} and
`{{math|''N''(''x'', ''y'')}}`{=mediawiki} are homogeneous functions of
the same degree `{{mvar|n}}`{=mediawiki}. That is, multiplying each
variable by a parameter `{{math|?}}`{=mediawiki}, we find

$$M(\lambda x, \lambda y) = \lambda^n M(x,y) \quad \text{and} \quad N(\lambda x, \lambda y) = \lambda^n N(x,y)\,.$$

Thus,

$$\frac{M(\lambda x, \lambda y)}{N(\lambda x, \lambda y)} = \frac{M(x,y)}{N(x,y)}\,.$$

\paragraph{Solution method}


In the quotient $\frac{M(tx,ty)}{N(tx,ty)} = \frac{M(x,y)}{N(x,y)}$, we
can let $t = \frac{1}{x}$ to simplify this quotient to a function
`{{mvar|f}}`{=mediawiki} of the single variable $\frac{y}{x}$:

$$\frac{M(x,y)}{N(x,y)} = \frac{M(tx,ty)}{N(tx,ty)} = \frac{M(1,y/x)}{N(1,y/x)}=f(y/x)\,.$$
That is

$$\frac{dy}{dx} = -f(y/x).$$

Introduce the change of variables
`{{math|1=''y'' = ''ux''}}`{=mediawiki}; differentiate using the product
rule:

$$\frac{dy}{dx}=\frac{d(ux)}{dx} = x\frac{du}{dx} + u\frac{dx}{dx} = x\frac{du}{dx} + u.$$

This transforms the original differential equation into the separable
form

:   $x\frac{du}{dx} = -f(u) - u,$

or

:   $\frac 1x\frac{dx}{du} = \frac {-1}{f(u) + u},$

which can now be integrated directly: `{{math|ln ''x''}}`{=mediawiki}
equals the antiderivative of the right-hand side (see ordinary
differential equation).

\paragraph{Special case}


A first order differential equation of the form
(`{{mvar|a}}`{=mediawiki}, `{{mvar|b}}`{=mediawiki},
`{{mvar|c}}`{=mediawiki}, `{{mvar|e}}`{=mediawiki},
`{{mvar|f}}`{=mediawiki}, `{{mvar|g}}`{=mediawiki} are all constants)

$$\left(ax + by + c\right) dx + \left(ex + fy + g\right) dy = 0$$ where
`{{math|''af'' ? ''be''}}`{=mediawiki} can be transformed into a
homogeneous type by a linear transformation of both variables
(`{{mvar|a}}`{=mediawiki} and `{{mvar|ß}}`{=mediawiki} are constants):

$$t = x + \alpha; \; \; z = y + \beta \,.$$

\paragraph{Homogeneous linear differential equations}


A linear differential equation is \textbf{homogeneous} if it is a homogeneous
linear equation in the unknown function and its derivatives. It follows
that, if `{{math|''f''(''x'')}}`{=mediawiki} is a solution, so is
`{{math|''cf''(''x'')}}`{=mediawiki}, for any (non-zero) constant
`{{mvar|c}}`{=mediawiki}. In order for this condition to hold, each
nonzero term of the linear differential equation must depend on the
unknown function or any derivative of it. A linear differential equation
that fails this condition is called \textbf{inhomogeneous.}

A linear differential equation can be represented as a linear operator
acting on `{{math|''y''(''x'')}}`{=mediawiki} where
`{{mvar|x}}`{=mediawiki} is usually the independent variable and
`{{mvar|y}}`{=mediawiki} is the dependent variable. Therefore, the
general form of a linear homogeneous differential equation is

$$L(y) = 0$$

where `{{mvar|L}}`{=mediawiki} is a differential operator, a sum of
derivatives (defining the \""0th derivative\"" as the original,
non-differentiated function), each multiplied by a function
`{{math|''f''''i''}}`{=mediawiki} of
`{{mvar|x}}`{=mediawiki}:

$$L = \sum_{i=0}^n f_i(x)\frac{d^i}{dx^i} \, ,$$ where
`{{math|''f''''i''}}`{=mediawiki} may be constants, but not
all `{{math|''f''''i''}}`{=mediawiki} may be zero.

For example, the following linear differential equation is homogeneous:

$$\sin(x) \frac{d^2y}{dx^2} + 4 \frac{dy}{dx} + y = 0 \,,$$

whereas the following two are inhomogeneous:

$$2 x^2 \frac{d^2y}{dx^2} + 4 x \frac{dy}{dx} + y = \cos(x) \,;$$

$$2 x^2 \frac{d^2y}{dx^2} - 3 x \frac{dy}{dx} + y = 2 \,.$$ The
existence of a constant term is a sufficient condition for an equation
to be inhomogeneous, as in the above example.

\paragraph{Second order equations}


In order to show how these equations are solved, lets start with the
most basic case - a second order equation

$$A\frac{d^2y}{dx^2}+B\frac{dy}{dx}+Cy=0$$

where A, B, and C are constants.

\paragraph{The Auxiliary Quadratic}


For a DE

$$a\frac{d^2y}{dx^2}+b\frac{dy}{dx}+cy=0,$$

Make the following substitution:

$$y=e^{mx}\,$$

This gives also

$$\frac{dy}{dx}=me^{mx}\,$$

$$\frac{d^2y}{dx^2}=m^2e^{mx}\,$$

The DE is now

$$am^2e^{mx}+bme^{mx}+ce^{mx}=0 \,$$

Dividing by $e^{mx}$ gives (note$$e^{mx}$$ can never equal zero)

$$am^2+bm+c=0 \,$$

This is the \textbf{auxiliary quadratic} (AQ) of the DE. There are four
classes of outcomes to the auxiliary quadratic:

1.  $b^2-4ac > 0, \,$, giving two distinct, real, roots.
2.  $b^2-4ac = 0, \,$, giving two coincident real, roots.
3.  $b^2-4ac < 0, \,$, giving complex roots.

:   \textit{a}: Purely imaginary roots.
:   \textit{b}: Complex-conjugate pair.

The method of solution of the DE depends on the class of AQ.

\paragraph{Class 1: Distinct, Real Roots}


Consider the DE

$$\frac{d^2y}{dx^2}+\frac{dy}{dx}-6y=0$$

The AQ is

$$m^2+m-6=0 \,$$

$$(m+3)(m-2)=0 \,$$

This gives us the following roots:

$$m=2, \ m=-3$$

Going back to the substitution we made to obtain the AQ, we have

$$y=e^{2x},\ y=e^{-3x}$$

as two distinct solutions to the DE. According to the superposition
principle, the general solution is therefore

$$y=Ae^{2x}+Be^{-3x} \,$$

\paragraph{General Solution to Class 1 DEs}


:   Generalizing, for the Second Order DE

$$a\frac{d^2y}{dx^2}+b\frac{dy}{dx}+cy=0$$

:   with the auxiliary quadratic given by

$$am^2+bm+c=0 \,$$

:   with roots \textit{a} and \textit{ß}, the general solution is

$$y=Ae^{\alpha x}+Be^{\beta x} \,$$

\paragraph{Class 2: Coincident, Real Roots}


Consider the DE

$$\frac{d^2y}{dx^2}-4\frac{dy}{dx}+4y=0$$

The AQ is

$$m^2-4m+4=0 \,$$

$$(m-2)(m-2)=0 \,$$

so, $e^{2x}\,$ is a solution. However, we cannot have it as both
solutions as the factor of two produced will be absorbed into the
constant, leaving us with only one constant, and therefore a DE without
a full solution.

For the other solution we will use the Method of Reduction of Order. To
do so we assume that it is in the form of:

$$y_2=u(x)y_1=u(x)e^{2x}\,$$ At the end we will check if our assumption
is correct. We will now substitute this equation and solve for u(x)

$$y_2''-4y_2'+4y_2=0\,$$

$$u''(x)e^{2x}+2u'(x)e^{2x}+2u'(x)e^{2x}+4u(x)e^{2x}-4(u'(x)e^{2x}+2u(x)e^{2x})+4u(x)e^{2x}=0\,$$

$$u''(x)e^{2x}=0\,$$

$$e^{2x}\,$$ is always non-zero so the only way the product can equal
zero is if:

$$u''(x)=0\,$$ Integrating twice offers

$$u(x)=a_1x + a_2\,$$ Therefore

$$y_2=(a_1x+a_2)y_1\,$$

$$y_2=(a_1x+a_2)e^{2x}\,$$ Our general solution is:

$$y=C_1y_1+C_2y_2\,$$

$$y=C_1e^{2x}+C_2(a_1x+a_2)e^{2x}\,$$

$$y=(C_1+a_2)e^{2x}+C_2a_1xe^{2x}\,$$ Because each constant is arbitrary
we can simply write

$$y=C_1e^{2x}+C_2xe^{2x}\,$$

The Method of Reduction of Order can be used on different equations and
u(x) does not always equal x. You can see below that $y=xe^{x}$ is a
valid solution.

$$y=xe^{2x}\,$$

$$\frac{dy}{dx}=2xe^{2x}+e^{2x}=e^{2x}(2x+1)\,$$

$$\frac{d^2y}{dx^2}=2e^{2x}+2e^{2x}(2x+1)=e^{2x}(4x+4)\,$$

To check substitute these into the original DE:

$$e^{2x}(4x+4)-4e^{2x}(2x+1)+4xe^{2x}=0 \,$$

$$4xe^{2x}+4e^{2x}-8xe^{2x}-4e^{2x}+4xe^{2x}=0 \,$$

$$0=0\,$$

Therefore, $y=xe^{2x}\,$ is a solution as well.

\paragraph{General Solution to Class 2 DEs}


:   Generalizing, for the Second Order DE

$$a\frac{d^2y}{dx^2}+b\frac{dy}{dx}+cy=0$$

:   with the coincident root \textit{a} of the AQ, the general solution is

$$y=(A+Bx)e^{\alpha x} \,$$

\paragraph{Class 3a: Purely Imaginary Roots}


To have complex roots, the AQ must have a discriminant less than zero,
so

$$b^2-4ac < 0 \,$$

Also, for the solution to be purely imaginary, the value of \textit{b} must be
exactly zero.

Therefore,

$$4ac>0\,$$

This means that \textit{a} and \textit{c} have to have the same signs: either \textit{a} and
\textit{c} are both positive or they are both negative. If we consider our
general second-order DE:

$$a\frac{d^2y}{dx^2}+b\frac{dy}{dx}+cy=0$$

Setting \textit{b} to zero gives

$$a\frac{d^2y}{dx^2}+cy=0$$

Dividing through by \textit{a} gives

$$\frac{d^2y}{dx^2}+\frac{c}{a}y=0$$.

Therefore, the \textit{y} term is always positive, and this can be represented
by

$$\frac{d^2y}{dx^2}+\omega^2y=0$$.

(I\'m using \textit{?} here as it is used for simple harmonic motion, which is
the primary use of this DE). There are now two paths to the solution of
the DE. The first relies on us spotting that we can use the cyclical
nature of trig. functions when derived. Substitute the following

$$y=\cos \omega x \,$$

$$\frac{dy}{dx}=-\omega \sin \omega x\,$$

$$\frac{d^2y}{dx^2}=-\omega^2 \cos \omega x\,$$

And check in our DE:

$$\frac{d^2y}{dx^2}+\omega^2 y=-\omega^2 \cos \omega x+\omega^2 \cos \omega x=0$$

This checks out, so $\cos \omega x\,$ is a solution. A similar result
holds true for the substitution using $y=\sin \omega x \,$.

Our solutions are therefore

$$y=\cos \omega x \,$$

$$y=\sin \omega x \,$$

So the general solution is

$$y=A\cos \omega x+B\sin \omega x. \,$$

The other method of solving this equation is to use Euler\'s Formula:

$$e^{\omega ix}=\cos \omega x +i \sin x \,$$

:   and

$$e^{-\omega ix}=\cos \omega x -i \sin x \,$$

From our original DE, we have an AQ of

$$m^2+\omega^2=0 \,$$

giving us roots of

$$m=\pm i \omega \,$$

so the general solution, similar to the Class 1 DEs, is

$$y=Ae^{\omega ix}+Be^{-\omega ix} \,$$

$$y=A(\cos \omega x +i \sin \omega x)+B(\cos \omega x -i \sin \omega x)\,$$

$$y=(A+B)\cos \omega x+i(A-B)\sin \omega x \,$$

Since A and B are arbitrary, we can set new constants for convenience,
letting our new A equal A+B and our new B equal \textit{i}(A-B).

Thus we have as our general solution

$$y=A\cos \omega x+B\sin \omega x \,$$

\paragraph{General Solution to Class 3a DEs}


:   Generalizing, for the Second Order DE

$$\frac{d^2y}{dx^2}+\omega^2 y=0$$

:   the general solution is

$$y=A\cos \omega x+B\sin \omega x \,$$

\paragraph{Class 3b: Complex Conjugate Roots}


Since it is a proven theorem that complex roots of polynomials always
occur in conjugate pairs, the only remaining class of AQ is the one with
complex conjugates for solutions.

Given that the solutions are complex, we know that in the AQ

$$am^2+bm+c=0 \,$$

$$b^2-4ac,<0,\ b \ne 0 \,$$ (see Class 3a).

The roots of this are in the form

$$(p \pm iq)\,$$

The general solution is then

$$y=Ae^{(p+iq)x}+Be^{(p-iq)x}\,$$

$$y=e^{px} \left( Ae^{iqx}+Be^{-iqx}\right)\,$$

From Euler\'s Formulas, we can now get

$$y=e^{px}\left( A\left(\cos q x + i \sin q x \right)+B\left(\cos q x - i \sin q x \right) \right)\,$$

$$y=e^{px}\left( (A+B)\cos q x+i(A-B)\sin q x \right)\,$$

As A and B are arbitrary, we can collapse them as in Class 3a, so that
we have the general solution

$$y=e^{px}\left( A\cos q x+B\sin q x \right)\,$$

\paragraph{General Solution to Class 3b DEs}


:   Generalizing, for the Second Order DE

$$a\frac{d^2y}{dx^2}+b\frac{dy}{dx}+cy=0$$

:   with an AQ with roots

$$m=p \pm iq,$$

:   The general solution is

$$y=e^{px}\left( A\cos q x+B\sin q x \right)\,$$

We have now covered all possible types of homogeneous second-order
differential equation, and we didn\'t even have to integrate anything!
We will now have a look at higher order equivalents.

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Homogeneous_differential_equation}{Homogeneous differential equation,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license
\begin{itemize}
    \item \href{https://en.wikibooks.org/wiki/Ordinary_Differential_Equations/Homogenous_1}{Homogeneous 1, Wikibooks: Ordinary Differential
\end{itemize}
    Equations}
    under a CC BY-SA license"

%%===============================================================
\section{Abel’s Theorem}
\label{sec:abel-s-theorem}
%%===============================================================

"In mathematics, \textbf{Abel\'s theorem} for power series relates a limit of
a power series to the sum of its coefficients. It is named after
Norwegian mathematician Niels Henrik Abel.

\paragraph{Theorem}


Let

$$G (x) = \sum_{k=0}^\infty a_k x^k$$

be a power series with real coefficients \textit{a}\textasciitilde{}\textit{k}\textasciitilde{} with radius of
convergence 1. Suppose that the series

$$\sum_{k=0}^\infty a_k$$

converges. Then \textit{G}(\textit{x}) is continuous from the left at $x=1$, i.e.

$$\lim_{x\to 1^-} G(x) = \sum_{k=0}^\infty a_k.$$

The same theorem holds for complex power series

$$G(z) = \sum_{k=0}^\infty a_k z^k,$$

provided that $z \to 1$ within a \textit{Stolz sector}, that is, a region of
the open unit disk where

:   $|1-z|\leq M(1-|z|)$

for some $M > 1$. Without this restriction, the limit may fail to exist:
for example, the power series

$$\sum_{n>0} \frac{z^{3^n}-z^{2\cdot 3^n}} n$$

converges to 0 at \textit{z} = 1, but is unbounded near any point of the form
\textit{e}^$\pi$i/3^\textit{n}^^, so the value at \textit{z} = 1 is not the limit as \textit{z}
tends to 1 in the whole open disk.

Note that \textit{G}(\textit{z}) is continuous on the real closed interval \[0, \textit{t}\]
for \textit{t} \< 1, by virtue of the uniform convergence of the series on
compact subsets of the disk of convergence. Abel\'s theorem allows us to
say more, namely that \textit{G}(\textit{z}) is continuous on \[0, 1\].

\paragraph{Remarks}


As an immediate consequence of this theorem, if \textit{z} is any nonzero
complex number for which the series

$$\sum_{k=0}^\infty a_k z^k$$

converges, then it follows that

$$\lim_{t\to 1^{-}} G(tz) = \sum_{k=0}^\infty a_kz^k$$

in which the limit is taken from below.

The theorem can also be generalized to account for sums which diverge to
infinity. If

$$\sum_{k=0}^\infty a_k = \infty$$

then

$$\lim_{z\to 1^{-}} G(z) \to \infty.$$

However, if the series is only known to be divergent, but for reasons
other than diverging to infinity, then the claim of the theorem may
fail: take, for example, the power series for

$$\frac{1}{1+z}.$$

At $z=1$ the series is equal to $1 - 1 + 1 - 1 + \cdots,$ but
$\tfrac{1}{1+1}=\tfrac{1}{2}.$

We also remark the theorem holds for radii of convergence other than
$R=1$: let

$$G(x)=\sum_{k=0}^\infty a_kx^k$$

be a power series with radius of convergence $R$, and suppose the series
converges at $x=R$. Then $G(x)$ is continuous from the left at $x=R$,
i.e.

$$\lim_{x\to R^-}G(x)=G(R)$$

\paragraph{Applications}


The utility of Abel\'s theorem is that it allows us to find the limit of
a power series as its argument (i.e. $z$) approaches 1 from below, even
in cases where the radius of convergence, $R$, of the power series is
equal to 1 and we cannot be sure whether the limit should be finite or
not. See e.g. the binomial series. Abel\'s theorem allows us to evaluate
many series in closed form. For example, when

$$a_k = \frac{(-1)^k}{k+1},$$

we obtain

$$G_a(z) = \frac{\ln(1+z)}{z}, \qquad 0 < z < 1,$$

by integrating the uniformly convergent geometric power series term by
term on $[-z, 0]$; thus the series

$$\sum_{k=0}^\infty \frac{(-1)^k}{k+1}$$

converges to $\ln(2)$ by Abel\'s theorem. Similarly,

$$\sum_{k=0}^\infty \frac{(-1)^k}{2k+1}$$

converges to $\arctan(1) = \tfrac{\pi}{4}.$

$G_a(z)$ is called the generating function of the sequence $a$. Abel\'s
theorem is frequently useful in dealing with generating functions of
real-valued and non-negative sequences, such as probability-generating
functions. In particular, it is useful in the theory of Galton--Watson
processes.

\paragraph{Outline of proof}


After subtracting a constant from $a_0$, we may assume that
$\sum_{k=0}^\infty a_k=0$. Let $s_n=\sum_{k=0}^n a_k\!$. Then
substituting $a_k=s_k-s_{k-1}$ and performing a simple manipulation of
the series (summation by parts) results in

$$G_a(z) = (1-z)\sum_{k=0}^{\infty} s_k z^k.$$

Given $\varepsilon > 0,$ pick \textit{n} large enough so that
$|s_k| < \varepsilon$ for all $k\ge n$ and note that

$$\left|(1-z)\sum_{k=n}^\infty s_kz^k \right| \le \varepsilon |1-z|\sum_{k=n}^\infty |z|^k = \varepsilon|1-z|\frac{|z|^n}{1-|z|} < \varepsilon M$$

when \textit{z} lies within the given Stolz angle. Whenever \textit{z} is sufficiently
close to 1 we have

$$\left|(1-z)\sum_{k=0}^{n-1} s_kz^k \right| < \varepsilon,$$

so that $|G_a(z)| < (M+1)\varepsilon$ when \textit{z} is both sufficiently
close to 1 and within the Stolz angle.

\paragraph{Related concepts}


Converses to a theorem like Abel\'s are called Tauberian theorems: There
is no exact converse, but results conditional on some hypothesis. The
field of divergent series, and their summation methods, contains many
theorems \textit{of abelian type} and \textit{of tauberian type}.

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Abel\%27s_theorem}{Abel\'s theorem,
\end{itemize}
    Wikipedia} under a
    CC BY-SA license"

%%===============================================================
\section{Fundamental Solutions}
\label{sec:fundamental-solutions}
%%===============================================================

Fundamental solutions to linear homogeneous equations

> 
>
> ``{=html}Theorem 1:``{=html} Let
> $\frac{d^2 y}{dt^2} + p(t) \frac{dy}{dt} + q(t)y = 0$ be a second
> order linear homogenous differential equation where $p$ and $q$ are
> continuous on an open interval $I$ such that $t_0 \in I$, and let
> $y = y_1(t)$ and $y = y_2(t)$ be two solutions to this differential
> equation. The set of all linear combinations of these two solutions,
> $y = Cy_1(t) + Dy_2(t)$ where $C$ and $D$ are constants contains all
> solutions to this differential equation if and only if there exists a
> point $t_0$ for which the Wronksian of $y_1$ and $y_2$ at $t_0$ is
> nonzero, that is $W(y_1, y_2) \bigg|_{t_0} \neq 0$.
>
> 

\begin{itemize}
    \item ``{=html}Proof:``{=html} Let $y = y_1(t)$ and
\end{itemize}
    $y = y_2(t)$ both be solutions to the differential equation
    $\frac{d^2 y}{dt^2} + p(t) \frac{dy}{dt} + q(t)y = 0$ and suppose
    that $y = \phi (t)$ is any arbitrary solution as well. We want to
    show that $\phi (t)$ is a linear combination of $y_1$ and $y_2$ for
    some constants $C$ and $D$.

<!-- -->

\begin{itemize}
    \item $\Leftarrow$ Let $t_0$ be such that the Wronskian of $y_1$ and $y_2$
\end{itemize}
    evaluated at $t_0$ is nonzero, that is:



$\begin{aligned} \quad W(y_1, y_2) \bigg|_{t_0} \neq 0 \end{aligned}$



\begin{itemize}
    \item Take this value $t_0$ and evaluate both $\phi$ and $\phi'$ at this
\end{itemize}
    point. Then $\phi (t_0) = y_0$ and $\phi'(t_0) = y'_0$ (since
    $y = \phi(t)$ is a solution to our differential equation). Now
    consider the initial value problem
    $\frac{d^2 y}{dt^2} + p(t) \frac{dy}{dt} + q(t) y = 0$ with the
    initial conditions $y(t_0) = y_0$ and $y'(t_0) = y'_0$. The function
    $\phi$ satisfies this differential equation. Since
    $W(y_1, y_2) \bigg|_{t_0} \neq 0$ then we have that there exists
    constants $C$ and $D$ such that $y = Cy_1(t) + Dy_2(t)$ satisfies
    this initial value problem. But since $p$ and $q$ are continuous on
    the open interval $I$ containing $t_0$ then this implies that a
    unique solution exists, and so:



$\begin{aligned} \phi(t) = Cy_1(t) + Dy_2(t) \end{aligned}$



\begin{itemize}
    \item So all solutions for this differential equation are a linear
\end{itemize}
    combination of the solutions $y = y_1(t)$ and $y = y_2(t)$.

<!-- -->

\begin{itemize}
    \item $\Rightarrow$ Suppose that every point $t_0 \in I$ is such that
\end{itemize}
    $W(y_1, y_2) \bigg|_{t_0} = 0$, that is, there exists no point $t_0$
    on $I$ where the Wronskian of $y_1$ and $y_2$ evaluated at $t_0$ is
    nonzero. Let $y_0$ and $y'_0$ be values for which the system
    $\left\{\begin{matrix} Cy_1(t_0) + Dy_2(t_0) = y_0 \\\\ Cy_1'(t_0) + Dy_2'(t_0) = y'_0 \end{matrix}\right.$
    has no solutions for a set of constants $C$ and $D$.

<!-- -->

\begin{itemize}
    \item Now since $p$ and $q$ are continuous on an open interval $I$
\end{itemize}
    containing $t_0$, such a solution $\phi(t)$ satisfies the initial
    conditions $y(t_0) = y_0$ and $y'(t_0) = y'_0$. Note though this
    solution is not a linear combination of $y_1$ and $y_2$ though which
    completes our proof. $\blacksquare$

Theorem 1 above implies that if we can find two solutions $y = y_1(t)$
and $y = y_2(t)$ for which the Wronskian $W(y_1, y_2) \neq 0$, then for
constants $C$ and $D$, all solutions of the second order linear
homogenous differential equation
$\frac{d^2 y}{dt^2} + p(t) \frac{dy}{dt} + q(t) y = 0$ are given by:



$\begin{aligned} \quad y = Cy_1(t) + Dy_2(t) \end{aligned}$



Also note that thus far we have not said that $y = y_1(t)$ and
$y = y_2(t)$ need to be distinct. However, with the Theorem above, we
see that if $y_1(t) = y_2(t)$ then the Wronskian
$W(y_1, y_2) = W(y_1, y_1) = W(y_2, y_2)$ is zero (as you should verify)
and so not all solutions to a second order linear homogenous
differential are given by the linear combination of just $y_1$.

> 
>
> ``{=html}Definition:``{=html} Let
> $\frac{d^2 y}{dt^2} + p(t) \frac{dy}{dt} + q(t)y = 0$ where $p$ and
> $q$ are continuous on an open interval $I$ such that $t_0 \in I$ and
> let $y = y_1(t)$ and $y = y_2(t)$ be solutions to this differential
> equation. If the Wronskian $W(y_1, y_2) \neq 0$ then the set of linear
> combinations of $y_1$ and $y_2$ is known as the
> ``{=html}Fundamental Set of Solutions``{=html} to
> this differential equation.
>
> 

From the definition above, we see that if we can find two solutions
$y = y_1(t)$ and $y = y_2(t)$ for which the Wronskian $W(y_1, y_2)$ is
nonzero, then $y_1$ and $y_2$ form a fundamental set of solutions. The
next question that we might pose is whether or not a second order linear
homogenous differential equation always has a fundamental set of
solutions.

> 
>
> ``{=html}Theorem 2:``{=html} Let
> $\frac{d^2 y}{dt} + p(t) \frac{dy}{dt} + q(t) y = 0$ be a second order
> linear homogenous differential equation where $p$ and $q$ are
> continuous on an open interval $I$ such that $t_0 \in I$. If
> $y = y_1(t)$ is a solution to this differential equation that
> satisfies the initial conditions $y_1(t_0) = 1$ and $y_1'(t_0) = 0$,
> and if $y = y_2(t)$ be a solution to this differential equation that
> satisfies the initial conditions $y_2(t_0) = 0$ and $y_2'(t_0) = 1$.
> Then $y_1$ and $y_2$ form a fundamental set of solutions for this
> differential equation.
>
> 

\begin{itemize}
    \item ``{=html}Proof:``{=html} We note that:
\end{itemize}



$\begin{aligned} \quad W(y_1, y_2) \bigg|_{t_0} = \begin{vmatrix} y_1(t_0)  y_2(t_0) \\\\ y_1'(t_0)  y_2'(t_0)\end{vmatrix} = \begin{vmatrix} 1  0\\\\ 0  1 \end{vmatrix} = 1 \neq 0 \end{aligned}$



\begin{itemize}
    \item Thus Theorem 1 implies that ALL solutions to this differential
\end{itemize}
    equation are given by $y = Cy_1(t) + Dy_2(t)$ where $C$ and $D$ are
    constants. Thus $y_1$ and $y_2$ form a fundamental set of solutions
    for this differential equation. $\blacksquare$

\paragraph{Fundamental sets of solutions}


> 
>
> ``{=html}Definition:``{=html} A
> ``{=html}Fundamental Set of Solutions``{=html} to the
> linear homogeneous system of first order ODEs
> $\mathbf{x}' = A(t) \mathbf{x}$ on $J = (a, b)$ is a set
> $\{ \phi^{[1]}, \phi^{[2]}, ..., \phi^{[n]} \}$ of linearly
> independent solutions to this system on $J$.
>
> 

For example, consider the following linear homogeneous system of $2$
first order ODEs:



$\begin{aligned} \quad \left\{\begin{matrix} x_1' = x_1\\\\ x_2' = 2x_2 \end{matrix}\right. \end{aligned}$



We can easily solve this system. For the first differential equation:



$\begin{aligned} \quad \frac{dx_1}{dt} &= x_1 \\\\ \frac{dx_1}{x_1} &= dt \\\\ \int \frac{1}{x_1}  dx &= \int  dt \\\\ \ln (x_1) &= t + C \\\\ x_1 &= e^{t + C} \\\\ x_1 &= C_1e^t \end{aligned}$



Where $C_1 = e^C > 0$.

For the second differential equation:



$\begin{aligned} \quad \frac{dx_2}{dt} &= 2x_2 \\\\ \frac{dx_2}{x_2} &= 2  dt \\\\ \int \frac{1}{x_2}  dx_2 &= \int 2  dt \\\\ \ln (x_2) &= 2t + C \\\\ x_2 &= e^{2t + C} \\\\ x_2 &= C_2e^{2t} \end{aligned}$



Where $C_2 = e^C > 0$.

Note that in fact $C_1, C_2$ can be any real numbers are we are not
simply restricted to $C_1, C_2 > 0$.

Now by taking $C_1 = 1$, $C_2 = 0$ we get that
$\phi^{[1]} = \begin{bmatrix} e^t\\\\ 0 \end{bmatrix}$ is a solution to
this system. Also, by taking $C_1 = 0$ and $C_2 = 1$ we get that
$\phi^{[2]} = \phi^{[1]} = \begin{bmatrix} 0\\\\ e^{2t} \end{bmatrix}$ is
a solution to this system.

We will now show that $\{ \phi^{[1]}, \phi^{[2]} \}$ is a Fundamental
set of solutions to this system on all of $\mathbb{R}$. Let
$\alpha, beta \in \mathbb{R}$ and consider the following equation:



$\begin{aligned} \alpha \phi^{[1]} + \beta \phi^{[2]} &= 0 \\\\ \alpha \begin{bmatrix} e^t\\\\ 0 \end{bmatrix} + \beta \begin{bmatrix} 0\\\\ e^{2t} \end{bmatrix} &= \begin{bmatrix} 0 \\\\ 0 \end{bmatrix} \\\\ \begin{bmatrix} \alpha e^t \\\\ \beta e^{2t} \end{bmatrix} &= \begin{bmatrix} 0 \\\\ 0 \end{bmatrix} \end{aligned}$



The equation above implies that $\alpha e^t = 0$ and $\beta e^{2t} = 0$
for all $t \in \mathbb{R}$. Since $e^t, e^{2t} > 0$ for all
$t \in \mathbb{R}$ this implies that $\alpha, \beta = 0$. So
$\{ \phi^{[1]}, \phi^{[2]} \}$ is a linearly independent set of
solutions to this system and so $\{ \phi^{[1]}, \phi^{[2]} \}$ is a
fundamental set of solutions to this system on $\mathbb{R}$.

\paragraph{Resources}


\begin{itemize}
    \item \href{https://tutorial.math.lamar.edu/Classes/DE/FundamentalSetsofSolutions.aspx}{Fundamental Sets of Solutions
\end{itemize}
    Notes}.
    Produced by Paul Dawkins, Lamar University
\begin{itemize}
    \item \href{https://www.cfm.brown.edu/people/dobrush/am33/Mathematica/ch4/fundset.html}{Fundamental Set of
\end{itemize}
    Solutions},
    Brown University

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{http://mathonline.wikidot.com/fundamental-solutions-to-linear-homogenous-differential-equa}{Fundamental Solutions to Linear Homogenous Differential Equations,
\end{itemize}
    mathonline.wikidot.com}
    under a CC BY-SA license
\begin{itemize}
    \item \href{http://mathonline.wikidot.com/fundamental-sets-of-solutions-to-a-linear-homogeneous-system}{Fundamental Sets of Solutions,
\end{itemize}
    mathonline.wikidot.com}
    under a CC BY-SA license"

%%===============================================================
\section{Linear Non-homogeneous Equations}
\label{sec:linear-non-homogeneous-equations}
%%===============================================================

"In mathematics, a \textbf{linear differential equation} is a differential
equation that is defined by a linear polynomial in the unknown function
and its derivatives, that is an equation of the form

$$a_0(x)y + a_1(x)y' + a_2(x)y'' \cdots + a_n(x)y^{(n)} = b(x)$$ where
\textit{a}\textasciitilde{}0\textasciitilde{}(\textit{x}), ..., \textit{a}\textasciitilde{}\textit{n}\textasciitilde{}(\textit{x}) and `{{math|''b''(''x'')}}`{=mediawiki}
are arbitrary differentiable functions that do not need to be linear,
and `{{math|''y''', …, ''y''(''n'') }}`{=mediawiki} are the
successive derivatives of an unknown function `{{mvar|y}}`{=mediawiki}
of the variable `{{mvar|x}}`{=mediawiki}.

Such an equation is an ordinary differential equation (ODE). A *linear
differential equation* may also be a linear partial differential
equation (PDE), if the unknown function depends on several variables,
and the derivatives that appear in the equation are partial derivatives.

A linear differential equation or a system of linear equations such that
the associated homogeneous equations have constant coefficients may be
solved by quadrature, which means that the solutions may be expressed in
terms of integrals. This is also true for a linear equation of order
one, with non-constant coefficients. An equation of order two or higher
with non-constant coefficients cannot, in general, be solved by
quadrature. For order two, Kovacic\'s algorithm allows deciding whether
there are solutions in terms of integrals, and computing them if any.

The solutions of linear differential equations with polynomial
coefficients are called holonomic functions. This class of functions is
stable under sums, products, differentiation, integration, and contains
many usual functions and special functions such as exponential function,
logarithm, sine, cosine, inverse trigonometric functions, error
function, Bessel functions and hypergeometric functions. Their
representation by the defining differential equation and initial
conditions allows making algorithmic (on these functions) most
operations of calculus, such as computation of antiderivatives, limits,
asymptotic expansion, and numerical evaluation to any precision, with a
certified error bound.

\paragraph{Basic terminology}


The highest order of derivation that appears in a (linear) differential
equation is the \textit{order} of the equation. The term
`{{math|''b''(''x'')}}`{=mediawiki}, which does not depend on the
unknown function and its derivatives, is sometimes called the *constant
term* of the equation (by analogy with algebraic equations), even when
this term is a non-constant function. If the constant term is the zero
function, then the differential equation is said to be \textit{homogeneous}, as
it is a homogeneous polynomial in the unknown function and its
derivatives. The equation obtained by replacing, in a linear
differential equation, the constant term by the zero function is the
\textit{associated homogeneous equation}. A differential equation has *constant
coefficients* if only constant functions appear as coefficients in the
associated homogeneous equation.

A \textit{solution} of a differential equation is a function that satisfies the
equation. The solutions of a homogeneous linear differential equation
form a vector space. In the ordinary case, this vector space has a
finite dimension, equal to the order of the equation. All solutions of a
linear differential equation are found by adding to a particular
solution any solution of the associated homogeneous equation.

\paragraph{Linear differential operator}


A \textit{basic differential operator} of order `{{mvar|i}}`{=mediawiki} is a
mapping that maps any differentiable function to its
`{{mvar|i}}`{=mediawiki}th derivative, or, in the case of several
variables, to one of its partial derivatives of order
`{{mvar|i}}`{=mediawiki}. It is commonly denoted

$$\frac{d^i}{dx^i}$$ in the case of univariate functions, and

$$\frac{\partial^{i_1+\cdots +i_n}}{\partial x_1^{i_1}\cdots \partial x_n^{i_n}}$$
in the case of functions of `{{mvar|n}}`{=mediawiki} variables. The
basic differential operators include the derivative of order 0, which is
the identity mapping.

A \textbf{linear differential operator} (abbreviated, in this article, as
\textit{linear operator} or, simply, \textit{operator}) is a linear combination of
basic differential operators, with differentiable functions as
coefficients. In the univariate case, a linear operator has thus the
form

$$a_0(x)+a_1(x)\frac{d}{dx} + \cdots +a_n(x)\frac{d^n}{dx^n},$$ where
`{{math|''a''0(''x''), …, ''a''''n''(''x'')}}`{=mediawiki}
are differentiable functions, and the nonnegative integer
`{{mvar|n}}`{=mediawiki} is the \textit{order} of the operator (if
`{{math|''a''''n''(''x'')}}`{=mediawiki} is not the zero
function).

Let `{{mvar|L}}`{=mediawiki} be a linear differential operator. The
application of `{{mvar|L}}`{=mediawiki} to a function
`{{mvar|f}}`{=mediawiki} is usually denoted
`{{math|''Lf''}}`{=mediawiki} or `{{math|''Lf''(''X'')}}`{=mediawiki},
if one needs to specify the variable (this must not be confused with a
multiplication). A linear differential operator is a linear operator,
since it maps sums to sums and the product by a scalar to the product by
the same scalar.

As the sum of two linear operators is a linear operator, as well as the
product (on the left) of a linear operator by a differentiable function,
the linear differential operators form a vector space over the real
numbers or the complex numbers (depending on the nature of the functions
that are considered). They form also a free module over the ring of
differentiable functions.

The language of operators allows a compact writing for differentiable
equations: if

$$L=a_0(x)+a_1(x)\frac{d}{dx} + \cdots +a_n(x)\frac{d^n}{dx^n},$$ is a
linear differential operator, then the equation

$$a_0(x)y +a_1(x)y' + a_2(x)y'' +\cdots +a_n(x)y^{(n)}=b(x)$$ may be
rewritten

$$Ly=b(x).$$

There may be several variants to this notation; in particular the
variable of differentiation may appear explicitly or not in
`{{mvar|y}}`{=mediawiki} and the right-hand and of the equation, such as
`{{math|1=''Ly''(''x'') = ''b''(''x'')}}`{=mediawiki} or
`{{math|1=''Ly'' = ''b''}}`{=mediawiki}.

The \textit{kernel} of a linear differential operator is its kernel as a linear
mapping, that is the vector space of the solutions of the (homogeneous)
differential equation `{{math|1=''Ly'' = 0}}`{=mediawiki}.

In the case of an ordinary differential operator of order
`{{mvar|n}}`{=mediawiki}, Carathéodory\'s existence theorem implies
that, under very mild conditions, the kernel of `{{mvar|L}}`{=mediawiki}
is a vector space of dimension `{{mvar|n}}`{=mediawiki}, and that the
solutions of the equation
`{{math|1=''Ly''(''x'') = ''b''(''x'')}}`{=mediawiki} have the form

$$S_0(x) + c_1S_1(x) + \cdots +c_nS_n(x),$$ where
`{{math|''c''1, …, ''c''''n''}}`{=mediawiki} are
arbitrary numbers. Typically, the hypotheses of Carathéodory\'s theorem
are satisfied in an interval `{{mvar|I}}`{=mediawiki}, if the functions
`{{math|''b'', ''a''0, …, ''a''''n''}}`{=mediawiki}
are continuous in `{{mvar|I}}`{=mediawiki}, and there is a positive real
number `{{mvar|k}}`{=mediawiki} such that \|\textit{a}\textasciitilde{}\textit{n}\textasciitilde{}(\textit{x})\| \> \textit{k} for
every `{{mvar|x}}`{=mediawiki} in `{{mvar|I}}`{=mediawiki}.

\paragraph{Homogeneous equation with constant coefficients}


A homogeneous linear differential equation has \textit{constant coefficients}
if it has the form

$$a_0y + a_1y' + a_2y'' + \cdots + a_n y^{(n)} = 0$$ where
`{{math|''a''1, …, ''a''''n''}}`{=mediawiki} are
(real or complex) numbers. In other words, it has constant coefficients
if it is defined by a linear operator with constant coefficients.

The study of these differential equations with constant coefficients
dates back to Leonhard Euler, who introduced the exponential function
`{{math|''e''''x''}}`{=mediawiki}, which is the unique
solution of the equation `{{math|1=''f''' = ''f''}}`{=mediawiki} such
that `{{math|1=''f''(0) = 1}}`{=mediawiki}. It follows that the
`{{mvar|n}}`{=mediawiki}th derivative of
`{{math|''e''''cx'' }}`{=mediawiki} is
`{{math|''c''''n''''e''''cx''}}`{=mediawiki}, and
this allows solving homogeneous linear differential equations rather
easily.

Let

$$a_0y + a_1y' + a_2y'' + \cdots + a_ny^{(n)} = 0$$ be a homogeneous
linear differential equation with constant coefficients (that is
`{{math|''a''0, …, ''a''''n''}}`{=mediawiki} are
real or complex numbers).

Searching solutions of this equation that have the form
`{{math|''e''''ax''}}`{=mediawiki} is equivalent to searching
the constants `{{mvar|a}}`{=mediawiki} such that

$$a_0e^{\alpha x} + a_1\alpha e^{\alpha x} + a_2\alpha^2 e^{\alpha x}+\cdots + a_n\alpha^n e^{\alpha x} = 0.$$
Factoring out `{{math|''e''''ax''}}`{=mediawiki} (which is
never zero), shows that `{{mvar|a}}`{=mediawiki} must be a root of the
\textit{characteristic polynomial}

$$a_0 + a_1t + a_2t^2 + \cdots + a_nt^n$$ of the differential equation,
which is the left-hand side of the characteristic equation

$$a_0 + a_1t + a_2t^2 + \cdots + a_nt^n = 0.$$

When these roots are all distinct, one has `{{mvar|n}}`{=mediawiki}
distinct solutions that are not necessarily real, even if the
coefficients of the equation are real. These solutions can be shown to
be linearly independent, by considering the Vandermonde determinant of
the values of these solutions at
`{{math|1=''x'' = 0, …, ''n'' – 1}}`{=mediawiki}. Together they form a
basis of the vector space of solutions of the differential equation
(that is, the kernel of the differential operator).

+----------------------------------------------------------------------+
| Example                                                              |
+======================================================================+
| $$y''''-2y'''+2y''-2y'+y=0$$ has the characteristic equation         |
|                                                                      |
| :   $z^4-2z^3+2z^2-2z+1=0.$                                          |
|                                                                      |
| This has zeros, `{{mvar|i}}`{=mediawiki},                            |
| `{{math|-''i''}}`{=mediawiki}, and `{{math|1}}`{=mediawiki}          |
| (multiplicity 2). The solution basis is thus                         |
|                                                                      |
| :   $e^{ix}, \; e^{-ix}, \; e^x, \; xe^x.$                              |
|                                                                      |
| A real basis of solution is thus                                     |
|                                                                      |
| :   $\cos x, \; \sin x, \; e^x, \; xe^x.$                               |
+----------------------------------------------------------------------+

In the case where the characteristic polynomial has only simple roots,
the preceding provides a complete basis of the solutions vector space.
In the case of multiple roots, more linearly independent solutions are
needed for having a basis. These have the form

$$x^ke^{\alpha x},$$ where `{{mvar|k}}`{=mediawiki} is a nonnegative
integer, `{{mvar|a}}`{=mediawiki} is a root of the characteristic
polynomial of multiplicity `{{mvar|m}}`{=mediawiki}, and
`{{math|''k'' < ''m''}}`{=mediawiki}. For proving that these functions
are solutions, one may remark that if `{{mvar|a}}`{=mediawiki} is a root
of the characteristic polynomial of multiplicity
`{{mvar|m}}`{=mediawiki}, the characteristic polynomial may be factored
as
`{{math|''P''(''t'') (''t'' - ''a'')''m''}}`{=mediawiki}.
Thus, applying the differential operator of the equation is equivalent
with applying first `{{mvar|m}}`{=mediawiki} times the operator
$\frac{d}{dx} - \alpha$, and then the operator that has
`{{mvar|P}}`{=mediawiki} as characteristic polynomial. By the
exponential shift theorem,

$$\left(\frac{d}{dx}-\alpha\right)\left(x^ke^{\alpha x}\right)= kx^{k-1}e^{\alpha x},$$

and thus one gets zero after `{{math|''k'' + 1}}`{=mediawiki}
application of $\frac{d}{dx} - \alpha$.

As, by the fundamental theorem of algebra, the sum of the multiplicities
of the roots of a polynomial equals the degree of the polynomial, the
number of above solutions equals the order of the differential equation,
and these solutions form a base of the vector space of the solutions.

In the common case where the coefficients of the equation are real, it
is generally more convenient to have a basis of the solutions consisting
of real-valued functions. Such a basis may be obtained from the
preceding basis by remarking that, if
`{{math|''a'' + ''ib''}}`{=mediawiki} is a root of the characteristic
polynomial, then `{{math|''a'' – ''ib''}}`{=mediawiki} is also a root,
of the same multiplicity. Thus a real basis is obtained by using
Euler\'s formula, and replacing
`{{math|''x''''k''''e''(''a''+''ib'')''x''}}`{=mediawiki}
and
`{{math|''x''''k''''e''(''a''-''ib'')''x''}}`{=mediawiki}
by
`{{math|''x''''k''''e''''ax'' cos(''bx'')}}`{=mediawiki}
and
`{{math|''x''''k''''e''''ax'' sin(''bx'')}}`{=mediawiki}.

\paragraph{Second-order case}


A homogeneous linear differential equation of the second order may be
written

$$y'' + ay' + by = 0,$$ and its characteristic polynomial is

$$r^2 + ar + b.$$

If `{{mvar|a}}`{=mediawiki} and `{{mvar|b}}`{=mediawiki} are real, there
are three cases for the solutions, depending on the discriminant
`{{math|1=''D'' = ''a''2 - 4''b''}}`{=mediawiki}. In all
three cases, the general solution depends on two arbitrary constants
`{{math|''c''1}}`{=mediawiki} and
`{{math|''c''2}}`{=mediawiki}.

\begin{itemize}
    \item If `{{math|''D'' > 0}}`{=mediawiki}, the characteristic polynomial
\end{itemize}
    has two distinct real roots `{{mvar|a}}`{=mediawiki}, and
    `{{mvar|ß}}`{=mediawiki}. In this case, the general solution is

$$c_1 e^{\alpha x} + c_2 e^{\beta x}.$$

\begin{itemize}
    \item If `{{math|1=''D'' = 0}}`{=mediawiki}, the characteristic polynomial
\end{itemize}
    has a double root `{{math|-''a''/2}}`{=mediawiki}, and the general
    solution is

$$(c_1 + c_2 x) e^{-ax/2}.$$

\begin{itemize}
    \item If `{{math|''D'' < 0}}`{=mediawiki}, the characteristic polynomial
\end{itemize}
    has two complex conjugate roots
    `{{math|''a'' ± ''ßi''}}`{=mediawiki}, and the general solution is

$$c_1 e^{(\alpha + \beta i)x} + c_2 e^{(\alpha - \beta i)x},$$

:   which may be rewritten in real terms, using Euler\'s formula as
    $$e^{\alpha x}  (c_1\cos(\beta x) + c_2 \sin(\beta x)).$$

Finding the solution `{{math|''y''(''x'')}}`{=mediawiki} satisfying
`{{math|1=''y''(0) = ''d''1}}`{=mediawiki} and
`{{math|1=''y'''(0) = ''d''2}}`{=mediawiki}, one equates the
values of the above general solution at `{{math|0}}`{=mediawiki} and its
derivative there to `{{math|''d''1}}`{=mediawiki} and
`{{math|''d''2}}`{=mediawiki}, respectively. This results in
a linear system of two linear equations in the two unknowns
`{{math|''c''1}}`{=mediawiki} and
`{{math|''c''2}}`{=mediawiki}. Solving this system gives the
solution for a so-called Cauchy problem, in which the values at
`{{math|0}}`{=mediawiki} for the solution of the DEQ and its derivative
are specified.

\paragraph{Non-homogeneous equation with constant coefficients}


A non-homogeneous equation of order `{{mvar|n}}`{=mediawiki} with
constant coefficients may be written

$$y^{(n)}(x) + a_1 y^{(n-1)}(x) + \cdots + a_{n-1} y'(x)+ a_ny(x) = f(x),$$
where `{{math|''a''1, …, ''a''''n''}}`{=mediawiki}
are real or complex numbers, `{{mvar|f}}`{=mediawiki} is a given
function of `{{mvar|x}}`{=mediawiki}, and `{{mvar|y}}`{=mediawiki} is
the unknown function (for sake of simplicity,
\""`{{math|(''x'')}}`{=mediawiki}\"" will be omitted in the following).

There are several methods for solving such an equation. The best method
depends on the nature of the function `{{mvar|f}}`{=mediawiki} that
makes the equation non-homogeneous. If `{{mvar|f}}`{=mediawiki} is a
linear combination of exponential and sinusoidal functions, then the
exponential response formula may be used. If, more generally,
`{{mvar|f}}`{=mediawiki} is a linear combination of functions of the
form `{{math|''x''''n''''e''''ax''}}`{=mediawiki},
`{{math|''x''''n'' cos(''ax'')}}`{=mediawiki}, and
`{{math|''x''''n'' sin(''ax'')}}`{=mediawiki}, where
`{{mvar|n}}`{=mediawiki} is a nonnegative integer, and
`{{mvar|a}}`{=mediawiki} a constant (which need not be the same in each
term), then the method of undetermined coefficients may be used. Still
more general, the annihilator method applies when
`{{mvar|f}}`{=mediawiki} satisfies a homogeneous linear differential
equation, typically, a holonomic function.

The most general method is the variation of constants, which is
presented here.

The general solution of the associated homogeneous equation

$$y^{(n)} + a_1 y^{(n-1)} + \cdots + a_{n-1} y'+ a_ny = 0$$ is

$$y=u_1y_1+\cdots+ u_ny_n,$$ where
`{{math|(''y''1, …, ''y''''n'')}}`{=mediawiki} is
a basis of the vector space of the solutions and
`{{math|''u''1, …, ''u''''n''}}`{=mediawiki} are
arbitrary constants. The method of variation of constants takes its name
from the following idea. Instead of considering
`{{math|''u''1, …, ''u''''n''}}`{=mediawiki} as
constants, they can considered as unknown functions that have to be
determined for making `{{mvar|y}}`{=mediawiki} a solution of the
non-homogeneous equation. For this purpose, one adds the constraints

$$\begin{aligned}
0 &= u'_1y_1 + u'_2y_2 + \cdots+u'_ny_n \\\0 &= u'_1y'_1 + u'_2y'_2 + \cdots + u'_n y'_n \\\  & \; \;\vdots \\\0 &= u'_1y^{(n-2)}_1+u'_2y^{(n-2)}_2 + \cdots + u'_n y^{(n-2)}_n,
\end{aligned}$$ which imply (by product rule and induction)

$$y^{(i)} = u_1 y_1^{(i)} + \cdots + u_n y_n^{(i)}$$ for
`{{math|1=''i'' = 1, …, ''n'' – 1}}`{=mediawiki}, and

$$y^{(n)} = u_1 y_1^{(n)} + \cdots + u_n y_n^{(n)} +u'_1y_1^{(n-1)}+u'_2y_2^{(n-1)}+\cdots+u'_ny_n^{(n-1)}.$$

Replacing in the original equation `{{mvar|y}}`{=mediawiki} and its
derivatives by these expressions, and using the fact that
`{{math|''y''1, …, ''y''''n''}}`{=mediawiki} are
solutions of the original homogeneous equation, one gets

$$f=u'_1y_1^{(n-1)} + \cdots + u'_ny_n^{(n-1)}.$$

This equation and the above ones with `{{math|0}}`{=mediawiki} as
left-hand side form a system of `{{mvar|n}}`{=mediawiki} linear
equations in
`{{math|''u'''1, …, ''u'''''n''}}`{=mediawiki}
whose coefficients are known functions (`{{mvar|f}}`{=mediawiki}, the
$y_i$, and their derivatives). This system can be solved by any method
of linear algebra. The computation of antiderivatives gives
`{{math|''u''1, …, ''u''''n''}}`{=mediawiki}, and
then
`{{math|1=''y'' = ''u''1''y''1 + ? + ''u''''n''''y''''n''}}`{=mediawiki}.

As antiderivatives are defined up to the addition of a constant, one
finds again that the general solution of the non-homogeneous equation is
the sum of an arbitrary solution and the general solution of the
associated homogeneous equation.

\paragraph{First-order linear equation}


The general form of a linear ordinary differential equation of order 1,
after dividing out the coefficient of
`{{math|''y'''(''x'')}}`{=mediawiki}, is:

$$y'(x) = f(x) y(x) + g(x).$$

If the equation is homogeneous, i.e.
`{{math|1=''g''(''x'') = 0}}`{=mediawiki}, one may rewrite and
integrate:

$$\frac{y'}{y}= f, \qquad \log y = k +F,$$ where
`{{mvar|k}}`{=mediawiki} is an arbitrary constant of integration and
$F=\textstyle\int f\,dx$ is any antiderivative of
`{{mvar|f}}`{=mediawiki}. Thus, the general solution of the homogeneous
equation is

$$y=ce^F,$$ where `{{math|1=''c'' = ''e''''k''}}`{=mediawiki}
is an arbitrary constant.

For the general non-homogeneous equation, one may multiply it by the
reciprocal `{{math|''e''-''F''}}`{=mediawiki} of a solution
of the homogeneous equation. This gives

$$y'e^{-F}-yfe^{-F}= ge^{-F}.$$ As
$-fe^{-F} = \frac{d}{dx} \left(e^{-F}\right),$ the product rule allows
rewriting the equation as

$$\frac{d}{dx}\left(ye^{-F}\right)= ge^{-F}.$$ Thus, the general
solution is

$$y=ce^F + e^F\int ge^{-F}dx,$$ where `{{mvar|c}}`{=mediawiki} is a
constant of integration, and `{{mvar|F}}`{=mediawiki} is any
antiderivative of `{{mvar|f}}`{=mediawiki} (changing of antiderivative
amounts to change the constant of integration).

\paragraph{Example}


Solving the equation

:   $y'(x) + \frac{y(x)}{x} = 3x.$

The associated homogeneous equation $y'(x) + \frac{y(x)}{x} = 0$ gives

$$\frac{y'}{y}=-\frac{1}{x},$$ that is

$$y=\frac{c}{x}.$$

Dividing the original equation by one of these solutions gives

:   $xy'+y=3x^2.$

That is

$$(xy)'=3x^2,$$

$$xy=x^3 +c,$$ and

$$y(x)=x^2+c/x.$$ For the initial condition

:   $y(1)=\alpha,$

one gets the particular solution

$$y(x)=x^2+\frac{\alpha-1}{x}.$$

\paragraph{Making Linear Equations from Non-Linear Equations}


Sometimes a non-linear equation, which is not solvable like this, can be
made linear, and more easily solvable, by applying a substitution.

\paragraph{Example}


$$y\frac{dy}{dx}+xy^2=5x$$

:   Let\'s make the following substitution:

$$v=y^2 \,$$

$$\frac{dv}{dx}=2y\frac{dy}{dx}$$

:   Plugging in, we get

$$\frac{1}{2}\frac{dv}{dx}+xv=5x$$

$$\frac{dv}{dx}+2xv=10x$$

:   We can then solve as a linear equation in \textit{v}, using the
    step-by-step method above:

<!-- -->

:   \textbf{Step 1}: Find the integrating factor:

$$e^{\int P(x)dx}$$

$$e^{\int 2x dx}=e^{x^2+C}$$

$$e^{\int P(x)dx}=Ce^{x^2}$$

:   Letting C=1 for convenience, we get $e^{x^2}$ as our integrating
    factor.

<!-- -->

:   \textbf{Step 2}: Multiply through

$$e^{x^2}\frac{dv}{dx}+e^{x^2}xv=10e^{x^2}x$$

:   \textbf{Step 3}: Recognize that the left hand is
    $\frac{d}{dx} e^{\int P(x)dx}v$

$$\frac{d}{dx} e^{x^2}v=10e^{x^2}x$$

:   \textbf{Step 4}: Integrate both sides w.r.t.\textit{x}.

$$\int (\frac{d}{dx} e^{x^2}v)dx=\int 10e^{x^2}xdx$$

$$e^{x^2}v=5e^{x^2}+C$$

:   \textbf{Step 5}: Solve for \textit{v}.

$$v=5+\frac{C}{e^{x^2}}$$

:   Now that we have \textit{v}, solve for \textit{y}.

$$v=y^2 \,$$

$$y^2=5+\frac{C}{e^{x^2}} \,$$

\paragraph{Higher order linear equations}


A linear ordinary equation of order one with variable coefficients may
be solved by quadrature, which means that the solutions may be expressed
in terms of integrals. This is not the case for order at least two. This
is the main result of Picard--Vessiot theory which was initiated by
Émile Picard and Ernest Vessiot, and whose recent developments are
called differential Galois theory.

The impossibility of solving by quadrature can be compared with the
Abel--Ruffini theorem, which states that an algebraic equation of degree
at least five cannot, in general, be solved by radicals. This analogy
extends to the proof methods and motivates the denomination of
differential Galois theory.

Similarly to the algebraic case, the theory allows deciding which
equations may be solved by quadrature, and if possible solving them.
However, for both theories, the necessary computations are extremely
difficult, even with the most powerful computers.

Nevertheless, the case of order two with rational coefficients has been
completely solved by Kovacic\'s algorithm.

\paragraph{Cauchy--Euler equation}


Cauchy--Euler equations are examples of equations of any order, with
variable coefficients, that can be solved explicitly. These are the
equations of the form

$$x^n y^{(n)}(x) + a_{n-1} x^{n-1} y^{(n-1)}(x) + \cdots + a_0 y(x) = 0,$$
where $a_0, \ldots, a_{n-1}$ are constant coefficients.

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Linear_differential_equation}{Linear differential equation,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license
\begin{itemize}
    \item \href{https://en.wikibooks.org/wiki/Ordinary_Differential_Equations/First_Order_Linear_1}{First Order Linear 1, Wikibooks: Differential
\end{itemize}
    Equations}
    under a CC BY-SA license"

%%===============================================================
\section{Variation of Parameters (2nd Order)}
\label{sec:variation-of-parameters-2nd-order}
%%===============================================================

"In mathematics, \textbf{variation of parameters}, also known as **variation
of constants**, is a general method to solve inhomogeneous linear
ordinary differential equations.

For first-order inhomogeneous linear differential equations it is
usually possible to find solutions via integrating factors or
undetermined coefficients with considerably less effort, although those
methods leverage heuristics that involve guessing and do not work for
all inhomogeneous linear differential equations.

Variation of parameters extends to linear partial differential equations
as well, specifically to inhomogeneous problems for linear evolution
equations like the heat equation, wave equation, and vibrating plate
equation. In this setting, the method is more often known as Duhamel\'s
principle, named after Jean-Marie Duhamel (1797--1872) who first applied
the method to solve the inhomogeneous heat equation. Sometimes variation
of parameters itself is called Duhamel\'s principle and vice versa.

\paragraph{Intuitive explanation}


Consider the equation of the forced dispersionless spring, in suitable
units:

$$x''(t) + x(t) = F(t).$$ Here `{{math|''x''}}`{=mediawiki} is the
displacement of the spring from the equilibrium $x = 0$, and
`{{math|''F''(''t'')}}`{=mediawiki} is an external applied force that
depends on time. When the external force is zero, this is the
homogeneous equation (whose solutions are linear combinations of sines
and cosines, corresponding to the spring oscillating with constant total
energy).

We can construct the solution physically, as follows. Between times
$t=s$ and $t=s+ds$, the momentum corresponding to the solution has a net
change $F(s)\,ds$ (see: Impulse (physics)). A solution to the
inhomogeneous equation, at the present time
`{{math|''t'' > 0}}`{=mediawiki}, is obtained by linearly superposing
the solutions obtained in this manner, for `{{math|''s''}}`{=mediawiki}
going between 0 and `{{math|t}}`{=mediawiki}.

The homogeneous initial-value problem, representing a small impulse
$F(s)\,ds$ being added to the solution at time $t=s$, is

$$x''(t)+x(t)=0,\quad x(s)=0,\ x'(s)=F(s)\,ds.$$ The unique solution to
this problem is easily seen to be $x(t) = F(s)\sin(t-s)\,ds$. The linear
superposition of all of these solutions is given by the integral:

$$x(t) = \int_0^t F(s)\sin(t-s)\,ds.$$

To verify that this satisfies the required equation:

$$x'(t)=\int_0^t F(s)\cos(t-s)\,ds$$

$$x''(t) = F(t) - \int_0^tF(s)\sin(t-s)\,ds = F(t)-x(t),$$ as required
(see: Leibniz integral rule).

The general method of variation of parameters allows for solving an
inhomogeneous linear equation

$$Lx(t)=F(t)$$ by means of considering the second-order linear
differential operator \textit{L} to be the net force, thus the total impulse
imparted to a solution between time \textit{s} and \textit{s}+\textit{ds} is \textit{F}(\textit{s})\textit{ds}.
Denote by $x_s$ the solution of the homogeneous initial value problem

$$Lx(t)=0, \quad x(s)=0,\ x'(s)=F (s)\,ds.$$ Then a particular solution
of the inhomogeneous equation is

$$x (t)=\int_0^t x_s (t)\,ds,$$ the result of linearly superposing the
infinitesimal homogeneous solutions. There are generalizations to higher
order linear differential operators.

In practice, variation of parameters usually involves the fundamental
solution of the homogeneous problem, the infinitesimal solutions $x_s$
then being given in terms of explicit linear combinations of linearly
independent fundamental solutions. In the case of the forced
dispersionless spring, the kernel
$\sin(t-s)=\sin t\cos s - \sin s\cos t$ is the associated decomposition
into fundamental solutions.

\paragraph{Description of method (Second Order)}


Consider a general second order linear nonhomogeneous differential
equation whose coefficient functions $p$, $q$, and $g$ are continuous:

$\begin{aligned} \quad \frac{d^2y}{dt^2} + p(t) \frac{dy}{dt} + q(t) y = g(t) \end{aligned}$

The corresponding second order linear homogeneous differential equation
is thus $\frac{d^2y}{dt^2} + p(t) \frac{dy}{dt} + q(t) y = 0$. Suppose
that we know the general solution to the corresponding second order
linear homogeneous differential equation in terms of two functions
$y_1(t)$ and $y_2(t)$ which form a fundamental set of solutions to the
corresponding second order linear homogeneous differential equation, say
$y_h(t) = Cy_1(t) + Dy_2(t)$.

We will now replace the constants $C$ and $D$ with functions, $u_1(t)$
and $u_2(t)$ to get:

$\begin{aligned} \quad Y(t) = u_1(t) y_1(t) + u_2(t) y_2(t) \end{aligned}$

We then want to try and determine what functions $u_1(t)$ and $u_2(t)$
make $Y(t) = u_1(t)y_1(t) + y_2(t)y_2(t)$ a particular solution to our
original second order linear nonhomogeneous differential equation. We
first differentiate $y$ and apply the product rule where appropriate to
get:

$\begin{aligned} \quad y' = u_1'(t)y_1(t) + u_1(t)y_1'(t) + u_2'(t)y_2(t) + u_2(t)y_2'(t) \end{aligned}$

Now we will set the terms containing the derivatives of the functions
$u_1'(t)$ and $u_2'(t)$ to equal zero, that is
$u_1'(t)y_1(t) + u_2'(t)y_2(t) = 0$. Note that this is a rather hefty
assumption, however, this assumption is not rash as we\'re looking only
for a particular solution, namely one for which this property holds.
We\'ll see that making this assumption does not lead to any
contradictions, and so:

$\begin{aligned} \quad y' = u_1(t)y_1'(t) + u_2(t)y_2'(t) \end{aligned}$

We now differentiate again by applying the product rule where
appropriate to get the second derivative of $y$:

$\begin{aligned} \quad y'' = u_1'(t)y_1'(t) + u_1(t)y_1''(t) + u_2'(t)y_2'(t) + u_2(t)y_2''(t) \end{aligned}$

We will now plug in $y$, $y'$, and $y''$ into our second order linear
nonhomogeneous differential equation to get that:

$\begin{aligned} \quad \frac{d^2y}{dt^2} + p(t) \frac{dy}{dt} + q(t)y = g(t) \\\\ \quad \left [ u_1'(t)y_1'(t) + u_1(t)y_1''(t) + u_2'(t)y_2'(t) + u_2(t)y_2''(t) \right ] + p(t) \left [ u_1(t)y_1'(t) + u_2(t)y_2'(t) \right ] + q(t) \left [ u_1(t) y_1(t) + u_2(t) y_2(t) \right ] = g(t) \\\\ \quad u_1(t) \underbrace{\left [ y_1''(t) + p(t)y_1'(t) + q(t)y_1(t) \right ]}_{=0} + u_2(t) \underbrace{\left [ y_2''(t) + p(t)y_2'(t) + q(t)y_2(t) \right ]}_{=0} + \left [ u_1'(t)y_1'(t) + u_2'(t)y_2'(t) \right ] = g(t) \\\\ \quad u_1'(t)y_1'(t) + u_2'(t)y_2'(t) = g(t) \end{aligned}$

Now recall that we supposed that $u_1'(t)y_1(t) + u_2'(t)y_2(t) = 0$. To
solve for $u_1'(t)$ and $u_2'(t)$, then all we need to do is solve the
following system of equations:

$\begin{aligned} u_1'(t)y_1(t) + u_2'(t)y_2(t) = 0 \\\\ u_1'(t)y_1'(t) + u_2'(t)y_2'(t) = g(t) \end{aligned}$

Recall that a unique solution exists provided that the determinant
$\begin{vmatrix} y_1(t) & y_2(t)\\\\ y_1'(t) & y_2'(t) \end{vmatrix}$
is nonzero. But this determinant is identically the Wronskian
$W(y_1, y_2)$, and it is assumed that this Wronskian is nonzero since
$y_1$ and $y_2$ form the general solution of the corresponding second
order linear homogeneous differential equation, and so by apply
Cramer\'s rule, the values of $u_1'(t)$ and $u_2'(t)$ are:

$\begin{aligned} \quad u_1'(t) = \frac{\begin{vmatrix}0 & y_2(t)\\\\ g(t) & y_2'(t) \end{vmatrix}}{W(y_1, y_2)} = - \frac{y_2(t)g(t)}{W(y_1, y_2)} \quad , \quad u_2'(t) = \frac{\begin{vmatrix} y_1(t) & 0\\\\ y_1'(t) & g(t) \end{vmatrix}}{W(y_1, y_2)} = \frac{y_1(t)g(t)}{W(y_1, y_2)} \end{aligned}$

We now integrate both sides of each equation above, and for constants
$A$ and $B$, we get $u_1(t)$ and $u_2(t)$:

$\begin{aligned} \quad u_1(t) = -\int \frac{y_2(t)g(t)}{W(y_1, y_2)}  dt + A \quad , \quad u_2(t) = \int \frac{y_1(t)g(t)}{W(y_1, y_2)}  dt + B \end{aligned}$

Therefore, a particular solution to our second order linear
nonhomogeneous differential equation is:

$\begin{aligned} \quad Y(t) = u_1(t)y_1(t) + u_2(t)y_2(t) \\\\ \quad Y(t) = y_1(t) \left ( -\int \frac{y_2(t)g(t)}{W(y_1, y_2)}  dt + A \right ) + y_2(t) \left ( \int \frac{y_1(t)g(t)}{W(y_1, y_2)}  dt + B \right ) \\\\ \quad Y(t) = -y_1(t) \int \frac{y_2(t)g(t)}{W(y_1, y_2)}  dt + y_2(t) \int \frac{y_1(t)g(t)}{W(y_1, y_2)}  dt  \underbrace{-Ay_1(t) + By_2(t)}_{=0} \\\\ \quad Y(t) = -y_1(t) \int \frac{y_2(t)g(t)}{W(y_1, y_2)}  dt + y_2(t) \int \frac{y_1(t)g(t)}{W(y_1, y_2)}  dt \end{aligned}$

And finally, the general solution to our differential equation will be:

$\begin{aligned} \quad y = Cy_1(t) + Dy_2(t) + Y(t) \\\\ \quad y = Cy_1(t) + Dy_2(t) -y_1(t) \int \frac{y_2(t)g(t)}{W(y_1, y_2)}  dt + y_2(t) \int \frac{y_1(t)g(t)}{W(y_1, y_2)}  dt \end{aligned}$

Note that the method of variation of parameters is useful provided that
the general solution to the corresponding second order linear
homogeneous differential equation is easy to solve, and provided that
the two integrals in the formula above are relatively simply to compute.

\paragraph{Examples}


\paragraph{First-order equation}


$$y' + p(x)y = q(x)$$ The general solution of the corresponding
homogeneous equation (written below) is the complementary solution to
our original (inhomogeneous) equation:

:   $y' + p(x)y = 0$.

This homogeneous differential equation can be solved by different
methods, for example separation of variables:

$$\frac{d}{dx} y + p(x)y = 0$$

$$\frac{dy}{dx}=-p(x)y$$

$${dy \over y} = -{p(x)\,dx},$$

$$\int \frac{1}{ y} \, dy = -\int p(x) \, dx$$

$$\ln |y| = -\int p(x) \, dx + C$$

$$y = \pm e^{-\int p(x) \, dx +C } = C_0 e^{-\int p(x) \, dx}$$ The
complementary solution to our original equation is therefore:

$$y_c = C_0 e^{-\int p(x) \, dx}$$ Now we return to solving the
non-homogeneous equation:

:   $y' + p(x)y = q(x)$

Using the method variation of parameters, the particular solution is
formed by multiplying the complementary solution by an unknown function
\textit{C}(\textit{x}):

$$y_p = C(x) e^{-\int p(x) \, dx}$$ By substituting the particular
solution into the non-homogeneous equation, we can find \textit{C}(\textit{x}):

:   $C' (x) e^{-\int p(x) \, dx} - C(x) p(x) e^{-\int p(x) \, dx} + p(x) C(x) e^{-\int p(x) \, dx} = q(x)$

<!-- -->

:   $C' (x) e^{-\int p(x) \, dx} = q(x)$

<!-- -->

:   $C' (x) = q(x) e^{\int p(x) \, dx}$

<!-- -->

:   $C(x) =\int q(x) e^{\int p(x) \, dx} \, dx + C_1$

We only need a single particular solution, so we arbitrarily select
$C_1=0$ for simplicity. Therefore the particular solution is:

$$y_p =e^{-\int p(x) \, dx} \int q(x) e^{\int p(x) \, dx} \, dx$$ The
final solution of the differential equation is:

$$\begin{aligned}
y &= y_c + y_p\\\&=C_0 e^{-\int p(x) \, dx} + e^{-\int p(x) \, dx} \int q(x) e^{\int p(x) \, dx} \, dx
\end{aligned}$$

This recreates the method of integrating factors.

\paragraph{Specific second-order equation}


Let us solve

$$y''+4y'+4y = \cosh x$$

We want to find the general solution to the differential equation, that
is, we want to find solutions to the homogeneous differential equation

:   $y''+4y'+4y=0.$

The characteristic equation is:

:   $\lambda^2+4\lambda+4=(\lambda+2)^2=0$

Since $\lambda=-2$ is a repeated root, we have to introduce a factor of
\textit{x} for one solution to ensure linear independence: \textit{u}\textasciitilde{}1\textasciitilde{} = \textit{e}^-2\textit{x}^
and \textit{u}\textasciitilde{}2\textasciitilde{} = \textit{xe}^-2\textit{x}^. The Wronskian of these two functions is

:   $W=\begin{vmatrix}
      e^{-2x} \& xe^{-2x} \\\    -2e^{-2x} \& -e^{-2x}(2x-1)\\\    \end{vmatrix} = -e^{-2x}e^{-2x}(2x-1)+2xe^{-2x}e^{-2x} = e^{-4x}.$

Because the Wronskian is non-zero, the two functions are linearly
independent, so this is in fact the general solution for the homogeneous
differential equation (and not a mere subset of it).

We seek functions \textit{A}(\textit{x}) and \textit{B}(\textit{x}) so
\textit{A}(\textit{x})\textit{u}\textasciitilde{}1\textasciitilde{} + \textit{B}(\textit{x})\textit{u}\textasciitilde{}2\textasciitilde{} is a particular solution of the
non-homogeneous equation. We need only calculate the integrals

$$A(x) = - \int {1\over W} u_2(x) b(x)\,\mathrm dx, \; B(x) = \int {1 \over W} u_1(x)b(x)\,\mathrm dx$$

Recall that for this example

$$b(x) = \cosh x$$

That is,

$$A(x) = - \int {1\over e^{-4x}} xe^{-2x} \cosh x \,\mathrm dx = - \int xe^{2x}\cosh x \,\mathrm dx = -{1\over 18}e^x\left(9(x-1)+e^{2x}(3x-1)\right)+C_1$$

$$B(x) = \int {1 \over e^{-4x}} e^{-2x} \cosh x \,\mathrm dx = \int e^{2x}\cosh x\,\mathrm dx ={1\over 6}e^x\left(3+e^{2x}\right)+C_2$$

where $C_1$ and $C_2$ are constants of integration.

\paragraph{General second-order equation}


We have a differential equation of the form

$$u''+p(x)u'+q(x)u=f(x)$$

and we define the linear operator

$$L=D^2+p(x)D+q(x)$$

where \textit{D} represents the differential operator. We therefore have to
solve the equation $L u(x)=f(x)$ for $u(x)$, where $L$ and $f(x)$ are
known.

We must solve first the corresponding homogeneous equation:

$$u''+p(x)u'+q(x)u=0$$

by the technique of our choice. Once we\'ve obtained two linearly
independent solutions to this homogeneous differential equation (because
this ODE is second-order) --- call them \textit{u}\textasciitilde{}1\textasciitilde{} and \textit{u}\textasciitilde{}2\textasciitilde{} --- we can
proceed with variation of parameters.

Now, we seek the general solution to the differential equation $u_G(x)$
which we assume to be of the form

$$u_G(x)=A(x)u_1(x)+B(x)u_2(x).$$

Here, $A(x)$ and $B(x)$ are unknown and $u_1(x)$ and $u_2(x)$ are the
solutions to the homogeneous equation. (Observe that if $A(x)$ and
$B(x)$ are constants, then $Lu_G(x)=0$.) Since the above is only one
equation and we have two unknown functions, it is reasonable to impose a
second condition. We choose the following:

$$A'(x)u_1(x)+B'(x)u_2(x)=0.$$

Now,

$$\begin{aligned}
u_G'(x) &= \left (A(x)u_1(x)+B(x)u_2(x) \right )' \\\&= \left (A(x)u_1(x) \right )'+ \left (B(x)u_2(x) \right )'\\\&=A'(x)u_1(x)+A(x)u_1'(x)+B'(x)u_2(x)+B(x)u_2'(x)\\\&=A'(x)u_1(x)+B'(x)u_2(x)+A(x)u_1'(x)+B(x)u_2'(x) \\\&= A(x)u_1'(x)+B(x)u_2'(x)
\end{aligned}$$

Differentiating again (omitting intermediary steps)

$$u_G''(x)=A(x)u_1''(x)+B(x)u_2''(x)+A'(x)u_1'(x)+B'(x)u_2'(x).$$

Now we can write the action of \textit{L} upon \textit{u}\textasciitilde{}\textit{G}\textasciitilde{} as

$$Lu_G=A(x)Lu_1(x)+B(x)Lu_2(x)+A'(x)u_1'(x)+B'(x)u_2'(x).$$

Since \textit{u}\textasciitilde{}1\textasciitilde{} and \textit{u}\textasciitilde{}2\textasciitilde{} are solutions, then

$$Lu_G=A'(x)u_1'(x)+B'(x)u_2'(x).$$

We have the system of equations

$$\begin{bmatrix}
u_1(x)  & u_2(x) \\\u_1'(x) & u_2'(x) \end{bmatrix}
\begin{bmatrix}
A'(x) \\\B'(x)\end{bmatrix} =
\begin{bmatrix} 0 \\\\ f \end{bmatrix}.$$

Expanding,

$$\begin{bmatrix}
A'(x)u_1(x)+B'(x)u_2(x)\\\A'(x)u_1'(x)+B'(x)u_2'(x) \end{bmatrix}
= \begin{bmatrix} 0\\\\f\end{bmatrix}.$$

So the above system determines precisely the conditions

$$A'(x)u_1(x)+B'(x)u_2(x)=0.$$

$$A'(x)u_1'(x)+B'(x)u_2'(x)=Lu_G=f.$$

We seek \textit{A}(\textit{x}) and \textit{B}(\textit{x}) from these conditions, so, given

$$\begin{bmatrix}
u_1(x)  & u_2(x) \\\u_1'(x) & u_2'(x)
\end{bmatrix}
\begin{bmatrix}
A'(x) \\\B'(x)\end{bmatrix} =
\begin{bmatrix}
0\\\f\end{bmatrix}$$

we can solve for (\textit{A}'(\textit{x}), \textit{B}'(\textit{x}))^T^, so

$$\begin{bmatrix} A'(x) \\\\ B'(x) \end{bmatrix} =
\begin{bmatrix}
u_1(x)  & u_2(x) \\\u_1'(x) & u_2'(x)
\end{bmatrix}^{-1}
\begin{bmatrix} 0\\\\ f \end{bmatrix} =\frac{1}{W} \begin{bmatrix}
u_2'(x)  & -u_2(x) \\\-u_1'(x) & u_1(x) \end{bmatrix}
\begin{bmatrix} 0\\\\ f \end{bmatrix},$$

where \textit{W} denotes the Wronskian of \textit{u}\textasciitilde{}1\textasciitilde{} and \textit{u}\textasciitilde{}2\textasciitilde{}. (We know that \textit{W}
is nonzero, from the assumption that \textit{u}\textasciitilde{}1\textasciitilde{} and \textit{u}\textasciitilde{}2\textasciitilde{} are linearly
independent.) So,

$$\begin{aligned}
A'(x) &= - {1\over W} u_2(x) f(x), & B'(x) &= {1 \over W} u_1(x)f(x) \\\A(x)  &= - \int {1\over W} u_2(x) f(x)\,\mathrm dx, & B(x) &= \int {1 \over W} u_1(x)f(x)\,\mathrm dx
\end{aligned}$$

While homogeneous equations are relatively easy to solve, this method
allows the calculation of the coefficients of the general solution of
the \textit{in} homogeneous equation, and thus the complete general solution of
the inhomogeneous equation can be determined.

Note that $A(x)$ and $B(x)$ are each determined only up to an arbitrary
additive constant (the constant of integration). Adding a constant to
$A(x)$ or $B(x)$ does not change the value of $Lu_G(x)$ because the
extra term is just a linear combination of \textit{u}\textasciitilde{}1\textasciitilde{} and \textit{u}\textasciitilde{}2\textasciitilde{}, which is a
solution of $L$ by definition.

\paragraph{Higher Order}


The method of variation of parameters can also be applied to higher
order differential equations. The process can be derived similarly.
Suppose that we have a higher order differential equation of the
following form:

$\begin{aligned} \quad \frac{d^ny}{dt^n} + p_1(t) \frac{d^{n-1}}{dt^{n-1}} + ... + p_n(t) y = g(t) \end{aligned}$

We first solve the corresponding homogeneous differential equation to
get $y_h(t) = C_1 y_1(t) + C_2y_2(t) + ... + C_ny_n(t)$. The functions
$y_1(t)$, $y_2(t)$, ..., $y_n(t)$ form a fundamental set. Assume a
particular solution to the nonhomogeneous differential equation is of
the form:

$\begin{aligned} \quad Y(t) = u_1(t)y_1(t) + u_2(t)y_2(t) + ... + u_n(t)y_n(t) \end{aligned}$

We then solve the following system of equations for the functions
$u_1'(t)$, $u_2'(t)$, ..., $u_n'(t)$.

$\begin{aligned} \quad \left\{\begin{matrix} u_1'(t)y_1(t) + u_2'(t)y_2(t) + ... + u_n'(t)y_n(t) = 0 \\\\ u_1'(t)y_1'(t) + u_2'(t)y_2'(t) + ... + u_n'(t)y_n'(t) = 0 \\\\ \vdots \\\\ u_1'(t)y_1^{(n-1)}(t) + u_2'(t)y_2^{(n-1)}(t) + ... + u_n'(t)y_n^{(n-1)}(t) = g(t) \end{matrix}\right. \end{aligned}$

Once again, a unique solution is guaranteed since $y_1(t)$, $y_2(t)$,
..., $y_n(t)$ form a fundamental set of solutions implies the Wronskian
$W(y_1, y_2, ..., y_n)$ is nonzero. Furthermore, it should be noted that
the system above can be solved for with row reduction (if the process is
simple) or more commonly by applying Cramer\'s rule once again.

We then integrate each of the functions $u_1'(t)$, $u_2'(t)$, ...,
$u_n'(t)$ to obtain $u_1(t)$, $u_2(t)$, ..., $u_n(t)$. Lastly, we obtain
that $Y(t) = u_1(t)y_1(t) + u_2(t)y_2(t) + ... + u_n(t)y_n(t)$ as our
particular solution.

We will now look at an example of using the method of variation of
parameters for higher order nonhomogeneous differential equations.

\paragraph{Example 1}


``{=html}Solve the third order linear nonhomogeneous
differential equation $y''' + y' = \tan t$ using the method of variation
of parameters.``{=html}

For the corresponding homogeneous differential equation $y''' + y' = 0$
we have that the characteristic equation is $r^3 + r = 0$ which can be
factored as $r(r^2 + 1) = 0$ and so $r = 0$ or $r = \pm i$. The solution
to the corresponding homogeneous differential equation is therefore:

$\begin{aligned} \quad y(t) = P + Q\cos t + R \sin t \end{aligned}$

Furthermore, $y_1 = e^{0t} = 1$, $y_2 = e^{0t} \cos t= \cos t$, and
$y_3 = e^{0t} \sin t = \sin t$.

Now let\'s try to find a particular solution to this differential
equation. We have that:

$\begin{aligned} \quad y_p = u_1y_1 + u_2y_2 + u_3y_3 \\\\ \quad y_p = u_1 + u_2 \cos t + u_3 \sin t \end{aligned}$

Thus we want to solve the following system of equations:

$\begin{aligned} \quad u_1' (1) + u_2' \cos t + u_3' \sin t = 0 \\\\ \quad u_1' (0) + u_2' (-\sin t) + u_3' (\cos t) = 0 \\\\ \quad u_1' (0) + u_2' (\cos t) + u_3' (\sin t) = \tan t \end{aligned}$

We will use Cramer\'s rule in order to solve for $u_1'$, $u_2'$, and
$u_3'$. We first find the corresponding Wronskian:

$\begin{aligned} \quad W = \begin{bmatrix} 1 & \cos t & \sin t \\\\ 0 & -\sin t & \cos t \\\\ 0 & -\cos t & -\sin t \end{bmatrix} = \begin{bmatrix} -\sin t & \cos t \\\\ -\cos t & -\sin t \end{bmatrix} = 1 \end{aligned}$

Now we have that:

$\begin{aligned} \quad u_1' = \frac{\begin{bmatrix} 0 & \cos t & \sin t \\\\ 0 & -\sin t & \cos t \\\\ \tan t & -\cos t & -\sin t \end{bmatrix}}{1} = \tan t \begin{bmatrix} \cos t & \sin t \\\\ -\sin t & \cos t \end{bmatrix} = \tan t \end{aligned}$

$\begin{aligned} \quad u_2' = \frac{\begin{bmatrix} 1 & 0 & \sin t \\\\ 0 & 0 & \cos t \\\\ 0 & \tan t & -\sin t \end{bmatrix}}{1} = - \begin{bmatrix} 1 & 0 & \sin t \\\\ 0 & \tan t & -\sin t \\\\ 0 & 0 & \cos t \end{bmatrix} = -\tan t \cos t = -\frac{\sin t}{\cos t} \cos t = -\sin t \end{aligned}$

$\begin{aligned} \quad u_3' = \frac{\begin{bmatrix} 1 & \cos t & 0 \\\\ 0 & -\sin t & 0 \\\\ 0 & -\cos t & \tan t \end{bmatrix}}{1} = \begin{bmatrix} -\sin t & 0 \\\\ -\cos t & \tan t \end{bmatrix} = -\sin t \tan t = -\frac{\sin^2 t}{\cos t} \end{aligned}$

We will now integrate $u_1'$, $u_2'$, and $u_3'$ to get:

$\begin{aligned} \quad \int u_1'(t)   dt = \int \tan t   dt = \ln \mid \sec t \mid + C \end{aligned}$

$\begin{aligned} \quad \int u_2'(t)   dt = \int - \sin t   dt = \cos t + D \end{aligned}$

$\begin{aligned} \quad \int u_3'(t)   dt = \int -\frac{\sin^2 t}{\cos t}   dt = \int \frac{\cos^2 t - 1}{\cos t}   dt = \int \left ( \cos t - \sec t \right )   dt = \sin t + \ln \mid \sec t + \tan t \mid + E \end{aligned}$

Thus we have that:

$\begin{aligned} \quad y(t) = \left ( \ln \mid \sec t \mid + C \right ) (1) + \left ( \cos t + D \right ) \cos t + \left ( \sin t + \ln \mid \sec t + \tan t \mid + E \right ) \sin t \end{aligned}$

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Variation_of_parameters}{Variation of parameters,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license
\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Variation_of_parameters}{The Method of Variation of Parameters,
\end{itemize}
    mathonline.wikidot.com}
    under a CC BY-SA license
\begin{itemize}
    \item \href{http://mathonline.wikidot.com/the-method-of-variation-of-parameters-for-higher-order-nonho}{Variation of Parameters for Higher Order,
\end{itemize}
    mathonline.wikidot.com}
    under a CC BY-SA license"

%%===============================================================
\section{Method of Undetermined Coefficients (2nd Order)}
\label{sec:method-of-undetermined-coefficients-2nd-order}
%%===============================================================

"In mathematics, the \textbf{method of undetermined coefficients} is an
approach to finding a particular solution to certain nonhomogeneous
ordinary differential equations and recurrence relations. It is closely
related to the annihilator method, but instead of using a particular
kind of differential operator (the annihilator) in order to find the
best possible form of the particular solution, a \""guess\"" is made as to
the appropriate form, which is then tested by differentiating the
resulting equation. For complex equations, the annihilator method or
variation of parameters is less time-consuming to perform.

Undetermined coefficients is not as general a method as variation of
parameters, since it only works for differential equations that follow
certain forms.

\paragraph{Description of the method}


Consider a linear non-homogeneous ordinary differential equation of the
form

$$\sum_{i=0}^n c_i y^{(i)} + y^{(n+1)} = g(x)$$

:   where $y^{(i)}$ denotes the i-th derivative of $y$, and $c_i$
    denotes a function of $x$.

The method of undetermined coefficients provides a straightforward
method of obtaining the solution to this ODE when two criteria are met:

1.  $c_i$ are constants.
2.  \textit{g}(\textit{x}) is a constant, a polynomial function, exponential function
    $e^{\alpha x}$, sine or cosine functions $\sin{\beta x}$ or
    $\cos{\beta x}$, or finite sums and products of these functions
    (${\alpha}$, ${\beta}$ constants).

The method consists of finding the general homogeneous solution $y_c$
for the complementary linear homogeneous differential equation

$$\sum_{i=0}^n c_i y^{(i)} + y^{(n+1)} = 0,$$

and a particular integral $y_p$ of the linear non-homogeneous ordinary
differential equation based on $g(x)$. Then the general solution $y$ to
the linear non-homogeneous ordinary differential equation would be

$$y = y_c + y_p.$$

If $g(x)$ consists of the sum of two functions $h(x) + w(x)$ and we say
that $y_{p_1}$ is the solution based on $h(x)$ and $y_{p_2}$ the
solution based on $w(x)$. Then, using a superposition principle, we can
say that the particular integral $y_p$ is

$$y_p = y_{p_1} + y_{p_2}.$$

\paragraph{Typical forms of the particular integral}


In order to find the particular integral, we need to \'guess\' its form,
with some coefficients left as variables to be solved for. This takes
the form of the first derivative of the complementary function. Below is
a table of some typical functions and the solution to guess for them.

  Function of \textit{x}                                                                                                           Form for \textit{y}
  ------------------------------------------------------------------------------------------------------------------------- ------------------------------------------------------------------------------------------------------------------
  $k e^{a x}\!$                                                                                                             $C e^{a x}\!$
  $k x^n, \; n = 0, 1, 2,\ldots\!$                                                                                           $\sum_{i=0}^n K_i x^i \!$
  $k \cos(a x) \text{ or } k \sin(a x) \!$                                                                                  $K \cos(a x) + M \sin(a x) \!$
  $k e^{a x} \cos(b x) \text{ or } ke^{a x} \sin(b x) \!$                                                                   $e^{a x} (K \cos(b x) + M \sin(b x)) \!$
  $\left(\sum_{i=0}^n k_i x^i\right) \cos(b x) \text{ or }\ \left(\sum_{i=0}^n k_i x^i\right) \sin(b x) \!$                 $\left(\sum_{i=0}^n Q_i x^i\right) \cos(b x) + \left(\sum_{i=0}^n R_i x^i\right) \sin(b x)$
  $\left(\sum_{i=0}^n k_i x^i\right) e^{a x} \cos(b x) \text{ or } \left(\sum_{i=0}^n k_i x^i\right) e^{a x} \sin(b x)\!$   $e^{a x} \left(\left(\sum_{i=0}^n Q_i x^i\right) \cos(b x) + \left(\sum_{i=0}^n R_i x^i\right) \sin(b x)\right)$

If a term in the above particular integral for \textit{y} appears in the
homogeneous solution, it is necessary to multiply by a sufficiently
large power of \textit{x} in order to make the solution independent. If the
function of \textit{x} is a sum of terms in the above table, the particular
integral can be guessed using a sum of the corresponding terms for \textit{y}.

\paragraph{Examples}


\paragraph{Example 1}


Find a particular integral of the equation

$$y'' + y = t \cos t.$$

The right side \textit{t} cos \textit{t} has the form

$$P_n e^{\alpha t} \cos{\beta t}$$

with \textit{n} = 2, \textit{a} = 0, and \textit{ß} = 1.

Since \textit{a} + \textit{iß} = \textit{i} is \textit{a simple root} of the characteristic equation

$$\lambda^2 + 1 = 0$$

we should try a particular integral of the form

$$\begin{aligned}
y_p &= t \left [F_1 (t) e^{\alpha t} \cos{\beta t} + G_1 (t) e^{\alpha t} \sin{\beta t} \right ] \\\&= t \left [F_1 (t) \cos t + G_1 (t) \sin t \right ] \\\&= t \left [ \left (A_0 t + A_1 \right ) \cos t + \left (B_0 t + B_1 \right ) \sin t \right ] \\\&= \left (A_0 t^2 + A_1 t \right ) \cos t + \left (B_0 t^2 + B_1 t \right) \sin t.
\end{aligned}$$

Substituting \textit{y}\textasciitilde{}\textit{p}\textasciitilde{} into the differential equation, we have the
identity

$$\begin{aligned}
t \cos t &= y_p'' + y_p \\\&= \left  [ \left(A_0 t^2 + A_1 t \right ) \cos t + \left (B_0 t^2 + B_1 t \right ) \sin t \right ]'' + \left[\left(A_0 t^2 + A_1 t \right ) \cos t + \left(B_0 t^2 + B_1 t \right ) \sin t \right ] \\\&=  \left [2A_0 \cos t + 2 \left (2A_0 t + A_1 \right )(-\sin t) + \left (A_0 t^2 + A_1 t \right )(-\cos t) + 2B_0 \sin t + 2 \left (2B_0 t + B_1 \right ) \cos t + \left (B_0 t^2 + B_1 t \right )(- \sin t) \right ] \\\&\qquad +\left[\left(A_0 t^2 + A_1 t \right ) \cos t + \left(B_0 t^2 + B_1 t \right ) \sin t \right ] \\\&= [4B_0 t + (2A_0 + 2B_1)] \cos t + [-4A_0 t + (-2A_1 + 2B_0)] \sin t.
\end{aligned}$$

Comparing both sides, we have

$$\begin{cases} 
1 = 4B_0\\\0 = 2A_0 + 2B_1  \\\0 = -4A_0 \\\0 = -2A_1 + 2B_0
\end{cases}$$

which has the solution

$$A_0 = 0, \quad A_1 = B_0 = \frac{1}{4}, \quad  B_1 = 0.$$

We then have a particular integral

$$y_p = \frac {1} {4} t \cos t + \frac {1}{4} t^2 \sin t.$$

\paragraph{Example 2}


Consider the following linear nonhomogeneous differential equation:

:   $\frac{dy}{dx} = y + e^x.$

This is like the first example above, except that the nonhomogeneous
part ($e^x$) is \textit{not} linearly independent to the general solution of
the homogeneous part ($c_1 e^x$); as a result, we have to multiply our
guess by a sufficiently large power of \textit{x} to make it linearly
independent.

Here our guess becomes:

$$y_p = A x e^x.$$

By substituting this function and its derivative into the differential
equation, one can solve for \textit{A}:

$$\frac{d}{dx} \left( A x e^x \right) = A x e^x + e^x$$

$$A x e^x + A e^x = A x e^x + e^x$$

$$A = 1.$$

So, the general solution to this differential equation is:

$$y = c_1 e^x + xe^x.$$

\paragraph{Example 3}


Find the general solution of the equation:

$$\frac{dy}{dt} = t^2 - y$$

$t^2$ is a polynomial of degree 2, so we look for a solution using the
same form,

$$y_p = A t^2 + B t + C,$$

Plugging this particular function into the original equation yields,

$$2 A t + B = t^2 - (A t^2 + B t + C),$$

$$2 A t + B =(1-A)t^2 -Bt -C,$$

$$(A-1)t^2 + (2A+B)t + (B+C) = 0.$$

which gives:

$$A-1 = 0, \quad 2A+B =0, \quad  B+C=0.$$

Solving for constants we get:

$$y_p = t^2 - 2 t + 2$$

To solve for the general solution,

$$y= y_p + y_c$$

where $y_c$ is the homogeneous solution $y_c = c_1 e^{-t}$, therefore,
the general solution is:

$$y= t^2 - 2 t + 2 + c_1 e^{-t}$$

\paragraph{Higher Order}


If we have a second order linear nonhomogenous differential equation
whose coefficients were constant, that is
$a \frac{d^2y}{dt^2} + b \frac{dy}{dt} + cy = g(t)$, then to solve this
differential equation, all we need to do is solve the corresponding
second order linear homogenous differential equation
$a \frac{d^2y}{dt^2} + b \frac{dy}{dt} + cy = 0$, and then find a
partial solution by assuming the form of the particular solution. More
precisely, if $g(t)$ was a polynomial, exponential, or sine/cosine
function (or a combination of these), then we could assume a form for
the particular solutions (see the linked page above for more details)
and solve for the coefficients of this form to obtain a particular
solution.

We will now look at this method for higher order linear nonhomogenous
differential equations. Consider the following $n^{\mathrm{th}}$ order
linear nonhomogenous differential equation with coefficients
$a_0, a_1, ..., a_n \in \mathbb{R}$:

$\begin{aligned} \quad a_0 \frac{d^ny}{dt^n} + a_1 \frac{d^{n-1}y}{dt^{n-1}} + ... + a_{n-1}\frac{dy}{dt} + a_n y = g(t) \end{aligned}$

Suppose that $g(t)$ is of a form containing a polynomial, exponential
function, or a sine/cosine function (like with when we were dealing with
the method of undetermined coefficients for second order linear
nonhomogenous differential equations). We can then find a particular
solution $Y(t)$ if we extend the assumed forms of combinations of
polynomials, exponential functions, or sine/cosine functions, and
multiplying by powers of $t$ to ensure our particular solution does not
contain part of a solution to the corresponding higher order linear
homogenous differential equation. We can then solve for these constants
by plugging our assumed form $Y(t)$ into our differential equation
above.

Providing a list of possible forms is trivial, so we will instead look
at some examples of apply this method.

\paragraph{Example 1}


``{=html}Find the general solution to the differential equation
$\frac{d^3y}{dt^3} + \frac{d^2y}{dt^2} + \frac{dy}{dt} + y = e^{-t} + 4t$.``{=html}

We will need to first solve the corresponding third order linear
homogenous differential equation
$\frac{d^3y}{dt^3} + \frac{d^2y}{dt^2} + \frac{dy}{dt} + y = 0$. This
characteristic equation to this differential equation is:

$\begin{aligned} \quad r^3 + r^2 + r + 1 = 0 \end{aligned}$

We can (by trial and error) see that $r_1 = -1$ is a solution to this
characteristic equation. Applying long division with the factor
$(r + 1)$ and we have that our characteristic equation can be written
as:

$\begin{aligned} \quad (r + 1)(r^2 + 1) = 0 \end{aligned}$

Therefore we can see that $r_2 = i$ and $r_3 = -i$. Therefore the
general solution to our third order linear homogenous differential
equation is:

$\begin{aligned} \quad y_h(t) = C_1e^{-t} + C_2\cos(t) + C_3\sin(t) \end{aligned}$

Now we note that $g(t) = e^{-t} + 4t$ has an exponential term and a
cosine term, so we expect the form of our particular solution to be
$Y(t) = Ae^{-t} ...$. We now compute the first, second, and third
derivatives of $Y$.

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Method_of_undetermined_coefficients}{Method of undetermined coefficients,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license"

%%===============================================================
\section{Non-linear 2nd Order ODEs}
\label{sec:non-linear-2nd-order-odes}
%%===============================================================

Non-linear 2nd Order ODEs

%%===============================================================
\section{Variation of Parameters (Higher Order)}
\label{sec:variation-of-parameters-higher-order}
%%===============================================================

"In mathematics, \textbf{variation of parameters}, also known as **variation
of constants**, is a general method to solve inhomogeneous linear
ordinary differential equations.

For first-order inhomogeneous linear differential equations it is
usually possible to find solutions via integrating factors or
undetermined coefficients with considerably less effort, although those
methods leverage heuristics that involve guessing and do not work for
all inhomogeneous linear differential equations.

Variation of parameters extends to linear partial differential equations
as well, specifically to inhomogeneous problems for linear evolution
equations like the heat equation, wave equation, and vibrating plate
equation. In this setting, the method is more often known as Duhamel\'s
principle, named after Jean-Marie Duhamel (1797--1872) who first applied
the method to solve the inhomogeneous heat equation. Sometimes variation
of parameters itself is called Duhamel\'s principle and vice versa.

\paragraph{Intuitive explanation}


Consider the equation of the forced dispersionless spring, in suitable
units:

$$x''(t) + x(t) = F(t).$$ Here `{{math|''x''}}`{=mediawiki} is the
displacement of the spring from the equilibrium $x = 0$, and
`{{math|''F''(''t'')}}`{=mediawiki} is an external applied force that
depends on time. When the external force is zero, this is the
homogeneous equation (whose solutions are linear combinations of sines
and cosines, corresponding to the spring oscillating with constant total
energy).

We can construct the solution physically, as follows. Between times
$t=s$ and $t=s+ds$, the momentum corresponding to the solution has a net
change $F(s)\,ds$ (see: Impulse (physics)). A solution to the
inhomogeneous equation, at the present time
`{{math|''t'' > 0}}`{=mediawiki}, is obtained by linearly superposing
the solutions obtained in this manner, for `{{math|''s''}}`{=mediawiki}
going between 0 and `{{math|t}}`{=mediawiki}.

The homogeneous initial-value problem, representing a small impulse
$F(s)\,ds$ being added to the solution at time $t=s$, is

$$x''(t)+x(t)=0,\quad x(s)=0,\ x'(s)=F(s)\,ds.$$ The unique solution to
this problem is easily seen to be $x(t) = F(s)\sin(t-s)\,ds$. The linear
superposition of all of these solutions is given by the integral:

$$x(t) = \int_0^t F(s)\sin(t-s)\,ds.$$

To verify that this satisfies the required equation:

$$x'(t)=\int_0^t F(s)\cos(t-s)\,ds$$

$$x''(t) = F(t) - \int_0^tF(s)\sin(t-s)\,ds = F(t)-x(t),$$ as required
(see: Leibniz integral rule).

The general method of variation of parameters allows for solving an
inhomogeneous linear equation

$$Lx(t)=F(t)$$ by means of considering the second-order linear
differential operator \textit{L} to be the net force, thus the total impulse
imparted to a solution between time \textit{s} and \textit{s}+\textit{ds} is \textit{F}(\textit{s})\textit{ds}.
Denote by $x_s$ the solution of the homogeneous initial value problem

$$Lx(t)=0, \quad x(s)=0,\ x'(s)=F (s)\,ds.$$ Then a particular solution
of the inhomogeneous equation is

$$x (t)=\int_0^t x_s (t)\,ds,$$ the result of linearly superposing the
infinitesimal homogeneous solutions. There are generalizations to higher
order linear differential operators.

In practice, variation of parameters usually involves the fundamental
solution of the homogeneous problem, the infinitesimal solutions $x_s$
then being given in terms of explicit linear combinations of linearly
independent fundamental solutions. In the case of the forced
dispersionless spring, the kernel
$\sin(t-s)=\sin t\cos s - \sin s\cos t$ is the associated decomposition
into fundamental solutions.

\paragraph{Description of method (Second Order)}


Consider a general second order linear nonhomogeneous differential
equation whose coefficient functions $p$, $q$, and $g$ are continuous:

$\begin{aligned} \quad \frac{d^2y}{dt^2} + p(t) \frac{dy}{dt} + q(t) y = g(t) \end{aligned}$

The corresponding second order linear homogeneous differential equation
is thus $\frac{d^2y}{dt^2} + p(t) \frac{dy}{dt} + q(t) y = 0$. Suppose
that we know the general solution to the corresponding second order
linear homogeneous differential equation in terms of two functions
$y_1(t)$ and $y_2(t)$ which form a fundamental set of solutions to the
corresponding second order linear homogeneous differential equation, say
$y_h(t) = Cy_1(t) + Dy_2(t)$.

We will now replace the constants $C$ and $D$ with functions, $u_1(t)$
and $u_2(t)$ to get:

$\begin{aligned} \quad Y(t) = u_1(t) y_1(t) + u_2(t) y_2(t) \end{aligned}$

We then want to try and determine what functions $u_1(t)$ and $u_2(t)$
make $Y(t) = u_1(t)y_1(t) + y_2(t)y_2(t)$ a particular solution to our
original second order linear nonhomogeneous differential equation. We
first differentiate $y$ and apply the product rule where appropriate to
get:

$\begin{aligned} \quad y' = u_1'(t)y_1(t) + u_1(t)y_1'(t) + u_2'(t)y_2(t) + u_2(t)y_2'(t) \end{aligned}$

Now we will set the terms containing the derivatives of the functions
$u_1'(t)$ and $u_2'(t)$ to equal zero, that is
$u_1'(t)y_1(t) + u_2'(t)y_2(t) = 0$. Note that this is a rather hefty
assumption, however, this assumption is not rash as we\'re looking only
for a particular solution, namely one for which this property holds.
We\'ll see that making this assumption does not lead to any
contradictions, and so:

$\begin{aligned} \quad y' = u_1(t)y_1'(t) + u_2(t)y_2'(t) \end{aligned}$

We now differentiate again by applying the product rule where
appropriate to get the second derivative of $y$:

$\begin{aligned} \quad y'' = u_1'(t)y_1'(t) + u_1(t)y_1''(t) + u_2'(t)y_2'(t) + u_2(t)y_2''(t) \end{aligned}$

We will now plug in $y$, $y'$, and $y''$ into our second order linear
nonhomogeneous differential equation to get that:

$\begin{aligned} \quad \frac{d^2y}{dt^2} + p(t) \frac{dy}{dt} + q(t)y = g(t) \\\\ \quad \left [ u_1'(t)y_1'(t) + u_1(t)y_1''(t) + u_2'(t)y_2'(t) + u_2(t)y_2''(t) \right ] + p(t) \left [ u_1(t)y_1'(t) + u_2(t)y_2'(t) \right ] + q(t) \left [ u_1(t) y_1(t) + u_2(t) y_2(t) \right ] = g(t) \\\\ \quad u_1(t) \underbrace{\left [ y_1''(t) + p(t)y_1'(t) + q(t)y_1(t) \right ]}_{=0} + u_2(t) \underbrace{\left [ y_2''(t) + p(t)y_2'(t) + q(t)y_2(t) \right ]}_{=0} + \left [ u_1'(t)y_1'(t) + u_2'(t)y_2'(t) \right ] = g(t) \\\\ \quad u_1'(t)y_1'(t) + u_2'(t)y_2'(t) = g(t) \end{aligned}$

Now recall that we supposed that $u_1'(t)y_1(t) + u_2'(t)y_2(t) = 0$. To
solve for $u_1'(t)$ and $u_2'(t)$, then all we need to do is solve the
following system of equations:

$\begin{aligned} u_1'(t)y_1(t) + u_2'(t)y_2(t) = 0 \\\\ u_1'(t)y_1'(t) + u_2'(t)y_2'(t) = g(t) \end{aligned}$

Recall that a unique solution exists provided that the determinant
$\begin{vmatrix} y_1(t) & y_2(t)\\\\ y_1'(t) & y_2'(t) \end{vmatrix}$
is nonzero. But this determinant is identically the Wronskian
$W(y_1, y_2)$, and it is assumed that this Wronskian is nonzero since
$y_1$ and $y_2$ form the general solution of the corresponding second
order linear homogeneous differential equation, and so by apply
Cramer\'s rule, the values of $u_1'(t)$ and $u_2'(t)$ are:

$\begin{aligned} \quad u_1'(t) = \frac{\begin{vmatrix}0 & y_2(t)\\\\ g(t) & y_2'(t) \end{vmatrix}}{W(y_1, y_2)} = - \frac{y_2(t)g(t)}{W(y_1, y_2)} \quad , \quad u_2'(t) = \frac{\begin{vmatrix} y_1(t) & 0\\\\ y_1'(t) & g(t) \end{vmatrix}}{W(y_1, y_2)} = \frac{y_1(t)g(t)}{W(y_1, y_2)} \end{aligned}$

We now integrate both sides of each equation above, and for constants
$A$ and $B$, we get $u_1(t)$ and $u_2(t)$:

$\begin{aligned} \quad u_1(t) = -\int \frac{y_2(t)g(t)}{W(y_1, y_2)}  dt + A \quad , \quad u_2(t) = \int \frac{y_1(t)g(t)}{W(y_1, y_2)}  dt + B \end{aligned}$

Therefore, a particular solution to our second order linear
nonhomogeneous differential equation is:

$\begin{aligned} \quad Y(t) = u_1(t)y_1(t) + u_2(t)y_2(t) \\\\ \quad Y(t) = y_1(t) \left ( -\int \frac{y_2(t)g(t)}{W(y_1, y_2)}  dt + A \right ) + y_2(t) \left ( \int \frac{y_1(t)g(t)}{W(y_1, y_2)}  dt + B \right ) \\\\ \quad Y(t) = -y_1(t) \int \frac{y_2(t)g(t)}{W(y_1, y_2)}  dt + y_2(t) \int \frac{y_1(t)g(t)}{W(y_1, y_2)}  dt  \underbrace{-Ay_1(t) + By_2(t)}_{=0} \\\\ \quad Y(t) = -y_1(t) \int \frac{y_2(t)g(t)}{W(y_1, y_2)}  dt + y_2(t) \int \frac{y_1(t)g(t)}{W(y_1, y_2)}  dt \end{aligned}$

And finally, the general solution to our differential equation will be:

$\begin{aligned} \quad y = Cy_1(t) + Dy_2(t) + Y(t) \\\\ \quad y = Cy_1(t) + Dy_2(t) -y_1(t) \int \frac{y_2(t)g(t)}{W(y_1, y_2)}  dt + y_2(t) \int \frac{y_1(t)g(t)}{W(y_1, y_2)}  dt \end{aligned}$

Note that the method of variation of parameters is useful provided that
the general solution to the corresponding second order linear
homogeneous differential equation is easy to solve, and provided that
the two integrals in the formula above are relatively simply to compute.

\paragraph{Examples}


\paragraph{First-order equation}


$$y' + p(x)y = q(x)$$ The general solution of the corresponding
homogeneous equation (written below) is the complementary solution to
our original (inhomogeneous) equation:

:   $y' + p(x)y = 0$.

This homogeneous differential equation can be solved by different
methods, for example separation of variables:

$$\frac{d}{dx} y + p(x)y = 0$$

$$\frac{dy}{dx}=-p(x)y$$

$${dy \over y} = -{p(x)\,dx},$$

$$\int \frac{1}{ y} \, dy = -\int p(x) \, dx$$

$$\ln |y| = -\int p(x) \, dx + C$$

$$y = \pm e^{-\int p(x) \, dx +C } = C_0 e^{-\int p(x) \, dx}$$ The
complementary solution to our original equation is therefore:

$$y_c = C_0 e^{-\int p(x) \, dx}$$ Now we return to solving the
non-homogeneous equation:

:   $y' + p(x)y = q(x)$

Using the method variation of parameters, the particular solution is
formed by multiplying the complementary solution by an unknown function
\textit{C}(\textit{x}):

$$y_p = C(x) e^{-\int p(x) \, dx}$$ By substituting the particular
solution into the non-homogeneous equation, we can find \textit{C}(\textit{x}):

:   $C' (x) e^{-\int p(x) \, dx} - C(x) p(x) e^{-\int p(x) \, dx} + p(x) C(x) e^{-\int p(x) \, dx} = q(x)$

<!-- -->

:   $C' (x) e^{-\int p(x) \, dx} = q(x)$

<!-- -->

:   $C' (x) = q(x) e^{\int p(x) \, dx}$

<!-- -->

:   $C(x) =\int q(x) e^{\int p(x) \, dx} \, dx + C_1$

We only need a single particular solution, so we arbitrarily select
$C_1=0$ for simplicity. Therefore the particular solution is:

$$y_p =e^{-\int p(x) \, dx} \int q(x) e^{\int p(x) \, dx} \, dx$$ The
final solution of the differential equation is:

$$\begin{aligned}
y &= y_c + y_p\\\&=C_0 e^{-\int p(x) \, dx} + e^{-\int p(x) \, dx} \int q(x) e^{\int p(x) \, dx} \, dx
\end{aligned}$$

This recreates the method of integrating factors.

\paragraph{Specific second-order equation}


Let us solve

$$y''+4y'+4y = \cosh x$$

We want to find the general solution to the differential equation, that
is, we want to find solutions to the homogeneous differential equation

:   $y''+4y'+4y=0.$

The characteristic equation is:

:   $\lambda^2+4\lambda+4=(\lambda+2)^2=0$

Since $\lambda=-2$ is a repeated root, we have to introduce a factor of
\textit{x} for one solution to ensure linear independence: \textit{u}\textasciitilde{}1\textasciitilde{} = \textit{e}^-2\textit{x}^
and \textit{u}\textasciitilde{}2\textasciitilde{} = \textit{xe}^-2\textit{x}^. The Wronskian of these two functions is

:   $W=\begin{vmatrix}
      e^{-2x} \& xe^{-2x} \\\    -2e^{-2x} \& -e^{-2x}(2x-1)\\\    \end{vmatrix} = -e^{-2x}e^{-2x}(2x-1)+2xe^{-2x}e^{-2x} = e^{-4x}.$

Because the Wronskian is non-zero, the two functions are linearly
independent, so this is in fact the general solution for the homogeneous
differential equation (and not a mere subset of it).

We seek functions \textit{A}(\textit{x}) and \textit{B}(\textit{x}) so
\textit{A}(\textit{x})\textit{u}\textasciitilde{}1\textasciitilde{} + \textit{B}(\textit{x})\textit{u}\textasciitilde{}2\textasciitilde{} is a particular solution of the
non-homogeneous equation. We need only calculate the integrals

$$A(x) = - \int {1\over W} u_2(x) b(x)\,\mathrm dx, \; B(x) = \int {1 \over W} u_1(x)b(x)\,\mathrm dx$$

Recall that for this example

$$b(x) = \cosh x$$

That is,

$$A(x) = - \int {1\over e^{-4x}} xe^{-2x} \cosh x \,\mathrm dx = - \int xe^{2x}\cosh x \,\mathrm dx = -{1\over 18}e^x\left(9(x-1)+e^{2x}(3x-1)\right)+C_1$$

$$B(x) = \int {1 \over e^{-4x}} e^{-2x} \cosh x \,\mathrm dx = \int e^{2x}\cosh x\,\mathrm dx ={1\over 6}e^x\left(3+e^{2x}\right)+C_2$$

where $C_1$ and $C_2$ are constants of integration.

\paragraph{General second-order equation}


We have a differential equation of the form

$$u''+p(x)u'+q(x)u=f(x)$$

and we define the linear operator

$$L=D^2+p(x)D+q(x)$$

where \textit{D} represents the differential operator. We therefore have to
solve the equation $L u(x)=f(x)$ for $u(x)$, where $L$ and $f(x)$ are
known.

We must solve first the corresponding homogeneous equation:

$$u''+p(x)u'+q(x)u=0$$

by the technique of our choice. Once we\'ve obtained two linearly
independent solutions to this homogeneous differential equation (because
this ODE is second-order) --- call them \textit{u}\textasciitilde{}1\textasciitilde{} and \textit{u}\textasciitilde{}2\textasciitilde{} --- we can
proceed with variation of parameters.

Now, we seek the general solution to the differential equation $u_G(x)$
which we assume to be of the form

$$u_G(x)=A(x)u_1(x)+B(x)u_2(x).$$

Here, $A(x)$ and $B(x)$ are unknown and $u_1(x)$ and $u_2(x)$ are the
solutions to the homogeneous equation. (Observe that if $A(x)$ and
$B(x)$ are constants, then $Lu_G(x)=0$.) Since the above is only one
equation and we have two unknown functions, it is reasonable to impose a
second condition. We choose the following:

$$A'(x)u_1(x)+B'(x)u_2(x)=0.$$

Now,

$$\begin{aligned}
u_G'(x) &= \left (A(x)u_1(x)+B(x)u_2(x) \right )' \\\&= \left (A(x)u_1(x) \right )'+ \left (B(x)u_2(x) \right )'\\\&=A'(x)u_1(x)+A(x)u_1'(x)+B'(x)u_2(x)+B(x)u_2'(x)\\\&=A'(x)u_1(x)+B'(x)u_2(x)+A(x)u_1'(x)+B(x)u_2'(x) \\\&= A(x)u_1'(x)+B(x)u_2'(x)
\end{aligned}$$

Differentiating again (omitting intermediary steps)

$$u_G''(x)=A(x)u_1''(x)+B(x)u_2''(x)+A'(x)u_1'(x)+B'(x)u_2'(x).$$

Now we can write the action of \textit{L} upon \textit{u}\textasciitilde{}\textit{G}\textasciitilde{} as

$$Lu_G=A(x)Lu_1(x)+B(x)Lu_2(x)+A'(x)u_1'(x)+B'(x)u_2'(x).$$

Since \textit{u}\textasciitilde{}1\textasciitilde{} and \textit{u}\textasciitilde{}2\textasciitilde{} are solutions, then

$$Lu_G=A'(x)u_1'(x)+B'(x)u_2'(x).$$

We have the system of equations

$$\begin{bmatrix}
u_1(x)  & u_2(x) \\\u_1'(x) & u_2'(x) \end{bmatrix}
\begin{bmatrix}
A'(x) \\\B'(x)\end{bmatrix} =
\begin{bmatrix} 0 \\\\ f \end{bmatrix}.$$

Expanding,

$$\begin{bmatrix}
A'(x)u_1(x)+B'(x)u_2(x)\\\A'(x)u_1'(x)+B'(x)u_2'(x) \end{bmatrix}
= \begin{bmatrix} 0\\\\f\end{bmatrix}.$$

So the above system determines precisely the conditions

$$A'(x)u_1(x)+B'(x)u_2(x)=0.$$

$$A'(x)u_1'(x)+B'(x)u_2'(x)=Lu_G=f.$$

We seek \textit{A}(\textit{x}) and \textit{B}(\textit{x}) from these conditions, so, given

$$\begin{bmatrix}
u_1(x)  & u_2(x) \\\u_1'(x) & u_2'(x)
\end{bmatrix}
\begin{bmatrix}
A'(x) \\\B'(x)\end{bmatrix} =
\begin{bmatrix}
0\\\f\end{bmatrix}$$

we can solve for (\textit{A}'(\textit{x}), \textit{B}'(\textit{x}))^T^, so

$$\begin{bmatrix} A'(x) \\\\ B'(x) \end{bmatrix} =
\begin{bmatrix}
u_1(x)  & u_2(x) \\\u_1'(x) & u_2'(x)
\end{bmatrix}^{-1}
\begin{bmatrix} 0\\\\ f \end{bmatrix} =\frac{1}{W} \begin{bmatrix}
u_2'(x)  & -u_2(x) \\\-u_1'(x) & u_1(x) \end{bmatrix}
\begin{bmatrix} 0\\\\ f \end{bmatrix},$$

where \textit{W} denotes the Wronskian of \textit{u}\textasciitilde{}1\textasciitilde{} and \textit{u}\textasciitilde{}2\textasciitilde{}. (We know that \textit{W}
is nonzero, from the assumption that \textit{u}\textasciitilde{}1\textasciitilde{} and \textit{u}\textasciitilde{}2\textasciitilde{} are linearly
independent.) So,

$$\begin{aligned}
A'(x) &= - {1\over W} u_2(x) f(x), & B'(x) &= {1 \over W} u_1(x)f(x) \\\A(x)  &= - \int {1\over W} u_2(x) f(x)\,\mathrm dx, & B(x) &= \int {1 \over W} u_1(x)f(x)\,\mathrm dx
\end{aligned}$$

While homogeneous equations are relatively easy to solve, this method
allows the calculation of the coefficients of the general solution of
the \textit{in} homogeneous equation, and thus the complete general solution of
the inhomogeneous equation can be determined.

Note that $A(x)$ and $B(x)$ are each determined only up to an arbitrary
additive constant (the constant of integration). Adding a constant to
$A(x)$ or $B(x)$ does not change the value of $Lu_G(x)$ because the
extra term is just a linear combination of \textit{u}\textasciitilde{}1\textasciitilde{} and \textit{u}\textasciitilde{}2\textasciitilde{}, which is a
solution of $L$ by definition.

\paragraph{Higher Order}


The method of variation of parameters can also be applied to higher
order differential equations. The process can be derived similarly.
Suppose that we have a higher order differential equation of the
following form:

$\begin{aligned} \quad \frac{d^ny}{dt^n} + p_1(t) \frac{d^{n-1}}{dt^{n-1}} + ... + p_n(t) y = g(t) \end{aligned}$

We first solve the corresponding homogeneous differential equation to
get $y_h(t) = C_1 y_1(t) + C_2y_2(t) + ... + C_ny_n(t)$. The functions
$y_1(t)$, $y_2(t)$, ..., $y_n(t)$ form a fundamental set. Assume a
particular solution to the nonhomogeneous differential equation is of
the form:

$\begin{aligned} \quad Y(t) = u_1(t)y_1(t) + u_2(t)y_2(t) + ... + u_n(t)y_n(t) \end{aligned}$

We then solve the following system of equations for the functions
$u_1'(t)$, $u_2'(t)$, ..., $u_n'(t)$.

$\begin{aligned} \quad \left\{\begin{matrix} u_1'(t)y_1(t) + u_2'(t)y_2(t) + ... + u_n'(t)y_n(t) = 0 \\\\ u_1'(t)y_1'(t) + u_2'(t)y_2'(t) + ... + u_n'(t)y_n'(t) = 0 \\\\ \vdots \\\\ u_1'(t)y_1^{(n-1)}(t) + u_2'(t)y_2^{(n-1)}(t) + ... + u_n'(t)y_n^{(n-1)}(t) = g(t) \end{matrix}\right. \end{aligned}$

Once again, a unique solution is guaranteed since $y_1(t)$, $y_2(t)$,
..., $y_n(t)$ form a fundamental set of solutions implies the Wronskian
$W(y_1, y_2, ..., y_n)$ is nonzero. Furthermore, it should be noted that
the system above can be solved for with row reduction (if the process is
simple) or more commonly by applying Cramer\'s rule once again.

We then integrate each of the functions $u_1'(t)$, $u_2'(t)$, ...,
$u_n'(t)$ to obtain $u_1(t)$, $u_2(t)$, ..., $u_n(t)$. Lastly, we obtain
that $Y(t) = u_1(t)y_1(t) + u_2(t)y_2(t) + ... + u_n(t)y_n(t)$ as our
particular solution.

We will now look at an example of using the method of variation of
parameters for higher order nonhomogeneous differential equations.

\paragraph{Example 1}


``{=html}Solve the third order linear nonhomogeneous
differential equation $y''' + y' = \tan t$ using the method of variation
of parameters.``{=html}

For the corresponding homogeneous differential equation $y''' + y' = 0$
we have that the characteristic equation is $r^3 + r = 0$ which can be
factored as $r(r^2 + 1) = 0$ and so $r = 0$ or $r = \pm i$. The solution
to the corresponding homogeneous differential equation is therefore:

$\begin{aligned} \quad y(t) = P + Q\cos t + R \sin t \end{aligned}$

Furthermore, $y_1 = e^{0t} = 1$, $y_2 = e^{0t} \cos t= \cos t$, and
$y_3 = e^{0t} \sin t = \sin t$.

Now let\'s try to find a particular solution to this differential
equation. We have that:

$\begin{aligned} \quad y_p = u_1y_1 + u_2y_2 + u_3y_3 \\\\ \quad y_p = u_1 + u_2 \cos t + u_3 \sin t \end{aligned}$

Thus we want to solve the following system of equations:

$\begin{aligned} \quad u_1' (1) + u_2' \cos t + u_3' \sin t = 0 \\\\ \quad u_1' (0) + u_2' (-\sin t) + u_3' (\cos t) = 0 \\\\ \quad u_1' (0) + u_2' (\cos t) + u_3' (\sin t) = \tan t \end{aligned}$

We will use Cramer\'s rule in order to solve for $u_1'$, $u_2'$, and
$u_3'$. We first find the corresponding Wronskian:

$\begin{aligned} \quad W = \begin{bmatrix} 1 & \cos t & \sin t \\\\ 0 & -\sin t & \cos t \\\\ 0 & -\cos t & -\sin t \end{bmatrix} = \begin{bmatrix} -\sin t & \cos t \\\\ -\cos t & -\sin t \end{bmatrix} = 1 \end{aligned}$

Now we have that:

$\begin{aligned} \quad u_1' = \frac{\begin{bmatrix} 0 & \cos t & \sin t \\\\ 0 & -\sin t & \cos t \\\\ \tan t & -\cos t & -\sin t \end{bmatrix}}{1} = \tan t \begin{bmatrix} \cos t & \sin t \\\\ -\sin t & \cos t \end{bmatrix} = \tan t \end{aligned}$

$\begin{aligned} \quad u_2' = \frac{\begin{bmatrix} 1 & 0 & \sin t \\\\ 0 & 0 & \cos t \\\\ 0 & \tan t & -\sin t \end{bmatrix}}{1} = - \begin{bmatrix} 1 & 0 & \sin t \\\\ 0 & \tan t & -\sin t \\\\ 0 & 0 & \cos t \end{bmatrix} = -\tan t \cos t = -\frac{\sin t}{\cos t} \cos t = -\sin t \end{aligned}$

$\begin{aligned} \quad u_3' = \frac{\begin{bmatrix} 1 & \cos t & 0 \\\\ 0 & -\sin t & 0 \\\\ 0 & -\cos t & \tan t \end{bmatrix}}{1} = \begin{bmatrix} -\sin t & 0 \\\\ -\cos t & \tan t \end{bmatrix} = -\sin t \tan t = -\frac{\sin^2 t}{\cos t} \end{aligned}$

We will now integrate $u_1'$, $u_2'$, and $u_3'$ to get:

$\begin{aligned} \quad \int u_1'(t)   dt = \int \tan t   dt = \ln \mid \sec t \mid + C \end{aligned}$

$\begin{aligned} \quad \int u_2'(t)   dt = \int - \sin t   dt = \cos t + D \end{aligned}$

$\begin{aligned} \quad \int u_3'(t)   dt = \int -\frac{\sin^2 t}{\cos t}   dt = \int \frac{\cos^2 t - 1}{\cos t}   dt = \int \left ( \cos t - \sec t \right )   dt = \sin t + \ln \mid \sec t + \tan t \mid + E \end{aligned}$

Thus we have that:

$\begin{aligned} \quad y(t) = \left ( \ln \mid \sec t \mid + C \right ) (1) + \left ( \cos t + D \right ) \cos t + \left ( \sin t + \ln \mid \sec t + \tan t \mid + E \right ) \sin t \end{aligned}$

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Variation_of_parameters}{Variation of parameters,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license
\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Variation_of_parameters}{The Method of Variation of Parameters,
\end{itemize}
    mathonline.wikidot.com}
    under a CC BY-SA license
\begin{itemize}
    \item \href{http://mathonline.wikidot.com/the-method-of-variation-of-parameters-for-higher-order-nonho}{Variation of Parameters for Higher Order,
\end{itemize}
    mathonline.wikidot.com}
    under a CC BY-SA license"

%%===============================================================
\section{Method of Undetermined Coefficients (Higher Order)}
\label{sec:method-of-undetermined-coefficients-higher-order}
%%===============================================================

"In mathematics, the \textbf{method of undetermined coefficients} is an
approach to finding a particular solution to certain nonhomogeneous
ordinary differential equations and recurrence relations. It is closely
related to the annihilator method, but instead of using a particular
kind of differential operator (the annihilator) in order to find the
best possible form of the particular solution, a \""guess\"" is made as to
the appropriate form, which is then tested by differentiating the
resulting equation. For complex equations, the annihilator method or
variation of parameters is less time-consuming to perform.

Undetermined coefficients is not as general a method as variation of
parameters, since it only works for differential equations that follow
certain forms.

\paragraph{Description of the method}


Consider a linear non-homogeneous ordinary differential equation of the
form

$$\sum_{i=0}^n c_i y^{(i)} + y^{(n+1)} = g(x)$$

:   where $y^{(i)}$ denotes the i-th derivative of $y$, and $c_i$
    denotes a function of $x$.

The method of undetermined coefficients provides a straightforward
method of obtaining the solution to this ODE when two criteria are met:

1.  $c_i$ are constants.
2.  \textit{g}(\textit{x}) is a constant, a polynomial function, exponential function
    $e^{\alpha x}$, sine or cosine functions $\sin{\beta x}$ or
    $\cos{\beta x}$, or finite sums and products of these functions
    (${\alpha}$, ${\beta}$ constants).

The method consists of finding the general homogeneous solution $y_c$
for the complementary linear homogeneous differential equation

$$\sum_{i=0}^n c_i y^{(i)} + y^{(n+1)} = 0,$$

and a particular integral $y_p$ of the linear non-homogeneous ordinary
differential equation based on $g(x)$. Then the general solution $y$ to
the linear non-homogeneous ordinary differential equation would be

$$y = y_c + y_p.$$

If $g(x)$ consists of the sum of two functions $h(x) + w(x)$ and we say
that $y_{p_1}$ is the solution based on $h(x)$ and $y_{p_2}$ the
solution based on $w(x)$. Then, using a superposition principle, we can
say that the particular integral $y_p$ is

$$y_p = y_{p_1} + y_{p_2}.$$

\paragraph{Typical forms of the particular integral}


In order to find the particular integral, we need to \'guess\' its form,
with some coefficients left as variables to be solved for. This takes
the form of the first derivative of the complementary function. Below is
a table of some typical functions and the solution to guess for them.

  Function of \textit{x}                                                                                                           Form for \textit{y}
  ------------------------------------------------------------------------------------------------------------------------- ------------------------------------------------------------------------------------------------------------------
  $k e^{a x}\!$                                                                                                             $C e^{a x}\!$
  $k x^n, \; n = 0, 1, 2,\ldots\!$                                                                                           $\sum_{i=0}^n K_i x^i \!$
  $k \cos(a x) \text{ or } k \sin(a x) \!$                                                                                  $K \cos(a x) + M \sin(a x) \!$
  $k e^{a x} \cos(b x) \text{ or } ke^{a x} \sin(b x) \!$                                                                   $e^{a x} (K \cos(b x) + M \sin(b x)) \!$
  $\left(\sum_{i=0}^n k_i x^i\right) \cos(b x) \text{ or }\ \left(\sum_{i=0}^n k_i x^i\right) \sin(b x) \!$                 $\left(\sum_{i=0}^n Q_i x^i\right) \cos(b x) + \left(\sum_{i=0}^n R_i x^i\right) \sin(b x)$
  $\left(\sum_{i=0}^n k_i x^i\right) e^{a x} \cos(b x) \text{ or } \left(\sum_{i=0}^n k_i x^i\right) e^{a x} \sin(b x)\!$   $e^{a x} \left(\left(\sum_{i=0}^n Q_i x^i\right) \cos(b x) + \left(\sum_{i=0}^n R_i x^i\right) \sin(b x)\right)$

If a term in the above particular integral for \textit{y} appears in the
homogeneous solution, it is necessary to multiply by a sufficiently
large power of \textit{x} in order to make the solution independent. If the
function of \textit{x} is a sum of terms in the above table, the particular
integral can be guessed using a sum of the corresponding terms for \textit{y}.

\paragraph{Examples}


\paragraph{Example 1}


Find a particular integral of the equation

$$y'' + y = t \cos t.$$

The right side \textit{t} cos \textit{t} has the form

$$P_n e^{\alpha t} \cos{\beta t}$$

with \textit{n} = 2, \textit{a} = 0, and \textit{ß} = 1.

Since \textit{a} + \textit{iß} = \textit{i} is \textit{a simple root} of the characteristic equation

$$\lambda^2 + 1 = 0$$

we should try a particular integral of the form

$$\begin{aligned}
y_p &= t \left [F_1 (t) e^{\alpha t} \cos{\beta t} + G_1 (t) e^{\alpha t} \sin{\beta t} \right ] \\\&= t \left [F_1 (t) \cos t + G_1 (t) \sin t \right ] \\\&= t \left [ \left (A_0 t + A_1 \right ) \cos t + \left (B_0 t + B_1 \right ) \sin t \right ] \\\&= \left (A_0 t^2 + A_1 t \right ) \cos t + \left (B_0 t^2 + B_1 t \right) \sin t.
\end{aligned}$$

Substituting \textit{y}\textasciitilde{}\textit{p}\textasciitilde{} into the differential equation, we have the
identity

$$\begin{aligned}
t \cos t &= y_p'' + y_p \\\&= \left  [ \left(A_0 t^2 + A_1 t \right ) \cos t + \left (B_0 t^2 + B_1 t \right ) \sin t \right ]'' + \left[\left(A_0 t^2 + A_1 t \right ) \cos t + \left(B_0 t^2 + B_1 t \right ) \sin t \right ] \\\&=  \left [2A_0 \cos t + 2 \left (2A_0 t + A_1 \right )(-\sin t) + \left (A_0 t^2 + A_1 t \right )(-\cos t) + 2B_0 \sin t + 2 \left (2B_0 t + B_1 \right ) \cos t + \left (B_0 t^2 + B_1 t \right )(- \sin t) \right ] \\\&\qquad +\left[\left(A_0 t^2 + A_1 t \right ) \cos t + \left(B_0 t^2 + B_1 t \right ) \sin t \right ] \\\&= [4B_0 t + (2A_0 + 2B_1)] \cos t + [-4A_0 t + (-2A_1 + 2B_0)] \sin t.
\end{aligned}$$

Comparing both sides, we have

$$\begin{cases} 
1 = 4B_0\\\0 = 2A_0 + 2B_1  \\\0 = -4A_0 \\\0 = -2A_1 + 2B_0
\end{cases}$$

which has the solution

$$A_0 = 0, \quad A_1 = B_0 = \frac{1}{4}, \quad  B_1 = 0.$$

We then have a particular integral

$$y_p = \frac {1} {4} t \cos t + \frac {1}{4} t^2 \sin t.$$

\paragraph{Example 2}


Consider the following linear nonhomogeneous differential equation:

:   $\frac{dy}{dx} = y + e^x.$

This is like the first example above, except that the nonhomogeneous
part ($e^x$) is \textit{not} linearly independent to the general solution of
the homogeneous part ($c_1 e^x$); as a result, we have to multiply our
guess by a sufficiently large power of \textit{x} to make it linearly
independent.

Here our guess becomes:

$$y_p = A x e^x.$$

By substituting this function and its derivative into the differential
equation, one can solve for \textit{A}:

$$\frac{d}{dx} \left( A x e^x \right) = A x e^x + e^x$$

$$A x e^x + A e^x = A x e^x + e^x$$

$$A = 1.$$

So, the general solution to this differential equation is:

$$y = c_1 e^x + xe^x.$$

\paragraph{Example 3}


Find the general solution of the equation:

$$\frac{dy}{dt} = t^2 - y$$

$t^2$ is a polynomial of degree 2, so we look for a solution using the
same form,

$$y_p = A t^2 + B t + C,$$

Plugging this particular function into the original equation yields,

$$2 A t + B = t^2 - (A t^2 + B t + C),$$

$$2 A t + B =(1-A)t^2 -Bt -C,$$

$$(A-1)t^2 + (2A+B)t + (B+C) = 0.$$

which gives:

$$A-1 = 0, \quad 2A+B =0, \quad  B+C=0.$$

Solving for constants we get:

$$y_p = t^2 - 2 t + 2$$

To solve for the general solution,

$$y= y_p + y_c$$

where $y_c$ is the homogeneous solution $y_c = c_1 e^{-t}$, therefore,
the general solution is:

$$y= t^2 - 2 t + 2 + c_1 e^{-t}$$

\paragraph{Higher Order}


If we have a second order linear nonhomogenous differential equation
whose coefficients were constant, that is
$a \frac{d^2y}{dt^2} + b \frac{dy}{dt} + cy = g(t)$, then to solve this
differential equation, all we need to do is solve the corresponding
second order linear homogenous differential equation
$a \frac{d^2y}{dt^2} + b \frac{dy}{dt} + cy = 0$, and then find a
partial solution by assuming the form of the particular solution. More
precisely, if $g(t)$ was a polynomial, exponential, or sine/cosine
function (or a combination of these), then we could assume a form for
the particular solutions (see the linked page above for more details)
and solve for the coefficients of this form to obtain a particular
solution.

We will now look at this method for higher order linear nonhomogenous
differential equations. Consider the following $n^{\mathrm{th}}$ order
linear nonhomogenous differential equation with coefficients
$a_0, a_1, ..., a_n \in \mathbb{R}$:

$\begin{aligned} \quad a_0 \frac{d^ny}{dt^n} + a_1 \frac{d^{n-1}y}{dt^{n-1}} + ... + a_{n-1}\frac{dy}{dt} + a_n y = g(t) \end{aligned}$

Suppose that $g(t)$ is of a form containing a polynomial, exponential
function, or a sine/cosine function (like with when we were dealing with
the method of undetermined coefficients for second order linear
nonhomogenous differential equations). We can then find a particular
solution $Y(t)$ if we extend the assumed forms of combinations of
polynomials, exponential functions, or sine/cosine functions, and
multiplying by powers of $t$ to ensure our particular solution does not
contain part of a solution to the corresponding higher order linear
homogenous differential equation. We can then solve for these constants
by plugging our assumed form $Y(t)$ into our differential equation
above.

Providing a list of possible forms is trivial, so we will instead look
at some examples of apply this method.

\paragraph{Example 1}


``{=html}Find the general solution to the differential equation
$\frac{d^3y}{dt^3} + \frac{d^2y}{dt^2} + \frac{dy}{dt} + y = e^{-t} + 4t$.``{=html}

We will need to first solve the corresponding third order linear
homogenous differential equation
$\frac{d^3y}{dt^3} + \frac{d^2y}{dt^2} + \frac{dy}{dt} + y = 0$. This
characteristic equation to this differential equation is:

$\begin{aligned} \quad r^3 + r^2 + r + 1 = 0 \end{aligned}$

We can (by trial and error) see that $r_1 = -1$ is a solution to this
characteristic equation. Applying long division with the factor
$(r + 1)$ and we have that our characteristic equation can be written
as:

$\begin{aligned} \quad (r + 1)(r^2 + 1) = 0 \end{aligned}$

Therefore we can see that $r_2 = i$ and $r_3 = -i$. Therefore the
general solution to our third order linear homogenous differential
equation is:

$\begin{aligned} \quad y_h(t) = C_1e^{-t} + C_2\cos(t) + C_3\sin(t) \end{aligned}$

Now we note that $g(t) = e^{-t} + 4t$ has an exponential term and a
cosine term, so we expect the form of our particular solution to be
$Y(t) = Ae^{-t} ...$. We now compute the first, second, and third
derivatives of $Y$.

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Method_of_undetermined_coefficients}{Method of undetermined coefficients,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license"

%%===============================================================
\section{Linear Differential Equations (Higher Order)}
\label{sec:linear-differential-equations-higher-order}
%%===============================================================

"In mathematics, a \textbf{linear differential equation} is a differential
equation that is defined by a linear polynomial in the unknown function
and its derivatives, that is an equation of the form

$$a_0(x)y + a_1(x)y' + a_2(x)y'' \cdots + a_n(x)y^{(n)} = b(x)$$ where
\textit{a}\textasciitilde{}0\textasciitilde{}(\textit{x}), ..., \textit{a}\textasciitilde{}\textit{n}\textasciitilde{}(\textit{x}) and `{{math|''b''(''x'')}}`{=mediawiki}
are arbitrary differentiable functions that do not need to be linear,
and `{{math|''y''', …, ''y''(''n'') }}`{=mediawiki} are the
successive derivatives of an unknown function `{{mvar|y}}`{=mediawiki}
of the variable `{{mvar|x}}`{=mediawiki}.

Such an equation is an ordinary differential equation (ODE). A *linear
differential equation* may also be a linear partial differential
equation (PDE), if the unknown function depends on several variables,
and the derivatives that appear in the equation are partial derivatives.

A linear differential equation or a system of linear equations such that
the associated homogeneous equations have constant coefficients may be
solved by quadrature, which means that the solutions may be expressed in
terms of integrals. This is also true for a linear equation of order
one, with non-constant coefficients. An equation of order two or higher
with non-constant coefficients cannot, in general, be solved by
quadrature. For order two, Kovacic\'s algorithm allows deciding whether
there are solutions in terms of integrals, and computing them if any.

The solutions of linear differential equations with polynomial
coefficients are called holonomic functions. This class of functions is
stable under sums, products, differentiation, integration, and contains
many usual functions and special functions such as exponential function,
logarithm, sine, cosine, inverse trigonometric functions, error
function, Bessel functions and hypergeometric functions. Their
representation by the defining differential equation and initial
conditions allows making algorithmic (on these functions) most
operations of calculus, such as computation of antiderivatives, limits,
asymptotic expansion, and numerical evaluation to any precision, with a
certified error bound.

\paragraph{Basic terminology}


The highest order of derivation that appears in a (linear) differential
equation is the \textit{order} of the equation. The term
`{{math|''b''(''x'')}}`{=mediawiki}, which does not depend on the
unknown function and its derivatives, is sometimes called the *constant
term* of the equation (by analogy with algebraic equations), even when
this term is a non-constant function. If the constant term is the zero
function, then the differential equation is said to be \textit{homogeneous}, as
it is a homogeneous polynomial in the unknown function and its
derivatives. The equation obtained by replacing, in a linear
differential equation, the constant term by the zero function is the
\textit{associated homogeneous equation}. A differential equation has *constant
coefficients* if only constant functions appear as coefficients in the
associated homogeneous equation.

A \textit{solution} of a differential equation is a function that satisfies the
equation. The solutions of a homogeneous linear differential equation
form a vector space. In the ordinary case, this vector space has a
finite dimension, equal to the order of the equation. All solutions of a
linear differential equation are found by adding to a particular
solution any solution of the associated homogeneous equation.

\paragraph{Linear differential operator}


A \textit{basic differential operator} of order `{{mvar|i}}`{=mediawiki} is a
mapping that maps any differentiable function to its
`{{mvar|i}}`{=mediawiki}th derivative, or, in the case of several
variables, to one of its partial derivatives of order
`{{mvar|i}}`{=mediawiki}. It is commonly denoted

$$\frac{d^i}{dx^i}$$ in the case of univariate functions, and

$$\frac{\partial^{i_1+\cdots +i_n}}{\partial x_1^{i_1}\cdots \partial x_n^{i_n}}$$
in the case of functions of `{{mvar|n}}`{=mediawiki} variables. The
basic differential operators include the derivative of order 0, which is
the identity mapping.

A \textbf{linear differential operator} (abbreviated, in this article, as
\textit{linear operator} or, simply, \textit{operator}) is a linear combination of
basic differential operators, with differentiable functions as
coefficients. In the univariate case, a linear operator has thus the
form

$$a_0(x)+a_1(x)\frac{d}{dx} + \cdots +a_n(x)\frac{d^n}{dx^n},$$ where
`{{math|''a''0(''x''), …, ''a''''n''(''x'')}}`{=mediawiki}
are differentiable functions, and the nonnegative integer
`{{mvar|n}}`{=mediawiki} is the \textit{order} of the operator (if
`{{math|''a''''n''(''x'')}}`{=mediawiki} is not the zero
function).

Let `{{mvar|L}}`{=mediawiki} be a linear differential operator. The
application of `{{mvar|L}}`{=mediawiki} to a function
`{{mvar|f}}`{=mediawiki} is usually denoted
`{{math|''Lf''}}`{=mediawiki} or `{{math|''Lf''(''X'')}}`{=mediawiki},
if one needs to specify the variable (this must not be confused with a
multiplication). A linear differential operator is a linear operator,
since it maps sums to sums and the product by a scalar to the product by
the same scalar.

As the sum of two linear operators is a linear operator, as well as the
product (on the left) of a linear operator by a differentiable function,
the linear differential operators form a vector space over the real
numbers or the complex numbers (depending on the nature of the functions
that are considered). They form also a free module over the ring of
differentiable functions.

The language of operators allows a compact writing for differentiable
equations: if

$$L=a_0(x)+a_1(x)\frac{d}{dx} + \cdots +a_n(x)\frac{d^n}{dx^n},$$ is a
linear differential operator, then the equation

$$a_0(x)y +a_1(x)y' + a_2(x)y'' +\cdots +a_n(x)y^{(n)}=b(x)$$ may be
rewritten

$$Ly=b(x).$$

There may be several variants to this notation; in particular the
variable of differentiation may appear explicitly or not in
`{{mvar|y}}`{=mediawiki} and the right-hand and of the equation, such as
`{{math|1=''Ly''(''x'') = ''b''(''x'')}}`{=mediawiki} or
`{{math|1=''Ly'' = ''b''}}`{=mediawiki}.

The \textit{kernel} of a linear differential operator is its kernel as a linear
mapping, that is the vector space of the solutions of the (homogeneous)
differential equation `{{math|1=''Ly'' = 0}}`{=mediawiki}.

In the case of an ordinary differential operator of order
`{{mvar|n}}`{=mediawiki}, Carathéodory\'s existence theorem implies
that, under very mild conditions, the kernel of `{{mvar|L}}`{=mediawiki}
is a vector space of dimension `{{mvar|n}}`{=mediawiki}, and that the
solutions of the equation
`{{math|1=''Ly''(''x'') = ''b''(''x'')}}`{=mediawiki} have the form

$$S_0(x) + c_1S_1(x) + \cdots +c_nS_n(x),$$ where
`{{math|''c''1, …, ''c''''n''}}`{=mediawiki} are
arbitrary numbers. Typically, the hypotheses of Carathéodory\'s theorem
are satisfied in an interval `{{mvar|I}}`{=mediawiki}, if the functions
`{{math|''b'', ''a''0, …, ''a''''n''}}`{=mediawiki}
are continuous in `{{mvar|I}}`{=mediawiki}, and there is a positive real
number `{{mvar|k}}`{=mediawiki} such that \|\textit{a}\textasciitilde{}\textit{n}\textasciitilde{}(\textit{x})\| \> \textit{k} for
every `{{mvar|x}}`{=mediawiki} in `{{mvar|I}}`{=mediawiki}.

\paragraph{Homogeneous equation with constant coefficients}


A homogeneous linear differential equation has \textit{constant coefficients}
if it has the form

$$a_0y + a_1y' + a_2y'' + \cdots + a_n y^{(n)} = 0$$ where
`{{math|''a''1, …, ''a''''n''}}`{=mediawiki} are
(real or complex) numbers. In other words, it has constant coefficients
if it is defined by a linear operator with constant coefficients.

The study of these differential equations with constant coefficients
dates back to Leonhard Euler, who introduced the exponential function
`{{math|''e''''x''}}`{=mediawiki}, which is the unique
solution of the equation `{{math|1=''f''' = ''f''}}`{=mediawiki} such
that `{{math|1=''f''(0) = 1}}`{=mediawiki}. It follows that the
`{{mvar|n}}`{=mediawiki}th derivative of
`{{math|''e''''cx'' }}`{=mediawiki} is
`{{math|''c''''n''''e''''cx''}}`{=mediawiki}, and
this allows solving homogeneous linear differential equations rather
easily.

Let

$$a_0y + a_1y' + a_2y'' + \cdots + a_ny^{(n)} = 0$$ be a homogeneous
linear differential equation with constant coefficients (that is
`{{math|''a''0, …, ''a''''n''}}`{=mediawiki} are
real or complex numbers).

Searching solutions of this equation that have the form
`{{math|''e''''ax''}}`{=mediawiki} is equivalent to searching
the constants `{{mvar|a}}`{=mediawiki} such that

$$a_0e^{\alpha x} + a_1\alpha e^{\alpha x} + a_2\alpha^2 e^{\alpha x}+\cdots + a_n\alpha^n e^{\alpha x} = 0.$$
Factoring out `{{math|''e''''ax''}}`{=mediawiki} (which is
never zero), shows that `{{mvar|a}}`{=mediawiki} must be a root of the
\textit{characteristic polynomial}

$$a_0 + a_1t + a_2t^2 + \cdots + a_nt^n$$ of the differential equation,
which is the left-hand side of the characteristic equation

$$a_0 + a_1t + a_2t^2 + \cdots + a_nt^n = 0.$$

When these roots are all distinct, one has `{{mvar|n}}`{=mediawiki}
distinct solutions that are not necessarily real, even if the
coefficients of the equation are real. These solutions can be shown to
be linearly independent, by considering the Vandermonde determinant of
the values of these solutions at
`{{math|1=''x'' = 0, …, ''n'' – 1}}`{=mediawiki}. Together they form a
basis of the vector space of solutions of the differential equation
(that is, the kernel of the differential operator).

+----------------------------------------------------------------------+
| Example                                                              |
+======================================================================+
| $$y''''-2y'''+2y''-2y'+y=0$$ has the characteristic equation         |
|                                                                      |
| :   $z^4-2z^3+2z^2-2z+1=0.$                                          |
|                                                                      |
| This has zeros, `{{mvar|i}}`{=mediawiki},                            |
| `{{math|-''i''}}`{=mediawiki}, and `{{math|1}}`{=mediawiki}          |
| (multiplicity 2). The solution basis is thus                         |
|                                                                      |
| :   $e^{ix}, \; e^{-ix}, \; e^x, \; xe^x.$                              |
|                                                                      |
| A real basis of solution is thus                                     |
|                                                                      |
| :   $\cos x, \; \sin x, \; e^x, \; xe^x.$                               |
+----------------------------------------------------------------------+

In the case where the characteristic polynomial has only simple roots,
the preceding provides a complete basis of the solutions vector space.
In the case of multiple roots, more linearly independent solutions are
needed for having a basis. These have the form

$$x^ke^{\alpha x},$$ where `{{mvar|k}}`{=mediawiki} is a nonnegative
integer, `{{mvar|a}}`{=mediawiki} is a root of the characteristic
polynomial of multiplicity `{{mvar|m}}`{=mediawiki}, and
`{{math|''k'' < ''m''}}`{=mediawiki}. For proving that these functions
are solutions, one may remark that if `{{mvar|a}}`{=mediawiki} is a root
of the characteristic polynomial of multiplicity
`{{mvar|m}}`{=mediawiki}, the characteristic polynomial may be factored
as
`{{math|''P''(''t'') (''t'' - ''a'')''m''}}`{=mediawiki}.
Thus, applying the differential operator of the equation is equivalent
with applying first `{{mvar|m}}`{=mediawiki} times the operator
$\frac{d}{dx} - \alpha$, and then the operator that has
`{{mvar|P}}`{=mediawiki} as characteristic polynomial. By the
exponential shift theorem,

$$\left(\frac{d}{dx}-\alpha\right)\left(x^ke^{\alpha x}\right)= kx^{k-1}e^{\alpha x},$$

and thus one gets zero after `{{math|''k'' + 1}}`{=mediawiki}
application of $\frac{d}{dx} - \alpha$.

As, by the fundamental theorem of algebra, the sum of the multiplicities
of the roots of a polynomial equals the degree of the polynomial, the
number of above solutions equals the order of the differential equation,
and these solutions form a base of the vector space of the solutions.

In the common case where the coefficients of the equation are real, it
is generally more convenient to have a basis of the solutions consisting
of real-valued functions. Such a basis may be obtained from the
preceding basis by remarking that, if
`{{math|''a'' + ''ib''}}`{=mediawiki} is a root of the characteristic
polynomial, then `{{math|''a'' – ''ib''}}`{=mediawiki} is also a root,
of the same multiplicity. Thus a real basis is obtained by using
Euler\'s formula, and replacing
`{{math|''x''''k''''e''(''a''+''ib'')''x''}}`{=mediawiki}
and
`{{math|''x''''k''''e''(''a''-''ib'')''x''}}`{=mediawiki}
by
`{{math|''x''''k''''e''''ax'' cos(''bx'')}}`{=mediawiki}
and
`{{math|''x''''k''''e''''ax'' sin(''bx'')}}`{=mediawiki}.

\paragraph{Second-order case}


A homogeneous linear differential equation of the second order may be
written

$$y'' + ay' + by = 0,$$ and its characteristic polynomial is

$$r^2 + ar + b.$$

If `{{mvar|a}}`{=mediawiki} and `{{mvar|b}}`{=mediawiki} are real, there
are three cases for the solutions, depending on the discriminant
`{{math|1=''D'' = ''a''2 - 4''b''}}`{=mediawiki}. In all
three cases, the general solution depends on two arbitrary constants
`{{math|''c''1}}`{=mediawiki} and
`{{math|''c''2}}`{=mediawiki}.

\begin{itemize}
    \item If `{{math|''D'' > 0}}`{=mediawiki}, the characteristic polynomial
\end{itemize}
    has two distinct real roots `{{mvar|a}}`{=mediawiki}, and
    `{{mvar|ß}}`{=mediawiki}. In this case, the general solution is

$$c_1 e^{\alpha x} + c_2 e^{\beta x}.$$

\begin{itemize}
    \item If `{{math|1=''D'' = 0}}`{=mediawiki}, the characteristic polynomial
\end{itemize}
    has a double root `{{math|-''a''/2}}`{=mediawiki}, and the general
    solution is

$$(c_1 + c_2 x) e^{-ax/2}.$$

\begin{itemize}
    \item If `{{math|''D'' < 0}}`{=mediawiki}, the characteristic polynomial
\end{itemize}
    has two complex conjugate roots
    `{{math|''a'' ± ''ßi''}}`{=mediawiki}, and the general solution is

$$c_1 e^{(\alpha + \beta i)x} + c_2 e^{(\alpha - \beta i)x},$$

:   which may be rewritten in real terms, using Euler\'s formula as
    $$e^{\alpha x}  (c_1\cos(\beta x) + c_2 \sin(\beta x)).$$

Finding the solution `{{math|''y''(''x'')}}`{=mediawiki} satisfying
`{{math|1=''y''(0) = ''d''1}}`{=mediawiki} and
`{{math|1=''y'''(0) = ''d''2}}`{=mediawiki}, one equates the
values of the above general solution at `{{math|0}}`{=mediawiki} and its
derivative there to `{{math|''d''1}}`{=mediawiki} and
`{{math|''d''2}}`{=mediawiki}, respectively. This results in
a linear system of two linear equations in the two unknowns
`{{math|''c''1}}`{=mediawiki} and
`{{math|''c''2}}`{=mediawiki}. Solving this system gives the
solution for a so-called Cauchy problem, in which the values at
`{{math|0}}`{=mediawiki} for the solution of the DEQ and its derivative
are specified.

\paragraph{Non-homogeneous equation with constant coefficients}


A non-homogeneous equation of order `{{mvar|n}}`{=mediawiki} with
constant coefficients may be written

$$y^{(n)}(x) + a_1 y^{(n-1)}(x) + \cdots + a_{n-1} y'(x)+ a_ny(x) = f(x),$$
where `{{math|''a''1, …, ''a''''n''}}`{=mediawiki}
are real or complex numbers, `{{mvar|f}}`{=mediawiki} is a given
function of `{{mvar|x}}`{=mediawiki}, and `{{mvar|y}}`{=mediawiki} is
the unknown function (for sake of simplicity,
\""`{{math|(''x'')}}`{=mediawiki}\"" will be omitted in the following).

There are several methods for solving such an equation. The best method
depends on the nature of the function `{{mvar|f}}`{=mediawiki} that
makes the equation non-homogeneous. If `{{mvar|f}}`{=mediawiki} is a
linear combination of exponential and sinusoidal functions, then the
exponential response formula may be used. If, more generally,
`{{mvar|f}}`{=mediawiki} is a linear combination of functions of the
form `{{math|''x''''n''''e''''ax''}}`{=mediawiki},
`{{math|''x''''n'' cos(''ax'')}}`{=mediawiki}, and
`{{math|''x''''n'' sin(''ax'')}}`{=mediawiki}, where
`{{mvar|n}}`{=mediawiki} is a nonnegative integer, and
`{{mvar|a}}`{=mediawiki} a constant (which need not be the same in each
term), then the method of undetermined coefficients may be used. Still
more general, the annihilator method applies when
`{{mvar|f}}`{=mediawiki} satisfies a homogeneous linear differential
equation, typically, a holonomic function.

The most general method is the variation of constants, which is
presented here.

The general solution of the associated homogeneous equation

$$y^{(n)} + a_1 y^{(n-1)} + \cdots + a_{n-1} y'+ a_ny = 0$$ is

$$y=u_1y_1+\cdots+ u_ny_n,$$ where
`{{math|(''y''1, …, ''y''''n'')}}`{=mediawiki} is
a basis of the vector space of the solutions and
`{{math|''u''1, …, ''u''''n''}}`{=mediawiki} are
arbitrary constants. The method of variation of constants takes its name
from the following idea. Instead of considering
`{{math|''u''1, …, ''u''''n''}}`{=mediawiki} as
constants, they can considered as unknown functions that have to be
determined for making `{{mvar|y}}`{=mediawiki} a solution of the
non-homogeneous equation. For this purpose, one adds the constraints

$$\begin{aligned}
0 &= u'_1y_1 + u'_2y_2 + \cdots+u'_ny_n \\\0 &= u'_1y'_1 + u'_2y'_2 + \cdots + u'_n y'_n \\\  & \; \;\vdots \\\0 &= u'_1y^{(n-2)}_1+u'_2y^{(n-2)}_2 + \cdots + u'_n y^{(n-2)}_n,
\end{aligned}$$ which imply (by product rule and induction)

$$y^{(i)} = u_1 y_1^{(i)} + \cdots + u_n y_n^{(i)}$$ for
`{{math|1=''i'' = 1, …, ''n'' – 1}}`{=mediawiki}, and

$$y^{(n)} = u_1 y_1^{(n)} + \cdots + u_n y_n^{(n)} +u'_1y_1^{(n-1)}+u'_2y_2^{(n-1)}+\cdots+u'_ny_n^{(n-1)}.$$

Replacing in the original equation `{{mvar|y}}`{=mediawiki} and its
derivatives by these expressions, and using the fact that
`{{math|''y''1, …, ''y''''n''}}`{=mediawiki} are
solutions of the original homogeneous equation, one gets

$$f=u'_1y_1^{(n-1)} + \cdots + u'_ny_n^{(n-1)}.$$

This equation and the above ones with `{{math|0}}`{=mediawiki} as
left-hand side form a system of `{{mvar|n}}`{=mediawiki} linear
equations in
`{{math|''u'''1, …, ''u'''''n''}}`{=mediawiki}
whose coefficients are known functions (`{{mvar|f}}`{=mediawiki}, the
$y_i$, and their derivatives). This system can be solved by any method
of linear algebra. The computation of antiderivatives gives
`{{math|''u''1, …, ''u''''n''}}`{=mediawiki}, and
then
`{{math|1=''y'' = ''u''1''y''1 + ? + ''u''''n''''y''''n''}}`{=mediawiki}.

As antiderivatives are defined up to the addition of a constant, one
finds again that the general solution of the non-homogeneous equation is
the sum of an arbitrary solution and the general solution of the
associated homogeneous equation.

\paragraph{First-order linear equation}


The general form of a linear ordinary differential equation of order 1,
after dividing out the coefficient of
`{{math|''y'''(''x'')}}`{=mediawiki}, is:

$$y'(x) = f(x) y(x) + g(x).$$

If the equation is homogeneous, i.e.
`{{math|1=''g''(''x'') = 0}}`{=mediawiki}, one may rewrite and
integrate:

$$\frac{y'}{y}= f, \qquad \log y = k +F,$$ where
`{{mvar|k}}`{=mediawiki} is an arbitrary constant of integration and
$F=\textstyle\int f\,dx$ is any antiderivative of
`{{mvar|f}}`{=mediawiki}. Thus, the general solution of the homogeneous
equation is

$$y=ce^F,$$ where `{{math|1=''c'' = ''e''''k''}}`{=mediawiki}
is an arbitrary constant.

For the general non-homogeneous equation, one may multiply it by the
reciprocal `{{math|''e''-''F''}}`{=mediawiki} of a solution
of the homogeneous equation. This gives

$$y'e^{-F}-yfe^{-F}= ge^{-F}.$$ As
$-fe^{-F} = \frac{d}{dx} \left(e^{-F}\right),$ the product rule allows
rewriting the equation as

$$\frac{d}{dx}\left(ye^{-F}\right)= ge^{-F}.$$ Thus, the general
solution is

$$y=ce^F + e^F\int ge^{-F}dx,$$ where `{{mvar|c}}`{=mediawiki} is a
constant of integration, and `{{mvar|F}}`{=mediawiki} is any
antiderivative of `{{mvar|f}}`{=mediawiki} (changing of antiderivative
amounts to change the constant of integration).

\paragraph{Example}


Solving the equation

:   $y'(x) + \frac{y(x)}{x} = 3x.$

The associated homogeneous equation $y'(x) + \frac{y(x)}{x} = 0$ gives

$$\frac{y'}{y}=-\frac{1}{x},$$ that is

$$y=\frac{c}{x}.$$

Dividing the original equation by one of these solutions gives

:   $xy'+y=3x^2.$

That is

$$(xy)'=3x^2,$$

$$xy=x^3 +c,$$ and

$$y(x)=x^2+c/x.$$ For the initial condition

:   $y(1)=\alpha,$

one gets the particular solution

$$y(x)=x^2+\frac{\alpha-1}{x}.$$

\paragraph{Making Linear Equations from Non-Linear Equations}


Sometimes a non-linear equation, which is not solvable like this, can be
made linear, and more easily solvable, by applying a substitution.

\paragraph{Example}


$$y\frac{dy}{dx}+xy^2=5x$$

:   Let\'s make the following substitution:

$$v=y^2 \,$$

$$\frac{dv}{dx}=2y\frac{dy}{dx}$$

:   Plugging in, we get

$$\frac{1}{2}\frac{dv}{dx}+xv=5x$$

$$\frac{dv}{dx}+2xv=10x$$

:   We can then solve as a linear equation in \textit{v}, using the
    step-by-step method above:

<!-- -->

:   \textbf{Step 1}: Find the integrating factor:

$$e^{\int P(x)dx}$$

$$e^{\int 2x dx}=e^{x^2+C}$$

$$e^{\int P(x)dx}=Ce^{x^2}$$

:   Letting C=1 for convenience, we get $e^{x^2}$ as our integrating
    factor.

<!-- -->

:   \textbf{Step 2}: Multiply through

$$e^{x^2}\frac{dv}{dx}+e^{x^2}xv=10e^{x^2}x$$

:   \textbf{Step 3}: Recognize that the left hand is
    $\frac{d}{dx} e^{\int P(x)dx}v$

$$\frac{d}{dx} e^{x^2}v=10e^{x^2}x$$

:   \textbf{Step 4}: Integrate both sides w.r.t.\textit{x}.

$$\int (\frac{d}{dx} e^{x^2}v)dx=\int 10e^{x^2}xdx$$

$$e^{x^2}v=5e^{x^2}+C$$

:   \textbf{Step 5}: Solve for \textit{v}.

$$v=5+\frac{C}{e^{x^2}}$$

:   Now that we have \textit{v}, solve for \textit{y}.

$$v=y^2 \,$$

$$y^2=5+\frac{C}{e^{x^2}} \,$$

\paragraph{Higher order linear equations}


A linear ordinary equation of order one with variable coefficients may
be solved by quadrature, which means that the solutions may be expressed
in terms of integrals. This is not the case for order at least two. This
is the main result of Picard--Vessiot theory which was initiated by
Émile Picard and Ernest Vessiot, and whose recent developments are
called differential Galois theory.

The impossibility of solving by quadrature can be compared with the
Abel--Ruffini theorem, which states that an algebraic equation of degree
at least five cannot, in general, be solved by radicals. This analogy
extends to the proof methods and motivates the denomination of
differential Galois theory.

Similarly to the algebraic case, the theory allows deciding which
equations may be solved by quadrature, and if possible solving them.
However, for both theories, the necessary computations are extremely
difficult, even with the most powerful computers.

Nevertheless, the case of order two with rational coefficients has been
completely solved by Kovacic\'s algorithm.

\paragraph{Cauchy--Euler equation}


Cauchy--Euler equations are examples of equations of any order, with
variable coefficients, that can be solved explicitly. These are the
equations of the form

$$x^n y^{(n)}(x) + a_{n-1} x^{n-1} y^{(n-1)}(x) + \cdots + a_0 y(x) = 0,$$
where $a_0, \ldots, a_{n-1}$ are constant coefficients.

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Linear_differential_equation}{Linear differential equation,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license
\begin{itemize}
    \item \href{https://en.wikibooks.org/wiki/Ordinary_Differential_Equations/First_Order_Linear_1}{First Order Linear 1, Wikibooks: Differential
\end{itemize}
    Equations}
    under a CC BY-SA license"

%%===============================================================
\section{Power Series Solutions}
\label{sec:power-series-solutions}
%%===============================================================

"In mathematics, the \textbf{power series method} is used to seek a power
series solution to certain differential equations. In general, such a
solution assumes a power series with unknown coefficients, then
substitutes that solution into the differential equation to find a
recurrence relation for the coefficients.

\paragraph{Method}


Consider the second-order linear differential equation

:   $a_2(z)f''(z)+a_1(z)f'(z)+a_0(z)f(z)=0. \; \!$

Suppose \textit{a}\textasciitilde{}2\textasciitilde{} is nonzero for all \textit{z}. Then we can divide throughout to
obtain

:   $f''+{a_1(z)\over a_2(z)}f'+{a_0(z)\over a_2(z)}f=0.$

Suppose further that \textit{a}\textasciitilde{}1\textasciitilde{}/\textit{a}\textasciitilde{}2\textasciitilde{} and \textit{a}\textasciitilde{}0\textasciitilde{}/\textit{a}\textasciitilde{}2\textasciitilde{} are analytic
functions.

The power series method calls for the construction of a power series
solution

$$f=\sum_{k=0}^\infty A_kz^k.$$

If \textit{a}\textasciitilde{}2\textasciitilde{} is zero for some \textit{z}, then the Frobenius method, a variation
on this method, is suited to deal with so called \""singular points\"".
The method works analogously for higher order equations as well as for
systems.

\paragraph{Example usage}


Let us look at the Hermite differential equation,

:   $f''-2zf'+\lambda f=0; \; \lambda=1$

We can try to construct a series solution

:   $f=\sum_{k=0}^\infty A_kz^k$
:   $f'=\sum_{k=1}^\infty kA_kz^{k-1}$
:   $f''=\sum_{k=2}^\infty k(k-1)A_kz^{k-2}$

Substituting these in the differential equation

:   $\begin{aligned}
    \& \sum_{k=2}^\infty k(k-1)A_kz^{k-2}-2z\sum_{k=1}^\infty kA_kz^{k-1}+\sum_{k=0}^\infty A_kz^k=0 \\\    = \& \sum_{k=2}^\infty k(k-1)A_kz^{k-2}-\sum_{k=1}^\infty 2kA_kz^k+\sum_{k=0}^\infty A_kz^k
    \end{aligned}$

Making a shift on the first sum

:   $\begin{aligned}
    \& = \sum_{k=0}^\infty (k+2)(k+1) A_{k+2} z^k - \sum_{k=1}^\infty 2k A_k z^k + \sum_{k=0}^\infty A_k z^k \\\    \& = 2 A_2 + \sum_{k=1}^\infty (k+2)(k+1) A_{k+2} z^k - \sum_{k=1}^\infty 2k A_k z^k + A_0 + \sum_{k=1}^\infty A_k z^k \\\    \& = 2 A_2 + A_0 + \sum_{k=1}^\infty \left( (k+2) (k+1) A_{k+2} + (-2k+1) A_k \right) z^k
    \end{aligned}$

If this series is a solution, then all these coefficients must be zero,
so for both \textit{k}=0 and \textit{k}\>0:

:   $(k+2)(k+1)A_{k+2}+(-2k+1)A_k=0 \; \!$

We can rearrange this to get a recurrence relation for \textit{A}\textasciitilde{}\textit{k}+2\textasciitilde{}.

:   $(k+2)(k+1)A_{k+2}=-(-2k+1)A_k \; \!$

<!-- -->

:   $A_{k+2}={(2k-1)\over (k+2)(k+1)}A_k \; \!$

Now, we have

:   $A_2 = {-1 \over (2)(1)}A_0={-1\over 2}A_0,\, A_3 = {1 \over (3)(2)} A_1={1\over 6}A_1$

We can determine \textit{A}\textasciitilde{}0\textasciitilde{} and \textit{A}\textasciitilde{}1\textasciitilde{} if there are initial conditions, i.e.
if we have an initial value problem.

So we have

:   $\begin{aligned}
    A_4 \& ={1\over 4}A_2 = \left({1\over 4}\right)\left({-1 \over 2}\right)A_0 = {-1 \over 8}A_0 \\\\[8pt]
    A_5 \& ={1\over 4}A_3  = \left({1\over 4}\right)\left({1 \over 6}\right)A_1 = {1 \over 24}A_1 \\\\[8pt]
    A_6 \& = {7\over 30}A_4 = \left({7\over 30}\right)\left({-1 \over 8}\right)A_0 = {-7 \over 240}A_0 \\\\[8pt]
    A_7 \& = {3\over 14}A_5 = \left({3\over 14}\right)\left({1 \over 24}\right)A_1 = {1 \over 112}A_1
    \end{aligned}$

and the series solution is

:   $\begin{aligned}
    f \& = A_0z^0 + A_1z^1 +A_2z^2 +A_3z^3 +A_4z^4 +A_5z^5 + A_6z^6 + A_7z^7+\cdots \\\\[8pt]
    \& = A_0z^0 + A_1z^1 + {-1\over 2}A_0z^2 + {1\over 6}A_1z^3 + {-1 \over 8}A_0z^4 + {1 \over 24}A_1z^5 + {-7 \over 240}A_0z^6 + {1 \over 112}A_1z^7 + \cdots \\\\[8pt]
    \& = A_0z^0 + {-1\over 2}A_0z^2 + {-1 \over 8}A_0z^4 + {-7 \over 240}A_0z^6 + A_1z + {1\over 6}A_1z^3 + {1 \over 24}A_1z^5 + {1 \over 112}A_1z^7 + \cdots
    \end{aligned}$

which we can break up into the sum of two linearly independent series
solutions:

:   $f=A_0 \left(1+{-1\over 2}z^2+{-1 \over 8}z^4+{-7 \over 240}z^6+\cdots\right) + A_1\left(z+{1\over 6}z^3+{1 \over 24}z^5+{1 \over 112}z^7+\cdots\right)$

which can be further simplified by the use of hypergeometric series.

\paragraph{A simpler way using Taylor series}


A much simpler way of solving this equation (and power series solution
in general) using the Taylor series form of the expansion. Here we
assume the answer is of the form

:   $f=\sum_{k=0}^\infty {A_kz^k\over {k!}}$

If we do this, the general rule for obtaining the recurrence
relationship for the coefficients is

$$y^{[n]} \to A_{k+n}$$ and

$$x^m y^{[n]} \to (k)(k-1)\cdots(k-m+1)A_{k+n-m}$$

In this case we can solve the Hermite equation in fewer steps:

:   $f''-2zf'+\lambda f=0; \; \lambda=1$

becomes

:   $A_{k+2} -2kA_k +\lambda A_k=0$

or

:   $A_{k+2} = (2k-\lambda) A_k$

in the series

:   $f=\sum_{k=0}^\infty {A_kz^k\over {k!}}$

\paragraph{Nonlinear equations}


The power series method can be applied to certain nonlinear differential
equations, though with less flexibility. A very large class of nonlinear
equations can be solved analytically by using the Parker--Sochacki
method. Since the Parker--Sochacki method involves an expansion of the
original system of ordinary differential equations through auxiliary
equations, it is not simply referred to as the power series method. The
Parker--Sochacki method is done before the power series method to make
the power series method possible on many nonlinear problems. An ODE
problem can be expanded with the auxiliary variables which make the
power series method trivial for an equivalent, larger system. Expanding
the ODE problem with auxiliary variables produces the same coefficients
(since the power series for a function is unique) at the cost of also
calculating the coefficients of auxiliary equations. Many times, without
using auxiliary variables, there is no known way to get the power series
for the solution to a system, hence the power series method alone is
difficult to apply to most nonlinear equations.

The power series method will give solutions only to initial value
problems (opposed to boundary value problems), this is not an issue when
dealing with linear equations since the solution may turn up multiple
linearly independent solutions which may be combined (by superposition)
to solve boundary value problems as well. A further restriction is that
the series coefficients will be specified by a nonlinear recurrence (the
nonlinearities are inherited from the differential equation).

In order for the solution method to work, as in linear equations, it is
necessary to express every term in the nonlinear equation as a power
series so that all of the terms may be combined into one power series.

As an example, consider the initial value problem

$$F F'' + 2 F'^2 + \eta F' = 0 \quad ; \quad F(1) = 0 \ , \ F'(1) = -\frac{1}{2}$$

which describes a solution to capillary-driven flow in a groove. There
are two nonlinearities: the first and second terms involve products. The
initial values are given at $\eta = 1$, which hints that the power
series must be set up as:

$$F(\eta) = \sum_{i = 0}^{\infty} c_i (\eta - 1)^i$$

since in this way

$$\frac{d^n F}{d \eta^n} \Bigg|_{\eta = 1} = n! \ c_n$$

which makes the initial values very easy to evaluate. It is necessary to
rewrite the equation slightly in light of the definition of the power
series,

$$F F'' + 2 F'^2 + (\eta - 1) F' + F' = 0 \quad ; \quad F(1) = 0 \ , \ F'(1) = -\frac{1}{2}$$

so that the third term contains the same form $\eta - 1$ that shows in
the power series.

The last consideration is what to do with the products; substituting the
power series in would result in products of power series when it\'s
necessary that each term be its own power series. This is where the
\href{Cauchy_product ""wikilink""}{Cauchy product}

$$\left(\sum_{i = 0}^{\infty} a_i x^i\right) \left(\sum_{i = 0}^{\infty} b_i x^i\right) = 
\sum_{i = 0}^{\infty} x^i \sum_{j = 0}^i a_{i - j} b_j$$

is useful; substituting the power series into the differential equation
and applying this identity leads to an equation where every term is a
power series. After much rearrangement, the recurrence

$$\sum_{j = 0}^i \left((j + 1) (j + 2) c_{i - j} c_{j + 2} + 2 (i - j + 1) (j + 1) c_{i - j + 1} c_{j + 1}\right) + i c_i + (i + 1) c_{i + 1} = 0$$

is obtained, specifying exact values of the series coefficients. From
the initial values, $c_0 = 0$ and $c_1 = -1/2$, thereafter the above
recurrence is used. For example, the next few coefficients:

$$c_2 = -\frac{1}{6} \quad ; \quad c_3 = -\frac{1}{108} \quad ; \quad c_4 = \frac{7}{3240} \quad ; \quad c_5 = -\frac{19}{48600} \ \dots$$

A limitation of the power series solution shows itself in this example.
A numeric solution of the problem shows that the function is smooth and
always decreasing to the left of $\eta = 1$, and zero to the right. At
$\eta = 1$, a slope discontinuity exists, a feature which the power
series is incapable of rendering, for this reason the series solution
continues decreasing to the right of $\eta = 1$ instead of suddenly
becoming zero.

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Power_series_solution_of_differential_equations}{Power series solution of differential equations,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license"

%%===============================================================
\section{Laplace Transform}
\label{sec:laplace-transform}
%%===============================================================

"In mathematics, the Laplace transform, named after its inventor
Pierre-Simon Laplace, is an integral transform that converts a function
of a real variable t (often time) to a function of a complex variable s
(complex frequency). The transform has many applications in science and
engineering because it is a tool for solving differential equations. In
particular, it transforms linear differential equations into algebraic
equations and convolution into multiplication.

For suitable functions f, the Laplace transform is the integral

$\mathcal{L}\{f\}(s) = \int_0^\infty f(t)e^{-st} \, dt.$

  $f(t)$              $F(s) = \mathcal{L}\{f\}(s)$
  ------------------- -----------------------------------------------------
  $1$                 $\frac{1}{s}$           $s > 0$
  $t$                 $\frac{1}{s^2}$           $s > 0$
  $t^{n}$             $\frac{n!}{s^{n+1}}$           $s > 0$
  $t^{n}e^{at}$       $\frac{n!}{s^{n+1}}$           $s > a$
  $e^{at}$            $\frac{1}{s-a}$           $s > a$
  $\sin{at}$          $\frac{a}{s^2+a^2}$           $s > 0$
  $\cos{at}$          $\frac{s}{s^2+a^2}$           $s > 0$
  $\sinh{at}$         $\frac{a}{s^2-a^2}$           $s > |a|$
  $\cosh{at}$         $\frac{s}{s^2-a^2}$           $s > |a|$
  $e^{at}\sin{bt}$    $\frac{b}{(s-a)^2+b^2}$           $s > a$
  $e^{at}\cos{bt}$    $\frac{s-a}{(s-a)^2+b^2}$           $s > a$
  $e^{at}\sinh{bt}$   $\frac{b}{(s-a)^2-b^2}$           $(s - a) > |b|$
  $e^{at}\cosh{bt}$   $\frac{s-a}{(s-a)^2-b^2}$           $(s - a) > |b|$

  : List of Common Laplace Transforms

\paragraph{Resources}


\begin{itemize}
    \item \href{https://tutorial.math.lamar.edu/classes/de/LaplaceIntro.aspx}{Laplace
\end{itemize}
    Transforms},
    Paul\'s Online Notes

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Laplace_transform}{Laplace transform,
\end{itemize}
    Wikipedia} under a
    CC BY-SA license"

%%===============================================================
\section{Inverse Laplace Transform}
\label{sec:inverse-laplace-transform}
%%===============================================================

"In mathematics, the \textbf{inverse Laplace transform} of a function \textit{F}(\textit{s})
is the piecewise-continuous and exponentially-restricted real function
\textit{f}(\textit{t}) which has the property:

$$\mathcal{L}\{f\}(s) = \mathcal{L}\{f(t)\}(s) = F(s),$$ where
$\mathcal{L}$ denotes the Laplace transform.

It can be proven that, if a function \textit{F}(\textit{s}) has the inverse Laplace
transform \textit{f}(\textit{t}), then \textit{f}(\textit{t}) is uniquely determined (considering
functions which differ from each other only on a point set having
Lebesgue measure zero as the same). This result was first proven by
Mathias Lerch in 1903 and is known as Lerch\'s theorem.

The Laplace transform and the inverse Laplace transform together have a
number of properties that make them useful for analysing linear
dynamical systems.

\paragraph{Mellin\'s inverse formula}


An integral formula for the inverse Laplace transform, called the
\textit{Mellin\'s inverse formula}, the \textit{Bromwich integral}, or the
\textit{Fourier-Mellin integral}, is given by the line integral:

$$f(t) = \mathcal{L}^{-1} \{F(s)\}(t) =  \frac{1}{2\pi i}\lim_{T\to\infty}\int_{\gamma-iT}^{\gamma+iT}e^{st}F(s)\,ds$$
where the integration is done along the vertical line Re(\textit{s}) = \textit{?} in
the complex plane such that \textit{?} is greater than the real part of all
singularities of \textit{F}(\textit{s}) and \textit{F}(\textit{s}) is bounded on the line, for
example if contour path is in the region of convergence. If all
singularities are in the left half-plane, or \textit{F}(\textit{s}) is an entire
function , then \textit{?} can be set to zero and the above inverse integral
formula becomes identical to the inverse Fourier transform.

In practice, computing the complex integral can be done by using the
Cauchy residue theorem.

\paragraph{Post\'s inversion formula}


\textbf{Post\'s inversion formula} for Laplace transforms, named after Emil
Post, is a simple-looking but usually impractical formula for evaluating
an inverse Laplace transform.

The statement of the formula is as follows: Let \textit{f}(\textit{t}) be a continuous
function on the interval \\href{http://reference.wolfram.com/mathematica/ref/InverseLaplaceTransform.html}{0, 8) of exponential order, i.e.

:   $\sup_{t>0} \frac{f(t)}{e^{bt}} < \infty$

for some real number \textit{b}. Then for all \textit{s} \> \textit{b}, the Laplace transform
for \textit{f}(\textit{t}) exists and is infinitely differentiable with respect to
\textit{s}. Furthermore, if \textit{F}(\textit{s}) is the Laplace transform of \textit{f}(\textit{t}), then
the inverse Laplace transform of \textit{F}(\textit{s}) is given by

:   $f(t) = \mathcal{L}^{-1} \{F(s)\}(t)
     = \lim_{k \to \infty} \frac{(-1)^k}{k!} \left( \frac{k}{t} \right) ^{k+1} F^{(k)} \left( \frac{k}{t} \right)$

for \textit{t} \> 0, where \textit{F}^(\textit{k})^ is the \textit{k}-th derivative of \textit{F} with
respect to \textit{s}.

As can be seen from the formula, the need to evaluate derivatives of
arbitrarily high orders renders this formula impractical for most
purposes.

With the advent of powerful personal computers, the main efforts to use
this formula have come from dealing with approximations or asymptotic
analysis of the Inverse Laplace transform, using the Grunwald--Letnikov
differintegral to evaluate the derivatives.

Post\'s inversion has attracted interest due to the improvement in
computational science and the fact that it is not necessary to know
where the poles of \textit{F}(\textit{s}) lie, which make it possible to calculate the
asymptotic behaviour for big \textit{x} using inverse Mellin transforms for
several arithmetical functions related to the Riemann hypothesis.

\paragraph{Software tools}


\begin{itemize}
    \item [InverseLaplaceTransform}
\end{itemize}
    performs symbolic inverse transforms in Mathematica
\begin{itemize}
    \item \href{http://library.wolfram.com/infocenter/MathSource/5026/}{Numerical Inversion of Laplace Transform with Multiple Precision
\end{itemize}
    Using the Complex
    Domain} in
    Mathematica gives numerical solutions
\begin{itemize}
    \item \href{http://www.mathworks.co.uk/help/symbolic/ilaplace.html}{ilaplace}
\end{itemize}
    performs symbolic inverse transforms in MATLAB
\begin{itemize}
    \item \href{http://www.mathworks.co.uk/matlabcentral/fileexchange/32824-numerical-inversion-of-laplace-transforms-in-matlab}{Numerical Inversion of Laplace Transforms in
\end{itemize}
    Matlab}
\begin{itemize}
    \item \href{https://www.mathworks.com/matlabcentral/fileexchange/71511-a-cme-based-numerical-inverse-laplace-transformation-method}{Numerical Inversion of Laplace Transforms based on concentrated
\end{itemize}
    matrix-exponential
    functions}
    in Matlab

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Inverse_Laplace_transform}{Inverse Laplace transform,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license"

%%===============================================================
\section{L-Transform to ODEs}
\label{sec:l-transform-to-odes}
%%===============================================================

"The Laplace transform is a powerful integral transform used to switch a
function from the time domain to the s-domain. The Laplace transform can
be used in some cases to solve linear differential equations with given
initial conditions.

First consider the following property of the Laplace transform:

$$\mathcal{L}\{f'\}=s\mathcal{L}\{f\}-f(0)$$

$$\mathcal{L}\{f''\}=s^2\mathcal{L}\{f\}-sf(0)-f'(0)$$

One can prove by induction that

$$\mathcal{L}\{f^{(n)}\}=s^n\mathcal{L}\{f\}-\sum_{i=1}^{n}s^{n-i}f^{(i-1)}(0)$$

Now we consider the following differential equation:

$$\sum_{i=0}^{n}a_if^{(i)}(t)=\phi(t)$$

with given initial conditions

$$f^{(i)}(0)=c_i$$

Using the linearity of the Laplace transform it is equivalent to rewrite
the equation as

$$\sum_{i=0}^{n}a_i\mathcal{L}\{f^{(i)}(t)\}=\mathcal{L}\{\phi(t)\}$$

obtaining

$$\mathcal{L}\{f(t)\}\sum_{i=0}^{n}a_is^i-\sum_{i=1}^{n}\sum_{j=1}^{i}a_is^{i-j}f^{(j-1)}(0)=\mathcal{L}\{\phi(t)\}$$

Solving the equation for $\mathcal{L}\{f(t)\}$ and substituting
$f^{(i)}(0)$ with $c_i$ one obtains

$$\mathcal{L}\{f(t)\}=\frac{\mathcal{L}\{\phi(t)\}+\sum_{i=1}^{n}\sum_{j=1}^{i}a_is^{i-j}c_{j-1}}{\sum_{i=0}^{n}a_is^i}$$

The solution for \textit{f}(\textit{t}) is obtained by applying the inverse Laplace
transform to $\mathcal{L}\{f(t)\}.$

Note that if the initial conditions are all zero, i.e.

$$f^{(i)}(0)=c_i=0\quad\forall i\in\{0,1,2,...\ n\}$$

then the formula simplifies to

$$f(t)=\mathcal{L}^{-1}\left\{{\mathcal{L}\{\phi(t)\}\over\sum_{i=0}^{n}a_is^i}\right\}$$

\paragraph{An example}


We want to solve

$$f''(t)+4f(t)=\sin(2t)$$

with initial conditions \textit{f}(0) = 0 and \textit{f'}(0)=0.

We note that

$$\phi(t)=\sin(2t)$$

and we get

$$\mathcal{L}\{\phi(t)\}=\frac{2}{s^2+4}$$

The equation is then equivalent to

$$s^2\mathcal{L}\{f(t)\}-sf(0)-f'(0)+4\mathcal{L}\{f(t)\}=\mathcal{L}\{\phi(t)\}$$

We deduce

$$\mathcal{L}\{f(t)\}=\frac{2}{(s^2+4)^2}$$

Now we apply the Laplace inverse transform to get

$$f(t)=\frac{1}{8}\sin(2t)-\frac{t}{4}\cos(2t)$$

\paragraph{Laplace Transform to Systems of ODEs}


View an example from \href{https://tutorial.math.lamar.edu/classes/de/SystemsLaplace.aspx}{Laplace Transforms, Paul\'s Online
Notes}

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Laplace_transform_applied_to_differential_equations}{Laplace transform applied to ODEs,
\end{itemize}
    Wikipedia}
    under a CC BY-SA license"

%%===============================================================
\section{L-Transform to Systems of ODEs}
\label{sec:l-transform-to-systems-of-odes}
%%===============================================================

L-Transform to Systems of ODEs
