%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Unit 5: Section 1.1 – Functions and Function Notation}
\label{chap:unit5}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%===============================================================
\section{Section 1.1 – Functions and Function Notation}
\label{sec:section-1-1-functions-and-function-notation}
%%===============================================================

\paragraph{Constant Function}


% [Image: Constant function]

For the constant function $f(x) = c$ the domain consists of all real numbers; there are no restrictions on the input. The only output value is the constant $c$, so the range is the set $\{c\}$ that contains this single element. In interval notation, this is written as $[c,c]$, the interval that both begins and ends with $c$.

Domain: $( -\infty ,\infty )$

Range: $[c, c]$

\paragraph{Identity Function}


% [Image: Identity function]

For the identity function $f(x) = x$ there is no restriction on $x$. Both the domain and range are the set of all real numbers.

Domain: $( -\infty ,\infty )$

Range: $( -\infty ,\infty )$

\paragraph{Absolute Value Function}


% [Image: Absolute value function]

For the absolute value function $f(x) = |x|$ there is no restriction on x. The outputs are $x$ for $x\ge 0$ and $-x$ for $x < 0$, so the range is all numbers greater than or equal to 0.

Domain: $( -\infty ,\infty )$

Range : $[0 ,\infty )$

\paragraph{Quadratic Function}


% [Image: Quadratic function]

For the quadratic function $f(x) = x^2$ the domain is all real numbers since the horizontal extent of the graph is the whole real number line. Because the graph does not include any negative values for the range, the range is only nonnegative real numbers.

Domain: $( -\infty ,\infty )$

Range : $[0 ,\infty )$

\paragraph{Cubic Function}


% [Image: Cubic function]

For the cubic function $f(x) = x^3$ the domain is all real numbers because the horizontal extent of the graph is the whole real number line. The same applies to the vertical extent of the graph, so the domain and range include all real numbers.

Domain: $( -\infty ,\infty )$

Range: $( -\infty ,\infty )$

\paragraph{Rational Function}


% [Image: Rational function]

For the rational function (also known as the reciprocal function) $f(x) = \frac{1}{x}$ we cannot divide by 0, so we must exclude 0 from the domain. Further, 1 divided by any value can never be 0, so the range also will not include 0.

Domain: $( -\infty , 0) \cup (0, \infty )$

Range: $( -\infty , 0) \cup (0, \infty )$

\paragraph{Squared Rational Function}


% [Image: Squared rational function]

For the squared rational function $f(x) = \frac{1}{x^2}$ we cannot divide by 0, so we must exclude 0 from the domain. Further, 1 divided by any value can never be 0, so the range also will not include 0. Also, since $x^2 > 0$ for all $x\neq 0$, the range will only consist of positive numbers.

Domain: $( -\infty , 0) \cup (0, \infty )$

Range: $(0, \infty )$

\paragraph{Square Root Function}


% [Image: Square root function]

For the square root function $f(x) = \sqrt{x}$ we cannot take the square root of a negative real number, so the domain must be 0 or greater. The range also excludes negative numbers because the square root of a positive number $x$ is defined to be positive, even though the square of the negative number $-\sqrt{x}$ also gives us $x$.

Domain: $[0, \infty )$

Range: $[0, \infty )$

\paragraph{Cube Root Function}


% [Image: Cube root function]

For the cube root function $f(x) = \sqrt[3]{x}$ the domain and range include all real numbers. Note that there is no problem taking a cube root, or any odd-integer root, of a negative number, and the resulting output is negative (it is an odd function).

Domain: $( -\infty ,\infty )$

Range: $( -\infty ,\infty )$

\paragraph{Resources}


\begin{itemize}
    \item \href{https://mathresearch.utsa.edu/wikiFiles/MAT1053/Domain_Range_and_Toolkit_Functions/MAT1053_M1.2Domain_Range_and_Toolkit_Functions.pdf}{Domain Range and Toolkit Functions}, Book Chapter
    \item \href{https://mathresearch.utsa.edu/wikiFiles/MAT1053/Domain_Range_and_Toolkit_Functions/MAT1053_M1.2Domain_Range_and_Toolkit_FunctionsGN.pdf}{Guided Notes}
    \item \href{https://courses.lumenlearning.com/cuny-hunter-collegealgebra/chapter/toolkit-functions/}{Toolkit Functions}, Lumen Learning
\end{itemize}

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://courses.lumenlearning.com/cuny-hunter-collegealgebra/chapter/toolkit-functions/}{Toolkit Functions, Lumen Learning: Hunter College MATH101} under a CC0 (public domain) license
\end{itemize}

\subsection{Learning Objectives}

\begin{enumerate}
    \item Determine whether a relation represents a function.
    \item Find the value of a function.
    \item Determine whether a function is one-to-one.
    \item Use the vertical line test to identify functions.
    \item Graph the functions listed in the library of function.
\end{enumerate}

\subsection{Assessment Instrument}

\begin{enumerate}
    \item \textbf{Assessment Instrument: Functions and Their Graphs Quiz} \textbf{Format:}
    \item \textbf{Type:} Mixed-format quiz (multiple-choice, short answer, and problem-solving questions)
    \item \textbf{Duration:} 45 minutes
    \item \textbf{Total Marks:} 30 points \textbf{Sections:}
    \item \textbf{Understanding Constant Functions (10 points):}
    \item \textbf{Question 1:} Define a constant function and describe its domain and range. Provide an example of a constant function and graph it.
    \item \textbf{Marks:} 5 points
    \item \textbf{Question 2:} Given the function $f(x) = 7$, identify the domain and range, and explain why this function is classified as a constant function.
    \item \textbf{Marks:} 5 points
    \item \textbf{Graphing Functions (10 points):}
    \item \textbf{Question 3:} Graph the following functions and state their domain and range:
    \item a. $f(x) = 4$
    \item b. $f(x) = |x|$
    \item \textbf{Marks:} 5 points (2.5 points for each graph)
    \item \textbf{Question 4:} Explain how the graph of the identity function $f(x) = x$ differs from the graph of the constant function $f(x) = 3$.
    \item \textbf{Marks:} 5 points
    \item \textbf{Applying Function Concepts to Real-World Problems (10 points):}
    \item \textbf{Question 5:} A company has a fixed monthly subscription fee of $50 for a service. Model this situation using a constant function and determine the total revenue if 100 customers subscribe to the service.
    \item \textbf{Marks:} 5 points
    \item \textbf{Question 6:} Explain why a temperature of 25°C recorded every hour over a 24-hour period can be modeled as a constant function. Represent this scenario graphically.
    \item \textbf{Marks:} 5 points \textbf{Scoring Rubric:}
    \item \textbf{Understanding Constant Functions:}
    \item 2.5 points for defining the constant function and explaining its domain and range.
    \item 2.5 points for providing a correct example and graph.
    \item 2.5 points for correctly identifying the domain and range of $f(x) = 7$.
    \item 2.5 points for explaining why $f(x) = 7$ is a constant function.
    \item \textbf{Graphing Functions:}
    \item 2.5 points for each correctly plotted graph, with an accurate description of the domain and range.
    \item 5 points for a clear explanation of the differences between the graphs of $f(x) = x$ and $f(x) = 3$.
    \item \textbf{Applying Function Concepts to Real-World Problems:}
    \item 5 points for correctly modeling the subscription scenario using a constant function and calculating the total revenue.
    \item 5 points for an accurate explanation of the temperature scenario and a correct graphical representation. This quiz will assess the students' understanding of constant functions, their ability to graph functions accurately, and their capability to apply function concepts to solve real-world problems.
\end{enumerate}

%%===============================================================
\section{Section 1.2 – Domain and Range}
\label{sec:section-1-2-domain-and-range}
%%===============================================================

\paragraph{Function}

% [Image: The image demonstrates a mapping of some element a (the circle) in A, the domain, to exactly one element b in B, the range.]

Figure: $A$ is the domain of the function while $B$ is the range. This transformation from set $A$ to $B$ is an example of one-to-one function.

To make it simple, for the function $f(x)$, all of the possible $x$ values constitute the domain, and all of the values $f(x)$ ($y$ on the x-y plane) constitute the range. To put it in more formal terms, a function $f$ is a mapping of some element $a \in A$, called the domain, to exactly one element $b \in B$, called the range, such that $f: A \to B$. The image below should help explain the modern definition of a function:

>A function is considered \textbf{one-to-one} if an element $a \in A$ from domain $A$ of function $f$, leads to exactly one element $b \in B$ from range $B$ of the function. By definition, since only one element $b$ is mapped by function $f$ from some element $a$, $f: A \to B$ implies that there exists only one element $b$ from the mapping. Therefore, there exists a one-to-one function because it complies with the definition of a function. This definition is similar to \textbf{Figure 1}.

\paragraph{The horizontal line and the algebraic 1-1 test}

Similarly, the horizontal line test, though does not test if an equation is a function, tests if a function is injective (one-to-one). If any horizontal line ever touches the graph at more than one point, then the function is not one-to-one; if the line always touches at most one point on the graph, then the function is one-to-one.

% [Image: A one-to-one function passes the horizontal line test]

Figure: A one-to-one function passes the horizontal line test

% [Image: This function does NOT pass the horizontal line test]

Figure: This function does NOT pass the horizontal line test

The algebraic 1-1 test is the non-geometric way to see if a function is one-to-one. The basic concept is that:

Assume there is a function $f$. If:
> $f(a) = f(b)$, and $a = b$,
> $\newline$ then function $f$ is one-to-one.

Example:
$$\text{prove that } f(x) = \frac{1-2x}{1+x} \text{ is injective}$$

Since the notation is the notation for a function, the equation is a function. So we only need to prove that it is injective. Let $a$ and $b$ be the inputs of the function and that $f(a) = f(b)$. Thus,

$$ \begin{aligned} && \frac{1-2a}{1+a} & = \frac{1-2b}{1+b} \\\\ \Leftrightarrow && (1+b)(1-2a) & = (1+a)(1-2b) \\\\ \Leftrightarrow && 1 - 2a + b - 2ab & = 1 - 2b + a - 2ab \\\\ \Leftrightarrow && 1 - 2a + b & = 1 - 2b + a \\\\ \Leftrightarrow && -2a + b & = -2b + a \\\\ \Leftrightarrow && 1 + 3b & = 1 + 3a \\\\ \Leftrightarrow && a & = b \end{aligned} $$


So, the result is $a=b$, proving that the function is injective.

Another example is proving that $g(x) = x^2$ is not injective.

Using the same method, one can find that $a = \pm b$, which is not $a = b$. So, the function is not injective.

\paragraph{Resources}

\begin{itemize}
    \item \href{https://mathresearch.utsa.edu/wikiFiles/MAT1093/One-to-one\%20functions/Esparza\%201093\%20Notes\%201.7.pdf}{One-to-one functions. Written notes created by Professor Esparza, UTSA}
    \item \href{https://en.wikibooks.org/wiki/Calculus/Functions}{Functions, Wikibooks: Calculus}
    \item \href{https://www.math.uh.edu/\textasciitilde{}jiwenhe/Math1432/lectures/lecture01_handout.pdf}{One-to-one Functions and Inverses, University of Houston}
\end{itemize}

\paragraph{Licensing}

Content obtained and/or adapted from:
\begin{itemize}
    \item \href{https://en.wikibooks.org/wiki/Calculus/Functions}{Functions, Wikibooks: Calculus} under a CC BY-SA license
\end{itemize}

\subsection{Learning Objectives}

\begin{enumerate}
    \item Find the domain of a function defined by an equation.
    \item Graph piecewise-defined functions.
\end{enumerate}

\subsection{Assessment Instrument}

\begin{enumerate}
    \item \textbf{Assessment Instrument: One-to-One Functions Quiz} \textbf{Format:}
    \item \textbf{Type:} Mixed-format quiz (multiple-choice, short answer, and problem-solving questions)
    \item \textbf{Duration:} 45 minutes
    \item \textbf{Total Marks:} 30 points \textbf{Sections:}
    \item \textbf{Understanding Functions (10 points):}
    \item \textbf{Question 1:} Define a function and explain the difference between the domain and range. (4 points)
    \item \textbf{Question 2:} Given the set $A = \{1, 2, 3\}$ and set $B = \{4, 5, 6\}$, which of the following mappings represent a valid function? Justify your answer. (6 points)
    \item a. $f(1) = 4, f(2) = 5, f(3) = 6$
    \item b. $f(1) = 5, f(2) = 5, f(3) = 4$
    \item c. $f(1) = 4, f(2) = 5, f(2) = 6$
    \item \textbf{Classifying and Analyzing One-to-One Functions (10 points):}
    \item \textbf{Question 3:} Which of the following functions are one-to-one? Justify your answer using the horizontal line test. (5 points)
    \item a. $f(x) = 2x + 3$
    \item b. $f(x) = x^2$
    \item \textbf{Question 4:} Explain why the function $f(x) = \frac{1-2x}{1+x}$ is one-to-one using the algebraic method. (5 points)
    \item \textbf{Proving Injectivity (10 points):}
    \item \textbf{Question 5:} Prove that the function $f(x) = 3x - 7$ is injective. Show all your work. (5 points)
    \item \textbf{Question 6:} Determine whether the function $g(x) = x^3 - x$ is injective. If it is, provide a proof. If it is not, explain why not. (5 points) \textbf{Scoring Rubric:}
    \item \textbf{Understanding Functions:} 4 points for a correct definition and explanation, 6 points for correctly identifying and justifying valid functions.
    \item \textbf{Classifying and Analyzing One-to-One Functions:} 2.5 points for each correctly identified and justified function (total 5 points), 5 points for a correct algebraic explanation of injectivity.
    \item \textbf{Proving Injectivity:} 5 points for each correct proof or explanation (total 10 points). This quiz will help assess students' understanding of the concept of one-to-one functions, their ability to classify functions using the horizontal line test, and their skill in proving injectivity algebraically.
\end{enumerate}

%%===============================================================
\section{Section 2.1 – Angles}
\label{sec:section-2-1-angles}
%%===============================================================

\paragraph{Angular Measurement and Circular Sectors}

A section of a circle is known as a sector. One side of the sector is the radius. The portion of the circumference that is included in the sector is known as an arc.

% [Image: SectorArc]

\paragraph{Angular Degree}

A circle has 360 degrees. Each degree can have 60 minutes (designated as ' ) and each minute can have 60 seconds (designated as " ). Since we can not convert minutes or seconds into radians directly we need to convert the minutes and seconds into a decimal number. Here is the formula:
Convert $X^{\circ} Y' Z"$ into degrees.

$$X + \left (Y \times \frac{1}{60} \right ) + \left (Z \times \frac{1}{3600} \right ) = X + \frac{Y}{60}+ \frac{Z}{3600}$$

For a practical example, convert $23^\circ$ 18' 38" into degrees.

$$23 + \frac{18}{60}+ \frac{38}{3600} = 23.3106$$

\paragraph{Classification of Angles by Degree Measure}

\textbf{Acute Angle}
\begin{itemize}
    \item an angle is said to be acute if it measures between 0 and 90 degrees, exclusive.
\end{itemize}

% [Image: Acute Angle]

\textbf{Right Angle}
\begin{itemize}
    \item an angle is said to be right if it measures 90 degrees.
    \item notice the small box placed in the corner of a right angle, unless the box is present it is not assumed the angle is 90 degrees.
    \item all right angles are congruent
\end{itemize}

% [Image: Right Angle]

\textbf{Obtuse Angle}
\begin{itemize}
    \item an angle is said to be obtuse if it measures between 90 and 180 degrees, exclusive.
\end{itemize}

% [Image: Obtuse Angle]

\paragraph{Angular Radian}

A radian is the angle subtended at the center of a circle by an arc of its circumference equal in length to the radius of the circle. Since we know that that the formula for the circumference of a circle is $C = 2 \pi r$ we can determine that there are $2 \pi$ radians in a circle. We abbreviate radians as rad.

\paragraph{Conversion Between Degrees and Radians}

Mathematics requires us to use radians for most angular measurements. Therefore we need to know how to convert from degrees to radians. Since we know that there are $360^\circ$ degrees or $2 \pi$ radians in a circle, we can determine these equations:
$$1^\circ = \frac {\pi }{180}$$

$$1\ \text{radian} = \frac {180}{\pi }$$

So we can write these general formulae.

$$X^\circ = \frac {X \times \pi }{180}$$

$$X\ \text{radian} \left(s \right ) = \frac {X \times 180}{\pi }$$

Here is an example to convert $20^\circ$ into radians

$$20^\circ = \frac {20 \times \pi }{180} = \frac{1}{9} \pi\ \text{rad}$$

Convert $\frac{1}{9} \pi$ into degrees.

$$\frac{1}{9} \pi\ \text{radian} \left(s \right ) = \frac {\frac{1}{9} \pi \times 180}{\pi } = 20^\circ $$

\paragraph{Arc Length}

In most cases, it is very difficult to measure the length of an arc with a ruler. Therefore we need to use a formula in order to determine the length of the arc. The formula that we use is:

$$\text{Arc Length } = ( \theta\ \text{ in radians})(\text{radius})$$ in symbols this is $s= \theta r$. Note: $\theta$ needs to be in radians

Here is an example, determine the length of the arc created by a sector with a 6 cm radius and an angle of $53^\circ$.

The first thing we need to do is convert $\theta$ from degrees to radians.

$$53^\circ = \frac {53 \pi}{180}\text{rad}$$

Now we can calculate the length of the arc.

$$s = \frac {53 \pi}{180} \times 6 \approx 5.55 \text{cm}$$

\paragraph{Area of a Sector}

The area of a sector can be found using this formula:

$$\text{Area } = \frac{1}{2}r^2 \theta$$ Note: $\theta$ needs to be in radians.

Calculate the area of a sector with a 3 cm radius and an angle of $2\pi$.

$$\text{Area } = \frac{1}{2}3^2 \times 2 \pi \approx 28.3\text{cm}^2$$

\paragraph{The unit circle}


% [Image: The unit circle]

Figure: The center of the unit circle is the origin $O$, point $A=(x\_A,y\_A)$. Creating line segments starting from point $A$ to $x$ on the $x$-axis will create two line segments. From this, we learn that, in terms of $\theta$, point $A=(\cos\theta,\sin\theta)$ and other properties.

A circle with the equation $x^2 + y^2 = 1$ is called a 'unit circle'. Imagine a circle with center at the origin of a Cartesian coordinate system.

Rotating a ray from the direction of the positive half of the $x$-axis by an angle $\theta$ (counterclockwise for $\theta > 0$, and clockwise for $\theta < 0$) yields intersection points of this ray with the unit circle: $\mathrm{A} = (x\_\mathrm{A},y\_\mathrm{A})$, and, by extending the ray to a line if necessary, with the "$x=1$": $\mathrm{B} = (x\_\mathrm{B},y\_\mathrm{B})$, and with the "y=1": $\mathrm{C} = (x\_\mathrm{C},y\_\mathrm{C})$. The tangent line to the unit circle in point $\mathrm{A}$, which is orthogonal to this ray, intersects the $y$- and $x$-axis at points $\mathrm{D} = (0,y\_\mathrm{D})$ and $\mathrm{E} = (x\_\mathrm{E},0)$. The coordinate values of these points give all the existing values of the trigonometric functions for arbitrary real values of $\theta$ in the following manner:

\textbf{Table: Trigonometric functions and the unit circle}
$$\begin{array}{|c|c|c|} \hline \textbf{Function} & \textbf{Right-triangle definition} & \textbf{Unit circle definition} \\\\ \hline \sin\theta & \frac{\text{opposite}}{\text{hypotenuse}} & y _A \\\\ \hline \cos\theta & \frac{\text{adjacent}}{\text{hypotenuse}} & x _A \\\\ \hline \tan\theta & \frac{\text{opposite}}{\text{adjacent}} & y _B = \frac{y _A}{x _A} \\\\ \hline \csc\theta & \frac{\text{hypotenuse}}{\text{opposite}} & y _D = \frac{1}{y _A} \\\\ \hline \sec\theta & \frac{\text{hypotenuse}}{\text{adjacent}} & x _E = \frac{1}{x _A} \\\\ \hline \cot\theta & \frac{\text{adjacent}}{\text{opposite}} & x _C = \frac{x _A}{y _A} \\\\ \hline \end{array}$$

The importance of the unit circle is that it expands the definitions of trigonometric functions. Before, those values were constrained within a right triangle. Now, the


The importance of the unit circle is that it expands the definitions of trigonometric functions. Before, those values were constrained within a right triangle. Now, the functions can be used on any radian as their input.

\paragraph{Resources}

\begin{itemize}
    \item \href{https://mathresearch.utsa.edu/wikiFiles/MAT1093/Angles\%20and\%20their\%20measure/Esparza\%201093\%20Notes\%202.1.pdf}{Angles and their measure. Written notes created by Professor Esparza, UTSA}
    \item \href{https://en.wikibooks.org/w/index.php?title=Calculus/Trigonometry\&action=edit§ion=5}{Trigonometry, Wikibooks: Calculus}
    \item \href{https://en.wikibooks.org/wiki/Geometry/Angles}{Angles, Wikibooks: Geometry}
    \item \href{https://en.wikibooks.org/wiki/A-level_Mathematics/OCR/C2/Circles_and_Angles}{Circles and Angles, Wikibooks: A-Level Mathematics}
\end{itemize}

\paragraph{Licensing}

Content obtained and/or adapted from:
\begin{itemize}
    \item \href{https://en.wikibooks.org/w/index.php?title=Calculus/Trigonometry\&action=edit§ion=5}{Trigonometry, Wikibooks: Calculus} under a CC BY-SA license
    \item \href{https://en.wikibooks.org/wiki/Geometry/Angles}{Angles, Wikibooks: Geometry} under a CC BY-SA license
    \item \href{https://en.wikibooks.org/wiki/A-level_Mathematics/OCR/C2/Circles_and_Angles}{Circles and Angles, Wikibooks: A-Level Mathematics} under a CC BY-SA license
\end{itemize}

\subsection{Learning Objectives}

\begin{enumerate}
    \item Draw angles in standard position.
    \item Convert between degrees and radians.
    \item Find coterminal angles.
    \item Find the length of a circular arc.
    \item Use linear and angular speed to describe motion on a circular path.
\end{enumerate}

\subsection{Assessment Instrument}

\begin{enumerate}
    \item \textbf{Assessment Instrument: Angles and Their Measure Quiz} \textbf{Format:}
    \item \textbf{Type:} Mixed-format quiz (multiple-choice, short answer, and problem-solving questions)
    \item \textbf{Duration:} 45 minutes
    \item \textbf{Total Marks:} 30 points \textbf{Sections:}
    \item \textbf{Classifying Angles (10 points):}
    \item \textbf{Question 1:} Identify whether the following angles are acute, right, or obtuse:
    \item a. $35^\circ$
    \item b. $90^\circ$
    \item c. $120^\circ$
    \item d. $89.5^\circ$
    \item e. $100^\circ$
    \item \textbf{Conversion Between Degrees and Radians (10 points):}
    \item \textbf{Question 2:} Convert the following angles from degrees to radians or radians to degrees:
    \item a. $45^\circ$ to radians
    \item b. $2 \pi$ radians to degrees
    \item c. $30^\circ$ to radians
    \item d. $\dfrac{\pi}{4}$ radians to degrees
    \item e. $120^\circ$ to radians
    \item \textbf{Problem Solving: Arc Length and Area of Sector (10 points):}
    \item \textbf{Question 3:} Solve the following problems:
    \item a. Determine the arc length of a sector with a radius of 4 cm and an angle of $60^\circ$ (3 points).
    \item b. Calculate the area of a sector with a radius of 5 cm and an angle of $1.5$ radians (3 points).
    \item c. Convert $53^\circ 18' 38"$ into decimal degrees (4 points). \textbf{Scoring Rubric:}
    \item \textbf{Classifying Angles:} 2 points for each correct classification (total 10 points).
    \item \textbf{Conversion Between Degrees and Radians:} 2 points for each correct conversion (total 10 points).
    \item \textbf{Problem Solving: Arc Length and Area of Sector:} 3 points for each correct arc length or area calculation, and 4 points for the correct conversion to decimal degrees (total 10 points). The quiz will assess students' understanding and application of angular measurement concepts, including classifying angles, converting between degrees and radians, and solving problems related to arc length and the area of a sector. It ensures that students can accurately perform these calculations and understand the underlying principles.
\end{enumerate}

%%===============================================================
\section{Section 2.2 – The Unit Circle: Sine and Cosine Functions}
\label{sec:section-2-2-the-unit-circle-sine-and-cosine-functions}
%%===============================================================

\paragraph{The Unit Circle}

Consider an angle in standard position, such that the point $(x, y)$ on the terminal side of the angle is a point on a circle with radius 1.

% [Image: Defining Trigonometric Functions Figure 4]

This circle is called the \textbf{unit circle}. With $r = 1$, we can define the trigonometric functions in the unit circle:

$$ \begin{aligned} \cos \theta & = \frac{x}{r} = \frac{x}{1} = x && \sec \theta = \frac{r}{x} = \frac{1}{x} \\\\ \sin \theta & = \frac{y}{r} = \frac{y}{1} = y && \csc \theta = \frac{r}{y} = \frac{1}{y} \\\\ \tan \theta & = \frac{y}{x} && \cot \theta = \frac{x}{y} \end{aligned} $$

Notice that in the unit circle, the sine and cosine of an angle are the \textit{x} and \textit{y} coordinates of the point on the terminal side of the angle. Now we can find the values of the trigonometric functions of any angle of rotation, even the quadrantal angles, which are not angles in triangles.

% [Image: Unit Circle Definitions of Six Trigonometric Functions]

The center of the unit circle is the origin $O$, point $A=(x\_A,y\_A)$. Creating line segments starting from point $A$ to $x$ on the $x$-axis will create two line segments. From this, we learn that, in terms of $\theta$, point $A=(\cos\theta,\sin\theta)$ and other properties.

% [Image: Labeled unit circle]

We can use the figure above to determine values of the trig functions for the quadrantal angles. For example,

$$\sin{90^\circ} = y = 1 \quad\quad \sec{45^\circ} = \frac{1}{x} = \frac{2}{\sqrt{2}} = \sqrt{2}$$

\paragraph{Example 1}

Use the unit circle above to find each value:

a. $\cos 90^\circ$

b. $\cot 180^\circ$

c. $\sec 0^\circ$

----

\textbf{Solution}:

a. $\cos 90^\circ = 0$

The ordered pair for this angle is (0, 1). The cosine value is the \textit{x} coordinate, 0.

b. $\cot 180^\circ$ is undefined

The ordered pair for this angle is (-1, 0). The ratio $\tfrac{x}{y}$ is $\tfrac{-1}{0}$, which is undefined.

c. $\sec 0^\circ = 1$

The ordered pair for this angle is (1, 0). The ratio is $\tfrac{1}{x}$ is $\tfrac{1}{1} = 1$.

There are several important angles in the unit circle that you will work with extensively in your study of trigonometry: 30°, 45°, and 60°. To find the values of the trigonometric functions of these angles, we need to know the ordered pairs. Let's begin with 30°.

The terminal side of the angle intersects the unit circle at the point $\tfrac{\sqrt{3}}{2}, \tfrac{1}{2}$. (You will prove this in one of the review exercises.). Therefore we can find the values of any of the trig functions of 30°. For example, the cosine value is the \textit{x}-coordinate, so $\cos (30^\circ) = \tfrac{\sqrt{3}}{2}$. Because the coordinates are fractions, we have to do a bit more work in order to find the tangent value:

$$\tan (30^\circ) = \frac{y}{x} = \frac{\frac{1}{2}}{\frac{\sqrt{3}}{2}} = \frac{1}{2} \cdot \frac{2}{\sqrt{3}} = \frac{1}{\sqrt{3}}$$

In the review exercises you will find the values of the remaining four trig functions of this angle. The table below summarizes the ordered pairs for 30°, 45°, and 60° on the unit circle.

$$ \begin{array}{|c|c|c|} \hline \textbf{Angle} & \textbf{x coordinate} & \textbf{y coordinate} \\\\ \hline 30^\circ & 3/2 & 1/2 \\\\ \hline 45^\circ & \sqrt{2} / 2 & \sqrt{2} / 2 \\\\ \hline 60^\circ & 1/2 & 3/2 \\\\ \hline \end{array} $$

We can use these values to find the values of any of the six trig functions of these angles.

\paragraph{Example 2}

Find the value of each function.

a. $\cos (45^\circ)$

b. $\sin (60^\circ)$

c. $\tan (45^\circ)$

----

\textbf{Solution}:

a. $\cos (45^\circ) = \tfrac{\sqrt{2}}{2}$

The cosine value is the \textit{x} coordinate of the point.

b. $\sin (60^\circ) = \tfrac{\sqrt{3}}{2}$

The sine value is the \textit{y} coordinate of the point.

c. $\tan (45^\circ) = 1$

The tangent value is the ratio of the \textit{y} coordinate to the \textit{x} coordinate. Because the \textit{x} and \textit{y} coordinates are the same for this angle, the tangent ratio is 1.

\paragraph{Resources}

\begin{itemize}
    \item \href{https://mathresearch.utsa.edu/wikiFiles/MAT1093/Trigonometric\%20Functions_\%20Unit\%20Circle\%20Approach/Esparza\%201093\%20Notes\%202.2.pdf}{Trigonometric Functions: Unit Circle Approach. Written notes created by Professor Esparza, UTSA}
    \item \href{https://en.wikibooks.org/wiki/High_School_Trigonometry/Defining_Trigonometric_Functions}{Defining Trigonometric Functions, WikiBooks: High School Trigonometry}
    \item \href{https://en.wikibooks.org/wiki/Trigonometry/The_Unit_Circle}{The Unit Circle, WikiBooks: Trigonometry}
\end{itemize}

\paragraph{Licensing}

Content obtained and/or adapted from:
\begin{itemize}
    \item \href{https://en.wikibooks.org/wiki/Trigonometry/The_Unit_Circle}{The Unit Circle, WikiBooks: Trigonometry} under a CC BY-SA license
    \item \href{https://en.wikibooks.org/wiki/High_School_Trigonometry/Defining_Trigonometric_Functions}{Defining Trigonometric Functions, WikiBooks: High School Trigonometry} under a CC BY-SA license
\end{itemize}

\subsection{Learning Objectives}

\begin{enumerate}
    \item Find function values for the sine and cosine of 30° or ( pi/6 ), 45° or ( pi/4 ), and 60° or (pi/3 ).
    \item Identify the domain and range of sine and cosine functions.
    \item Use reference angles to evaluate trigonometric functions.
\end{enumerate}

\subsection{Assessment Instrument}

\begin{enumerate}
    \item \textbf{Assessment Instrument: Trigonometric Functions: Unit Circle Approach Quiz} \textbf{Format:}
    \item \textbf{Type:} Mixed-format quiz (multiple-choice, short answer, and problem-solving questions)
    \item \textbf{Duration:} 45 minutes
    \item \textbf{Total Marks:} 30 points \textbf{Sections:}
    \item \textbf{Understanding the Unit Circle (10 points):}
    \item \textbf{Question 1 (Multiple Choice, 5 points):} Which of the following statements is true about the unit circle?
    \item a. The radius of the unit circle is 2.
    \item b. The sine of an angle in the unit circle corresponds to the \textit{x} coordinate.
    \item c. The cosine of an angle in the unit circle corresponds to the \textit{y} coordinate.
    \item d. The sine of an angle in the unit circle corresponds to the \textit{y} coordinate.
    \item \textbf{Question 2 (Short Answer, 5 points):} Explain why the unit circle is essential in defining trigonometric functions.
    \item \textbf{Calculating Trigonometric Values (10 points):}
    \item \textbf{Question 3 (Problem Solving, 4 points):} Using the unit circle, find the exact values of $\cos(45^\circ)$ and $\sin(60^\circ)$.
    \item \textbf{Question 4 (Problem Solving, 3 points):} Determine the value of $\tan(30^\circ)$ using the coordinates from the unit circle.
    \item \textbf{Question 5 (Problem Solving, 3 points):} Given that the ordered pair on the unit circle for an angle of $120^\circ$ is $(-\frac{1}{2}, \frac{\sqrt{3}}{2})$, find the values of $\cos(120^\circ)$ and $\sin(120^\circ)$.
    \item \textbf{Analyzing Quadrantal Angles (10 points):}
    \item \textbf{Question 6 (Multiple Choice, 5 points):} Which of the following values correctly represents $\cos(90^\circ)$?
    \item a. 0
    \item b. 1
    \item c. -1
    \item d. Undefined
    \item \textbf{Question 7 (Problem Solving, 5 points):} For each of the following quadrantal angles, determine the value of the given trigonometric function:
    \item a. $\sec(0^\circ)$
    \item b. $\cot(180^\circ)$ \textbf{Scoring Rubric:}
    \item \textbf{Understanding the Unit Circle:} 5 points for each correct response (total 10 points)
    \item \textbf{Calculating Trigonometric Values:} 4 points for each correct problem in Question 3, 3 points each for Questions 4 and 5 (total 10 points)
    \item \textbf{Analyzing Quadrantal Angles:} 5 points for each correct response (total 10 points) The quiz will assess the students' understanding of the unit circle and their ability to apply this knowledge to determine trigonometric values for both standard and quadrantal angles. This assessment will also evaluate their comprehension of the geometric significance of trigonometric functions.
\end{enumerate}

%%===============================================================
\section{Section 2.3 – The Other Trigonometric Functions}
\label{sec:section-2-3-the-other-trigonometric-functions}
%%===============================================================

\paragraph{Identities Based on the Pythagorean Theorem}

Some of the most fundamental trigonometric identities are those derived from the Pythagorean Theorem. These are defined using a right triangle:

% [Image: Trigonometry Triangle]

Figure: Trigonometry Triangle

At this point in the course, it is important to be comfortable with the Pythagorean Theorem. If it is not yet familiar, it is recommended to learn it now.
$$a^2+b^2=c^2$$
$a$ and $b$ of course are the legs or the adjacent and opposite edges, and $c$ is the hypotenuse, the longest side, the side that does not include the right angle. This formula only works for a right angle triangle. If the angle shown as a right angle in the diagram were obtuse, larger than a right angle, then $c^2$ would be larger than the Pythagorean sum. If the angle shown as a right angle were smaller than a right angle, then $c^2$ would be smaller than $a^2+b^2$.

\paragraph{Pythagorean Theorem in Terms of \textit{sine} and \textit{cosine}}

The Pythagorean Theorem is the same thing as
$$\sin^2(A)+\cos^2(A)=1$$
We know this from an earlier lesson where we introduced $\rm sine$ and $\rm cosine$.
\begin{itemize}
    \item The $\rm cosine$ is the adjacent side when the hypotenuse is one.
    \item The $\rm sine$ is the opposite side when the hypotenuse is one.
\end{itemize}

The $\sin^2(A)+\cos^2(A)=1$ identity should be second nature to you too, but you should also be able to derive it from the Pythagorean relation.

We can see it's true from a right triangle that has a hypotenuse that is \textit{c} rather than one. Dividing the Pythagorean relation through by $c^2$ gives us

$$\left(\frac{a}{c}\right)^2+\left(\frac{b}{c}\right)^2=\left(\frac{c}{c}\right)^2=1$$

And we have already seen from soh-cah-toa that the sine of $A$ is $\left(\frac{a}{c}\right)$ and the cosine of $A$ is $\left(\frac{b}{c}\right)$.

\paragraph{Identities Based on Addition Formula}

We saw the addition formula earlier.

\paragraph{Letters are Arbitrary}

The letters in these equations $A$ and $B$ are arbitrary. We could equally use $\omega$ and $\theta$. We're pointing this out because we happen to have a diagram on this page that has angles $A$, $B$ and $C$ in it, and we want to make it clear that that was for Pythagoras and we're not talking about that diagram anymore. In these equations you can also always substitute actual values, e.g., replace $A$ by $\pi$ and they'd still be true. Or you could replace $A$ by $2t$ or $\pi-t$ and they'd still be true.

\paragraph{Addition Formula}

The addition formula for cosine is:
$$\cos(A+B)=\cos(A)\cos(B)-\sin(A)\sin(B)$$
and is worth learning. But don't learn the next one:

\paragraph{'Subtraction' Formula}

$$\cos(A-B)=\cos(A)\cos(B)+\sin(A)\sin(B)$$
Learning the second formula is extra effort that does not really gain you anything. You can get it instantly by replacing $B$ by $-B$ in the first equation. The only things on the right-hand side that will change are the terms that had $B$ in them. The $\cos(B)$ is unchanged because $\cos(-B)=\cos(B)$. The minus sign before the second term changes because $\sin(-B)=-\sin(B)$.

\paragraph{Checking the 'Subtraction' Formula}

Now do a quick check. If $A=B$ then $\cos(A-B)$ is $\cos(0)$ which is one, right? And on the right-hand side we have $\cos^2(A)+\sin^2(A)$, which is also one. Looks good.

It is much better to remember it this way, because you are doing less rote learning and you're also becoming more fluent with the algebra.

\paragraph{Double Angle Formula}

\begin{itemize}
    \item $\sin(2x) = 2\sin(x)\cos(x)$
    \item $$ \begin{aligned} \cos(2x) & = \cos^2(x) - \sin^2(x) \hspace{44em}\\\\ & = 2\cos^2(x) - 1 \\\\ & = 1 - 2\sin^2(x) \end{aligned} $$
    \item $\tan(2x) = \frac{2\tan(x)}{1 - \tan^2(x)}$
\end{itemize}

\paragraph{Half Angle Formula}

\begin{itemize}
    \item $\sin\left(\frac{x}{2}\right) = \pm \sqrt{\frac{1 - \cos(x)}{2}}$
    \item $\cos\left(\frac{x}{2}\right) = \pm \sqrt{\frac{1 + \cos(x)}{2}}$
    \item $$ \begin{aligned} \tan\left(\frac{x}{2}\right) & = \pm \sqrt{\frac{1 - \cos(x)}{1 + \cos(x)}} \hspace{44em} \\\\ & = \frac{\sin(x)}{1 + \cos(x)} \\\\ & = \frac{1 - \cos(x)}{\sin(x)} \end{aligned} $$
\end{itemize}

\paragraph{Product to Sum Formulas}

$$ \begin{aligned} \sin(u)\sin(v) & = \frac{1}{2}[\cos(u-v) - \cos(u+v)] \hspace{44em} \\\\ \cos(u)\cos(v) & = \frac{1}{2}[\cos(u-v) + \cos(u+v)] \\\\ \sin(u)\cos(v) & = \frac{1}{2}[\sin(u+v) + \sin(u-v)] \\\\ \cos(u)\sin(v) & = \frac{1}{2}[\sin(u+v) - \sin(u-v)] \end{aligned} $$

\paragraph{Sum to Product Formula}

$$ \begin{aligned} \sin(u) + \sin(v) & = 2\sin\left(\frac{u+v}{2}\right)\cos\left(\frac{u-v}{2}\right) \hspace{44em} \\\\ \sin(u) - \sin(v) & = 2\cos\left(\frac{u+v}{2}\right)\sin\left(\frac{u-v}{2}\right) \\\\ \cos(u) + \cos(v) & = 2\cos\left(\frac{u+v}{2}\right)\cos\left(\frac{u-v}{2}\right) \\\\ \cos(u) - \cos(v) & = -2\sin\left(\frac{u+v}{2}\right)\sin\left(\frac{u-v}{2}\right) \end{aligned} $$

\paragraph{Sum of Two Shifted Cosines}

For the next formula, you could just plough through the algebra...
$$\cos(x+\theta)+\cos(x-\theta)=\dots$$

\paragraph{Waves in and Out of Phase}

Before we do, notice that we are adding two cosine waves, the first shifted left and the second shifted right. We've seen this before when we looked at waves in and out of phase. We're going to get another sinusoidal wave. The expression is fairly symmetric, and actually we can quickly show that the expression gives an even function, that is that the value for $-x$ is the same as for $x$:
$$\cos(-x+\theta)+\cos(-x-\theta)=\cos(x-\theta)+\cos(x+\theta)$$
using $\cos(x)=\cos(-x)$ we get to:
$$\cos(x-\theta)+\cos(x+\theta)$$
and now swapping the two terms:
$$\cos(x+\theta)+\cos(x-\theta)$$
which is the original expression.

With practice, you'll be able to see those steps and that the expression is an even function immediately.

A sinusoidal wave that is an even function - well it is something based on cosine. We're expecting the formula will simplify to give us something like:
$$A\cos(x)$$
Where $A$ will depend on $\theta$.

\paragraph{Values to Check With}

When $\theta=0^\circ$ (try it in the original formula) we expect $A$ to be 2. When $\theta=90^\circ$ we expect $A$ to be 0, as the two cosine waves are $180^\circ$ out of phase. So now to plough through the algebra:

$$\cos(x+\theta)+\cos(x-\theta)=$$

$$\cos(x)\cos(\theta)-\sin(x)\sin(\theta)+\cos(x)\cos(-\theta)-\sin(x)\sin(-\theta)=$$

$$\cos(x)\cos(\theta)-\sin(x)\sin(\theta)+\cos(x)\cos(\theta)+\sin(x)\sin(\theta)=$$

$$\cos(x)\cos(\theta)+\cos(x)\cos(\theta)=$$

$$2\cos(\theta)\cos(x)$$

And to check we try with $\theta=0^\circ$ and get $2\cos(x)$, and we check with $\theta=90^\circ$ and get $0\times\cos(x)$, as we hoped.

Knowing roughly what we expected made it easier to get the algebraic steps right. We knew where we were going. We knew we weren't going to end up with four separate terms, some would have to cancel or combine in some other way.

\paragraph{Identities Based on Sine and Cosine Rules}

\paragraph{The Law of Sines}

$$\frac{a}{\sin(A)}=\frac{b}{\sin(B)}=\frac{c}{\sin(C)}$$
This is easy to remember because it is so symmetrical. You do not even need to remember which way up the ratios are as this is also true:
$$\frac{\sin(A)}{a}=\frac{\sin(B)}{b}=\frac{\sin(C)}{c}$$
What you do need to remember is how the triangle is labeled for it to be true.

\paragraph{The Law of Cosines}

With the cosine law formula:
$$a^2+b^2-2ab\cos(\theta)=c^2$$
It is best to think of it as a more general version of Pythagoras' theorem. The $a$ and $b$ have to be on equal footing, so $ab$ as a multiplier is reasonable.

If you are familiar with 'units' in physics, then use the fact that units must match up. The quantities $a,b,c$ are measurements of length. It is no good adding $a^2$ to $-2(a+b)\cos(\theta)$. One might have units Km², the other units of Km. Adding $a^2$ to $-2ab\cos(\theta)$ the units, e.g., Km², do match.

In some more detail, it has got to be minus for $-2ab\cos(\theta)$. That's because for $\theta<90^\circ$ we need $c^2$ to be less than for Pythagoras' theorem.

How can you remember and be sure the 2 is right? Think of an equilateral triangle with each side of length 1. The angle $\theta=60^\circ$ and $\cos(60^\circ)=0.5$. We need the 2 to get the right answer 1 + 1 + 2x1x1x0.5 = 1.

\paragraph{Ratio Identities}

$$\tan(A)=\frac{\sin(A)}{\cos(A)}$$
This is one definition of tan, and is something you should just learn.

\paragraph{Identities Based on Symmetry}

These symmetry identities are best remembered by remembering the graphs for these functions. Have a look at the graphs again if you don't remember them.

$$0,\frac{\pi}{2},\pi,\frac{3\pi}{2},2\pi$$ are landmarks on the graphs of sine and cosine, and you should know what happens for both those functions at those landmarks. When you have that knowledge the following symmetry identities are easy to write down because you can see them visually.

\paragraph{Shifting Left}

$$\cos(\pi+A)=-\cos(A)\quad\sin(\pi+A)=-\sin(A)\quad\tan(\pi+A)=\tan(A)$$
The identities above are visualized as shifting the graph left by $\pi$. It's a half cycle, and both cosine and sine change signs - and hence tan does not change sign, because when it is expressed as a ratio of sine over cosine both the top and bottom change signs.

\paragraph{Reflecting in $x=\frac{\pi}{2}$}

$$\cos(\pi-A)=-\cos(A)\quad\sin(\pi-A)=\sin(A)\quad\tan(\pi-A)=-\tan(A)$$
In the above, we are reflecting the graph about the vertical line $x=\frac{\pi}{2}$. Spend enough time looking at this on the graph to see that that is true. It is important to do that to remember this trick for remembering what happens.

Now sine has a maximum at $x=\frac{\pi}{2}$ (it has the value one) and it is symmetric about that x value, so
$$\sin(\pi-A)=\sin(A)$$

Cosine is zero at $x=\frac{\pi}{2}$. Reflecting cosine in the line $x=\frac{\pi}{2}$ reverses the sign. 
$$\cos(\pi-A)=-\cos(A)$$

\paragraph{Co-function Identities}

$$ \begin{aligned} \cos(A) & = \sin\left(\frac{\pi}{2} - A\right) \hspace{44em} \\\\ \sin(A) & = \cos\left(\frac{\pi}{2} - A\right) \end{aligned} $$

The above two formulas are most easily seen from the right-triangle definition of cosine and sine. The two acute angles in such a triangle add up to $\frac{\pi}{2}$, and we are just giving the ratios of sides lengths in terms of a different angle's trig function.
$$ \begin{aligned} \cos(A) & = \sin\left(\frac{\pi}{2} + A\right) \hspace{44em} \\\\ \sin(A) & = -\cos\left(\frac{\pi}{2} + A\right) \end{aligned} $$
These two formulae are 'the same' as the previous formulae. We get there in two steps. First subtracted pi from the angle on the right-hand side, which inverts the sign, and inverted the sign to compensate. Next invert the sign of the angle on the right-hand side. For cosine on the right, we are done. For sine on the right, we have to invert its sign since inverting the angle's sign changed the sign of the result.

Yes - keeping track of signs is tricky, and always needs care.

\paragraph{Resources}

\begin{itemize}
    \item \href{https://mathresearch.utsa.edu/wikiFiles/MAT1093/Properties\%20of\%20the\%20Trigonometric\%20Functions/Esparza\%201093\%20Notes\%20Week\%203.pdf}{Properties of the Trigonometric. Written notes created by Professor Esparza, UTSA}
\end{itemize}

\paragraph{Licensing}

Content obtained and/or adapted from:
\begin{itemize}
    \item \href{https://en.wikibooks.org/wiki/Trigonometry/Remembering_Trig_Formulae}{Remembering Trig Formulae, Wikibooks: Trigonometry} under a CC BY-SA license
\end{itemize}

\subsection{Learning Objectives}

\begin{enumerate}
    \item Find exact values of the trigonometric functions secant, cosecant, tangent, and cotangent of pi/3, pi/4, and pi/6
    \item Use reference angles to evaluate the trigonometric functions secant, cosecant, tangent, and cotangent.
    \item Use properties of even and odd trigonometric functions.
    \item Recognize and use fundamental identities.
    \item Evaluate trigonometric functions with a calculator.
\end{enumerate}

\subsection{Assessment Instrument}

\begin{enumerate}
    \item \textbf{Assessment Instrument: Properties of the Trig Functions Quiz} \textbf{Format:}
    \item \textbf{Type:} Mixed-format quiz (multiple-choice, short answer, and problem-solving questions)
    \item \textbf{Duration:} 45 minutes
    \item \textbf{Total Marks:} 30 points \textbf{Sections:}
    \item \textbf{Pythagorean Identity (10 points):}
    \item \textbf{Question 1:} Given that $\sin(A) = \frac{3}{5}$ and $A$ is in the first quadrant, find $\cos(A)$ using the Pythagorean identity $\sin^2(A) + \cos^2(A) = 1$ (5 points).
    \item \textbf{Question 2:} Verify if the following identity is true or false: $\cos^2(A) = 1 - \sin^2(A)$ Explain your reasoning (5 points).
    \item \textbf{Simplification of Trigonometric Expressions (10 points):}
    \item \textbf{Question 3:} Simplify the following trigonometric expression using the Pythagorean identity: $\cos^2(A) - \sin^2(A)$ (5 points).
    \item \textbf{Question 4:} If $\cos(A) = \frac{4}{5}$ and $A$ is an acute angle, calculate the value of $\sin^2(A) + \cos^2(A)$ and determine if the identity holds true (5 points).
    \item \textbf{Problem Solving Using Trigonometric Identities (10 points):}
    \item \textbf{Question 5:} Prove the identity $\tan^2(A) + 1 = \sec^2(A)$ using the Pythagorean identity (5 points).
    \item \textbf{Question 6:} Solve for $A$ if $\cos^2(A) = \frac{3}{4}$. Find the possible values of $\sin(A)$, considering the Pythagorean identity (5 points). \textbf{Scoring Rubric:}
    \item \textbf{Pythagorean Identity:} 5 points for each correct solution (total 10 points)
    \item \textbf{Simplification of Trigonometric Expressions:} 5 points for each correctly simplified expression (total 10 points)
    \item \textbf{Problem Solving Using Trigonometric Identities:} 5 points for each correct solution and proof (total 10 points) The quiz assesses the students' understanding and application of the Pythagorean identity and related trigonometric properties. It ensures they can simplify expressions, prove identities, and solve trigonometric equations accurately.
\end{enumerate}

%%===============================================================
\section{Section 2.4 – Right Triangle Trigonometry}
\label{sec:section-2-4-right-triangle-trigonometry}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Use right triangles to evaluate trigonometric functions.
    \item Find function values for  30?° (pi/6), 45?° (pi/4), and 60?° (pi/3).
    \item Use cofunctions of complementary angles.
    \item Use the de?nitions of trigonometric functions of any angle.
    \item Use right triangle trigonometry to solve applied problems.
\end{enumerate}

%%===============================================================
\section{Section 3.1 – Graphs of Sine and Cosine Functions}
\label{sec:section-3-1-graphs-of-sine-and-cosine-functions}
%%===============================================================

\paragraph{Frequency}


% [Image: Some waves of different frequencies]

Figure: Some waves of different frequencies

Here are some sinusoidal (that is 'sine like') graphs. These show sinusoidal waves with different frequency. The frequency increases as you go down.

\textbf{Waves and Light} In physics, light is a kind of wave, and different frequencies of light give the different colors of the spectrum.

\begin{itemize}
    \item Does red light have a higher or lower frequency than blue light?
    \item In a rainbow is red on the outside or the inside of the rainbow? What about a double rainbow? (use the Internet)
    \item Are radio waves higher or lower frequency than light waves? (use the Internet)
\end{itemize}

\textbf{Waves and Sound} In physics sound is a kind of wave. It's physically a different kind of wave from light, but its vibrations can still be described using sines and cosines. Light is an electromagnetic wave and sound is a wave involving change in air pressure. That difference is the underlying reason why light can travel through the vacuum of space, but sound cannot.

High pitched notes are high-frequency sound waves. Low pitched notes are low-frequency sound waves.

\begin{itemize}
    \item Find out what the highest frequency of hearing is for humans and for bats. (use the Internet)
    \item Explain in your own words what 'Hertz' means, and what the K in the abbreviation KHz means.
    \item What vibration frequencies are typical in an Earthquake?
\end{itemize}

\paragraph{Sine and Cosine Graphs}


And here in this graph, the colors we've chosen to draw the waves in have no relation to the frequency. Both waves have the same frequency, but we've used different colors.

% [Image: ]

Notice that in the graph of sine and cosine above that the two graphs have the same shape. If we slide the sine graph slightly to the left, it coincides exactly with the cosine graph. Using the terminology used to describe sinusoidal waves, they have the same amplitude, the same frequency and different phases.

\paragraph{Definitions}


A sinusoidal wave is characterized by three parameters: \textbf{amplitude}, \textbf{frequency} and \textbf{phase}.

\begin{itemize}
    \item The \textbf{amplitude} is the amount the function varies, positively or negatively, from zero in the y direction.
    \item The \textbf{frequency} is how many complete cycles there are of the wave in unit distance on the x axis (which often measures time)
    \item The \textbf{phase} is relevant when comparing two waves of the same frequency. It is how much (measured in degrees or radians) one is shifted relative to the other on the x axis.
\end{itemize}

This terminology comes from sound engineering where higher frequency sounds have higher pitch and waves of greater amplitude are louder.

As an alternative to specifying the frequency, the number of cycles in unit distance, we can instead specify the \textbf{wavelength}, the length of one cycle. The higher the frequency, the shorter the wavelength. The lower the frequency the longer the wavelength.

You can check this on the diagram of waves of different frequencies at the top of the page.

\paragraph{In More Detail}


To fully understand how each of the following definitions relate, look at the graph of the function
$$f(t)=4\sin\left(\frac{\pi}{6}t\right)+40 \quad \text{ below (left)}$$
For the figure representing (a)-(b), read the corresponding definitions.

% [Image: The graph drawn is courtesy of the Desmos online graphing calculator.]

Figure: The graph drawn is courtesy of the Desmos online graphing calculator.

\paragraph{Amplitude}


The \textbf{amplitude} is the maximum amount that the wave differs from the \textit{sinusoidal axis value}, $c$ – the value by which the function is shifted by the average of the maximum to the minimum range of the function,
$$\frac{y _{max}+y _{min}}{2}$$
The amplitude comes from two different locations: the maximum value minus $c$ or the minimum value minus $c$. The positive magnitude of the value is taken (absolute value). Look at the data below for an example.

$\begin{array}{|c|c|} \hline x & g(x) \\\\ \hline 0 & 20 \\\\ \hline 2 & 0 \\\\ \hline 4 & -20 \\\\ \hline 6 & 0 \\\\ \hline 8 & 20 \\\\ \hline \vdots & \vdots \end{array} $

Let us not worry about finding the function for this set of data for now. Instead, focus on finding the amplitude. The maximum $y$-value $y _{max}=20$. The minimum $y$-value $y _{min}=-20$. Therefore, $\frac{20+(-20)}{2}=\frac{0}{2}=0$. The maximum difference comes from two different locations: the maximum value minus the sinusoidal axis value or the minimum value minus the sinusoidal axis value. In this example, we will choose to use the difference between the maximum value to the sinusoidal axis value, which is $20-0=20$. Therefore, the amplitude of the above data is $20$.

In general, for $\sin(b\theta)+c$ and $\cos(b\theta)+c$ (where $b>0$) the amplitude is 1.
$$\begin{aligned} \text{For } a\sin(b\theta)+c \quad \text{ and } \quad a\cos(b\theta)+c \, (\text{where } a > 0 \text{ and } b>0), \text{ the amplitude is } a. \hspace{44em} \end{aligned}$$
The difference between the greatest and least value is called the \textbf{double amplitude} or \textbf{peak to peak amplitude}. Knowing this, the way to find the amplitude given the greatest and least value of the range is to take the difference between the two and then divide by 2:

$$a=\frac{y _{max}-y _{min}}{2}$$.

\paragraph{Frequency}


The \textbf{frequency} (sometimes $f$) is the number of cycles a wave goes through in a standard distance or time. If this standard distance is $360^\circ$ or $2\pi$ radians, the frequency of $\sin(\theta)$ and $\cos(\theta)$ is $1$, and that of $\sin(b\theta)$ and $\cos(b\theta)$ is $b$. A graph of a high frequency wave shows more complete cycles in the same horizontal span as a low frequency wave. Notice $b$ is the frequency value, $f$, so $b=f$.

The \textbf{wavelength} (or \textbf{period}), $\lambda$, decreases as the frequency increases. Mathematically, this relationship is written as $\lambda\propto\frac{1}{f}$ (the wavelength is inversely proportional to the frequency). It is the distance over which a complete cycle occurs, that is, the wave goes from $c$ to its maximum positive value, back to $c$, down to the lowest negative value, then back up to $c$.

The functions $\sin(\theta)$ and $\cos(\theta)$ both go through a complete cycle as the angle increases by $360^\circ$ or $2\pi$ radians, hence this is the wavelength. The functions $\sin(b\theta)$ and $\cos(b\theta)$ (where $b$ is any positive number) go through one cycle as the angle $\theta$ increases by $\frac{360^\circ}{b}$ or $\frac{2\pi}{b}$ , hence their wavelength $\lambda=\frac{360^\circ}{|b|}$ or $\lambda=\frac{2\pi}{|b|}$ . This makes sense, because if $b=f$ and $\lambda=\frac{2\pi}{|b|}$ , then $\lambda\propto\frac{1}{b}$ .

Finally, using wavelength, one can directly find the frequency. If $\lambda=\frac{2\pi}{|b|}$, then

$$ \begin{aligned} \Rightarrow \lambda \times \frac{1}{\lambda} \times b & = \frac{2\pi}{b} \times b \times \frac{1}{\lambda} \\\\ \Rightarrow b & = \frac{2\pi}{\lambda} \end{aligned} $$

\paragraph{Phase Shift}


The \textbf{phase} is the fraction of the cycle that you have reached at any given point. If the point is not specified, it may be assumed to be 0°. The start of the cycle is the point where the wave is 0 and going from negative to positive. (If the wave is not a sine curve, there may be more than one such point in a complete cycle so the starting point may be arbitrary.) For $\sin(\theta)$ the phase at 0° is 0; for $\cos(\theta)$ it is ¼. If the cycle has wavelength $360^\circ$ or $2\pi$ radians, the phase is often expressed as an angle corresponding to the given fraction of the wavelength, so the phase of $\cos(\theta)$ at 0° is $90^\circ$ or $\tfrac{\pi}{2}$ radians.

It is assumed that the average value of the wave is 0. If the wave is say $\sin(\theta)+4$ or $2\pi$ , it is never 0 so to define the phase we would need the point where it reaches its average value (4 in this case).

\paragraph{Examples}


\textbf{Larger Amplitude}

$10\cos(\theta)$ has ten times the amplitude of $\displaystyle \cos \theta$

The graph oscillates between 10 and -10 on the y axis.

\textbf{Larger Frequency}

$\cos(100\theta)$ has a higher frequency than $\cos(\theta)$

There will be 100 times as many cycles in the same distance along the x axis. The amplitude is unchanged.

The wavelength is also decreased. The distance along the x axis for a complete cycle is one hundredth of what it was before.

\textbf{Different Phase}

$\sin(\theta)$ has a different phase from $\cos(\theta)$ . The phase difference happens to be 90°, because $\sin(\theta+90^\circ)=\cos(\theta)$ . If two waves are 180° out of phase then where one is positive the other has the same value only negative. $-\sin(\theta)$ is 180° out of phase with $\sin(\theta)$

\paragraph{Offset}


Two sinusoidal waves may also have a different offset to each other, i.e. their average values are different, but they still have the same shape. Their graphs are y shifted relative to each other. In sound engineering this is called \textbf{DC Bias}.

\textbf{Different Offset}

$1+\cos(\theta)$ has a different offset to $\cos(\theta)$ . Whereas $\cos(\theta)$ takes both positive and negative values, $1+\cos(\theta)$ is never below 0.

This offset is based off the \textbf{vertical shift}, also known as the \textbf{sinusoidal axis value}. Keep in mind, unlike other functions, this vertical shift does $\underline{\text{not}}$ denote the y-intercept. The only way to determine the y-intercept correctly is to substitute zero into a function. Quick quiz: what is $\cos (0)$? In case you forgot, it is $1$. Therefore, the y-intercept for $f(\theta)=1+\cos(\theta)$ is $f(0)=1+\cos(0)=1+1=2$.

\paragraph{The Graphs and Their Differences}


Comparing and contrasting two different items may seem like a pointless task to the student, at first. However, the purpose of such an activity is to make students analyze information and help cement differences between the two ideas. Here, it is imperative that the student understands the differences between cosine and sine because otherwise, the student is not able to model situations correctly! While this Wikibooks will not make the student analyze the two functions themselves, it may be a good idea to remember the differences.

\paragraph{The parent functions}


% [Image: The sine and cosine curve are both graphed over the period 2pi with a frequency of 1)]

Figure: The sine and cosine curve are both graphed over the period ($\lambda=2\pi$ with a frequency $f=1$)

The sine and cosine functions are shown above, where $0\le x\le 2\pi$. Notice that the frequency and period of the functions are both identical. Given that these are also parent functions, the vertical shift is zero. However, the similarities start to end. First, look at the pattern of the function. Over $\lambda=2\pi$, the periodic function $\sin(x)$ follows the basic pattern of "middle," "top," "middle," "bottom," "middle." That is, the value of $f(x)=\sin(x)$ starts at $f(0)=0$. However, $f\left(\frac{\pi}{2}\right)=1$. This is the highest possible value of sine when looking at the positive values, because the unit circle is peaked when $\theta=\frac{\pi}{2}$. At $x=\pi$, $f\left(\pi\right)=0$. This has again to do with the unit circle. If you remembered the patterns you identified in the circle (or even the unit circle itself) you can infer how$f(x)=\sin(x)$will look like. The minimum value of the range is at $-1$ precisely for this reason. As such, the pattern we identified is always true.

By contrast, the cosine function follows a more simplistic pattern. For $g(x)=\cos(x)$, you already start at the maximum value because $g(0)=\cos(0)=1$ is the largest $x$-value on the circle, while$g\left(\pi\right)=\cos\left(\pi\right)=-1$ is the lowest $x$-value on the circle. Because of the unit circle, the simplistic pattern results: "maximum," "medium," "minimum," "medium," "maximum."

The patterns mentioned above are things that you should keep in the back of your mind. These patterns are the main ways in which a person can easily identify when to use one particular function over another. When a phase shift is applied, choosing to use a cosine function or a sine function is best left to student judgement.

From the information we have so far, we get the following information:

$$ \begin{aligned} \text{The vertical shift } c &= \frac{y _{\text{max}} + y _{\text{min}}}{2} = \frac{1 + (-1)}{2} = \frac{0}{2} = 0. \\\\ \text{The amplitude } a &= \frac{y _{\text{max}} - y _{\text{min}}}{2} = \frac{1 - (-1)}{2} = \frac{2}{2} = 1. \\\\ \text{The frequency } b &= \frac{2\pi}{\lambda} = \frac{2\pi}{2\pi} = 1. \end{aligned} $$

\textbf{The general form of the sinusoidal function is}

$A\sin(B(x-D))+C$

\begin{itemize}
    \item The sine can be substituted for cosine and it would work just the same.
    \item $A$ is the amplitude.
    \item $B$ is the frequency.
    \item $C$ is the vertical shift (sinusoidal axis value).
    \item $D$ is the phase (horizontal) shift.
\end{itemize}

We have described the general pattern for the cosine and sine functions. However, it is important to step back and appreciate the underlying math for which the cosine and sine curves rely on, the unit circle. While these functions are beautiful, the underlying math behind why these functions behave the way they do is the reason for this beauty, and in a way, makes it more beautiful than the end result.

% [Image: ]

Figure: Increase in radius

Draw two circles with radii $r$ and $R$ centered at the origin, where $r<R$. If a line extending from the origin reaches an endpoint of each circle at the angle $\theta$, then $\cos\theta=\frac{x _{1}}{r}$, and $\cos\theta=\frac{x _{2}}{R}$. As such,

\begin{itemize}
    \item $x _{1}=r\cos\theta$
    \item $x _{2}=R\cos\theta$
\end{itemize}

Because $\cos\theta$ is multiplied directly to $r$ or $R$, for $r<R$, $r\cos\theta<R\cos\theta$.

Let $f(\theta)=r\cos\theta$, $g(\theta)=R\cos\theta$, and $r=1$. Because we know how to graph $f(\theta)=\cos\theta$, $R$ is multiplied directly to $\cos\theta$, and $g(\theta)=Rf(\theta)$, the resulting function must be "expanded" over the $y$-axis. This expansion must be related to the amplitude of the function, because $R y _{max}$ and $R y _{min}$ gives $\frac{R y _{max}-R y _{min}}{2}=\frac{R\left( y _{max}-y _{min}\right)}{2}=R\cdot\frac{y _{max}-y _{min}}{2}=Ra=R$. (Recall $a=1$).

\begin{itemize}
    \item The amplitude of the function is given by the radius of the circle.
    \item Changing the radius of the circle does not change the frequency, the vertical shift, or the phase shift. (\textbf{Exercise}: show that this is true for at least the frequency and the vertical shift.)
\end{itemize}

This result will be useful when working with problems later.

As of now, increasing the radius is all that we could connect these sinusoidal functions to the circle. We need to understand more about additions and such before one can address the beauty of this math. Nevertheless, always keep this result locked in your heart.

\paragraph{Modeling Situations}


The sinusoidal functions are very useful in real life applications, where these \textit{periodic} (repeating) functions model situations that follow a pattern over a set amount of time, such as in modeling day length, planetary motion, sound waves, energy waves, etcetera.

\textbf{Fish length}

\begin{itemize}
    \item Marine biologists determined the average length of $N=50,000$ fish in a pond in millimeters (mm) is dependent on the time, in years. The table is shown below, where $t$ is the time, in years, and $l(t)$ is the average length of the fish, in millimeters. Find the sinusoidal function that best models the information provided.
\end{itemize}

$\begin{array}{|c|c|} \hline t & l(t) \\\\ \hline 0 & 44 \\\\ \hline 3 & 40 \\\\ \hline 6 & 36 \\\\ \hline 9 & 40 \\\\ \hline 12 & 44 \\\\ \hline \vdots & \vdots \end{array} $

\begin{itemize}
    \item The situation is given to be periodic and sinusoidal. Therefore, it is best to use the sinusoidal function and its properties: $a\text{ trig}(b(x+d))+c$.
\end{itemize}

\begin{itemize}
    \item Find the vertical shift, $c$. Recall that $c=\frac{y _{max}+y _{min}}{2}$. According to the table, the maximum y-value is 44, while the minimum value is 36. Therefore, the vertical shift is
\end{itemize}

$$\frac{44+36}{2}=40$$

\begin{itemize}
    \item Find the amplitude, $a$. Recall that $a=\frac{y _{max}-y _{min}}{2}$. According to the table, the maximum y-value is 44, while the minimum value is 36. Therefore, the amplitude is
\end{itemize}

$$\frac{44-36}{2}=4$$

\begin{itemize}
    \item Find the frequency, $b$. Recall that $b=\frac{2\pi}{\lambda}$. If the wavelength is the distance whereby a complete cycle occurs, then from $t=0\text{ sec}$ to $t=12\text{ sec}$, the cycle is complete. As such, $\lambda=12-0=12$. Given this information,
\end{itemize}

$$b=\frac{2\pi}{12}=\frac{\pi}{6}$$

\begin{itemize}
    \item For now, let us ignore the phase shift, $d$. Substitute the information we already know into the function above: $4\text{ trig}\left(\frac{\pi}{6}t\right)+40$. Recall that a sinusoidal function that has a maximum value reappear again over a wavelength (from maximum to medium to minimum to medium to maximum) is cosine. As such $4\cos\left(\frac{\pi}{6}t\right)+40$ models the situation sufficiently.
\end{itemize}

\begin{itemize}
    \item Equivalently, assuming we want to keep it in terms of sine, we know the following is true:
    \item $\cos\left(\theta\right)=\sin\left(\frac{\pi}{2}-\theta\right)=\sin\left(\frac{\pi}{2}+\theta\right)=-\sin\left(\theta-\frac{\pi}{2}\right)$
    \item Therefore, $4\cos\left(\frac{\pi}{6}t\right)+40=4\sin\left(\frac{\pi}{6}\left(\frac{\pi}{2}-t\right)\right)+40=4\sin\left(\frac{\pi}{6}\left(\frac{\pi}{2}+t\right)\right)+40=-4\sin\left(\frac{\pi}{6}\left(t-\frac{\pi}{2}\right)\right)+40$.`{{Robox/Close}}`{=mediawiki}
\end{itemize}

The above question is comparatively easier to other similar problems like this. In the real world, you are less likely to see a steady increase/decrease in fish length because of two factors:

\begin{itemize}
    \item \textbf{Evolution}: bigger fish are usually seen as predators, which are often avoided. Smaller fish are more likely to be seen as weaker.
    \item \textbf{Internal Competition}: size among females may be seen as more desirable as mates, and therefore more likely to pass on offspring with those traits. As such, competition may exist among females for desirable mates.
\end{itemize}

There are many questions a person may be asking: where is the above circle in the question? Why use radians as opposed to degrees? . There are a few reasons, but it requires us to delve more deeply into mathematics that is usually considered advanced (polar coordinates). Unfortunately, Book 1 will not delve into this topic. However, for those that are curious, Book 3 will have polar coordinates that fully explain this behavior.

\paragraph{Resources}


\begin{itemize}
    \item \href{https://mathresearch.utsa.edu/wikiFiles/MAT1093/Phase\%20shift\%20and\%20Applications/Esparza\%201093\%20Notes\%202.6.pdf}{Phase shift and Applications}. Written notes created by Professor Esparza, UTSA.
    \item \href{https://en.wikibooks.org/wiki/Trigonometry/Phase_and_Frequency}{Phase and Frequencies}, Wikibooks: Trigonometry/Calculus
\end{itemize}

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikibooks.org/wiki/Trigonometry/Phase_and_Frequency}{Phase and Frequencies, Wikibooks: Trigonometry/Calculus} under a CC BY-SA license
\end{itemize}

\subsection{Learning Objectives}

\begin{enumerate}
    \item Graph variations of y=sin(x) and y=cos(x).
    \item Use phase shifts of sine and cosine curves.
\end{enumerate}

%%===============================================================
\section{Section 3.2 – Graphs of Other Trigonometric Functions}
\label{sec:section-3-2-graphs-of-other-trigonometric-functions}
%%===============================================================

\paragraph{Tangent}


% [Image: Graph of tangent]

Figure: Graph of tangent

The \textbf{tangent} of an angle is equivalent to the sine of that angle divided by the cosine of that angle.

$$\tan x = \frac{\sin{x}}{\cos{x}}$$

The range of the function $y = \tan{x}$ is $(-\infty ,\infty)$. Note that since $\tan{x} = \frac{\sin{x}}{\cos{x}}$, $y = \tan{x}$ is not defined whenever $\cos{x} = 0$; that is, when $x = \pi k + \frac{\pi}{2}$ for any integer k. We can see that there is a vertical asymptote at each of these values in the graph of $y = \tan{x}$.

\paragraph{Cotangent}


% [Image: Graph of cotangent]

Figure: Graph of cotangent

The \textbf{cotangent} of an angle is the reciprocal of its tangent.

$$\cot x = \frac{1}{\tan x}$$

The range of the function $y = \cot{x}$ is $(-\infty ,\infty)$, the same as the range of its reciprocal $\tan{x}$. 
$$\cot{x} = \frac{1}{\tan{x}} = \frac{\cos{x}}{\sin{x}}\text{,}$$
 so $y = \cot{x}$ is not defined whenever $\sin{x} = 0$; that is, when $x = \pi k$ for any integer k. We can see that there is a vertical asymptote at each of these values in the graph of $y = \cot{x}$.

\paragraph{Secant}


% [Image: Graph of secant]

Figure: Graph of secant

The \textbf{secant} of an angle is the reciprocal of its cosine.

$$\sec x = \frac{1}{\cos x}$$

Like $y=\cos{x}$, $y = \sec{x}$ is an even function; that is, $\sec{(x)} = \sec{(-x)}$. The range of the function $y = \cos{x}$ is $[-1, 1]$, so the range of $\sec{x}$ is $(-\infty , -1] \cup [1, \infty )$, which is all possible outputs of 1/x for $-1\leq x\leq 1$. Since $\frac{1}{0}$ is not defined, $y = \sec{x}$ is not defined whenever $\cos{x} = 0$; that is, when $x = \pi k + \frac{\pi}{2}$ for any integer k. We can see that there is a vertical asymptote at each of these values in the graph of $y = \sec{x}$.

\paragraph{Cosecant}


% [Image: Graph of cosecant]

Figure: Graph of cosecant

The \textbf{cosecant} of an angle is the reciprocal of its sine.

$\csc{x} = \frac{1}{\sin{x}}$

Like $y=\sin{x}$, $y = \csc{x}$ is an odd function; that is, $\csc{(x)} = -\csc{(-x)}$. The range of the function $y = \csc{x}$ is the same as the range for $\sec{x}$ ($(-\infty , -1] \cup [1, \infty )$). $y = \csc{x}$ is not defined whenever $\sin{x} = 0$; that is, when $x = \pi k$ for any integer k. We can see that there is a vertical asymptote at each of these values in the graph of $y = \csc{x}$.

\paragraph{Resources}


\begin{itemize}
    \item \href{https://mathresearch.utsa.edu/wikiFiles/MAT1093/Graphs\%20of\%20the\%20Tangent,\%20Cotangent,\%20Cosecant\%20and\%20Secant\%20Functions/Esparza\%201093\%20Notes\%202.5.pdf}{Graphs of the Tangent, Cotangent, Cosecant and Secant Functions}. Written notes created by Professor Esparza, UTSA.
    \item \href{https://en.wikibooks.org/wiki/A-level_Mathematics/CIE/Pure_Mathematics_2/Trigonometry}{Trigonometry}, WikiBooks: Pure Mathematics 2
\end{itemize}

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikibooks.org/wiki/A-level_Mathematics/CIE/Pure_Mathematics_2/Trigonometry}{Trigonometry, WikiBooks: Pure Mathematics 2} under a CC BY-SA license
\end{itemize}

\subsection{Learning Objectives}

\begin{enumerate}
    \item Analyze the graph of y=tan x.
    \item Graph variations of y=tan x.
    \item Analyze the graphs of y=sec x and y=csc x.
    \item Graph variations of y=sec x and y=csc x.
    \item Analyze the graph of y=cot x.
    \item Graph variations of y=cot x.
\end{enumerate}

\subsection{Assessment Instrument}

\begin{enumerate}
    \item \textbf{Assessment Instrument: Graphs of the Tangent, Cotangent, Secant, and Cosecant Functions Quiz} \textbf{Format:}
    \item \textbf{Type:} Mixed-format quiz (multiple-choice, short answer, and problem-solving questions)
    \item \textbf{Duration:} 45 minutes
    \item \textbf{Total Marks:} 30 points \textbf{Sections:}
    \item \textbf{Understanding the Tangent and Cotangent Functions (10 points):}
    \item \textbf{Question 1:} (Multiple Choice) Which of the following correctly represents the function for the tangent of an angle?
    \item a. $\tan x = \frac{\cos{x}}{\sin{x}}$
    \item b. $\tan x = \frac{\sin{x}}{\cos{x}}$
    \item c. $\tan x = \frac{1}{\cos{x}}$
    \item d. $\tan x = \frac{1}{\sin{x}}$
    \item \textbf{Question 2:} (Short Answer) At which points is the tangent function undefined? Provide the general form for these points.
    \item \textbf{Question 3:} (Problem-Solving) Given the function $y = \tan x$, sketch its graph over the interval $[-\pi, \pi]$, indicating the vertical asymptotes and the key points where the function crosses the x-axis.
    \item \textbf{Understanding the Secant and Cosecant Functions (10 points):}
    \item \textbf{Question 4:} (Multiple Choice) The secant function is the reciprocal of which trigonometric function?
    \item a. Tangent
    \item b. Sine
    \item c. Cosine
    \item d. Cotangent
    \item \textbf{Question 5:} (Short Answer) At which points is the secant function undefined? Provide the general form for these points.
    \item \textbf{Question 6:} (Problem-Solving) Sketch the graph of $y = \sec x$ over the interval $[-\pi, \pi]$. Mark the vertical asymptotes and indicate the range of the function.
    \item \textbf{Analysis and Interpretation of Trigonometric Graphs (10 points):}
    \item \textbf{Question 7:} (Multiple Choice) The cotangent function has vertical asymptotes at which of the following points?
    \item a. $x = \frac{\pi}{2} + \pi k$ for any integer $k$
    \item b. $x = \pi k$ for any integer $k$
    \item c. $x = \frac{\pi}{2} k$ for any integer $k$
    \item d. $x = \pi + \pi k$ for any integer $k$
    \item \textbf{Question 8:} (Short Answer) Explain why the cosecant function has the same range as the secant function.
    \item \textbf{Question 9:} (Problem-Solving) Sketch the graph of $y = \csc x$ over the interval $[-\pi, \pi]$. Mark the vertical asymptotes and indicate the points where the function is undefined. \textbf{Scoring Rubric:}
    \item \textbf{Understanding the Tangent and Cotangent Functions:}
    \item Multiple Choice: 2 points for each correct answer (total 2 points)
    \item Short Answer: 2 points for correctly identifying the undefined points (total 2 points)
    \item Problem-Solving: 6 points for accurately sketching the graph, marking the vertical asymptotes, and identifying key points (total 6 points)
    \item \textbf{Understanding the Secant and Cosecant Functions:}
    \item Multiple Choice: 2 points for each correct answer (total 2 points)
    \item Short Answer: 2 points for correctly identifying the undefined points (total 2 points)
    \item Problem-Solving: 6 points for accurately sketching the graph, marking the vertical asymptotes, and indicating the range (total 6 points)
    \item \textbf{Analysis and Interpretation of Trigonometric Graphs:}
    \item Multiple Choice: 2 points for each correct answer (total 2 points)
    \item Short Answer: 2 points for correctly explaining the range (total 2 points)
    \item Problem-Solving: 6 points for accurately sketching the graph, marking the vertical asymptotes, and identifying points where the function is undefined (total 6 points)
\end{enumerate}

%%===============================================================
\section{Section 3.3 – Inverse Functions}
\label{sec:section-3-3-inverse-functions}
%%===============================================================

The function $g$ is the inverse of the one-to-one function $f$ if and only if the following are true:
$$ \begin{aligned} g(f(x)) & = x \hspace{44em} \\\\ f(g(x)) & = x \end{aligned} $$

The inverse of function $f$ is denoted as $f^{-1}$.

Geometrically $f^{-1}$ is the reflection of $f$ across the line $y = x$.
Conceptually, using the \textit{box} analogy, a function's inverse box \textit{undoes} what the function's regular box does.

% [Image: Fruit function and inverse]

Figure: Fruit function and inverse

Example:

$$ \begin{aligned} f(x) & = 2x \hspace{44em} \\\\ f^{-1}(x) & = \frac{1}{2}x \\\\ f(f^{-1}(x)) & = f\left(\frac{1}{2}x\right) \\\\ f(f^{-1}(x)) & = 2\left(\frac{1}{2}x\right) \\\\ f(f^{-1}(x)) & = x \\\\ f^{-1}(f(x)) & = f^{-1}(2x) \\\\\ f^{-1}(f(x)) & = \frac{1}{2}(2x) \\\\ f^{-1}(f(x)) & = x \end{aligned} $$

To find the inverse of a function, remember that when we use $f^{-1}(x)$ as an input to $f$ the result is $x$. So start by writing $x = f(f^{-1}(x))$ and solve for $f^{-1}$.

Example:

$$ \begin{aligned} \text{Suppose: } && f(x) & = 2x + 1 \hspace{44em} \\\\ \text{Then } && x & = f(f^{-1}(x)) \\\\ && x & = 2f^{-1}(x) + 1 \\\\ && x - 1 & = 2f^{-1}(x) \\\\ && f^{-1}(x) & = \frac{x - 1}{2} \end{aligned} $$

The Domain of an inverse function is exactly the same as the Range of the original function.  If the Range of the original function is limited in some way, the inverse of a function will require a restricted domain.

Example:

$$ \begin{aligned} f(x) & = \sqrt{x - 1} \hspace{44em} \\\\ x & = f(f^{-1}(x)) \\\\ x & = \sqrt{f^{-1}(x) - 1} \\\\ x^2 & = \left(\sqrt{f^{-1}(x) - 1}\right)^2 \\\\ x^2 & = f^{-1}(x) - 1 \\\\ f^{-1}(x) & = x^2 + 1 \end{aligned} $$

The Range of $f(x)$ is $f(x) \ge 0$. So the Domain of $f^{-1}(x)$ is $x \ge 0$.

\paragraph{One-to-one function}

A function that for every input there exists an output unique to that input.

Equivalently, we may say that a function $f$ is called \textit{one-to-one} if for all
$x, x' \in A, f(x) = f(x')$ implies that $x = x'$ where \textit{A} is the domain set of \textit{f} and both \textit{x} and \textit{x}' are members of that set.

\textbf{Horizontal Line Test}
If no horizontal line intersects the graph of a function in more than one place then the function is a one-to-one function.

\paragraph{The Existence of Inverse Functions}

A function has an inverse if and only if it is one-to-one: that is, if for each $y$-value there is only \textbf{one} corresponding $x$-value. To test whether or not a function is one-to-one, we can draw multiple horizontal lines through the graph of the function. If any of these horizontal lines intersects with the graph of the function more than once, then the function is NOT one-to-one.

For example, take the graph of the function $f: \mathbb{R} \to \mathbb{R}, f(x) = x^2$.

We can draw a horizontal line anywhere on the graph (except for the turning point at $x = 0$) which will intersect with the graph twice. Therefore \textit{f} is NOT a one-to-one function, and will NOT have an inverse function.

However, consider the function $g: \mathbb{R} \to \mathbb{R}, g(x) = x^3$.

This function IS one-to-one, and will therefore have an inverse function, which we label $g^{-1}$.

\paragraph{Resources}

\begin{itemize}
    \item \href{https://mathresearch.utsa.edu/wikiFiles/MAT1093/Inverse\%20functions/Esparza\%201093\%20Notes\%201.7.pdf}{Inverse Functions. Written notes created by Professor Esparza, UTSA}
    \item \href{https://en.wikibooks.org/wiki/Algebra/Functions}{Functions, Wikibooks: Algebra}
    \item \href{https://en.wikibooks.org/wiki/VCE_Mathematical_Methods/Inverse_functions}{Inverse Functions, Wikibooks: VCE Mathematical Methods}
\end{itemize}

\paragraph{Licensing}

Content obtained and/or adapted from:
\begin{itemize}
    \item \href{https://en.wikibooks.org/wiki/Algebra/Functions}{Functions, Wikibooks: Algebra} under a CC BY-SA license
    \item \href{https://en.wikibooks.org/wiki/VCE_Mathematical_Methods/Inverse_functions}{Inverse Functions, Wikibooks: VCE Mathematical Methods} under a CC BY-SA license
\end{itemize}

\subsection{Learning Objectives}

\begin{enumerate}
    \item Verify inverse functions.
    \item Determine the domain and range of an inverse function, and restrict the domain of a function to make it one-to-one.
    \item Find or evaluate the inverse of a function.
    \item Use the graph of a one-to-one function to graph its inverse function on the same axes.
\end{enumerate}

\subsection{Assessment Instrument}

\begin{enumerate}
    \item \textbf{Assessment Instrument: Inverse Functions Quiz} \textbf{Format:}
    \item \textbf{Type:} Mixed-format quiz (multiple-choice, short answer, and problem-solving questions)
    \item \textbf{Duration:} 45 minutes
    \item \textbf{Total Marks:} 30 points \textbf{Sections:}
    \item \textbf{Understanding Inverse Functions (10 points):}
    \item \textbf{Question 1 (Multiple Choice):} Which of the following is the correct inverse function of $f(x) = 3x + 4$?
    \item a) $f^{-1}(x) = \frac{x - 4}{3}$
    \item b) $f^{-1}(x) = 3x - 4$
    \item c) $f^{-1}(x) = \frac{3x + 4}{x}$
    \item d) $f^{-1}(x) = \frac{x + 4}{3}$
    \item \textbf{Question 2 (Short Answer):} Given the function $f(x) = \frac{2}{x} + 3$, find the inverse function $f^{-1}(x)$.
    \item \textbf{Geometric Interpretation of Inverse Functions (10 points):}
    \item \textbf{Question 3 (Problem-Solving):} Sketch the graph of the function $f(x) = x^2 + 1$ for $x \ge 0$. Then, sketch the graph of its inverse function $f^{-1}(x)$. Explain how the graphs of $f(x)$ and $f^{-1}(x)$ are related geometrically.
    \item \textbf{Question 4 (Short Answer):} Explain why the function $f(x) = x^3 - 2$ is its own inverse. Provide a brief justification using the properties of inverse functions.
    \item \textbf{Determining the Existence of Inverse Functions (10 points):}
    \item \textbf{Question 5 (Multiple Choice):} Which of the following functions has an inverse?
    \item a) $f(x) = x^2$
    \item b) $f(x) = 2x + 5$
    \item c) $f(x) = |x|$
    \item d) $f(x) = \sin(x)$
    \item \textbf{Question 6 (Problem-Solving):} Use the Horizontal Line Test to determine if the function $f(x) = x^3 - 6x + 1$ has an inverse. Justify your answer with a brief explanation. \textbf{Scoring Rubric:}
    \item \textbf{Understanding Inverse Functions:}
    \item Question 1: 3 points for correct selection
    \item Question 2: 7 points for correctly solving the inverse function
    \item \textbf{Geometric Interpretation of Inverse Functions:}
    \item Question 3: 5 points for accurate sketches and explanation
    \item Question 4: 5 points for correct explanation and justification
    \item \textbf{Determining the Existence of Inverse Functions:}
    \item Question 5: 3 points for correct selection
    \item Question 6: 7 points for correct use of the Horizontal Line Test and explanation This quiz will assess students' understanding of the fundamental concepts of inverse functions, including their algebraic and geometric interpretations and the criteria for their existence.
\end{enumerate}

%%===============================================================
\section{Section 3.4 – Inverse Trigonometric Functions}
\label{sec:section-3-4-inverse-trigonometric-functions}
%%===============================================================

In mathematics, the \textbf{inverse trigonometric functions} (occasionally also called \textbf{arcus functions}, \textbf{antitrigonometric functions} or \textbf{cyclometric functions}) are the inverse functions of the trigonometric functions (with suitably restricted domains). Specifically, they are the inverses of the sine, cosine, tangent, cotangent, secant, and cosecant functions, and are used to obtain an angle from any of the angle''s trigonometric ratios. Inverse trigonometric functions are widely used in engineering, navigation, physics, and geometry.

\paragraph{Notation}

Several notations for the inverse trigonometric functions exist. The most common convention is to name inverse trigonometric functions using an arc- prefix: $ \arcsin(x) $, $ \arccos(x) $, $ \arctan(x) $, etc. This notation arises from the following geometric relationships: when measuring in radians, an angle of $\theta$ radians will correspond to an arc whose length is $ r\theta $, where $r$ is the radius of the circle. Thus in the unit circle, "the arc whose cosine is $x$" is the same as "the angle whose cosine is $x$", because the length of the arc of the circle in radii is the same as the measurement of the angle in radians. In computer programming languages, the inverse trigonometric functions are often called by the abbreviated forms asin, acos, atan.

The notations $ \sin^{-1}(x) $, $ \cos^{-1}(x) $, $ \tan^{-1}(x) $, etc., as introduced by John Herschel in 1813, are often used as well in English-language sources—conventions consistent with the notation of an inverse function. This might appear to conflict logically with the common semantics for expressions such as $ \sin^2(x) $, which refer to numeric power rather than function composition, and therefore may result in confusion between multiplicative inverse or reciprocal and compositional inverse. The confusion is somewhat mitigated by the fact that each of the reciprocal trigonometric functions has its own name—for example, $ (\cos(x))^{-1} = \sec(x) $. Nevertheless, certain authors advise against using it for its ambiguity. Another convention used by a few authors is to use an uppercase first letter, along with a $ ^{-1} $ superscript: $ \sin^{-1}(x) $, $ \cos^{-1}(x) $, $ \tan^{-1}(x) $, etc. This potentially avoids confusion with the multiplicative inverse, which should be represented by $ \sin^{-1}(x) $, $ \cos^{-1}(x) $, etc.

Since 2009, the ISO 80000-2 standard has specified solely the "arc" prefix for the inverse functions.

\paragraph{Basic concepts}


\paragraph{Principal values}

Since none of the six trigonometric functions are one-to-one, they must be restricted in order to have inverse functions. Therefore, the result ranges of the inverse functions are proper subsets of the domains of the original functions.

For example, using function in the sense of multivalued functions, just as the square root function $ y = \sqrt{x} $ could be defined from $ y^2 = x $, the function $ y = \arcsin(x) $ is defined so that $ \sin(y) = x $. For a given real number $ x $, with $ -1 \leq x \leq 1 $, there are multiple (in fact, countably infinite) numbers $ y $ such that $ \sin(y) = x $; for example, $ \sin(0) = 0 $, but also $ \sin(\pi) = 0 $, $ \sin(2 \pi) = 0 $, etc. When only one value is desired, the function may be restricted to its principal branch. With this restriction, for each $ x $ in the domain, the expression $ \arcsin(x) $ will evaluate only to a single value, called its principal value. These properties apply to all the inverse trigonometric functions.

The principal inverses are listed in the following table.


\paragraph{Table 1: Name, Usual Notation, Definition, Domain of \(x\) for Real Result}


$$ \begin{array}{|c|c|c|c|} \hline \textbf{Name} & \textbf{Usual notation} & \textbf{Definition} & \textbf{Domain of } x \textbf{ for real result} \\\\ \hline \text{arcsine} & y = \arcsin(x) & x = \sin(y) & -1 \leq x \leq 1 \\\\ \hline \text{arccosine} & y = \arccos(x) & x = \cos(y) & -1 \leq x \leq 1 \\\\ \hline \text{arctangent} & y = \arctan(x) & x = \tan(y) & \text{all real numbers} \\\\ \hline \text{arccotangent} & y = \text{arccot}(x) & x = \cot(y) & \text{all real numbers} \\\\ \hline \text{arcsecant} & y = \text{arcsec}(x) & x = \sec(y) & x \leq -1 \text{ or } x \geq 1 \\\\ \hline \text{arccosecant} & y = \text{arccsc}(x) & x = \csc(y) & x \leq -1 \text{ or } x \geq 1 \\\\ \hline \end{array} $$

\paragraph{Table 2: Range of Usual Principal Value (Radians), Range of Usual Principal Value (Degrees)}


$$ \begin{array}{|c|c|c|} \hline \textbf{Name} & \textbf{Range of usual principal value (radians)} & \textbf{Range of usual principal value (degrees)} \\\\ \hline \text{arcsine} & -\frac{\pi}{2} \leq y \leq \frac{\pi}{2} & -90^{\circ} \leq y \leq 90^{\circ} \\\\ \hline \text{arccosine} & 0 \leq y \leq \pi & 0^{\circ} \leq y \leq 180^{\circ} \\\\ \hline \text{arctangent} & -\frac{\pi}{2} < y < \frac{\pi}{2} & -90^{\circ} < y < 90^{\circ} \\\\ \hline \text{arccotangent} & 0 < y < \pi & 0^{\circ} < y < 180^{\circ} \\\\ \hline \text{arcsecant} & 0 \leq y < \frac{\pi}{2} \text{ or } \frac{\pi}{2} < y \leq \pi & 0^{\circ} \leq y < 90^{\circ} \text{ or } 90^{\circ} < y \leq 180^{\circ} \\\\ \hline \text{arccosecant} & -\frac{\pi}{2} \leq y < 0 \text{ or } 0 < y \leq \frac{\pi}{2} & -90^{\circ} \leq y < 0^{\circ} \text{ or } 0^{\circ} < y \leq 90^{\circ} \\\\ \hline \end{array} $$

Note: Some authors define the range of arcsecant to be $ \big( 0 \leq y < \tfrac{\pi}{2} \text{ or } \pi < y \leq \tfrac{3 \pi}{2} \big) $, because the tangent function is nonnegative on this domain. This makes some computations more consistent. For example, using this range, $ \tan(\text{arcsec}(x)) = \sqrt{x^2 - 1} $, whereas with the range $ \big( 0 \leq y < \tfrac{\pi}{2} \text{ or } \tfrac{\pi}{2} < y \leq \pi \big) $, we would have to write $ \tan(\text{arcsec}(x)) = \pm \sqrt{x^2 - 1} $, since tangent is nonnegative on $ 0 \leq y < \tfrac{\pi}{2} $, but nonpositive on $ \tfrac{\pi}{2} < y \leq \pi $. For a similar reason, the same authors define the range of arccosecant to be $ \big(- \pi < y \leq - \tfrac{\pi}{2} \text{ or } 0 < y \leq \tfrac{\pi}{2} \big) $.

If $ x $ is allowed to be a complex number, then the range of $ y $ applies only to its real part.

\paragraph{Solutions to elementary trigonometric equations}


Each of the trigonometric functions is periodic in the real part of its argument, running through all its values twice in each interval of $ 2 \pi $:

\begin{itemize}
    \item Sine and cosecant begin their period at $ 2 \pi k - \frac{\pi}{2} $ (where $ k $ is an integer), finish it at $ 2 \pi k + \frac{\pi}{2} $, and then reverse themselves over $ 2 \pi k + \frac{\pi}{2} $, to $ 2 \pi k + \frac{3\pi}{2} $.
    \item Cosine and secant begin their period at $ 2 \pi k $, finish it at $ 2 \pi k + \pi $. and then reverse themselves over $ 2 \pi k + \pi $ to $ 2 \pi k + 2 \pi $.
    \item Tangent begins its period at $ 2 \pi k - \frac{\pi}{2} $, finishes it at $ 2 \pi k + \frac{\pi}{2} $, and then repeats it (forward) over $ 2 \pi k + \frac{\pi}{2} $ to $ 2 \pi k + \frac{3 \pi}{2} $.
    \item Cotangent begins its period at $ 2 \pi k $, finishes it at $ 2 \pi k + \pi $, and then repeats it (forward) over $ 2 \pi k + \pi $ to $ 2 \pi k + 2 \pi $.
\end{itemize}

This periodicity is reflected in the general inverses, where $ k $ is some integer.

For example, if $ \cos \theta = -1 $ then $ \theta = \pi + 2 \pi k = -\pi + 2 \pi (1 + k) $ for some $ k \in \mathbb{Z} $. While if $ \sin \theta = \pm 1 $ then $ \theta = \frac{\pi}{2} + \pi k = - \frac{\pi}{2} + \pi (k + 1) $ for some $ k \in \mathbb{Z} $, where $ k $ will be even if $ \sin \theta = 1 $ and it will be odd if $ \sin \theta = -1 $. The equations $ \sec \theta = -1 $ and $ \csc \theta = \pm 1 $ have the same solutions as $ \cos \theta = -1 $ and $ \sin \theta = \pm 1 $, respectively. In all equations above except for those just solved (i.e. except for $ \sin/\csc \theta = \pm 1 $ and $ \cos/\sec \theta = - 1 $), the integer $ k $ in the solution''s formula is uniquely determined by $ \theta $ (for fixed $ r, s, x $, and $ y $).

\paragraph{Detailed example and explanation of the "plus or minus" symbol $ \pm $}


The solutions to $ \cos \theta = x $ and $ \sec \theta = x $ involve the "plus or minus" symbol $ \,\pm,\, $ whose meaning is now clarified. Only the solution to $ \cos \theta = x $ will be discussed since the discussion for $ \sec \theta = x $ is the same.
We are given $ x $ between $ -1 \leq x \leq 1 $ and we know that there is an angle $ \theta $ in some give interval that satisfies $ \cos \theta = x $. We want to find this $ \theta $. The formula for the solution involves:
$$ \pm \arccos x. $$
If $ \,\arccos x = 0\, $ (which only happens when $ x = 1 $) then $ \,+ \arccos x = 0\, $ and $ \,- \arccos x = 0\, $ so either way, $ \,\pm \arccos x\, $ can only be equal to $ 0 $. But if $ \,\arccos x \neq 0,\, $ which will now be assumed, then the solution to $ \cos \theta = x $, which is written above as
$$ \,\theta = \pm \arccos x + 2 \pi k\, \quad \text{ for some }k \in \mathbb{Z} $$
is shorthand for the following statement:
Either
\begin{itemize}
    \item $ \,\theta = \arccos x + 2 \pi k\, $ for some integer $ k $,
\end{itemize}
or else
\begin{itemize}
    \item $ \,\theta = - \arccos x + 2 \pi k\, $ for some integer $ k $.
\end{itemize}
Because $ \,\arccos x \neq 0\, $ and $ \,0 < \arccos x \leq \pi\, $ exactly one of these two equalities can hold. Additional information about $ \theta $ is needed to determine which one holds. For example, suppose that $ x = 0 $ and that all that is known about $ \theta $ is that $ \,- \pi \leq \theta \leq \pi\, $ (and nothing more is known). Then
$$ \arccos x = \arccos 0 = \frac{\pi}{2} $$
and moreover, in this particular case $ k = 0 $ (for both the $ \,+\, $ case and the $ \,-\, $ case) and so consequently,
$$ \theta ~=~ \pm \arccos x + 2 \pi k ~=~ \pm \left(\frac{\pi}{2}\right) + 2\pi (0) ~=~ \pm \frac{\pi}{2}. $$
This means that $ \theta $ could be either $ \,\pi/2\, $ or $ \,- \pi/2. $ Without additional information it is not possible to determine which of these values $ \theta $ has.
An example of some additional information that could determine the value of $ \theta $ would be knowing that the angle is above the $ x $-axis (in which case $ \theta = \pi/2 $) or alternatively, knowing that it is below the $ x $-axis (in which case $ \theta = - \pi/2 $).

\paragraph{Transforming equations}


The equations above can be transformed by using the identities

$$ \begin{array}{|c|c|c|c|c|} \hline \textbf{func} & \textbf{func}(-\theta) & \textbf{func}\left(\frac{\pi}{2} - \theta\right) & \textbf{func}(\pi - \theta) & \textbf{func}(2\pi - \theta) \\\\ \mid\mid & \mid\mid & \mid\mid & \mid\mid & \mid\mid \\\\ \hline \textbf{sin} & -\sin \theta & \cos \theta & \sin \theta & -\sin \theta \\\\ \hline \textbf{cos} & \cos \theta & \sin \theta & -\cos \theta & \cos \theta \\\\ \hline \textbf{tan} & -\tan \theta & \cot \theta & -\tan \theta & -\tan \theta \\\\ \hline \textbf{csc} & -\csc \theta & \sec \theta & \csc \theta & -\csc \theta \\\\ \hline \textbf{sec} & \sec \theta & \csc \theta & -\sec \theta & \sec \theta \\\\ \hline \textbf{cot} & -\cot \theta & \tan \theta & -\cot \theta & -\cot \theta \\\\ \hline \end{array} $$

So for example, by using the equality $ \sin \left(\frac{\pi}{2} - \theta\right) = \cos \theta $, the equation $ \cos \theta = x $ can be transformed into $ \sin \left(\frac{\pi}{2} - \theta\right) = x $, which allows for the solution to the equation $ \;\sin \varphi = x\; $ (where $ \varphi := \frac{\pi}{2} - \theta $) to be used; that solution being:
$$ \varphi  = (-1)^k \arcsin (x) + \pi k \; \text{ for some } k \in \mathbb{Z}, $$
which becomes:
$$ \frac{\pi}{2} - \theta ~=~ (-1)^k \arcsin (x) + \pi k \quad \text{ for some } k \in \mathbb{Z} $$
where using the fact that $ (-1)^{k} = (-1)^{-k} $ and substituting $ h := - k $ proves that another solution to $ \;\cos \theta = x\; $ is:
$$ \theta ~=~ (-1)^{h+1} \arcsin (x) + \pi h + \frac{\pi}{2} \quad \text{ for some } h \in \mathbb{Z}. $$
The substitution $ \;\arcsin x = \frac{\pi}{2} - \arccos x\; $ may be used express the right hand side of the above formula in terms of $ \;\arccos x\; $ instead of $ \;\arcsin x.\; $

\paragraph{Relationships between trigonometric functions and inverse trigonometric functions}


Trigonometric functions of inverse trigonometric functions are tabulated below. A quick way to derive them is by considering the geometry of a right-angled triangle, with one side of length 1 and another side of length $ x $, then applying the Pythagorean theorem and definitions of the trigonometric ratios. Purely algebraic derivations are longer. It is worth noting that for arcsecant and arccosecant, the diagram assumes that $ x $ is positive, and thus the result has to be corrected through the use of absolute values and the signum (sgn) operation.

$$ \begin{array}{|c|c|c|c|} \hline \theta & \sin(\theta) & \cos(\theta) & \tan(\theta) \\\\ \hline \arcsin(x) & \sin(\arcsin(x)) = x & \cos(\arcsin(x)) = \sqrt{1-x^2} & \tan(\arcsin(x)) = x / \sqrt{1-x^2} \\\\ \hline \arccos(x) & \sin(\arccos(x)) = \sqrt{1-x^2} & \cos(\arccos(x)) = x & \tan(\arccos(x)) = \sqrt{1-x^2} / x \\\\ \hline \arctan(x) & \sin(\arctan(x)) = x / \sqrt{1+x^2} & \cos(\arctan(x)) = 1 / \sqrt{1+x^2} & \tan(\arctan(x)) = x \\\\ \hline \text{arccot}(x) & \sin(\text{arccot}(x)) = 1 / \sqrt{1+x^2} & \cos(\text{arccot}(x)) = x / \sqrt{1+x^2} & \tan(\text{arccot}(x)) = 1/x \\\\ \hline \text{arcsec}(x) & \sin(\text{arcsec}(x)) = \sqrt{x^2-1} / |x| & \cos(\text{arcsec}(x)) = 1 / x & \tan(\text{arcsec}(x)) = \text{sgn}(x)\sqrt{x^2-1} \\\\ \hline \text{arccsc}(x) & \sin(\text{arccsc}(x)) = 1 / x & \cos(\text{arccsc}(x)) = \sqrt{x^2-1} / |x| & \tan(\text{arccsc}(x)) = \text{sgn}(x) / \sqrt{x^2-1} \\\\ \hline \end{array} $$

\paragraph{Relationships among the inverse trigonometric functions}


% [Image: The usual principal values of the arcsin and arccos functions]

Figure: The usual principal values of the arcsin(x) ($\textcolor{red}{\text{red}}$) and arccos(x) ($\textcolor{blue}{\text{blue}}$) functions graphed on the cartesian plane.

% [Image: The usual principal values of the arctan and arccot functions]

Figure: The usual principal values of the arctan(x) and arccot(x) functions graphed on the cartesian plane.

% [Image: Principal values of the arcsec and arccsc functions]

Figure: Principal values of the arcsec(x) and arccsc(x) functions graphed on the cartesian plane.

Complementary angles:
$$ \begin{aligned} \arccos(x) & = \frac{\pi}{2} - \arcsin(x) \hspace{44em} \\\\ \text{arccot}(x) & = \frac{\pi}{2} - \arctan(x) \\\\ \text{arccsc}(x) & = \frac{\pi}{2} - \text{arcsec}(x) \end{aligned} $$

Negative arguments:
$$ \begin{aligned} \arcsin(-x) & = -\arcsin(x) \hspace{44em} \\\\ \arccos(-x) & = \pi - \arccos(x) \\\\ \arctan(-x) & = -\arctan(x) \\\\ \text{arccot}(-x) & = \pi - \text{arccot}(x) \\\\ \text{arcsec}(-x) & = \pi - \text{arcsec}(x) \\\\ \text{arccsc}(-x) & = -\text{arccsc}(x) \end{aligned} $$

Reciprocal arguments:
$$ \begin{aligned} \arccos\left(\frac{1}{x}\right) & = \text{arcsec}(x) && \hspace{44em} \\\\ \arcsin\left(\frac{1}{x}\right) & = \text{arccsc}(x) \\\\ \arctan\left(\frac{1}{x}\right) & = \frac{\pi}{2} - \arctan(x) = \text{arccot}(x)  && \text{ if } x > 0 \\\\ \arctan\left(\frac{1}{x}\right) & = -\frac{\pi}{2} - \arctan(x) = \text{arccot}(x) - \pi && \text{ if } x < 0 \\\\ \text{arccot}\left(\frac{1}{x}\right) & = \frac{\pi}{2} - \text{arccot}(x) = \arctan(x)  && \text{ if } x > 0 \\\\ \text{arccot}\left(\frac{1}{x}\right) & = \frac{3\pi}{2} - \text{arccot}(x) = \pi + \arctan(x)  && \text{ if } x < 0 \\\\ \text{arcsec}\left(\frac{1}{x}\right) & = \arccos(x) \\\\ \text{arccsc}\left(\frac{1}{x}\right) & = \arcsin(x) \end{aligned} $$

Useful identities if one only has a fragment of a sine table:
$$ \begin{aligned} \arccos(x) & = \arcsin\left(\sqrt{1 - x^2}\right) && \text{ if } 0 \leq x \leq 1 \hspace{44em} \\\\ \\\\ \arccos(x) & = \arcsin\left(\frac{2x}{1 + x^2}\right) && \text{ if } 0 \leq x \leq 1 \\\\ \\\\ \arcsin(x) & = \frac{\pi}{2} - \text{sgn}(x)\arcsin(x) \\\\ \\\\ \arccos(x) & = \frac{1}{2}\arccos\left(2x^2-1\right) && \text{ if } 0 \leq x \leq 1 \\\\ \\\\ \arcsin(x) & = \frac{1}{2}\arccos\left(1-2x^2\right) && \text{ if } 0 \leq x \leq 1 \\\\ \\\\ \arcsin(x) & = \arctan\left(\frac{x}{\sqrt{1 - x^2}}\right) \\\\ \\\\ \arccos(x) & = \arctan\left(\frac{\sqrt{1 - x^2}}{x}\right) \\\\ \\\\ \arctan(x) & = \arcsin\left(\frac{x}{\sqrt{1 + x^2}}\right) \\\\ \\\\ \text{arccot}(x) & = \arccos\left(\frac{x}{\sqrt{1 + x^2}}\right) \end{aligned} $$

Whenever the square root of a complex number is used here, we choose the root with the positive real part (or positive imaginary part if the square was negative real).

A useful form that follows directly from the table above is

$$ \arctan\left(x\right) = \arccos\left(\sqrt{\frac{1}{1+x^2}}\right)\, , \text{ if } x\geq 0. $$

It is obtained by recognizing that $ \cos\left(\arctan\left(x\right)\right) = \sqrt{\frac{1}{1+x^2}} = \cos\left(\arccos\left(\sqrt{\frac{1}{1+x^2}}\right)\right) $.

From the half-angle formula, $$\tan\left(\tfrac{\theta}{2}\right) = \tfrac{\sin(\theta)}{1 + \cos(\theta)},$$ we get:
$$ \begin{aligned} \arcsin(x) & = 2 \arctan\left(\frac{x}{1 + \sqrt{1 - x^2}}\right) && \hspace{44em} \\\\ \arccos(x) & = 2 \arctan\left(\frac{\sqrt{1 - x^2}}{1 + x}\right) && \text{ if } -1 < x \leq  1 \\\\ \arctan(x) & = 2 \arctan\left(\frac{x}{1 + \sqrt{1 + x^2}}\right) \end{aligned} $$

\paragraph{Arctangent addition formula}

$$ \arctan(u) \pm \arctan(v) = \arctan\left(\frac{u \pm v}{1 \mp uv}\right) \pmod \pi \, , \quad u v \ne 1 \, . $$
This is derived from the tangent addition formula
$$ \tan(\alpha \pm \beta) = \frac{\tan(\alpha) \pm \tan(\beta)}{1 \mp \tan(\alpha)\tan(\beta)} \, , $$
by letting
$$ \alpha = \arctan(u) \, , \quad \beta = \arctan(v) \, . $$

\paragraph{In calculus}


\paragraph{Derivatives of inverse trigonometric functions}


The derivatives for complex values of $ z $ are as follows:
$$ \begin{aligned} \frac{d}{dz} \arcsin(z) &{} = \frac{1}{\sqrt{1-z^2}} ; && z \neq -1, +1 & \hspace{44em} \\\\ \\\\ \frac{d}{dz} \arccos(z) &{} = -\frac{1}{\sqrt{1-z^2}} ; && z \neq -1, +1 \\\\ \\\\ \frac{d}{dz} \arctan(z) &{} = \frac{1}{1+z^2} ; && z \neq -i, +i \\\\ \\\\ \frac{d}{dz} \text{arccot}(z) &{} = -\frac{1}{1+z^2} ; && z \neq -i, +i \\\\ \\\\ \frac{d}{dz} \text{arcsec}(z) &{} = \frac{1}{z^2 \sqrt{1 - \frac{1}{z^{2}}}} ; && z \neq -1, 0, +1 \\\\ \\\\ \frac{d}{dz} \text{arccsc}(z) &{} = -\frac{1}{z^2 \sqrt{1 - \frac{1}{z^{2}}}} ; && z \neq -1, 0, +1 \end{aligned} $$
Only for real values of $ x $:
$$ \begin{aligned} \frac{d}{dx} \text{arcsec}(x) &{} = \frac{1}{|x| \sqrt{x^2-1}}  \quad && |x| > 1 & \hspace{44em} \\\\ \frac{d}{dx} \text{arccsc}(x) &{} = -\frac{1}{|x| \sqrt{x^2-1}}  && |x| > 1 \end{aligned} $$

For a sample derivation: if $ \theta = \arcsin(x) $, we get:
$$ \frac{d \arcsin(x)}{dx} = \frac{d \theta}{d \sin(\theta)} = \frac{d \theta}{\cos(\theta) \, d \theta} = \frac{1}{\cos(\theta)} = \frac{1}{\sqrt{1-\sin^2(\theta)}} = \frac{1}{\sqrt{1-x^2}} $$

\paragraph{Expression as definite integrals}

Integrating the derivative and fixing the value at one point gives an expression for the inverse trigonometric function as a definite integral:
$$ \begin{aligned} \arcsin(x) &{}= \int _0^x \frac{1}{\sqrt{1 - z^2}} \, dz && |x| \leq 1 & \hspace{44em} \\\\ \arccos(x) &{}= \int _x^1 \frac{1}{\sqrt{1 - z^2}} \, dz && |x| \leq 1 \\\\ \arctan(x) &{}= \int _0^x \frac{1}{z^2 + 1} \, dz \\\\ \text{arccot}(x) &{}= \int _x^\infty \frac{1}{z^2 + 1} \, dz \\\\ \text{arcsec}(x) &{}= \int _1^x \frac{1}{z \sqrt{z^2 - 1}} \, dz = \pi + \int _x^{-1} \frac{1}{z \sqrt{z^2 - 1}} \, dz && x \geq 1 \\\\ \text{arccsc}(x) &{}= \int _x^\infty \frac{1}{z \sqrt{z^2 - 1}} \, dz = \int _{-\infty}^x \frac{1}{z \sqrt{z^2 - 1}} \, dz && x \geq 1 \\ \end{aligned} $$
When $ x $ equals 1, the integrals with limited domains are improper integrals, but still well-defined.

\paragraph{Indefinite integrals of inverse trigonometric functions}


For real and complex values of $ z $:
$$ \begin{aligned} \int \arcsin(z) \, dz &{}= z \, \arcsin(z) + \sqrt{1 - z^2} + C \hspace{44em} \\\\ \\\\ \int \arccos(z) \, dz &{}= z \, \arccos(z) - \sqrt{1 - z^2} + C \\\\ \\\\ \int \arctan(z) \, dz &{}= z \, \arctan(z) - \frac{1}{2} \ln \left( 1 + z^2 \right) + C \\\\ \\\\ \int \text{arccot}(z) \, dz &{}= z \, \text{arccot}(z) + \frac{1}{2} \ln \left( 1 + z^2 \right) + C \\\\ \\\\ \int \text{arcsec}(z) \, dz &{}= z \, \text{arcsec}(z) - \ln \left[ z \left( 1 + \sqrt{ \frac{z^2-1}{z^2} } \right) \right] + C \\\\ \\\\ \int \text{arccsc}(z) \, dz &{}= z \, \text{arccsc}(z) + \ln \left[ z \left( 1 + \sqrt{ \frac{z^2-1}{z^2} } \right) \right] + C \end{aligned} $$

For real $ x \geq 1 $:
$$ \begin{aligned} \int \text{arcsec}(x) \, dx &{}= x \, \text{arcsec}(x) - \ln \left( x + \sqrt{x^2-1} \right) + C \hspace{44em} \\\\ \int \text{arccsc}(x) \, dx &{}= x \, \text{arccsc}(x) + \ln \left( x + \sqrt{x^2-1} \right) + C \end{aligned} $$

For all real $ x $ not between -1 and 1:
$$ \begin{aligned} \int \text{arcsec}(x) \, dx &{}= x \, \text{arcsec}(x) - \text{sgn}(x) \ln\left(\left| x + \sqrt{x^2-1}\right|\right) + C \hspace{44em} \\\\ \int \text{arccsc}(x) \, dx &{}= x \, \text{arccsc}(x) + \text{sgn}(x) \ln\left(\left| x + \sqrt{x^2-1}\right|\right) + C \end{aligned} $$

The absolute value is necessary to compensate for both negative and positive values of the arcsecant and arccosecant functions. The signum function is also necessary due to the absolute values in the derivatives of the two functions, which create two different solutions for positive and negative values of x. These can be further simplified using the logarithmic definitions of the inverse hyperbolic functions:
$$ \begin{aligned} \int \text{arcsec}(x) \, dx &{}= x \, \text{arcsec}(x) - \operatorname{arcosh}(|x|) + C \hspace{44em} \\\\ \int \text{arccsc}(x) \, dx &{}= x \, \text{arccsc}(x) + \operatorname{arcosh}(|x|) + C \\\\ \end{aligned} $$

The absolute value in the argument of the arcosh function creates a negative half of its graph, making it identical to the signum logarithmic function shown above.

All of these antiderivatives can be derived using integration by parts and the simple derivative forms shown above.

\paragraph{Example}

Using $ \int u \, dv = u v - \int v \, du $ (i.e. integration by parts), set

$$ \begin{aligned}
u &= \arcsin(x) & dv &= dx \\\du &= \frac{dx}{\sqrt{1-x^2}} & v &= x
\end{aligned} $$

Then

$$ \int \arcsin(x) \, dx = x \arcsin(x) - \int \frac{x}{\sqrt{1-x^2}} \, dx, $$

which by the simple substitution $ w=1-x^2,\ dw = -2x\,dx $ yields the final result:

$$ \int \arcsin(x) \, dx = x \arcsin(x) + \sqrt{1-x^2} + C $$

\paragraph{Resources}

\paragraph{Review}

\begin{itemize}
    \item \href{https://www.khanacademy.org/math/geometry/hs-geo-trig/hs-geo-solve-for-an-angle/a/inverse-trig-functions-intro}{Introduction to Inverse Trig Functions}, Khan Academy
    \item \href{https://www.youtube.com/watch?v=jt7p-mCC0ng}{Evaluating Inverse Trigonometric Functions}, The Organic Chemistry Tutor
\end{itemize}

\paragraph{Calculus}

\begin{itemize}
    \item \href{https://www.youtube.com/watch?v=KbYW9FDm-Zk}{Derivatives of Inverse Trigonometric Functions}, The Organic Chemistry Tutor
\end{itemize}

\paragraph{Licensing}

Content obtained and/or adapted from:
\begin{itemize}
    \item \href{https://en.wikipedia.org/wiki/Inverse_trigonometric_functions}{Inverse trigonometric functions, Wikipedia} under a CC BY-SA license
\end{itemize}

\subsection{Learning Objectives}

\begin{enumerate}
    \item Understand and use the inverse sine, cosine, and tangent functions.
    \item Find the exact value of expressions involving the inverse sine, cosine, and tangent functions.
    \item Find exact values of composite functions with inverse trigonometric functions.
\end{enumerate}

\subsection{Assessment Instrument}

\begin{enumerate}
    \item \textbf{Assessment Instrument: Inverse Sine, Cosine, and Tangent Functions Quiz} \textbf{Format:}
    \item \textbf{Type:} Mixed-format quiz (multiple-choice, short answer, and problem-solving questions)
    \item \textbf{Duration:} 45 minutes
    \item \textbf{Total Marks:} 30 points \textbf{Sections:}
    \item \textbf{Understanding Inverse Trigonometric Functions (10 points):}
    \item \textbf{Question 1 (Multiple Choice):} Which of the following is the correct definition of $\arcsin(x)$?
    \item a. The angle whose cosine is $x$
    \item b. The angle whose tangent is $x$
    \item c. The angle whose sine is $x$
    \item d. The reciprocal of $\sin(x)$
    \item \textbf{Question 2 (Short Answer):} State the principal value range for the inverse sine function, $\arcsin(x)$.
    \item \textbf{Question 3 (Multiple Choice):} If $\arccos(x) = \frac{\pi}{3}$, what is $x$?
    \item a. $\frac{1}{2}$
    \item b. $-\frac{1}{2}$
    \item c. $\frac{\sqrt{3}}{2}$
    \item d. $-\frac{\sqrt{3}}{2}$
    \item \textbf{Computing Inverse Trigonometric Functions (10 points):}
    \item \textbf{Question 4 (Problem Solving):} Calculate the value of $\arctan\left(\frac{1}{\sqrt{3}}\right)$ and provide your answer in radians.
    \item \textbf{Question 5 (Short Answer):} Determine the value of $\arcsin\left(-\frac{1}{2}\right)$ in radians.
    \item \textbf{Question 6 (Multiple Choice):} If $\arcsin(x) = 0$, what is $x$?
    \item a. $1$
    \item b. $0$
    \item c. $-\frac{1}{2}$
    \item d. $\frac{\sqrt{2}}{2}$
    \item \textbf{Application of Inverse Trigonometric Functions (10 points):}
    \item \textbf{Question 7 (Problem Solving):} A ladder leans against a wall forming an angle $\theta$ with the ground. If the ladder is 10 meters long and the distance from the base of the ladder to the wall is 6 meters, find the angle $\theta$ using the appropriate inverse trigonometric function. Express your answer in degrees.
    \item \textbf{Question 8 (Problem Solving):} In a right-angled triangle, if the adjacent side to angle $\theta$ is 4 units and the opposite side is 3 units, calculate the value of $\theta$ using the inverse tangent function, and express your answer in degrees.
    \item \textbf{Question 9 (Short Answer):} Explain how the domain restrictions on the sine function affect the definition of its inverse, $\arcsin(x)$. \textbf{Scoring Rubric:}
    \item \textbf{Understanding Inverse Trigonometric Functions:} 3 points for each correct answer in Question 1, 2 points for Question 2, and 3 points for Question 3 (total 10 points).
    \item \textbf{Computing Inverse Trigonometric Functions:} 4 points for correct computation in Question 4, 3 points for the correct answer in Question 5, and 3 points for the correct answer in Question 6 (total 10 points).
    \item \textbf{Application of Inverse Trigonometric Functions:} 4 points for correct solution in Question 7, 4 points for correct solution in Question 8, and 2 points for a clear and correct explanation in Question 9 (total 10 points). The quiz will assess students' understanding of inverse trigonometric functions, their ability to compute values using these functions, and their application in solving real-world problems involving angles and trigonometric ratios.
\end{enumerate}

%%===============================================================
\section{Section 4.1 – Solving Trigonometric Equations with Identities}
\label{sec:section-4-1-solving-trigonometric-equations-with-identities}
%%===============================================================

\paragraph{Identities Based on the Pythagorean Theorem}


Some of the most fundamental trigonometric identities are those derived from the Pythagorean Theorem. These are defined using a right triangle:

% [Image: Right triangle]

Figure: Right triangle

At this stage in the course the Pythagorean Theorem should be second nature to you. If you have not yet learned it, learn it now.

$$a^2+b^2=c^2$$ $a$ and $b$ of course are the legs or the adjacent and opposite edges, and $c$ is the hypotenuse, the longest side, the side that does not include the right angle. This formula only works for a right angle triangle. If the angle shown as a right angle in the diagram were obtuse, larger than a right angle, then $c^2$ would be larger than the Pythagorean sum. If the angle shown as a right angle were smaller than a right angle, then $c^2$ would be smaller than $a^2+b^2$ .

\paragraph{Pythagorean Theorem in Terms of \textit{sine} and \textit{cosine}}


The Pythagorean Theorem is the same thing as

$$\sin^2(A)+\cos^2(A)=1$$ We know this from the earlier in the book where we introduced $\rm sine$ and $\rm cosine$ .

\begin{itemize}
    \item The $\rm cosine$ is the adjacent side when the hypotenuse is one.
    \item The $\rm sine$ is the opposite side when the hypotenuse is one.
\end{itemize}

The $\sin^2(A)+\cos^2(A)=1$ identity should be second nature to you too, but you should also be able to derive it from the Pythagorean relation.

We can see it's true from a right triangle that has a hypotenuse that is 'c' rather than one. Dividing the Pythagorean relation through by $c^2$ gives us

$$\left(\frac{a}{c}\right)^2+\left(\frac{b}{c}\right)^2=\left(\frac{c}{c}\right)^2=1$$

And we have already seen from soh-cah-toa that the sine of \textit{A} is $\left(\frac{a}{c}\right)$ and the cosine of \textit{A} is $\left(\frac{b}{c}\right)$ .

\paragraph{Identities based on Addition Formula}


We saw the addition formula earlier.

\paragraph{Letters are Arbitrary}


The letters in these equations $A$ and $B\,$ are arbitrary. We could equally use $\omega$ and $\theta$ . We're pointing this out because we happen to have a diagram on this page that has angles $A$ , $B$ and $C$ in it, and we want to make it clear that that was for Pythagoras and we're not talking about that diagram any more. In these equations you can also always substitute actual values, e.g. replace $A$ by $\pi$ and they'd still be true. Or you could replace $A$ by $2t$ or $\pi-t$and they'd still be true.

\paragraph{Addition Formula}


The addition formula for cosine is:

$$\begin{aligned} \cos(A+B)=\cos(A)\cos(B)-\sin(A)\sin(B) \hspace{44em} \end{aligned}$$
The formula above is important to understand. However, memorizing the following one is not necessary:

\paragraph{'Subtraction' Formula}


$$\begin{aligned} \cos(A-B)=\cos(A)\cos(B)+\sin(A)\sin(B) \hspace{44em} \end{aligned}$$

This formula can be derived by substituting $B$ with $-B$ in the previous equation. The $\cos(B)$ is unchanged because $\cos(-B)=\cos(B)$ . The minus sign before the second term changes because $\sin(-B)=-\sin(B)$ .

\paragraph{Checking the 'subtraction' formula}


Now do a quick check. If $A=B$ then $cos(A-B)$ is $\cos(0)$ which is one, right? And on the right hand side we have $\cos^2(A)+\sin^2(A)$ , which is also one. Looks good.

It is much better to remember it this way, because you are doing less rote learning and you're also becoming more fluent with the algebra.

\paragraph{Double Angle Formula}


\begin{itemize}
    \item $$\begin{aligned} \sin{(2x)} = 2\sin{x}\cos{x} & \hspace{44em} \end{aligned}$$
\end{itemize}

\begin{itemize}
    \item $$ \begin{aligned} \cos{(2x)} &= \cos^2{x} - \sin^2{x} & \hspace{44em} \\\\ &= 2\cos^2{x} - 1 \\\\ &= 1 - 2\sin^2{x} \end{aligned} $$
\end{itemize}

\begin{itemize}
    \item $$\begin{aligned} \tan{(2x)} = \frac{2\tan{x}}{1 - \tan^2{x}} & \hspace{44em} \end{aligned}$$
\end{itemize}


\paragraph{Half Angle Formula}


\begin{itemize}
    \item $$\begin{aligned} \sin{\left(\frac{x}{2}\right)} = \pm \sqrt{\frac{1 - \cos{x}}{2}} \hspace{44em} \end{aligned}$$
\end{itemize}

\begin{itemize}
    \item $$\begin{aligned} \cos{\left(\frac{x}{2}\right)} = \pm \sqrt{\frac{1 + \cos{x}}{2}} \hspace{44em} \end{aligned}$$
\end{itemize}

\begin{itemize}
    \item $$\begin{aligned} \tan{\left(\frac{x}{2}\right)} &= \pm \sqrt{\frac{1 - \cos{x}}{1 + \cos{x}}} & \hspace{44em}\\\\ &= \frac{\sin{x}}{1 + \cos{x}} \\\\ &= \frac{1 - \cos{x}}{\sin{x}} \end{aligned} $$
\end{itemize}

\paragraph{Product to Sum Formulas}


$$ \begin{aligned} \sin(u) \sin(v) &= \frac{1}{2}[\cos(u-v) - \cos(u+v)] \hspace{44em} \\\\ \cos(u) \cos(v) &= \frac{1}{2}[\cos(u-v) + \cos(u+v)] \\\\ \sin(u) \cos(v) &= \frac{1}{2}[\sin(u+v) + \sin(u-v)] \\\\ \cos(u) \sin(v) &= \frac{1}{2}[\sin(u+v) - \sin(u-v)] \end{aligned} $$

\paragraph{Sum to Product Formula}


$$ \begin{aligned} \sin(u) + \sin(v) &= 2\sin\left(\frac{u+v}{2}\right)\cos\left(\frac{u-v}{2}\right) \hspace{44em} \\\\ \sin(u) - \sin(v) &= 2\cos\left(\frac{u+v}{2}\right)\sin\left(\frac{u-v}{2}\right) \\\\ \cos(u) + \cos(v) &= 2\cos\left(\frac{u+v}{2}\right)\cos\left(\frac{u-v}{2}\right) \\\\ \cos(u) - \cos(v) &= -2\sin\left(\frac{u+v}{2}\right)\sin\left(\frac{u-v}{2}\right) \end{aligned} $$

\paragraph{Sum of two shifted Cosines}


The next formula may be approached by systematically working through the algebraic process.

$$\begin{aligned} \cos(x+\theta)+\cos(x-\theta)=\dots \hspace{44em} \end{aligned}$$

\paragraph{Waves in and out of phase}


Before we do, notice that we are adding two cosine waves, the first shifted left and the second shifted right. We've seen this before when we looked at waves in and out of phase. We're going to get another sinusoidal wave. The expression is fairly symmetric, and actually we can quickly show that the expression gives an even function, that is that the value for $-x$ is the same as for $x$ :

$$ \begin{aligned} \cos(-x+\theta) + \cos(-x-\theta) &= \cos(x-\theta) + \cos(x+\theta) \quad && \text{(since } \cos(-x) = \cos(x)) \hspace{44em} \\\\ &= \cos(x+\theta) + \cos(x-\theta) \quad && \text{(by swapping the terms)} \\\\ &= \cos(x + \theta) + \cos(x-\theta) \quad &&\text{the original expression.} \end{aligned} $$

with practice you'll be able to see those steps and that the expression is an even function immediately.

A sinusoidal wave that is an even function - well it is something based on cosine. We're expecting the formula will simplify to give us something like:

$$\begin{aligned} A\cos(x) \quad \text{ where $A$ will depend on $\theta$} \hspace{44em} \end{aligned}$$

\paragraph{Values to Check With}


When $\theta=0^\circ$ (try it in the original formula) we expect A to be 2. When $\theta=90^\circ$ we expect A to be 0, as the two cosine waves are $180^\circ$ out of phase. So now to plough through the algebra:

$$ \begin{aligned} \cos(x+\theta) + \cos(x-\theta) &= \cos(x)\cos(\theta) - \sin(x)\sin(\theta) + \cos(x)\cos(-\theta) - \sin(x)\sin(-\theta) & \hspace{44em} \\\\ &= \cos(x)\cos(\theta) - \sin(x)\sin(\theta) + \cos(x)\cos(\theta) + \sin(x)\sin(\theta) \\\\ &= \cos(x)\cos(\theta) + \cos(x)\cos(\theta) \\\\ &= 2\cos(x)\cos(\theta) \\\\ &= 2\cos(\theta)\cos(x) \end{aligned} $$

And to check we try with $\theta=0^\circ$ and get $2\cos(x)$ , and we check with $\theta=90^\circ$ and get $0\times\cos(x)$ , as we hoped.

Knowing roughly what we expected made it easier to get the algebraic steps right. We knew where we were going. We knew we weren't going to end up with four separate terms, some would have to cancel or combine in some other way.

\paragraph{Identities based on Sine and Cosine rules}


\paragraph{The Law of Sines}


$$\begin{aligned} \frac{a}{\sin(A)}=\frac{b}{\sin(B)}=\frac{c}{\sin(C)} \hspace{44em} \end{aligned}$$
This is easy to remember because it is so symmetrical. You do not even need to remember which way up the ratios are as this is also true:

$$\begin{aligned} \frac{\sin(A)}{a}=\frac{\sin(B)}{b}=\frac{\sin(C)}{c} \hspace{44em} \end{aligned}$$
What you do need to remember is how the triangle is labelled for it to be true.

\paragraph{The Law of Cosines}


With the cosine law formula:

$$\begin{aligned} a^2+b^2-2ab\cos(\theta)=c^2 \hspace{44em} \end{aligned}$$
It is best to think of it as a more general version of Pythagoras' theorem. The a and b have to be on equal footing, so ab as a multiplier is reasonable.

> If you are familiar with 'units' in physics, then use the fact that units must match up. The quantities $a,b,c$ are measurements of length. It is no good adding $a^2$ to $-2(a+b)\cos(\theta)$ . One might have units $\text{km}^2$ the other units of Km. Adding $a^2$ to $-2ab\cos(\theta)$ the units, e.g: $\text{km}^2$, do match.

In some more detail, it has got to be minus for $-2ab\cos(\theta)$ . That's because for $\theta<90^\circ$ we need $c^2$ to be less than for Pythagoras' theorem.

How can you remember and be sure the 2 is right? Think of an equilateral triangle with each side of length 1. The angle $\theta=60^\circ$ and $\cos(60^\circ)=0.5$. We need the 2 to get the right answer 1 + 1 + 2x1x1x0.5 = 1.

\paragraph{Ratio identities}


$$\begin{aligned} \tan(A)=\frac{\sin(A)}{\cos(A)} \hspace{44em} \end{aligned}$$
This is one definition of tan, and is something you should just learn.

\paragraph{Identities based on symmetry}


These symmetry identities are best remembered by remembering the graphs for these functions. Have a look at the graphs again if you don't remember them.

$0,\frac{\pi}{2},\pi,\frac{3\pi}{2},2\pi$ are landmarks on the graphs of sine and cosine, and you should know what happens for both those functions at those landmarks. When you have that knowledge the following symmetry identities are easy to write down, because you can see them visually.

\paragraph{Shifting Left}


$$\begin{aligned} \cos(\pi+A)  &=-\cos(A) & \hspace{44em} \\\\ \sin(\pi+A) &=-\sin(A) & \\\\ \tan(\pi+A) &=\tan(A) & \end{aligned}$$
The identities above are visualised as shifting the graph left by $\pi$ . It's a half cycle, and both cosine and sine change signs - and hence tan does not change sign, because when it is expressed as a ratio of sine over cosine both the top and bottom change signs.

\paragraph{Reflecting in $x=\frac{\pi}{2}$}


$$\begin{aligned}  \cos(\pi-A) &=-\cos(A) & \hspace{44em} \\\\ \sin(\pi-A) &=\sin(A) & \\\\ \tan(\pi-A) &=-\tan(A) & \end{aligned}$$
In the above we are reflecting the graph about the vertical line $x=\frac{\pi}{2}$. Spend enough time looking at this on the graph to see that that is true. It is important to do that to remember this trick for remembering what happens.

Now sine has a maximum at $x=\frac{\pi}{2}$ (it has the value one) and it is symmetric about that x value, so
$\newline$ $\sin(\pi-A)=\sin(A)$ . Cosine is zero at $x=\frac{\pi}{2}$ . Reflecting cosine in the line $x=\frac{\pi}{2}$ reverses the sign. $\newline$ $\cos(\pi-A)=-\cos(A)$

\paragraph{Co-function identities}


 $$\begin{aligned} \cos(A) &=\sin\left(\frac{\pi}{2}-A\right) & \hspace{44em} \\\\ \sin(A) &=\cos\left(\frac{\pi}{2}-A\right) & \end{aligned}$$

The above two formulas are most easily seen from the right-triangle definition of cosine and sine. The two acute angles in such a triangle add up to $\frac{\pi}{2}$ , and we are just giving the ratios of sides lengths in terms of a different angle's trig function.

 $$\begin{aligned} \cos(A) &=\sin\left(\frac{\pi}{2}+A\right) & \hspace{44em} \\\\ \sin(A) &=-\cos\left(\frac{\pi}{2}+A\right) & \end{aligned}$$
These two formulae are 'the same' as the previous formulae. We get there in two steps. First subtracted pi from the angle on the right hand side, which inverts the sign, and inverted the sign to compensate. Next invert the sign of the angle on the right hand side. For cosine on the right we are done. For sine on the right we have to invert its sign since inverting the angle's sign changed the sign of the result.

Yes - keeping track of signs is tricky, and always needs care.

\paragraph{Resources}


\begin{itemize}
    \item \href{https://mathresearch.utsa.edu/wikiFiles/MAT1093/Properties\%20of\%20the\%20Trigonometric\%20Functions/Esparza\%201093\%20Notes\%20Week\%203.pdf}{Properties of the Trigonometric}. Written notes created by Professor Esparza, UTSA.
\end{itemize}

\paragraph{Licensing}


Content obtained and/or adapted from:

\begin{itemize}
    \item \href{https://en.wikibooks.org/wiki/Trigonometry/Remembering_Trig_Formulae}{Remembering Trig Formulae, Wikibooks: Trigonometry} under a CC BY-SA license
\end{itemize}

\subsection{Learning Objectives}

\begin{enumerate}
    \item Verify the fundamental trigonometric identities.
    \item Simplify trigonometric expressions using algebra and the identities.
\end{enumerate}

\subsection{Assessment Instrument}

\begin{enumerate}
    \item \textbf{Assessment Instrument: Trigonometric Identities Quiz} \textbf{Format:}
    \item \textbf{Type:} Mixed-format quiz (multiple-choice, short answer, and problem-solving questions)
    \item \textbf{Duration:} 45 minutes
    \item \textbf{Total Marks:} 30 points \textbf{Sections:}
    \item \textbf{Pythagorean Identity (10 points):}
    \item \textbf{Question 1:} (Multiple Choice) Which of the following is the correct Pythagorean identity?
    \item a. $\sin^2(A) + \cos^2(A) = 1$
    \item b. $\sin(A) \cdot \cos(A) = 1$
    \item c. $\tan(A) = \dfrac{\sin(A)}{\cos(A)}$
    \item d. $\sin(A) + \cos(A) = 1$
    \item \textbf{Question 2:} (Short Answer) Derive the identity $\sin^2(A) + \cos^2(A) = 1$ from the Pythagorean Theorem. Show your steps clearly.
    \item \textbf{Question 3:} (Problem Solving) Given $\sin(A) = \dfrac{3}{5}$ and $A$ is in the first quadrant, use the Pythagorean identity to find $\cos(A)$.
    \item \textbf{Trigonometric Functions on the Unit Circle (10 points):}
    \item \textbf{Question 4:} (Multiple Choice) On the unit circle, the coordinates of a point corresponding to an angle $A$ are:
    \item a. $(\sin(A), \cos(A))$
    \item b. $(\cos(A), \sin(A))$
    \item c. $(\tan(A), 1)$
    \item d. $(1, \tan(A))$
    \item \textbf{Question 5:} (Short Answer) Explain why $\sin^2(A) + \cos^2(A) = 1$ holds true for any angle $A$ on the unit circle.
    \item \textbf{Question 6:} (Problem Solving) If the point on the unit circle corresponding to an angle $B$ is $\left(\dfrac{1}{2}, \dfrac{\sqrt{3}}{2}\right)$, what are the sine and cosine of $B$? Verify that they satisfy the Pythagorean identity.
    \item \textbf{Generalizing the Pythagorean Identity (10 points):}
    \item \textbf{Question 7:} (Multiple Choice) Which of the following represents a generalized form of the Pythagorean identity for any right triangle with hypotenuse $c$?
    \item a. $\dfrac{a^2 + b^2}{c^2} = 1$
    \item b. $\left(\dfrac{a}{c}\right)^2 + \left(\dfrac{b}{c}\right)^2 = 1$
    \item c. $a^2 + b^2 = c^2$
    \item d. $\left(\dfrac{\sin(A)}{c}\right)^2 + \left(\dfrac{\cos(A)}{c}\right)^2 = 1$
    \item \textbf{Question 8:} (Short Answer) Show how the Pythagorean identity $\sin^2(A) + \cos^2(A) = 1$ can be derived for a right triangle with a hypotenuse of length $c$, using the ratios of the triangle’s sides.
    \item \textbf{Question 9:} (Problem Solving) For a right triangle with legs of lengths 6 and 8, find the sine and cosine of the angle opposite the side of length 6. Verify that these values satisfy the Pythagorean identity. \textbf{Scoring Rubric:}
    \item \textbf{Pythagorean Identity:}
    \item \textbf{Question 1:} 2 points for correct identification of the Pythagorean identity (total 2 points)
    \item \textbf{Question 2:} 4 points for a correct and complete derivation of the identity (total 4 points)
    \item \textbf{Question 3:} 4 points for correctly finding $\cos(A)$ and explaining the process (total 4 points)
    \item \textbf{Trigonometric Functions on the Unit Circle:}
    \item \textbf{Question 4:} 2 points for correct selection of the coordinates (total 2 points)
    \item \textbf{Question 5:} 4 points for a correct explanation of why the identity holds true (total 4 points)
    \item \textbf{Question 6:} 4 points for correctly finding the sine and cosine of $B$ and verifying the identity (total 4 points)
    \item \textbf{Generalizing the Pythagorean Identity:}
    \item \textbf{Question 7:} 2 points for correct identification of the generalized identity (total 2 points)
    \item \textbf{Question 8:} 4 points for a correct and complete derivation (total 4 points)
    \item \textbf{Question 9:} 4 points for correctly finding the sine and cosine of the angle and verifying the identity (total 4 points) This quiz will assess the students' understanding and application of trigonometric identities, especially those derived from the Pythagorean Theorem. It ensures that students can derive, interpret, and generalize these identities in various mathematical contexts.
\end{enumerate}

%%===============================================================
\section{Section 4.2 – Sum and Difference Identities}
\label{sec:section-4-2-sum-and-difference-identities}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Use sum and difference formulas for cosine.
    \item Use sum and difference formulas for sine.
    \item Use sum and difference formulas for tangent.
    \item Use sum and difference formulas for cofunctions.
    \item Use sum and difference formulas to verify identities.
\end{enumerate}

%%===============================================================
\section{Section 4.3 – Double-Angle, Half-Angle, and Reduction Formulas}
\label{sec:section-4-3-double-angle-half-angle-and-reduction-formulas}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Use double-angle formulas to find exact values.
    \item Use double-angle formulas to verify identities.
    \item Use reduction formulas to simplify an expression.
    \item Use half-angle formulas to find exact values.
\end{enumerate}

%%===============================================================
\section{Section 4.4 – Sum-to-Product and Product-to-Sum Formulas}
\label{sec:section-4-4-sum-to-product-and-product-to-sum-formulas}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Express products as sums.
    \item Express sums as products.
\end{enumerate}

%%===============================================================
\section{Section 4.5 – Solving Trigonometric Equations}
\label{sec:section-4-5-solving-trigonometric-equations}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Solve linear trigonometric equations in sine and cosine.
    \item Solve equations involving a single trigonometric function.
    \item Solve trigonometric equations using a calculator.
    \item Solve trigonometric equations that are quadratic in form.
    \item Solve trigonometric equations using fundamental identities.
    \item Solve trigonometric equations with multiple angles.
    \item Solve right triangle problems.
\end{enumerate}

%%===============================================================
\section{Section 5.1 – Non-Right Triangles: Law of Sines}
\label{sec:section-5-1-non-right-triangles-law-of-sines}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Use the Law of Sines to solve oblique triangles.
    \item Find the area of an oblique triangle using the sine function.
    \item Solve applied problems using the Law of Sines.
\end{enumerate}

%%===============================================================
\section{Section 5.2 – Non-Right Triangles: Law of Cosines}
\label{sec:section-5-2-non-right-triangles-law-of-cosines}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Use the Law of Cosines to solve oblique triangles.
    \item Solve applied problems using the Law of Cosines.
    \item Use Heron’s formula to ?nd the area of a triangle.
\end{enumerate}

%%===============================================================
\section{Section 5.3 – Polar Coordinates}
\label{sec:section-5-3-polar-coordinates}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Plot points using polar coordinates.
    \item Convert from polar coordinates to rectangular coordinates.
    \item Convert from rectangular coordinates to polar coordinates.
    \item Transform equations between polar and rectangular forms.
    \item Identify and graph polar equations by converting to rectangular equations.
\end{enumerate}

%%===============================================================
\section{Section 5.4 – Polar Coordinates: Graphs}
\label{sec:section-5-4-polar-coordinates-graphs}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Test polar equations for symmetry.
    \item Graph polar equations by plotting points.
\end{enumerate}

%%===============================================================
\section{Section 5.5 – The Polar Form of Complex Numbers}
\label{sec:section-5-5-the-polar-form-of-complex-numbers}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Plot complex numbers in the complex plane.
    \item Find the absolute value of a complex number.
    \item Write complex numbers in polar form.
    \item Convert a complex number from polar to rectangular form.
    \item Find products of complex numbers in polar form.
    \item Find quotients of complex numbers in polar form.
    \item Find powers of complex numbers in polar form.
    \item Find roots of complex numbers in polar form.
\end{enumerate}

%%===============================================================
\section{Section 6.1 – Exponential Functions and Their Graphs}
\label{sec:section-6-1-exponential-functions-and-their-graphs}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Evaluate exponential functions.
    \item Find the equation of an exponential function.
    \item Use compound interest formulas.
    \item Evaluate exponential functions with base e.
    \item Graph exponential functions.
    \item Graph exponential functions using transformations.
\end{enumerate}

%%===============================================================
\section{Section 6.2 – Logarithmic Functions and Their Graphs}
\label{sec:section-6-2-logarithmic-functions-and-their-graphs}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Convert from logarithmic to exponential form.
    \item Convert from exponential to logarithmic form.
    \item Evaluate logarithms.
    \item Use common logarithms.
    \item Use natural logarithms.
    \item Identify the domain and range of a logarithmic function.
    \item Graph logarithmic functions.
\end{enumerate}

%%===============================================================
\section{Section 6.3 – Logarithmic Properties}
\label{sec:section-6-3-logarithmic-properties}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Use the product rule for logarithms.
    \item Use the quotient rule for logarithms.
    \item Use the power rule for logarithms.
    \item Expand logarithmic expressions.
    \item Condense logarithmic expressions.
    \item Use the change-of-base formula for logarithms.
\end{enumerate}

%%===============================================================
\section{Section 6.4 – Exponential and Logarithmic Equations}
\label{sec:section-6-4-exponential-and-logarithmic-equations}
%%===============================================================

\subsection{Learning Objectives}

\begin{enumerate}
    \item Use like bases to solve exponential equations.
    \item Use logarithms to solve exponential equations.
    \item Use the definition of a logarithm to solve logarithmic equations.
    \item Use the one-to-one property of logarithms to solve logarithmic equations.
    \item Solve applied problems involving exponential and logarithmic equations
\end{enumerate}
